### 2016-02-25 paper/root.tex
ini pap struct

% Markierungen: TODO
\documentclass{llncs}

\pagestyle{plain} % turn on page numbers
\usepackage[utf8]{inputenc}
\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
\usepackage{mathpartir}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\usepackage[boxed]{algorithm}
\usepackage{enumerate}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{color}
\usepackage[noend]{algpseudocode}
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}

\usepackage{multirow}

\input{macros}

\input{lstisabelle}
\lstset{captionpos=b}
\lstset{numberbychapter=false}

% \overfullrule=8pt

\begin{document}

\title{Working title: Verifying the Edmonds-Karp Algorithm}
\subtitle{A Proof Pearl}

\author{S.~Reza Sefidgar, Peter Lammich}

\institute{Technische Universit\"at M\"unchen, \email{\{sefidgar,lammich\}@in.tum.de}}

\maketitle
\begin{abstract}
the Ford-Fulkerson algorithm is a generic method that computes the maximum flow in a flow network. This paper outlines a formal proof of the Ford-Fulkerson theorem in Isabelle/HOL which follows the informal proof found in standard algorithm text books. This formalization is then used to develop an implementation of the algorithm in Standard ML. Refinement techniques are used to transform an abstract definition of the algorithm into an executable code.
\end{abstract}

\section{Introduction}
The Ford-Fulkerson algorithm [cite] is one of the important results of the graph theory and is used to find a solution to the maximum flow problem in flow networks. Many important mathematical problems like the maximum-bipartite-matching problem, the edge-disjoint-paths problem, the circulation-demand problem, and many other scheduling and resource allocating problems can be reduced to the maximum flow problem. Hence, the Ford-Fulkerson algorithm has major application in the field of mathematical optimization.

Despite its importance, no formalization of the Ford-Fulkerson algorithm has been developed in modern proof assistants. The only similar work that the authors are aware of is formalization of the Ford-Fulkerson algorithm in Mizar [cite]. This formalization defines and proves correctness of the algorithm at the level of graph manipulations without providing concrete implementation of the algorithm. Providing such an implementation is specially important, as it could provide the basis for many verfied programs with practical importance.

This paper present a formalization of the correctness proof of the Ford-Fulkerson algorithm in Isabelle/HOL. Our proof is based on the informal proof of the algorithm which is presented in the book "Introduction to algorithms" [cite]. Due to practical importance of the algorithm, we also present a verified implementation of the algorithm. Isabelle/HOL provides some automation for generating the code corresponding to a verified algorithm, however, in order to use the code generating features, one needs to do the formalization with executability in mind. Being limited only to executable concepts in the formalization has the disadvantage of cluttering the proofs with implementation details. Such approach makes the proofs more complicated, and may even render proofs of medium complex algorithms unmanageable.

One solution for the aforementioned problem is refinement [cite]. In order to generate executable code of an algorithm using refinement, we first formulate the algorithm on an abstract level. The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm. Next, we refine the abstract definition of the algorithm towards an executable implementation in possibly multiple refinement steps. During each step, we only need to prove the correctness of the implementation of a particular abstract concept. Hence, we can be sure about the correctness of the resulting executable program as each refinement step preserves the correctness.

There are several approaches to data refinement in Isabelle/HOL. We will be using the Autoref tool [cite], which has been used for proving more complex results such as formalized implementation of Hopcroft’s DFA-minimization algorithm [cite]. Given an algorithm phrased over abstract concepts like sets and maps, it automatically synthesizes a concrete, executable algorithm and the corresponding refinement theorem. To make it applicable for the development of actual algorithms, Autoref is integrated with the Isabelle Refinement Framework [cite] and the Isabelle Collection Framework [cite].

\section{Short Background on MinCutMaxFlow and FoFu}
  GOAL: 
    1) Remind the reader of FoFu. The educated reader should have the feeling:
      Yes, now I recall FoFu, and understand (again) what it does.

    2) Set the field for further, more detailed descriptions in next section.

  Short description of Network: Finite graph over nodes V, 
    edge (u,v) annotated with capacity c(u,v).
    Assuming distinct nodes s and t. 
    Note on additional assumptions: 
        No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
        Can transform any Network to match these assumptions, preserving the flow \cite{???}.

  s-t Flow: Annotation of edges with values, such that: 
    1) Values smaller than capacities.
    2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t

  Value of flow: Incoming flow - outgoing flow of s

  Max-Flow problem. 
    Rpt. why important?

  Cuts, intuition of min-cut >= max-flow

  Theorem: Min-Cut = Max-Flow
  Proven via augmenting flow in residual graph. 
    Present 1,2,3 of min-cut max-flow equivalences

  This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
    --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
      Termination? In General: Only for integer capacities.
      Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.

\section{Formalizing MinCutMaxFlow and Fofu}
  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
    Reader should think: Yeah, they did cool stuff!

  [???* Thematize definition of flows: On Graphs vs. on Networks. You find both in the literature. 
      We decides for [...]. This is better suited, as is gives nice and elegant proofs, even formally, bla bla bla]

  Proof of MinCut-MaxFlow: Follows the textbook proof.
    Uses Isar to even look like a textbook proof, being comprehensible even without running
      Isabelle/HOL.

    For example, augment_flow_presv_cap.
      Present formal proof text. Perhaps oppose it to textbook proof?.

  Abstract Algo looks like pseudocode presented in texbooks.
    Oppose our algo to textbook pseudocode.

\section{Refinement to executable code}
  GOAL: Same as previous section

  Introduction to refinement. (On abstract level)

  TODO: Show better complexity bound for FoFu with shortest path! --> Edmonds-Karp algo!

  FoFu-Scheme, can be instantiated with pathfinders now.
  These are developed independently, they are just search algorithms on graphs.
    We did BFS and DFS.
      DFS took from Refinement Framework examples.
      BFS: Based on our existing formalization for VSTTE12 verification competition, 
        but extended to return shortest path, and refined to imperative code.

        ??? Highlights of the BFS formalization?

  Down to executable code:
    Implement things that where not yet specified/specified in a non-executable way:
      Residual graph: By combination of c and f, and pred-succ. TODO: Directly!
        All we need is successor function. Implemented by tabulating 
          adjacent nodes of each node (pred-succ), and using c and f to filter out actual successor nodes. 
      augpath-spec by BFS/DFS (which work on succ-functions)
      bottleneck and augment: Give iterative impl.

    Use efficient data structures:
      Arrays for c,f, and ps
      List for augmenting path. [TODO/Future work: Iterator scheme]

  NetCheck

\section{Benchmarking}
\section{Conclusion} 
  ... and related work

Contributions

  Formal proof of mincut maxflow
  fofo-scheme
  inst to edmonds karp
  refinement down to executable code. Roughly 5 times slower than Java. (What about OCaml)
    + NetCheck

What shines (its a Pearl)
  Min-Cut Max-Flow: Textboook like formal reasoning: Comprehensible proof, BUT machine checked
    (Present one (carefully worked) example in paperI. We could use lemma augment_flow_presv_cap)

  Refinement based approach: Fofu-Scheme, instantiation to EdsKa. 
    +1: Abstract Algo looks almost like pseudo-code you would expect in textbook.
    +2: Fofu-Scheme proved correct for all aug-path finders. EdsKa is instantiation of it.
    +3: Modularity: Fofu-scheme and pathfinder developed+proved independently of each other.

  Down to executable code, plugging in efficient data structures.

Some minor contributions:
  Reusable BFS algorithm
  Imperative matrix data structure (really minor).

\bibliographystyle{abbrv}
\bibliography{root}

\end{document}

### 2016-02-25 paper/root.tex
More summary for refine-sectioon

@@
     Use efficient data structures:
-      Arrays for c,f, and ps
+      Arrays for c,f, and ps. c and f in row-major indexing.
+
       List for augmenting path. [TODO/Future work: Iterator scheme]
 
+      In BFS, we use lists for the queues (we have split the queue into one for 
+        the current level, and one for the next level), and an array for the predecessor map.
+
+      The refinement is performed using the Sepref tool.
+
   NetCheck
+    We also implemented an algorithm that reads a list of edges and a source and target node, 
+    converts it to a capacity matrix and an adjacency map, and checks whether the resulting graph satisfies our network assumptions.
+    We proved that this algorithm returns the corresponding Network and adjacency function iff the input describes a valid network,
+    and returns a failure value otherwise.
+
+  Combining the implementation of the Edmonds-Karp algorithm with the Network checker yields our final algorithm,
+  for which we can export code, and have proved the theorem: [...].
 

### 2016-02-25 paper/root.tex
first section complete

@@
 \section{Short Background on MinCutMaxFlow and FoFu}
+\begin{comment}
   GOAL: 
@@
       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
+  \end{comment}
+A flow network is a weighted digraph where the set of vertices (nodes) V is finite, and each edge (u, v) has a real capacity c (u, v) >=0.
+There are two distinct vertices s (source) and t (sink) in the flow network.
+In this case the network is called an s-t network.
+To make the formalization simple, some additional assumptions are added to the definition as following: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge (u, v) then there is no edge (v, u) in the reverse direction; and 3) every vertex of the flow network must be on a path from s to t.
+Although these assumptions seems restrictive, any network may be transformed such that it satisfies all the aformentioned properties.
+
+An s-t flow on a weighted digraph is a function f: E -> R which satisfies the following conditions. 1) (capacity constraint) the flow of
+each edge is a non-negative value which is smaller or equal to the edge capacity; 2) (conservation constraint) For all vertices except s and t, the sum of flows over incoming edges of the vertex is equal to the sum of flows over outgoing edges of the vertex.
+
+The value of an s-t flow f is denoted by |f |, and is computed by subtracting the sum of the flows over all incoming edges of s from the sum of the flows over all outgoing edges of s.
+Given an s-t network G, the maximum flow problem involves finding a flow f in G such that the value of f is greater or equal to the value of any other s-t flow in G.
+(Note that the value of the flow in our model of a flow network equals sum of flows over edges going out of s.
+Because the model prohibits edges coming into the source.)
+
+One of the important results in mathematical optimization is the corelation between flows and cuts in flow networks.
+An s-t cut in a flow network with source s and sink t is a partition of vertices that puts s and t in different subsets.
+Then the capacity of a cut is defined as sum of the edge capacities over all edges going from the source side to the sink side.
+Consider a valid flow in a flow network G.
+For any cut in G, the flow that crosses between the two subsets of the cut cannot exceed the capacity of the cut.
+So cuts define an upper bound for any flow in the flow network.
+The Ford-Fulkerson theorem tightens this bound and states that the value of the maximum flow is equal to the capacity of the minimum cut.
+
+The Ford-Fulkerson algorithm is the direct consequence of the Ford-Fulkerson theorem.
+It solves the maximum flow problem in a greedy approach.
+Assume an instance of the maximum flow problem in an s-t network.
+The algorithm starts with a zero flow: f ( e ) = 0 for all the edges e.
+In each iteration of the algorithm the value of f is increased by pushing flow along a path from s to t up to the limits imposed by the available edge capacities.
+The process would continue until no more paths can be found to push additional flow from s to t.
+
+The total flow value is increased by manipulating the flow values along different edges.
+However, the flow along specific edges might decrease during this process.
+In order to perform these operations, the Ford-Fulkerson algorithm defines the residual graph of the flow network.
+Intuitively, the residual graph corresponding to a flow in a given flow network, has the same vertices as the flow network.
+It also and consists of edges with capacities representing how flow can be changed along the edges of the flow network.
+
+In each iteration of the algorithm, the residual graph G_f corresponding to the current flow in the network is constructed.
+Then the algorithm looks for an augmenting path--- a simple path from source to the sink--- in the residual graph.
+If such a path could be found, it will be used to construct a flow in the residual graph.
+This flow provides a road map for modifying the current flow in the flow network.
+The process of increasing the current flow using an augmenting flow is called augmentation.
+Let f be the current flow in an s-t network G = ( V, E ) with capacity function c, and f be a flow in the corresponding residual graph.
+The augmentation of f using f' is a function from V × V to N, defined as following:
+(???? definition of augment function)
+
+The intuition behind the above definition is based on the definition of the residual graph. The augment procedure increases the flow on ( u, v ) by f ( u, v ) but decreases it
+by f ( v, u ) .
+This is due to the fact that pushing flow on the reverse edge on the residual graph signifies decreasing the flow in the original network.
+In each iteration of the algorithm we replace the current flow with the result of the augment function, and the algorithm terminates if no more augmenting paths can be found in the residual graph.
+According to the Ford-Fulkerson theorem, the flow computed in this states is indeed the maximum flow in the flow network.
+The theorem states that following statements are equivalent:
+
+1- f is a maximum s-t flow in flow network G.
+2- there is no augmenting path in the residual graph G_f corresponding to flow f and G.
+3- there is an s-t cut C in G such that capacity of C is equal to the value of f.
+
+The termination of the Ford-Fulkerson algorithm depends on how the augmenting path is found.
+For the networks with irrational edge capacities the algorithm might even fail to terminate.
+In practice, the maximum-flow problem often arises with integral capacities.
+Moreover, we may convert rational edge capacities to integral values by multiplying them with a big enough integer.
+If f* denotes a maximum-flow in the transformed network, then the algorithm will iterate at most |f*| times, since the augmentation procedure increases the flow value by at least one unit in each iteration.
+Assuming that we use breadth-first-search (BFS) or depth-first-search (DFS) for finding augmenting paths makes the total execution time of the Ford-Fulkerson algorithm O ( E | f ∗ |), as BFS and DFS run in O (E + V).
+
+The Ford-Fulkerson algorithm is sometimes called a method instead of an algorithm because it does not fully specify the procedure for finding the augmenting path in the residual graph.
+There are several implementations with different execution times.
+For instance, the Edmond-Karp algorithm uses breadth-first-search (BFS) for finding augmenting paths and has O ( VE 2 ) running-time.
+The augmentings paths that are computed using BFS are also shortest paths connecting source to the sink in the residual graph.
+Edmond and Karp showed that in the Ford–Fulkerson algorithm, if each augmenting path is the shortest one, the length of the augmenting paths is non-decreasing and it always terminates.
+Which finally gave the polynomial algorithm for solving the maximum flow problem in generic case of real edge capacities.
 

### 2016-02-26 paper/root.tex
changes

@@
 %\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
-\usepackage{mathpartir}
+% \usepackage{mathpartir}
 \DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
@@
 \section{Short Background on MinCutMaxFlow and FoFu}
-\begin{comment}
-  GOAL: 
-    1) Remind the reader of FoFu. The educated reader should have the feeling:
-      Yes, now I recall FoFu, and understand (again) what it does.
-
-    2) Set the field for further, more detailed descriptions in next section.
-
-  Short description of Network: Finite graph over nodes V, 
-    edge (u,v) annotated with capacity c(u,v).
-    Assuming distinct nodes s and t. 
-    Note on additional assumptions: 
-        No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
-        Can transform any Network to match these assumptions, preserving the flow \cite{???}.
-
-  s-t Flow: Annotation of edges with values, such that: 
-    1) Values smaller than capacities.
-    2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
-
-  Value of flow: Incoming flow - outgoing flow of s
-
-  Max-Flow problem. 
-    Rpt. why important?
-
-  Cuts, intuition of min-cut >= max-flow
-
-  Theorem: Min-Cut = Max-Flow
-  Proven via augmenting flow in residual graph. 
-    Present 1,2,3 of min-cut max-flow equivalences
-
-  This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
-    --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
-      Termination? In General: Only for integer capacities.
-      Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
-  \end{comment}
+% \begin{comment}
+%   GOAL: 
+%     1) Remind the reader of FoFu. The educated reader should have the feeling:
+%       Yes, now I recall FoFu, and understand (again) what it does.
+%       
+%     2) Set the field for further, more detailed descriptions in next section.
+%   
+% 
+%   Short description of Network: Finite graph over nodes V, 
+%     edge (u,v) annotated with capacity c(u,v).
+%     Assuming distinct nodes s and t. 
+%     Note on additional assumptions: 
+%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
+%         Can transform any Network to match these assumptions, preserving the flow \cite{???}.
+%         
+%   s-t Flow: Annotation of edges with values, such that: 
+%     1) Values smaller than capacities.
+%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
+%   
+%   Value of flow: Incoming flow - outgoing flow of s
+%   
+%   Max-Flow problem. 
+%     Rpt. why important?
+%   
+%   Cuts, intuition of min-cut >= max-flow
+%   
+%   Theorem: Min-Cut = Max-Flow
+%   Proven via augmenting flow in residual graph. 
+%     Present 1,2,3 of min-cut max-flow equivalences
+%   
+%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
+%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
+%       Termination? In General: Only for integer capacities.
+%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
+%   \end{comment}
 A flow network is a weighted digraph where the set of vertices (nodes) V is finite, and each edge (u, v) has a real capacity c (u, v) >=0.
@@
 
-In each iteration of the algorithm, the residual graph G_f corresponding to the current flow in the network is constructed.
+In each iteration of the algorithm, the residual graph $G_f$ corresponding to the current flow in the network is constructed.
 Then the algorithm looks for an augmenting path--- a simple path from source to the sink--- in the residual graph.
@@
 Let f be the current flow in an s-t network G = ( V, E ) with capacity function c, and f be a flow in the corresponding residual graph.
-The augmentation of f using f' is a function from V × V to N, defined as following:
+The augmentation of f using f' is a function from $V \times V$ to N, defined as following:
 (???? definition of augment function)
@@
 1- f is a maximum s-t flow in flow network G.
-2- there is no augmenting path in the residual graph G_f corresponding to flow f and G.
+2- there is no augmenting path in the residual graph $G_f$ corresponding to flow f and G.
 3- there is an s-t cut C in G such that capacity of C is equal to the value of f.
@@
 Moreover, we may convert rational edge capacities to integral values by multiplying them with a big enough integer.
-If f* denotes a maximum-flow in the transformed network, then the algorithm will iterate at most |f*| times, since the augmentation procedure increases the flow value by at least one unit in each iteration.
-Assuming that we use breadth-first-search (BFS) or depth-first-search (DFS) for finding augmenting paths makes the total execution time of the Ford-Fulkerson algorithm O ( E | f ∗ |), as BFS and DFS run in O (E + V).
+If f* denotes a maximum-flow in the transformed network, then the algorithm will iterate at most $|f^*|$ times, since the augmentation procedure increases the flow value by at least one unit in each iteration.
+Assuming that we use breadth-first-search (BFS) or depth-first-search (DFS) for finding augmenting paths makes the total execution time of the Ford-Fulkerson algorithm $O ( E | f^* |)$, as BFS and DFS run in O (E + V).
 
@@
 The augmentings paths that are computed using BFS are also shortest paths connecting source to the sink in the residual graph.
-Edmond and Karp showed that in the Ford–Fulkerson algorithm, if each augmenting path is the shortest one, the length of the augmenting paths is non-decreasing and it always terminates.
+Edmond and Karp showed that in the Ford-Fulkerson algorithm, if each augmenting path is the shortest one, the length of the augmenting paths is non-decreasing and it always terminates.
 Which finally gave the polynomial algorithm for solving the maximum flow problem in generic case of real edge capacities.
@@
 
-    For example, augment_flow_presv_cap.
+    For example, \isai{augment_flow_presv_cap}.
       Present formal proof text. Perhaps oppose it to textbook proof?.
@@
   ... and related work
+    Mizar-Formalization: What did they do.
 
@@
   Min-Cut Max-Flow: Textboook like formal reasoning: Comprehensible proof, BUT machine checked
-    (Present one (carefully worked) example in paperI. We could use lemma augment_flow_presv_cap)
+    (Present one (carefully worked) example in paperI. We could use lemma \isai{augment_flow_presv_cap})
 

### 2016-02-26 paper/root.tex
papaer modfied

@@
 %   \end{comment}
-A flow network is a weighted digraph where the set of vertices (nodes) V is finite, and each edge (u, v) has a real capacity c (u, v) >=0.
-There are two distinct vertices s (source) and t (sink) in the flow network.
-In this case the network is called an s-t network.
-To make the formalization simple, some additional assumptions are added to the definition as following: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge (u, v) then there is no edge (v, u) in the reverse direction; and 3) every vertex of the flow network must be on a path from s to t.
-Although these assumptions seems restrictive, any network may be transformed such that it satisfies all the aformentioned properties.
+A flow network is a weighted digraph where the set of vertices (nodes) $V$ is finite, and each edge $(u, v)$ has a real capacity $c (u, v) \geq 0$.
+There are two distinct vertices $s$ (source) and $t$ (sink) in the flow network.
+In this case the network is called an $s$-$t$ network.
+To make the formalization simple, some additional assumptions are added to the definition as following: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no edge $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$.
+Although these assumptions seems restrictive, any network may be transformed such that it satisfies all the aforementioned properties.
 
-An s-t flow on a weighted digraph is a function f: E -> R which satisfies the following conditions. 1) (capacity constraint) the flow of
-each edge is a non-negative value which is smaller or equal to the edge capacity; 2) (conservation constraint) For all vertices except s and t, the sum of flows over incoming edges of the vertex is equal to the sum of flows over outgoing edges of the vertex.
+An $s$-$t$ flow on a weighted digraph is a function $\colon V \times V \rightarrow \mathbb{R}$ which satisfies the following conditions.
+1) \textbf{Capacity constraint}: the flow of each edge is a non-negative value which is smaller or equal to the edge capacity; 2) \textbf{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over incoming edges of the vertex is equal to the sum of flows over outgoing edges of the vertex.
 
-The value of an s-t flow f is denoted by |f |, and is computed by subtracting the sum of the flows over all incoming edges of s from the sum of the flows over all outgoing edges of s.
-Given an s-t network G, the maximum flow problem involves finding a flow f in G such that the value of f is greater or equal to the value of any other s-t flow in G.
-(Note that the value of the flow in our model of a flow network equals sum of flows over edges going out of s.
-Because the model prohibits edges coming into the source.)
+The value of an $s$-$t$ flow f is denoted by $|f|$, and is computed by subtracting the sum of the flows over all incoming edges of $s$ from the sum of the flows over all outgoing edges of $s$.
+Given an $s$-$t$ network $G$, the maximum flow problem involves finding a flow $f$ in $G$ such that the value of $f$ is greater or equal to the value of any other $s$-$t$ flow in $G$~\footnote{Note that the value of the flow in our model of a flow network equals sum of flows over edges going out of $s$.
+Because the model prohibits edges coming into the source.}.
 
-One of the important results in mathematical optimization is the corelation between flows and cuts in flow networks.
-An s-t cut in a flow network with source s and sink t is a partition of vertices that puts s and t in different subsets.
-Then the capacity of a cut is defined as sum of the edge capacities over all edges going from the source side to the sink side.
-Consider a valid flow in a flow network G.
-For any cut in G, the flow that crosses between the two subsets of the cut cannot exceed the capacity of the cut.
+One of the important results in mathematical optimization is the correlation between flows and cuts in flow networks.
+An $s$-$t$ cut in a flow network with source $s$ and sink $t$ is a partition of vertices that puts $s$ and $t$ in different subsets.
+Then the capacity of a cut is defined as sum of the edge capacities over all edges going from the source's side to the sink's side.
+Consider a valid flow in a flow network $G$.
+For any cut in $G$, the flow that crosses between the two subsets of the cut cannot exceed the capacity of the cut.
 So cuts define an upper bound for any flow in the flow network.
@@
 It solves the maximum flow problem in a greedy approach.
-Assume an instance of the maximum flow problem in an s-t network.
-The algorithm starts with a zero flow: f ( e ) = 0 for all the edges e.
-In each iteration of the algorithm the value of f is increased by pushing flow along a path from s to t up to the limits imposed by the available edge capacities.
-The process would continue until no more paths can be found to push additional flow from s to t.
+Assume an instance of the maximum flow problem in an $s$-$t$ network.
+The algorithm starts with a zero flow: $f(e) = 0$ for all the edges $e$.
+In each iteration of the algorithm the value of $f$ is increased by pushing flow along a path from $s$ to $t$ up to the limits imposed by the available edge capacities.
+The process would continue until no more paths can be found to push additional flow from $s$ to $t$.
 
@@
 
-In each iteration of the algorithm, the residual graph $G_f$ corresponding to the current flow in the network is constructed.
+Assume a flow $f$ in an $s$-$t$ network with capacity function $c$.
+An edge $(u, v)$ of the network can admit an additional amount of flow equal to the edge’s capacity minus the flow along that edge.
+If this value is positive, then a $(u, v)$ edge is added to the residual graph with a residual capacity $c_f (u, v) = c (u, v) - f (u, v)$.
+So, the residual graph only contains those edges of the original flow network that have left-over capacities.
+The residual graph may also contain some edges that do not exist in the original flow network.
+For each edge $(u, v)$ in the flow network which has a positive flow value, a $(v, u)$ edge is added to the residual graph with the residual capacity $c_f (v, u) = f (u, v)$.
+The edge $(v, u)$ indicates the possibility for pushing the flow backward along the edge $(u, v)$ in the original flow network.
+
+After each update of the flow $f$, the algorithm considers the new residual graph $G_f$ corresponding to that flow.
 Then the algorithm looks for an augmenting path--- a simple path from source to the sink--- in the residual graph.
@@
 The process of increasing the current flow using an augmenting flow is called augmentation.
-Let f be the current flow in an s-t network G = ( V, E ) with capacity function c, and f be a flow in the corresponding residual graph.
-The augmentation of f using f' is a function from $V \times V$ to N, defined as following:
-(???? definition of augment function)
+Let $f$ be the current flow in an $s$-$t$ network $G = (V, E)$ with capacity function $c$, and $f$ be a flow in the corresponding residual graph.
+The augmentation of $f$ using $f'$ is a function from $V \times V$ to $\mathbb{R}$, defined as following:
+\[ (f \uparrow f') (u, v) = 
+	\begin{cases}
+	f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\	
+	0 \hfill & \text{ otherwise} \\
+	\end{cases} 
+\]
 
-The intuition behind the above definition is based on the definition of the residual graph. The augment procedure increases the flow on ( u, v ) by f ( u, v ) but decreases it
-by f ( v, u ) .
+The intuition behind the above definition is based on the definition of the residual graph.
+The augment procedure increases the flow on $(u, v)$ by $f ( u, v )$ but decreases it by $f (v, u)$.
 This is due to the fact that pushing flow on the reverse edge on the residual graph signifies decreasing the flow in the original network.
@@
 
-1- f is a maximum s-t flow in flow network G.
-2- there is no augmenting path in the residual graph $G_f$ corresponding to flow f and G.
-3- there is an s-t cut C in G such that capacity of C is equal to the value of f.
+\begin{itemize}
+\item $f$ is a maximum $s$-$t$ flow in flow network $G$.
+\item there is no augmenting path in the residual graph $G_f$.
+\item there is an $s$-$t$ cut $C$ in $G$ such that capacity of $C$ is equal to the value of $f$.
+\end{itemize}
 
@@
 Moreover, we may convert rational edge capacities to integral values by multiplying them with a big enough integer.
-If f* denotes a maximum-flow in the transformed network, then the algorithm will iterate at most $|f^*|$ times, since the augmentation procedure increases the flow value by at least one unit in each iteration.
-Assuming that we use breadth-first-search (BFS) or depth-first-search (DFS) for finding augmenting paths makes the total execution time of the Ford-Fulkerson algorithm $O ( E | f^* |)$, as BFS and DFS run in O (E + V).
+If $f^*$ denotes a maximum-flow in the transformed network, then the algorithm will iterate at most $|f^*|$ times, since the augmentation procedure increases the flow value by at least one unit in each iteration.
+Assuming that we use breadth-first-search (BFS) or depth-first-search (DFS) for finding augmenting paths makes the total execution time of the Ford-Fulkerson algorithm $O ( E | f^* |)$, as BFS and DFS run in $O (E + V)$.
 
@@
 There are several implementations with different execution times.
-For instance, the Edmond-Karp algorithm uses breadth-first-search (BFS) for finding augmenting paths and has O ( VE 2 ) running-time.
+For instance, the Edmond-Karp algorithm uses breadth-first-search (BFS) for finding augmenting paths and has $O (VE^2)$ running-time.
 The augmentings paths that are computed using BFS are also shortest paths connecting source to the sink in the residual graph.
@@
 \section{Formalizing MinCutMaxFlow and Fofu}
-  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
-    Reader should think: Yeah, they did cool stuff!
-
-  [???* Thematize definition of flows: On Graphs vs. on Networks. You find both in the literature. 
-      We decides for [...]. This is better suited, as is gives nice and elegant proofs, even formally, bla bla bla]
-
-  Proof of MinCut-MaxFlow: Follows the textbook proof.
-    Uses Isar to even look like a textbook proof, being comprehensible even without running
-      Isabelle/HOL.
-
-    For example, \isai{augment_flow_presv_cap}.
-      Present formal proof text. Perhaps oppose it to textbook proof?.
-
-  Abstract Algo looks like pseudocode presented in texbooks.
-    Oppose our algo to textbook pseudocode.
+%  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
+%    Reader should think: Yeah, they did cool stuff!
+%
+%  [???* Thematize definition of flows: On Graphs vs. on Networks. You find both in the literature. 
+%      We decides for [...]. This is better suited, as is gives nice and elegant proofs, even formally, bla bla bla]
+%    
+%  Proof of MinCut-MaxFlow: Follows the textbook proof.
+%    Uses Isar to even look like a textbook proof, being comprehensible even without running
+%      Isabelle/HOL.
+%      
+%    For example, \isai{augment_flow_presv_cap}.
+%      Present formal proof text. Perhaps oppose it to textbook proof?.
+%    
+%  Abstract Algo looks like pseudocode presented in texbooks.
+%    Oppose our algo to textbook pseudocode.
+%
 

### 2016-02-29 paper/root.tex
not compiling
Darfts of impl/refine sections

@@
 % \usepackage{mathpartir}
-\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
+%\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
 \usepackage[boxed]{algorithm}
@@
 
+\usepackage{packages/isabelle}
+\usepackage{packages/isabelletags}
+\usepackage{packages/isabellesym}
+\usepackage{packages/comment}
+
 \input{macros}
@@
 \lstset{numberbychapter=false}
+
+% Include snippets
+\newcommand{\DefineSnippet}[2]{%
+   \expandafter\newcommand\csname snippet--#1\endcsname{%
+     \begin{quote}
+     \begin{isabelle}
+     #2
+     \end{isabelle}
+     \end{quote}}}
+\newcommand{\Snippet}[1]{\csname snippet--#1\endcsname}
+\input{snippets/snippets}
 
@@
 
+\Snippet{isar-lem}
+
+\section{Refinement in Isabelle}
+Purpose: Separation of concerns.
+
+Short intro on Refinement Framework:
+
+Programs phrased as error-set monad. With order structure. 
+Refinement: Less possible results (refinement ordering). Error is greatest element (refined by everything).
+
+Building chains of refinement, at the end there should be a deterministic algorithm (yielding exactly one result), which is
+contained (by transitivity) in the specification.
+
+Standard combinators (bind, recursion, if, etc) are monotonic wrt. refinement ordering, which allows for modularity:
+Replacing part of a program by a refined version, yields a refined program.
+
+Data refinement: Changing the representation of data. Relation between concrete and abstract representation. 
+$\Uparrow m$ is biggest concrete program that refines $m$ wrt.\ relation $R$.
+
+The Isabelle Refinement Framework provides the nres-monad, and tools to reason about program refinement, 
+the most important one being a VCG.
+
+\section{Edmonds-Karp Algorithm}
+  The general Ford-Fulkerson method can only be shown to terminate if the edge capacities are integer numbers. 
+  An improvement over that is the Edmonds-Karp algorithm~\cite{}, which is obtained when always choosing 
+  a shortest augmenting path. In this case, it can be shown that only $O(|V||E|)$ augmentations are performed until the
+  algorithm terminates, even for real-valued capacities. A shortest augmenting path can be obtained by breadth first search in time $O(|E|)$,
+  yielding an overall complexity of $O(|V||E|^2) $ algorithm.
+
+  In our formalization, we refine the specification of an augmenting path to a specification of a shortest augmenting path, which immediately yields 
+  an abstract version of the Edmonds-Karp algorithm which refines the Ford-Fulkerson method.
+
+  The larger part of the formalization is spent on proving the complexity bound. Note that the refinement framework does not 
+  have a notion of computational complexity, so we cannot even define the runtime of an algorithm. However, we can instrument 
+  the while-loop of the algorithm with a counter, which is incremented on each iteration, and prove an upper bound on this counter.
+  Moreover, this also yields a termination argument for real-valued capacities.\footnote{As we restricted the algorithm 
+  to integer valued capacities at the very abstract level, this is actually not used.}
+
+  The idea of the complexity proof is as follows: Note that edges that are reverse to edges on a shortest augmenting path cannot lie 
+  on a shortest path itself.
+  On every augmentation, at least one edge that lies on a shortest path is flipped. Thus, either the length of the shortest path increases,
+  or the number of edges that lie on a shortest path decreases. As the length of a shortest path is at most $|V|$, there are no more than $O(|E||V|)$ iterations.
+
+  Formalizing the above intuitive argument is more tricky than it seems on first glance. 
+  While it easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
+  to a graph transformation which may add reverse edges and removes at least original edge, is tricky. Note that a straightforward induction
+  on the length of the augmenting path must fail, as after flipping the first edge, the path is no longer augmenting.
+
+  We decided to generalize the statement as follows. Assume we have an original graph $cf$ with a shortest augmenting path $p$, and a transformed graph $cf'$ that has been created from
+  $cf$ by adding some flipped edges from $p$ and removing some edges from $p$.
+  Then, we consider an edge $(u,v) \in p$, a path $p_1$ from $s$ to $v$ in the original graph, and a path $p_2$ from $u$ to $t$ in the transformed graph.
+
+  By induction on the number of reversed edges in $p_2$, we 
+  show that $|p_1|+|p_2| > |p|$. In the induction step, we use the proof idea for the single flipped edge $(v,u)$, and,
+  if $p_2$ contains more flipped edges, we split $p_2$ at the first flipped edge. Then, the initial segment of $p_2$ contains no flipped edges,
+  and thus also is a path in the original graph, and we can apply the induction hypothesis.
+
+  Proving that augmentation with a shortest augmenting path actually only adds flipped edges, and removes at least one original edge, required some more work,
+  but was straightforward and yielded no surprises.
+
+  Finally, we phrase the complexity argument as a measure function $m = l*2|E| + n$, where $l$ is the length of the shortest augmenting path,
+  and $n$ is the number of edges that lie on any shortest path. We show that $m$ is decreased by the loop body.
+  Adding a special case for loop termination (there are no augmenting paths), and observing that $m$ is bounded by $2|V||E|$, yields 
+  the desired upper bound on loop iterations. Refinement of the algorithm to add an explicit loop counter, and asserting the upper bound, is then straightforward.
+
 \section{Refinement to executable code}
@@
 
-  Introduction to refinement. (On abstract level)
+  In the previous sections, we have described the Ford-Fulkerson and the Edmonds-Karp algorithm abstractly,  
+  leaving open how to obtain a (shortest) augmenting path. A standard way to find a shortest path in a graph 
+  is breadth first search (BFS). Luckily, we had already formalized a BFS algorithm as an example for the refinement framework.
+  However, the existing formalization could only compute the minimum distance between two nodes, without returning an actual path.
+  We briefly report on our adapted formalization here, which is displayed in Figure~\fig{TODO: Snippet}: 
+  TODO: Reference to figure (line numbers?)
+  The abstract algorithm keeps track of a set $C$ of current nodes, a set $N$ of new nodes, and a partial map $P$ from already visited nodes to predecessor nodes.
+  Initially, only the start node is in $C$ and $N$ is empty. In each iteration of the loop, a node $u\in C$ is picked, and its successors $v$ that have not yet been 
+  visited are added to $N$, and $P~v$ is set to $u$. If $C$ becomes empty, $C$ and $N$ are swapped. If the target node is encountered, the algorithm immediately terminates.
+  Note that this implementation is a generalized version of the usual queue implementation of BFS: While a queue enforces a complete order on the encountered nodes, our 
+  implementation only enforces an ordering between nodes on the current level and nodes on the next level.
 
-  TODO: Show better complexity bound for FoFu with shortest path! --> Edmonds-Karp algo!
+  For the actual algorithm, we wrap this algorithm by a procedure to handle the special case where source and target nodes are the same, and to extract a shortest path 
+  from $P$ upon termination. Moreover, we implement the abstract specification of adding the successor nodes of a node by a loop.
+
+  Finally, we prove the following theorem: %TODO: Not actually true, the impls are oly done in bfs2!
+%     theorem
+%       assumes "src∈V" 
+%       shows "bfs src dst ≤ (spec p. 
+%         case p of None ⇒ ¬conn src dst | Some p ⇒ shortestPath src p dst)"
 
@@
     Implement things that where not yet specified/specified in a non-executable way:
-      Residual graph: By combination of c and f, and pred-succ. TODO: Directly!
+
+      Residual graph: All operations of the algorithm can also be expressed as only modifying the residual graph. 
+      Although we can compute the edges of the residual graph from the capacities and the current flow on the fly,
+      this still requires 2 or 3 accesses to adjacency matrices per edge.
+
+      Thus, we refine the algorithm to directly operate on residual graphs. The shortest path specification and bottleneck computation are already 
+      phrased on residual graphs. What remains is the computation of the initial residual graph (which equals the Network),
+      the augmentation of the residual graph, given an augmenting path and its bottleneck capacity, and the computation of the 
+      resulting maximum flow from the residual graph, at the end of the algorithm.
+
+      In order to refine these operations, we note that the residual graph uniquely determines the flow (and vice versa). 
+%         definition (in Network) "flow_of_cf cf e ≡ (if (e∈E) then c e - cf e else 0)"
+
+      For a valid flow $f$, we have that
+%         lemma (in NFlow)
+%           "flow_of_cf (residualGraph f) = f"
+%           by (rule fo_rg_inv)
+
+      Thus, augmentation can be expressed directly on the residual graph:
+%         "residualGraph (augment (augmentingFlow p)) (u,v) = (
+%           if (u,v)∈set p then (residualGraph f (u,v) - bottleNeck p)
+%           else if (v,u)∈set p then (residualGraph f (u,v) + bottleNeck p)
+%           else residualGraph f (u,v))"
+
+      Having these characterizations, a straightforward refinement yields an algorithm that operates on residual graphs.
+
+      In a next step, we have to specify iterative implementations for finding the bottleneck capacity and doing the augmentation.
+      This is straightforward by folding over the path, exploiting that an augmenting path is simple, and thus never contains an edge 
+      and its flipped edge at the same time.
+
+      By combination of c and f, and pred-succ. TODO: Directly!
         All we need is successor function. Implemented by tabulating 
@@
 
+      In order to find an augmenting path, the BFS algorithm has to compute the successors of a node. 
+      Although this can be implemented on the capacity matrix by iterating over all nodes, this implementation tends to be inefficient for sparser graphs.
+      A common optimization is to pre-compute a map from nodes to adjacent nodes in the network. As an edge in the residual graph is either parallel or reverse to 
+      a network edge, it is enough to iterate over the adjacent nodes in the network, and check whether they are actually successors in the residual graph.
+      It is straightforward to show that this implementation actually returns the successors in the residual graph:
+%         lemma (in RGraph) rg_succ_ref:
+%           assumes A: "is_pred_succ ps c"
+%           assumes B: "u∈V"
+%           shows "rg_succ2 ps cf u ≤ SPEC (λl. (l,cf.E``{u}) ∈ ⟨Id⟩list_set_rel)"
+
     Use efficient data structures:
+      Now we have an abstract version of the algorithm that only relies on executable concepts.
+      However, to yield an efficient algorithm, we have to choose efficient data structures.
+
+      We assume that the nodes are natural numbers less than $N$. 
+      Then we implement the capacity matrix of the residual graph by an array with row-major indexing. 
+      The adjacency map for network nodes is implemented as an array of lists. 
+      In the BFS algorithm, we use two lists, one that holds the nodes of the current level, and another 
+      in which nodes of the next level are collected. The predecessor map is implemented as an array.
+
+      The input of the algorithm is still a function mapping network edges to capacities, which is tabulated once 
+      to obtain the initial residual graph. This ensures some flexibility in using the algorithm, without preventing 
+      efficient implementations (The function can be based on array lookup itself\footnote{Due to technical limitations of our tools, this function 
+        cannot depend on heap content itself, but it can use the efficient functional and pseudo-functional datastructures provided by the Collection Framework.}).
+
+      The output of the algorithm is just the residual graph. The user can decide how he computes the maximum flow from it. For example,
+      in order to compute the maximum flow value, only the outgoing edges of the source node have to be computed, which is typically less 
+      expensive than computing the complete flow matrix. The correctness theorem of the algorithm abstractly states how to obtain the maximum flow from the output.
+
+      The last refinement step was performed using the Sepref tool, which supports data refinements from functional to imperative algorithms in Isabelle/HOL.
+      [Few sentences on Sepref]
+
       Arrays for c,f, and ps. c and f in row-major indexing.
@@
 \section{Benchmarking}
+  What implementations did we compare: Authors? Where do implementations come from? 
+  [We must convince the reader that we did not intentionally chose bad implementations to compare us against!]
+
+  Comparsion of algorithms (Modified data structure: Flow vs Flow-like vs ResGraph)
+    Computing successors in BFS: Filter by visited set first.
+      -> Technically more challenging, when computing on flow: 1 access first vs 2 accesses first. On resGraph: No advantage in memory accesses (However, plain array access vs. matrix access)
+
 \section{Conclusion} 
@@
     Mizar-Formalization: What did they do.
+    Future Work: Dinic. (Actually, our abstract scheme [almost] covers Dinic's algorithm!)
 
@@
   inst to edmonds karp
-  refinement down to executable code. Roughly 5 times slower than Java. (What about OCaml)
+  complexity analysis of edmonds karp
+  refinement down to executable code. Roughly 2.5 times slower than Java. (What about OCaml)
     + NetCheck

### 2016-02-29 paper/root.tex
Abstract

@@
 
-\title{Working title: Verifying the Edmonds-Karp Algorithm}
-\subtitle{A Proof Pearl}
+\title{Formalizing the Edmonds-Karp Algorithm}
+% \subtitle{}
 
@@
 \begin{abstract}
-the Ford-Fulkerson algorithm is a generic method that computes the maximum flow in a flow network.
-This paper outlines a formal proof of the Ford-Fulkerson theorem in Isabelle/HOL which follows the informal proof found in standard algorithm text books.
-This formalization is then used to develop an implementation of the algorithm in Standard ML.
-Refinement techniques are used to transform an abstract definition of the algorithm into an executable code.
+We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network. 
+The formal proofs are done in Isabelle/HOL. They closely follow a standard textbook proof and are relatively easy 
+to read, even without being an Isabelle expert.
+
+We then use stepwise refinement to obtain the Edmonds-Karp algorithm and prove its upper complexity bound of $O(|V||E|^2)$.
+Further refinement yields a verified implementation, which is only 2.5 times slower than an unverified  
+reference implementation in Java.
+
+% 
+% and a verified imperative implementation of it. Our implementation is roughly 2.5 times slower than an unverified  
+% reference implementation in Java. 
+% 
+% Finally, we prove the upper complexity bound of $O(|V||E|^2)$ for the Edmonds-Karp algorithm.
 \end{abstract}
@@
   However, the existing formalization could only compute the minimum distance between two nodes, without returning an actual path.
-  We briefly report on our adapted formalization here, which is displayed in Figure~\fig{TODO: Snippet}: 
+  We briefly report on our adapted formalization here, which is displayed in Figure~\ref{TODO: Snippet}: 
   TODO: Reference to figure (line numbers?)

### 2016-02-29 formalization/Peter_OldAlgo/evaluation/data/data-gen.c
Moved in Edka based on old graphs

@@ -1,0 +1,183 @@
+#include "stdbool.h"
+#include "string.h"
+#include "time.h"
+#include "stdio.h"
+#include "stdlib.h"
+
+/* This function creates a square matrix to store our graph,
+ * It also initializes all the capacities to zero */
+int** graph_init (int size) {
+	int** graph = malloc(size * sizeof(int*));
+
+	int i;
+	for (i = 0; i < size; i++)
+		graph[i] = calloc(size, sizeof(int));
+
+	return graph;
+}
+
+/* As our matrix is a int** we have to clean it up in a loop */
+void graph_free (int** graph, int size) {
+	int i;
+	for (i = 0; i < size; i++)
+		free(graph[i]);
+
+	free(graph);
+}
+
+/* After execution the "visited" array shows all those edges which where reached
+ * during the dfs execution on "graph". Note that the "reverse" flag indicates
+ * if we have to use the original graph or its reverse. We use this function to
+ * computing the set of nodes that are reachable from source, and the set of no-
+ * des that are reaching sink. */
+void dfs (int** graph, bool* visited, int start, int size, bool reverse) {
+	visited[start] = true;
+
+	int i;
+	for (i = 0; i < size; i++) {
+		int edge_cap = (!reverse) ? graph[start][i] : graph[i][start];
+
+		if (edge_cap != 0 && !visited[i])
+			dfs(graph, visited, i, size, reverse);
+	}
+}
+
+/* This function produces a random network which has no self loops. Also, there
+ * will be no parallel edges in the output network.*/
+void net_init (int** graph, int size, int try_count, int max_capacity) {
+	srand(time(NULL));
+
+	/* We assume the source is in index 0, and sink is in index size - 1. We put
+	 * random edges between any vertex except these two. and we connect them later */
+	int i;
+	for (i = 0; i < try_count; i++) {
+		int u = 1 + (rand() % (size - 2));
+		int v = 1 + (rand() % (size - 2));
+		int c = 1 + (rand() % max_capacity);
+
+		if (u != v && graph[v][u] == 0)
+			graph[u][v] = c;
+	}
+
+	/* compute the number of vertices that are connected to source and sink, we
[…]

### 2016-02-29 formalization/Peter_OldAlgo/evaluation/fofu-C++/fofu.cpp

@@ -1,0 +1,161 @@
+
+#include "time.h"
+#include <cstdio>
+#include <queue>
+#include <cstring>
+#include <vector>
+#include <iostream>
+#include <fstream>
+
+#define G_SIZE 10000
+
+using namespace std;
+
+vector<int> graph[G_SIZE];
+int capacities[G_SIZE][G_SIZE];
+int flowPassed[G_SIZE][G_SIZE];
+int parentsList[G_SIZE];
+int currentPathCapacity[G_SIZE];
+
+//int** graph_init (int size) {
+//	int** GRAPH = new int* [size];
+//
+//	int i;
+//	for (i = 0; i < size; i++){
+//		GRAPH[i] = new int [size];
+//	}
+//
+//	return GRAPH;
+//}
+//
+///* As our matrix is a int** we have to clean it up in a loop */
+//void graph_free (int** GRAPH, int size) {
+//	int i;
+//	for (i = 0; i < size; i++) {
+//		delete GRAPH[i];
+//		for (int j = 0; j < size; j++)
+//			GRAPH[i][j] = 0;
+//	}
+//
+//	delete GRAPH;
+//}
+
+int stat_counter_ga = 0;
+int stat_counter_mget = 0;
+
+int bfs(int startNode, int endNode)
+{
+    memset(parentsList, -1, sizeof(parentsList));
+    memset(currentPathCapacity, 0, sizeof(currentPathCapacity));
+
+    queue<int> q;
+    q.push(startNode);
+
+    parentsList[startNode] = -2;
+    currentPathCapacity[startNode] = 999;
+
+    while(!q.empty())
+    {
+        int currentNode = q.front();
+        q.pop();
+
+        for(int i=0; i<graph[currentNode].size(); i++)
+        {
+            stat_counter_mget+=3;
+            int to = graph[currentNode][i];
+            if(capacities[currentNode][to] - flowPassed[currentNode][to] > 0)
+            {
+                stat_counter_ga++;
+                if(parentsList[to] == -1)
+                {
+                    parentsList[to] = currentNode;
+                    currentPathCapacity[to] = min(currentPathCapacity[currentNode],
+                    capacities[currentNode][to] - flowPassed[currentNode][to]);
+                    if(to == endNode)
+                    {
+                        return currentPathCapacity[endNode];
+                    }
+                    q.push(to);
+                }
+            }
+        }
+    }
[…]

### 2016-02-29 paper/root.tex
Section on refinement framework

@@
 
-\section{Refinement in Isabelle}
-Purpose: Separation of concerns.
+\section{Refinement in Isabelle/HOL}
+When formalizing algorithms, there is often a trade off between verifiability 
+and efficiency: An efficient algorithm that uses elaborate data structures and optimizations is often harder to 
+verify than a simple but straightforward algorithm.
 
-Short intro on Refinement Framework:
+The idea of stepwise refinement \cite{Wirth} is to first specify a simple, abstract version of an algorithm, which can be easily verified,
+and then refine this algorithm towards an efficient implementation. Here, each refinement step can focus on a single aspect of the algorithm or 
+its implementation, independently of the other refinement steps. This results in a greatly increased modularity of proofs. 
+Refinement calculi \cite{Back} are used to systematically perform refinement in a Hoare-logic setting, and are particularly well-suited for 
+implementation in theorem provers.
 
-Programs phrased as error-set monad. With order structure. 
-Refinement: Less possible results (refinement ordering). Error is greatest element (refined by everything).
+Note that it is important to support nondeterminism in a refinement setting, as this allows to defer 
+implementation choices to the later refinement steps. For example, if we want to specify an algorithm 
+that returns a shortest path between two nodes of a graph, we do not want to fix a particular shortest path.
+Only later, when we decide to implement the specification by, e.g., breadth first search, we fix a particular path.
 
-Building chains of refinement, at the end there should be a deterministic algorithm (yielding exactly one result), which is
-contained (by transitivity) in the specification.
+In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12}. 
+It features a refinement calculus for programs phrased in a nondeterminism monad. 
+The monad's type is a set of results plus an additional value that indicates a failure:
+\begin{lstlisting}
+  datatype \<alpha> nres = res \<alpha> set | fail
+\end{lstlisting}
+The return operation \isai{return x} of the monad describes the single result \isai{x}, and the bind 
+operation \isai{bind m f} nondeterministically picks a result from $m$ and executes $f$ on it. 
+If either \isai{m = fail}, or \isai{f} may fail for a result in $m$, the bind operation fails.
 
-Standard combinators (bind, recursion, if, etc) are monotonic wrt. refinement ordering, which allows for modularity:
-Replacing part of a program by a refined version, yields a refined program.
+On nondeterministic results we define the \emph{refinement ordering} by lifting the subset ordering with \isai{fail} being the greatest element.
+Intuitively, $m\le m'$ means that $m$ is a refinement of $m'$, i.e., all possible results of $m$ are 
+also possible results of $m'$. 
+Note that this ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
+Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
+language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define constants for loop constructs (while, foreach), 
+assertions and specifications, and use a Haskell-like do-notation:
+\begin{lstlisting}
+  assert \<Phi> = if \<Phi> then return () else fail
+  spec x. P x = RES {x. P x}
+  do {x <- m; f x} = bind m f
+  do {m; m'} = bind m (%_. m')
+\end{lstlisting}
 
-Data refinement: Changing the representation of data. Relation between concrete and abstract representation. 
-$\Uparrow m$ is biggest concrete program that refines $m$ wrt.\ relation $R$.
+Correctness of a program $m$ with precondition $P$ and postcondition $Q$ is expressed as \isai{P ==> m <= spec r. Q r}, which
+means that, if $P$ holds, $m$ does not fail and all possible results of $m$ satisfy $Q$. Note that we provide different recursion constructs
+for partial and total correctness: A nonterminating total correct recursion yields \isai{fail}, while a nonterminating partial correct 
+recursion yields \isai|res {}|, which is the least element of the refinement ordering.
 
-The Isabelle Refinement Framework provides the nres-monad, and tools to reason about program refinement, 
-the most important one being a VCG.
+The Isabelle Refinement Framework also supports data refinement, changing the representation of results according to a \emph{refinement relation}, 
+which relates concrete with abstract results: Given a relation $R$, \isai{\<Down> R m} is the largest set of concrete results that are related to an 
+abstract result in $m$ by $R$. If \isai{m=fail}, then also \isai{\<Down> R m = fail}.
+
+Finally, the Isabelle Refinement Framework provides a refinement calculus, which comes with a verification condition 
+generator to simplify its usage.
+
+In a typical program development using the Isabelle refinement framework, one first comes up with an initial version $m_0$ of
+the algorithm and its specification $P,Q$, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions $m_i$ of the algorithm,
+proving \isai{$m_i$ <= \<Down>$R_i$ $m_{i-1}$}. Using transitivity and composability of data refinement, one 
+gets \isai{P ==> $m_i$ <= \<Down>$R_i\ldots R_1$ spec Q}, showing the correctness of the refined algorithm.
+
+xxx, ctd here ...
+Refinement itself can be modular ... sub-chains of refinement to refine one part of algo. Monotonicity to combine.
+
+% Each refinement step is shown to preserve correctness
+% 
+% Refinement techniques \cite{Wirth,Back} structure the verification of an algorithm into
+% modular steps, such that one first verifies the abstract algorithmic ideas, and then 
+% 
+% 
+% 
+% Purpose: Separation of concerns.
+% 
+% Short intro on Refinement Framework:
+% 
+% Programs phrased as error-set monad. With order structure. 
+% Refinement: Less possible results (refinement ordering). Error is greatest element (refined by everything).
+% 
+% Building chains of refinement, at the end there should be a deterministic algorithm (yielding exactly one result), which is
+% contained (by transitivity) in the specification.
+% 
+% Standard combinators (bind, recursion, if, etc) are monotonic wrt. refinement ordering, which allows for modularity:
+% Replacing part of a program by a refined version, yields a refined program. 
+% 
+% Data refinement: Changing the representation of data. Relation between concrete and abstract representation. 
+% $\Uparrow m$ is biggest concrete program that refines $m$ wrt.\ relation $R$. 
+% 
+% The Isabelle Refinement Framework provides the nres-monad, and tools to reason about program refinement, 
+% the most important one being a VCG.
 
@@
 
-\section{Refinement to executable code}
+\section{Refinement to Executable Code}
   GOAL: Same as previous section
@@
 
-  FoFu-Scheme, can be instantiated with pathfinders now.
-  These are developed independently, they are just search algorithms on graphs.
-    We did BFS and DFS.
-      DFS took from Refinement Framework examples.
-      BFS: Based on our existing formalization for VSTTE12 verification competition, 
-        but extended to return shortest path, and refined to imperative code.
-
-        ??? Highlights of the BFS formalization?
+  Obviously, the BFS algorithm refines the specification for obtaining a shortest path.
+  Using this to refine the Edmonds-Karp formalization yields an algorithm that algorithmically specifies all major operations.
 
   Down to executable code:
-    Implement things that where not yet specified/specified in a non-executable way:
 
-      Residual graph: All operations of the algorithm can also be expressed as only modifying the residual graph. 
-      Although we can compute the edges of the residual graph from the capacities and the current flow on the fly,
-      this still requires 2 or 3 accesses to adjacency matrices per edge.
+  In a next step, we observe that the algorithm is phrased in terms of a flow which is updated until it is maximal. However,
+  in order to update the flow, the residual graph is used, which is a combination of the current flow and the network.
+  Obviously, computing the complete residual graph each time before searching for an augmenting path would be a bad idea. 
+  One solution to this problem is to compute the edges of the residual graph on demand from the network and the current flow.
+  Although this solution seems to be common, it has the disadvantage that for each edge of the residual graph, two (or even three) 
+  edges of the network and the flow have to be accessed. As edges of the residual graph are accessed in the inner loop, during 
+  the BFS, these operations are especially time critical.
 
-      Thus, we refine the algorithm to directly operate on residual graphs. The shortest path specification and bottleneck computation are already 
+  Thus, we chose to let the algorithm operate on a representation of the residual graph directly.
+
+      The shortest path specification and bottleneck computation are already 
       phrased on residual graphs. What remains is the computation of the initial residual graph (which equals the Network),
@@
       -> Technically more challenging, when computing on flow: 1 access first vs 2 accesses first. On resGraph: No advantage in memory accesses (However, plain array access vs. matrix access)
+    Computing bottleneck during BFS. Saves one iteration over the path, at the cost of one extra memory access per discovered node.
+    For our benchmarks, we observed that the BFS discovers considerably more nodes than the length of the ultimately returned path, 
+    and the overall code was faster with the extra iteration for bottleneck computation.
 
@@
     Mizar-Formalization: What did they do.
+
+    History of the formalization: 
+      Impl: From first impl that was hardly able to compute a 100 nodes graph, to current efficient one.
+
     Future Work: Dinic. (Actually, our abstract scheme [almost] covers Dinic's algorithm!)

### 2016-03-01 paper/root.tex
Adjusted runtime difference in abstract
Finished Fofu formalization section

@@
 We then use stepwise refinement to obtain the Edmonds-Karp algorithm and prove its upper complexity bound of $O(|V||E|^2)$.
-Further refinement yields a verified implementation, which is only 2.5 times slower than an unverified  
+Further refinement yields a verified implementation, which is almost as fast as an unverified  
 reference implementation in Java.
@@
 
+A formal proof contains many technical details that are not presented in an informal proof.
+While dealing with the technical details in a proof causes distraction, leaving out such details makes it hard to follow the proofs.
+Our formal proof of the Ford-Fulkerson method closely follows the informal proof described by Cormen et al.
+[cite].
+So, readers may use this informal proof as a guide to better understand the relevant thoughts in our formal proof.
+Moreover using the proof language Isar, we made the formal proof readable even for readers that are not familiar with Isabelle.
+
+As an example consider the following formal proof.
+Let $cf$ be the residual graph corresponding to a flow $f$ in an $s$-$t$ network $c$, and let $f'$ be a flow in $cf$.
+Consider the definition of the augment function that was described in previous section.
+The following lemma states that the result of augmenting $f$ with $f'$ is another function which satisfies the capacity constraint of a $s$-$t$ flow in $c$.
 \Snippet{isar-lem}
+
+Most of the reference books in the field of algorithm provide pseudocodes instead of executable implementations. However, implementing an abstract presentation of 
+an algorithm involves many technical details that are never presented in the books.
+We also tried to fill in this gap.
+First, we present the Ford-Fulkerson method using the syntax of Autoref[cite] tool which is quite similar to pseudocode of the algorithm.
+Next, we refine the abstract representation into an executable implementation.
+For instance, the following Isabelel definition abstractly presents the Ford-Fulkerson method.
+\Snippet{6.1-1_0}
+\Snippet{6.1-1}
 

### 2016-03-01 paper/root.tex
snippets
Draft sections on refinement completed

@@
 
+\def\isadelimproof{}
+\def\endisadelimproof{}
+\def\isatagproof{}
+\def\endisatagproof{}
+\def\isafoldproof{}
+\def\isadelimproof{}
+\def\endisadelimproof{}
+
 \input{macros}
@@
      \begin{quote}
-     \begin{isabelle}
+     \begin{isabelle}\footnotesize
      #2
@@
 The following lemma states that the result of augmenting $f$ with $f'$ is another function which satisfies the capacity constraint of a $s$-$t$ flow in $c$.
-\Snippet{isar-lem}
+\Snippet{augment_flow_presv_cap}
 
@@
 For instance, the following Isabelel definition abstractly presents the Ford-Fulkerson method.
-\Snippet{6.1-1_0}
-\Snippet{6.1-1}
+% \Snippet{6.1-1_0}
+% \Snippet{6.1-1}
 
@@
 
-xxx, ctd here ...
-Refinement itself can be modular ... sub-chains of refinement to refine one part of algo. Monotonicity to combine.
+Monotonicity of the standard combinators also allows for modular refinement: Replacing part of a program by a refined version
+results in a program that refines the original program.
+
+In Section~\ref{sec:fofu}, we presented the abstract Ford-Fulkerson algorithm as a program in the Isabelle Refinement Framework.
+In the next sections, we describe its stepwise refinement down to executable code.
 
@@
 
+  TODO: Relate to Cormen-proof: Critical edge, removed by augmentation, bound on how often an edge can get critical.  
+    Same idea, but yields a more precise bound. On the other hand, we found our proof to be simpler to formalize, and its enough to
+    get the $O(|V||E|)$ bound.
+
   Formalizing the above intuitive argument is more tricky than it seems on first glance. 
@@
 \section{Refinement to Executable Code}
-  GOAL: Same as previous section
+  \subsection{Using Breadth First Search}
 
@@
   Obviously, the BFS algorithm refines the specification for obtaining a shortest path.
-  Using this to refine the Edmonds-Karp formalization yields an algorithm that algorithmically specifies all major operations.
+  Using this to refine the Edmonds-Karp formalization yields a program that algorithmically specifies all major operations.
 
-  Down to executable code:
-
+  \subsection{Manipulating Residual Graphs Directly}    
   In a next step, we observe that the algorithm is phrased in terms of a flow which is updated until it is maximal. However,
@@
   edges of the network and the flow have to be accessed. As edges of the residual graph are accessed in the inner loop, during 
-  the BFS, these operations are especially time critical.
+  the BFS, these operations are time critical.
 
@@
 
-      In a next step, we have to specify iterative implementations for finding the bottleneck capacity and doing the augmentation.
-      This is straightforward by folding over the path, exploiting that an augmenting path is simple, and thus never contains an edge 
-      and its flipped edge at the same time.
+%   Note: Straightforward, so omitted here!
+%   In a next step, we have to specify iterative implementations for finding the bottleneck capacity and doing the augmentation.
+%   This is straightforward by folding over the path, exploiting that an augmenting path is simple, and thus never contains an edge 
+%   and its flipped edge at the same time.
 
-      By combination of c and f, and pred-succ. TODO: Directly!
-        All we need is successor function. Implemented by tabulating 
-          adjacent nodes of each node (pred-succ), and using c and f to filter out actual successor nodes. 
-      augpath-spec by BFS/DFS (which work on succ-functions)
-      bottleneck and augment: Give iterative impl.
+  \subsection{Computing Successors}
+  In order to find an augmenting path, the BFS algorithm has to compute the successors of a node in the residual graph. 
+  Although this can be implemented on the capacity matrix by iterating over all nodes, this implementation tends to be inefficient for sparse graphs,
+  where we would have to iterate over many nodes just to find that there is no edge.
 
-      In order to find an augmenting path, the BFS algorithm has to compute the successors of a node. 
-      Although this can be implemented on the capacity matrix by iterating over all nodes, this implementation tends to be inefficient for sparser graphs.
   A common optimization is to pre-compute a map from nodes to adjacent nodes in the network. As an edge in the residual graph is either parallel or reverse to 
@@
 
-    Use efficient data structures:
-      Now we have an abstract version of the algorithm that only relies on executable concepts.
-      However, to yield an efficient algorithm, we have to choose efficient data structures.
+  \subsection{Using Efficient Data Structures}    
+  In a final step, we have to chose efficient data structures for the algorithm. The objects that the algorithm works with are
+  capacities, nodes, edges, the graph of the input network, residual graphs, the adjacency map, paths, and the resulting maximum flow.
 
-      We assume that the nodes are natural numbers less than $N$. 
-      Then we implement the capacity matrix of the residual graph by an array with row-major indexing. 
-      The adjacency map for network nodes is implemented as an array of lists. 
-      In the BFS algorithm, we use two lists, one that holds the nodes of the current level, and another 
-      in which nodes of the next level are collected. The predecessor map is implemented as an array.
+  We implement capacities as (arbitrary precision) integer numbers. Note that an implementation as fixed precision numbers would also be possible,
+  but requires additional checks on the network to ensure that no overflows can happen. 
+  % TODO: Can we trick the Java reference implementation on integer overflow?
 
-      The input of the algorithm is still a function mapping network edges to capacities, which is tabulated once 
-      to obtain the initial residual graph. This ensures some flexibility in using the algorithm, without preventing 
-      efficient implementations (The function can be based on array lookup itself\footnote{Due to technical limitations of our tools, this function 
-        cannot depend on heap content itself, but it can use the efficient functional and pseudo-functional datastructures provided by the Collection Framework.}).
+  We implement nodes as natural numbers less than an upper bound $N$, and residual graphs are implemented by their capacity matrices, which, in turn,
+  are realized as arrays of size $N*N$ with row-major indexing.
+  The adjacency map of the network is implemented as an array of lists of nodes.
 
-      The output of the algorithm is just the residual graph. The user can decide how he computes the maximum flow from it. For example,
+  For the BFS algorithm, we implement the sets $C$ and $N$ by lists, and the predecessor map $P$ as an array of \isai{node option}. 
+  The successors of a node are represented as a list with no duplicate elements. The resulting path, which is extracted from $P$ upon termination of
+  the BFS, is represented as a list of edges.
+
+  The input network of the algorithm is represented as a function mapping network edges to capacities, which is tabulated into an array
+  to obtain the initial residual graph.
+
+  This allows flexibility in using the algorithm, as any capacity matrix representation can be converted into a function easily, and 
+  without loosing efficiency for read-access\footnote{Due to technical limitations of our tools, this function 
+  cannot depend on heap content, but it can use the efficient functional and pseudo-functional datastructures provided by the Isabelle Collection Framework.}.
+
+  The output flow of the algorithm is represented as the residual graph. The user can decide how he computes the maximum flow from it. For example,
   in order to compute the maximum flow value, only the outgoing edges of the source node have to be computed, which is typically less 
@@
 
-      The last refinement step was performed using the Sepref tool, which supports data refinements from functional to imperative algorithms in Isabelle/HOL.
-      [Few sentences on Sepref]
+  Note that their is still some optimization potential left: For example, the BFS algorithm computes the predecessor map $P$.
+  Then, it iterates over $P$ to get the shortest path into a list. Then, this list is iterated again to compute the 
+  bottleneck, and, finally, another iteration over the list is done to do the augmentation. This calls for a deforestation optimization,
+  getting rid of the intermediate list altogether, and iterate two times over the predecessor map: A first time to compute the bottleneck, and a second time
+  to do the augmentation. Fortunately, iteration over the shortest path does not significantly contribute to the runtime of the algorithm, so that we leave this
+  optimization for future work.
 
-      Arrays for c,f, and ps. c and f in row-major indexing.
+  Note that we performed the last refinement step using our Sepref tool~\cite{La15}, which provides tool support for data refinements from 
+  purely functional programs of the Isabelle Refinement Framework into imperative programs expressed in Imperative/HOL~\cite{BKHEM08}.
+  The formalization of this refinement step consists of setting up the mappings between the abstract and concrete data structures,
+  and then using the tools provided by Sepref to synthesize the Imperative/HOL programs and their refinement proofs.
 
-      List for augmenting path. [TODO/Future work: Iterator scheme]
+  Finally, Imperative/HOL provides a heap-exception monad and comes with a setup for the Isabelle code generator to generate efficient imperative programs
+  in OCaml, SML, Scala, and Haskell.
 
-      In BFS, we use lists for the queues (we have split the queue into one for 
-        the current level, and one for the next level), and an array for the predecessor map.
-
-      The refinement is performed using the Sepref tool.
-
-  NetCheck
-    We also implemented an algorithm that reads a list of edges and a source and target node, 
-    converts it to a capacity matrix and an adjacency map, and checks whether the resulting graph satisfies our network assumptions.
-    We proved that this algorithm returns the corresponding Network and adjacency function iff the input describes a valid network,
+  \subsection{Network Checker}
+  We also implemented an algorithm that takes as input a list of edges, a source node, and a target node.
+  It converts these to a capacity matrix and an adjacency map, and checks whether the resulting graph satisfies our network assumptions.
+  We proved that this algorithm returns the correct capacity matrix and adjacency map iff the input describes a valid network,
   and returns a failure value otherwise.
 
-  Combining the implementation of the Edmonds-Karp algorithm with the Network checker yields our final algorithm,
-  for which we can export code, and have proved the theorem: [...].
+  Combining the implementation of the Edmonds-Karp algorithm with the network checker yields our final algorithm,
+  for which we can export code, and have proved the theorem: [...Snippet! edmonds-karp-correct]. 
+%     theorem edmonds_karp_correct:
+%     "<emp> edmonds_karp el s t <λ
+%         None ⇒ ↑(¬ln_invar el ∨ ¬Network (ln_α el) s t)
+%       | Some (N,fi) ⇒ ∃⇩Af. Network.is_rflow (ln_α el) N f fi * ↑(isMaxFlow (ln_α el) s t f)
+%           * ↑(ln_invar el ∧ Network (ln_α el) s t ∧ Graph.V (ln_α el) ⊆ {0..<N})
+%     >⇩t"
+  Note that this theorem is stated as a Hoare triple, using separation logic assertions. There are no preconditions on the input.
+  If the algorithm returns \isai{None}, then the edge list was malformed or described a graph that does not satisfy the network assumptions.
+  Here, \isai{ln_invar} describes well-formed edge lists, i.e., edge lists that have no duplicate edges and no edges with zero capacity,
+  and \isai{ln_\<alpha>} describes the mapping from (well-formed) edge lists to capacity matrices.
+  If the algorithm returns some number $N$ and residual graph $fi$, then the input was a well-formed edge list that describes a valid network.
+  Moreover, the returned residual graph describes a flow $f$ in the network, which is maximal. TODO: Swap in thm!
+  As there are no other possible return values, this theorem completely describes the correctness of the algorithm. 
+  Note that Isabelle/HOL does not have a notion of execution, thus total correctness of the code finally produced by the code 
+  generator cannot be defined. However, the program is phrased in a heap-exception monad, thus introducing some (coarse grained) notion of
+  computation. On this level, termination can be ensured, and, indeed, the above theorem implies that all the recursions stated by recursion 
+  combinators in the monad must terminate. However, it does not guarantee that we have not injected spurious code equations: E.g. $f~x = f~x$,
+  is provable by reflexivity, but causes the generated program to diverge.
 

### 2016-03-01 paper/root.tex
modfied first section

@@
 
-\section{Short Background on MinCutMaxFlow and FoFu}
+\section{A short background on the Ford-Fulkerson method}
 % \begin{comment}
@@
 
-Assume a flow $f$ in an $s$-$t$ network with capacity function $c$.
+%Assume a flow $f$ in an $s$-$t$ network with capacity function $c$.
 An edge $(u, v)$ of the network can admit an additional amount of flow equal to the edge’s capacity minus the flow along that edge.
@@
 
-After each update of the flow $f$, the algorithm considers the new residual graph $G_f$ corresponding to that flow.
+There are two kinds of edges in the residual graph: 1) those of the residual graph that present the edges of the flow network with left-over capacities; and 2) those of the residual graph that do not occur in the flow network, and indicate the possibility for pushing the flow backward.
+Formally speaking, consider a flow $f$ on an $s$-$t$ network $G = (V, E)$ with the capacity function $c$.
+A residual capacity function $c_f \colon V \times V \rightarrow \mathbb{R}$ is defined as following:
+\[ c_f (u, v) = 
+	\begin{cases}
+	c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
+	f (v, u) \hfill & \text{ if $(v, u) \in E$} \\	
+	0 \hfill & \text{ otherwise} \\
+	\end{cases} 
+\]
+
+Then the residual graph corresponding to $f$ and $G$ is indicated by $G_f = (V, E_f)$, where the residual edge set $E_f$ is defined as following:
+\[ E_f = \{(u,v) \in V \times V \colon c_f (u, v) > 0\}\]
+
+After each flow update, the algorithm considers the residual graph corresponding to the new flow.
 Then the algorithm looks for an augmenting path--- a simple path from source to the sink--- in the residual graph.
 If such a path could be found, it will be used to construct a flow in the residual graph.
+Let $p$ be such a path.
+The residual capacity of $p$ is defined as $c_f(p) = min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}$.
+Then an augmenting flow $f_p \colon V \times V \rightarrow \mathbb{N}$  in the residual graph can be defined using $p$ as following:
+\[ f_p (u, v) = 
+	\begin{cases}
+	c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\	
+	0 \hfill & \text{ otherwise} \\
+	\end{cases} 
+\]
+
 This flow provides a road map for modifying the current flow in the flow network.
 The process of increasing the current flow using an augmenting flow is called augmentation.
-Let $f$ be the current flow in an $s$-$t$ network $G = (V, E)$ with capacity function $c$, and $f$ be a flow in the corresponding residual graph.
+Let $f$ be the current flow in an $s$-$t$ network $G = (V, E)$ with capacity function $c$, and $f'$ be a flow in the corresponding residual graph.
 The augmentation of $f$ using $f'$ is a function from $V \times V$ to $\mathbb{R}$, defined as following:
@@
 
-\section{Formalizing MinCutMaxFlow and Fofu}
+\section{Formalizing the Ford-Fulkerson method}
 %  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.

### 2016-03-01 paper/root.tex
modfied first section
defined algorithm snippets
finished defining fofu

@@
 The residual capacity of $p$ is defined as $c_f(p) = min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}$.
-Then an augmenting flow $f_p \colon V \times V \rightarrow \mathbb{N}$  in the residual graph can be defined using $p$ as following:
+Then an augmenting flow $f_p \colon V \times V \rightarrow \mathbb{R}$  in the residual graph can be defined using $p$ as following:
 \[ f_p (u, v) = 
@@
 
-A formal proof contains many technical details that are not presented in an informal proof.
+%A formal proof contains many technical details that are not presented in an informal proof.
 While dealing with the technical details in a proof causes distraction, leaving out such details makes it hard to follow the proofs.
-Our formal proof of the Ford-Fulkerson method closely follows the informal proof described by Cormen et al.
-[cite].
+.
 So, readers may use this informal proof as a guide to better understand the relevant thoughts in our formal proof.
@@
 
-As an example consider the following formal proof.
+Informal proofs focus on the relevant thoughts by leaving out technical details.
+However, in a formal proof each step has to be precisely specified as a valid application of some inference rule which makes formal proofs often lengthy and hard to follow.
+In the case that an interactive theorem prover is used, readers may even need to replay the proof to understand the idea behind it.
+In Isabelle proof assistant, the Isar proof language bridges the semantic gap between internal notions of proof given by interactive theorem proving systems and an appropriate level of abstraction for user-level work.
+Isar makes formal proofs readable even for those who are not familiar with Isabelle concepts.
+We have used such a functionality to present our formal proof of the Ford-Fulkerson method which resembels the informal proof described by Cormen et al.
+[cite].
+
+As an example consider the following proof obtained form Cormen's algorithm book.
 Let $cf$ be the residual graph corresponding to a flow $f$ in an $s$-$t$ network $c$, and let $f'$ be a flow in $cf$.
 Consider the definition of the augment function that was described in previous section.
-The following lemma states that the result of augmenting $f$ with $f'$ is another function which satisfies the capacity constraint of a $s$-$t$ flow in $c$.
+Then, the result of augmenting $f$ with $f'$ is another function which satisfies the capacity constraint of a $s$-$t$ flow in $c$.
+
+\emph{\textbf{Proof.}} first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
+	\begin{align*}
+	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of augment)} \\
+	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
+	& = f'(u, v) \\
+	& \geq 0.
+	\end{align*}
+In addition,
+	\begin{align*}
+	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of augment)} \\
+	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
+	& \leq f(u, v) + c_f(u, v) &&  \text{(capacity constraint)} \\
+	& = f'(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
+	& = c (u, v).
+	\end{align*}
+
+Following we present the equivalent Isar proof.
+Note that we have to additionally consider the case that $(u, v) \notin E$.
+Moreover in line 6, we have used the assumption that there are no parallel edges in a flow network.
+This important property is not mentioned in Cormen's proof.
 \Snippet{augment_flow_presv_cap}
 
-Most of the reference books in the field of algorithm provide pseudocodes instead of executable implementations. However, implementing an abstract presentation of 
-an algorithm involves many technical details that are never presented in the books.
-We also tried to fill in this gap.
-First, we present the Ford-Fulkerson method using the syntax of Autoref[cite] tool which is quite similar to pseudocode of the algorithm.
-Next, we refine the abstract representation into an executable implementation.
-For instance, the following Isabelel definition abstractly presents the Ford-Fulkerson method.
-% \Snippet{6.1-1_0}
-% \Snippet{6.1-1}
+In reference books, it is a convention to use pseudocodes for presenting the algorithms.
+The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm and leaves out the implementation details.
+We present the Ford-Fulkerson method using the syntax of Autoref[cite] tool which is quite similar to pseudocode of the algorithm:
+\Snippet{ford_fulkerson_algo}
+
+In the pseudocode of the Ford-Fulkerson method, the path found while evaluating the loop condition is later used for augmenting the flow inside the loop body.
+The best way for implementing such requirement using syntax of the refinement framework is moving the path lookup inside the loop body.
+To do so, we modified the while-loop by introducing a break flag.
+The flag is initially off, and is set if no augmenting path is found in the residual graph.
 

### 2016-03-02 formalization/evaluation/fofu-Java/fofu.java
Merger complete

@@ -1,0 +1,38 @@
+import lib.FlowNetwork;
+import lib.FordFulkerson;
+
+public class fofu {
+    // test client that reads input network, solves max flow, and prints results
+    public static void main(String[] args) throws Exception {
+    	if (args.length == 0) {
+    		System.out.println("Usage: <GRAPH-PATH>");
+    	}
+    	else {
+	        FlowNetwork G = new FlowNetwork(args[0]);
+	        int V_size = G.V();
+	        int E_size = G.E();
+	
+	        if (V_size > 0 && V_size <= 10000) {
+            // Warm up
+	          new FordFulkerson(G, 0, V_size - 1);
+	          
+	          G = new FlowNetwork(args[0]);
+	        
+	        	long startTime = System.currentTimeMillis();
+	        	
+	        	FordFulkerson maxflow = new FordFulkerson(G, 0, V_size - 1);
+	        
+	        	long endTime   = System.currentTimeMillis();
+	        
+	        	System.out.format("@@@ Execution Time: %d milliseconds\n", endTime - startTime);
+	        	System.out.format("\t[Input (V E): %d %d]\n", V_size, E_size);
+	        	System.out.format("\t=> Maximum flow value: %d\n", (int)maxflow.value());
+	        	System.out.format("\touter_c: %d\n\tinner_c: %d\n", maxflow.outer_c, maxflow.inner_c);
+	        }
+	        else {
+	        	System.out.println("Input Graph is invalid");	        	
+	        }
+    	}
+    }
+
+}

### 2016-03-02 paper/root.tex

@@
 	& \leq f(u, v) + c_f(u, v) &&  \text{(capacity constraint)} \\
-	& = f'(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
+	& = f(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
 	& = c (u, v).

### 2016-03-02 formalization/evaluation/data/data-gen.c
benchmarks

@@ -45,7 +45,7 @@ void dfs (int** graph, bool* visited, int start, int size, bool reverse) {
 /* This function produces a random network which has no self loops. Also, there
  * will be no parallel edges in the output network.*/
 void net_init (int** graph, int size, int try_count, int max_capacity) {
-	srand(time(NULL));
+	srand(341546752);
 
 	/* We assume the source is in index 0, and sink is in index size - 1. We put
 	 * random edges between any vertex except these two. and we connect them later */

### 2016-03-02 formalization/evaluation/fofu-C++/fofu.cpp

@@ -138,9 +138,9 @@ int main(int argc, char** argv) {
 			clock_t tStart = clock();
 			int maxFlow = edmondsKarp(0, V_size -1);
 
-			printf("@@@ Execution Time: %.4f milliseconds\n", ((double)(clock() - tStart)/CLOCKS_PER_SEC) * 1000);
-			printf("\t[Input (V E): %d %d]\n", V_size, E_size);
-			printf("\t=> Maximum flow value: %d\n", maxFlow);
+			printf("@@@time: %.0f ms\n", ((double)(clock() - tStart)/CLOCKS_PER_SEC) * 1000);
+			printf("[Input (V E): %d %d]\n", V_size, E_size);
+			printf("@@@max-flow: %d\n", maxFlow);
       printf("stat_outer_c: %d\n", stat_outer_c);
       printf("stat_counter_ga: %d\n", stat_counter_ga);
       printf("stat_counter_mget: %d\n", stat_counter_mget);

### 2016-03-02 formalization/evaluation/fofu-Java/fofu.java

@@ -2,36 +2,68 @@ import lib.FlowNetwork;
 import lib.FordFulkerson;
 
 public class fofu {
+    static FordFulkerson maxflow;
+
+    public static long doTest(FlowNetwork G) {
+      int V_size = G.V();
+      int E_size = G.E();
+
+      System.gc();
+
+      long startTime = System.currentTimeMillis();
+      maxflow = new FordFulkerson(G, 0, V_size - 1);
+      long endTime   = System.currentTimeMillis();
+      return (endTime - startTime);
+    }
+
+
     // test client that reads input network, solves max flow, and prints results
     public static void main(String[] args) throws Exception {
-    	if (args.length == 0) {
-    		System.out.println("Usage: <GRAPH-PATH>");
+    	if (args.length > 2 || args.length < 1) {
+    		System.out.println("Usage: [<WARMUP-PATH>] <GRAPH-PATH>");
     	}
     	else {
-	        FlowNetwork G = new FlowNetwork(args[0]);
-	        int V_size = G.V();
-	        int E_size = G.E();
-	
-	        if (V_size > 0 && V_size <= 10000) {
+          int i=0;
+          
+          if (args.length == 2) {
             // Warm up
-	          new FordFulkerson(G, 0, V_size - 1);
-	          
-	          G = new FlowNetwork(args[0]);
+            FlowNetwork G = new FlowNetwork(args[i]);
+            doTest(G);
+            doTest(G);
+            doTest(G);
+            doTest(G);
+            i++;
+          }
+
+	        FlowNetwork G = new FlowNetwork(args[i]);
+	        long time = doTest(G);
+          System.out.format("@@@time: %d ms\n", time);
+          System.out.format("[Input (V E): %d %d]\n", G.V(), G.E());
+          System.out.format("@@@max-flow: %d\n", (int)maxflow.value());
+          System.out.format("outer_c: %d\n\tinner_c: %d\n", maxflow.outer_c, maxflow.inner_c);
 	        
-	        	long startTime = System.currentTimeMillis();
-	        	
-	        	FordFulkerson maxflow = new FordFulkerson(G, 0, V_size - 1);
-	        
-	        	long endTime   = System.currentTimeMillis();
-	        
-	        	System.out.format("@@@ Execution Time: %d milliseconds\n", endTime - startTime);
-	        	System.out.format("\t[Input (V E): %d %d]\n", V_size, E_size);
[…]

### 2016-03-02 formalization/evaluation/makeall.sh

@@ -1,0 +1,10 @@
+#!/bin/bash
+
+platforms=`cat platforms`
+
+for p in $platforms; do
+  echo "Compiling $p"
+  cd "fofu-$p"
+  ./build
+  cd ..
+done

### 2016-03-02 formalization/evaluation/data/data-gen.c
Scripting benchmarking complete. Modified data generator to generate graphs of specified target density

@@ -44,7 +44,7 @@ void dfs (int** graph, bool* visited, int start, int size, bool reverse) {
 
 /* This function produces a random network which has no self loops. Also, there
  * will be no parallel edges in the output network.*/
-void net_init (int** graph, int size, int try_count, int max_capacity) {
+void net_init (int** graph, int size, int try_count, float ss_factor, int max_capacity) {
 	srand(341546752);
 
 	/* We assume the source is in index 0, and sink is in index size - 1. We put
@@ -61,10 +61,14 @@ void net_init (int** graph, int size, int try_count, int max_capacity) {
 
 	/* compute the number of vertices that are connected to source and sink, we
 	 *	have assumed this number is at most half of the size of  the vertex set */
-	int sout_count = 1 + (rand() % (size / 2));
-	int tin_count = 1 + (rand() % (size / 2));
+// 	int sout_count = 1 + (rand() % (size / 2));
+// 	int tin_count = 1 + (rand() % (size / 2));
 
-	// onnecting source to some of the vertices
+  int sout_count = (int)(ss_factor * size);
+  int tin_count = (int)(ss_factor * size);
+  
+  
+	// connecting source to some of the vertices
 	for (i = 0; i < sout_count; i++) {
 		int u = 1 + (rand() % (size - 2));
 		int c = 1 + (rand() % max_capacity);
@@ -100,14 +104,17 @@ void net_clean (int** graph, int size, bool* acc_nodes) {
 }
 
 int main (int argc, char** argv) {
-	if (argc < 2) {
-		printf("Usage: <V_SIZE>\n");
+	if (argc != 5) {
+		printf("Usage: <V_SIZE> <target-density> <name-suffix> <correction>\n");
 	}
 	else {
 		int v_count = atoi(argv[1]);
+    float target_density = atof(argv[2]);
+    char *name_suffix = argv[3];
+    float correction = atof(argv[4]);
 
-		if (v_count > 0) {
-			int e_tries = v_count * (v_count / 5);
+		if (v_count > 0 && 0<target_density && target_density <= 1) {
+			int e_tries = (int)(target_density * v_count * (v_count - 1) * correction);
 			int max_c = v_count;
 
 			char buffer[20];
@@ -117,7 +124,7 @@ int main (int argc, char** argv) {
 			bool* acc_nodes = calloc(v_count, sizeof(bool));
 			int* index_chg = calloc(v_count, sizeof(int));
 
[…]

### 2016-03-02 formalization/evaluation/makeall.sh

@@ -1,5 +1,9 @@
 #!/bin/bash
 
+cd data
+./build
+cd ..
+
 platforms=`cat platforms`
 
 for p in $platforms; do

### 2016-03-02 paper/root.tex

@@
 \section{Introduction}
-The Ford-Fulkerson algorithm [cite] is one of the important results of the graph theory and is used to find a solution to the maximum flow problem in flow networks.
+The Ford-Fulkerson algorithm~\cite{FF56} is one of the important results of the graph theory and is used to find a solution to the maximum flow problem in flow networks.
 Many important mathematical problems like the maximum-bipartite-matching problem, the edge-disjoint-paths problem, the circulation-demand problem, and many other scheduling and resource allocating problems can be reduced to the maximum flow problem.
@@
 Despite its importance, no formalization of the Ford-Fulkerson algorithm has been developed in modern proof assistants.
-The only similar work that the authors are aware of is formalization of the Ford-Fulkerson algorithm in Mizar [cite].
+The only similar work that the authors are aware of is formalization of the Ford-Fulkerson algorithm in Mizar~\cite{Lee05}.
 This formalization defines and proves correctness of the algorithm at the level of graph manipulations without providing concrete implementation of the algorithm.
@@
 
-This paper present a formalization of the correctness proof of the Ford-Fulkerson algorithm in Isabelle/HOL.
-Our proof is based on the informal proof of the algorithm which is presented in the book "Introduction to algorithms" [cite].
+This paper present a formalization of the correctness proof of the Ford-Fulkerson algorithm in Isabelle/HOL~\cite{NPW02}.
+Our proof is based on the informal proof of the algorithm which is presented in the book \emph{Introduction to algorithms}~\cite{CLRS09}.
 Due to practical importance of the algorithm, we also present a verified implementation of the algorithm.
@@
 
-One solution for the aforementioned problem is refinement [cite].
+One solution for the aforementioned problem is refinement~\cite{Morr87}.
 In order to generate executable code of an algorithm using refinement, we first formulate the algorithm on an abstract level.
@@
 There are several approaches to data refinement in Isabelle/HOL.
-We will be using the Autoref tool [cite], which has been used for proving more complex results such as formalized implementation of Hopcroft’s DFA-minimization algorithm [cite].
+We will be using the Autoref tool~\cite{La13}, which has been used for proving more complex results such as formalized implementation of Hopcroft’s DFA-minimization algorithm~\cite{LaTu12}.
 Given an algorithm phrased over abstract concepts like sets and maps, it automatically synthesizes a concrete, executable algorithm and the corresponding refinement theorem.
-To make it applicable for the development of actual algorithms, Autoref is integrated with the Isabelle Refinement Framework [cite] and the Isabelle Collection Framework [cite].
+To make it applicable for the development of actual algorithms, Autoref is integrated with the Isabelle Refinement Framework~\cite{La12} and the Isabelle Collection Framework~\cite{LL10}.
 
@@
 Isar makes formal proofs readable even for those who are not familiar with Isabelle concepts.
-We have used such a functionality to present our formal proof of the Ford-Fulkerson method which resembels the informal proof described by Cormen et al.
-[cite].
+We have used such a functionality to present our formal proof of the Ford-Fulkerson method which resembels the informal proof described by Cormen et al.~\cite{CLRS09}.
 
@@
 The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm and leaves out the implementation details.
-We present the Ford-Fulkerson method using the syntax of Autoref[cite] tool which is quite similar to pseudocode of the algorithm:
+We present the Ford-Fulkerson method using the syntax of Autoref~\cite{La13} tool which is quite similar to pseudocode of the algorithm:
 \Snippet{ford_fulkerson_algo}

### 2016-03-02 paper/root.tex
modifed abstract intro

@@
 We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
-The formal proofs are done in Isabelle/HOL. They closely follow a standard textbook proof and are relatively easy 
-to read, even without being an Isabelle expert.
+Our formal proof closely follows a standard textbook proof, so it is easy to read even for those who 
+are not expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
 
-We then use stepwise refinement to obtain the Edmonds-Karp algorithm and prove its upper complexity bound of $O(|V||E|^2)$.
-Further refinement yields a verified implementation, which is almost as fast as an unverified  
-reference implementation in Java.
+We then use stepwise refinement to obtain the Edmonds-Karp algorithm and we formally prove its termination.
+Further refinement yields a verified implementation, which is almost as fast as an unverified reference implementation in Java.
+
+%We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network. 
+%The formal proofs are done in Isabelle/HOL. They closely follow a standard textbook proof and are relatively easy 
+%to read, even without being an Isabelle expert.
+
+%We then use stepwise refinement to obtain the Edmonds-Karp algorithm and prove its upper complexity bound of $O(|V||E|^2)$.
+%Further refinement yields a verified implementation, which is almost as fast as an unverified  
+%reference implementation in Java.
 
@@
 
-This paper present a formalization of the correctness proof of the Ford-Fulkerson algorithm in Isabelle/HOL~\cite{NPW02}.
-Our proof is based on the informal proof of the algorithm which is presented in the book \emph{Introduction to algorithms}~\cite{CLRS09}.
-Due to practical importance of the algorithm, we also present a verified implementation of the algorithm.
+This paper present a formalization of the correctness proof of the Ford-Fulkerson method in Isabelle/HOL~\cite{NPW02}.
+Our formalization is based on the informal proof presented in the book \emph{Introduction to algorithms}~\cite{CLRS09}.
+We then provide a verified implementation of the Edmons-Karp's algorithm~\cite{EK72}.
+This algorithm instantiates the Ford-Fulkerson method with breadth-first-search and provides a polynomial algorithm for solving the maximum flow problem in generic case--- maximum flow on networks with real edge capacities.
+
 Isabelle/HOL provides some automation for generating the code corresponding to a verified algorithm, however, in order to use the code generating features, one needs to do the formalization with executability in mind.

### 2016-03-02 paper/root.tex
modified FoFu

@@
 To make the formalization simple, some additional assumptions are added to the definition as following: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no edge $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$.
-Although these assumptions seems restrictive, any network may be transformed such that it satisfies all the aforementioned properties.
+Although these assumptions seems restrictive, any network may be transformed such that it satisfies all the aforementioned properties~\cite{CLRS09}.
 
 An $s$-$t$ flow on a weighted digraph is a function $\colon V \times V \rightarrow \mathbb{R}$ which satisfies the following conditions.
-1) \textbf{Capacity constraint}: the flow of each edge is a non-negative value which is smaller or equal to the edge capacity; 2) \textbf{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over incoming edges of the vertex is equal to the sum of flows over outgoing edges of the vertex.
+1) \emph{Capacity constraint}: the flow of each edge is a non-negative value which is smaller or equal to the edge capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over incoming edges of the vertex is equal to the sum of flows over outgoing edges of the vertex.
 
@@
 In order to perform these operations, the Ford-Fulkerson algorithm defines the residual graph of the flow network.
-Intuitively, the residual graph corresponding to a flow in a given flow network, has the same vertices as the flow network.
-It also and consists of edges with capacities representing how flow can be changed along the edges of the flow network.
+Intuitively, the residual graph corresponding to a flow in a given flow network, has the same vertices as the flow network and consists of edges with capacities representing how flow can be changed along the edges of the flow network.
 
@@
 This is due to the fact that pushing flow on the reverse edge on the residual graph signifies decreasing the flow in the original network.
-In each iteration of the algorithm we replace the current flow with the result of the augment function, and the algorithm terminates if no more augmenting paths can be found in the residual graph.
-According to the Ford-Fulkerson theorem, the flow computed in this states is indeed the maximum flow in the flow network.
-The theorem states that following statements are equivalent:
+
+In each iteration of the algorithm we replace the current flow with the result of the augment function--- another flow on the network but with greater value.
+The algorithm terminates if no more augmenting paths can be found in the residual graph.
+According to the Ford-Fulkerson theorem, the flow computed after termination is indeed the maximum flow in the flow network.
+According to the theorem, the following statements are equivalent:
 
@@
 The termination of the Ford-Fulkerson algorithm depends on how the augmenting path is found.
-For the networks with irrational edge capacities the algorithm might even fail to terminate.
+For the networks with irrational edge capacities the algorithm might even fail to terminate~\cite{Zwick95}.
 In practice, the maximum-flow problem often arises with integral capacities.
@@
 The augmentings paths that are computed using BFS are also shortest paths connecting source to the sink in the residual graph.
-Edmond and Karp showed that in the Ford-Fulkerson algorithm, if each augmenting path is the shortest one, the length of the augmenting paths is non-decreasing and it always terminates.
-Which finally gave the polynomial algorithm for solving the maximum flow problem in generic case of real edge capacities.
+Edmond and Karp showed that in the Ford-Fulkerson algorithm, if each augmenting path is the shortest one, the length of the augmenting paths is non-decreasing and the algorithm always terminates.
+Using this statement they could finally present a polynomial algorithm for solving the maximum flow problem in generic case of real edge capacities.
 

### 2016-03-02 paper/root.tex
eval-section, started concl

--- 2016-03-02 paper/root.tex
@@
 
-  \subsection{Manipulating Residual Graphs Directly}    
+  \subsection{Manipulating Residual Graphs Directly}\label{sec:impl_res_graph}
   In a next step, we observe that the algorithm is phrased in terms of a flow which is updated until it is maximal. However,
@@
 
-  \subsection{Using Efficient Data Structures}    
+  \subsection{Using Efficient Data Structures}\label{sec:impl_datastructures}
   In a final step, we have to chose efficient data structures for the algorithm. The objects that the algorithm works with are
@@
 \section{Benchmarking}
-  What implementations did we compare: Authors? Where do implementations come from? 
-  [We must convince the reader that we did not intentionally chose bad implementations to compare us against!]
+  We have compared the running time of our algorithm against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}.
+  We have done the comparison on randomly generated sparse and dense networks, the sparse networks having a \emph{density} ($\frac{|E|}{|V|(|V| - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
+For sparse networks, we varied the number of nodes between $1000$ and $5000$, for dense networks between $1000$ and $1400$.
+  The results are shown in Table~\ref{tab:benchmark}. 
+  \begin{wrapfigure}{o}{.4\textwidth}
+  \vspace{-20pt}
+  \begin{center}
+  \begin{tabular}{|c|c|c|}
+  \hline
+  Name        & SML       & Java   \\
+  \hline\hline
+  1000-sparse & 204       & 15     \\
+  2000-sparse & 1430      & 231    \\
+  3000-sparse & 4928      & 828    \\
+  4000-sparse & 9299      & 2145   \\
+  5000-sparse & 23625     & 4868   \\
+  1000-dense  & 6575      & 4011   \\
+  1100-dense  & 8067      & 5487   \\
+  1200-dense  & 10892     & 7011   \\
+  1300-dense  & 14070     & 9815   \\
+  1400-dense  & 17317     & 12156  \\
+  \hline
+  \end{tabular}
+  \end{center}
+  \vspace{-15pt}
+  \caption{Dijkstra benchmark}\label{tab:benchmark}
+  \vspace{-20pt}
+  \end{wrapfigure}
 
-  Comparsion of algorithms (Modified data structure: Flow vs Flow-like vs ResGraph)
-    Computing successors in BFS: Filter by visited set first.
-      -> Technically more challenging, when computing on flow: 1 access first vs 2 accesses first. On resGraph: No advantage in memory accesses (However, plain array access vs. matrix access)
-    Computing bottleneck during BFS. Saves one iteration over the path, at the cost of one extra memory access per discovered node.
-    For our benchmarks, we observed that the BFS discovers considerably more nodes than the length of the ultimately returned path, 
-    and the overall code was faster with the extra iteration for bottleneck computation.
+  We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of 6, while for dense graphs, our implementation comes quite close 
+  to the Java implementation. We have no full explanation for this phenomenon. Note that the Java implementation operates on flows, while our implementation 
+  operates on residual graphs (\cf Section~\ref{sec:impl_res_graph}). Moreover, the Java implementation does not store the augmenting 
+  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_datastructures}).
+
+  Finally note that a carefully optimized C++ implementation of the algorithm is only slightly faster than the Java implementation for sparse graphs,
+  but roughly one order of magnitude faster for dense graphs. We leave it to future work to investigate this issue, and conclude that we were able to produce
+  a reasonably fast verified implementation, without doing any low-level optimizations, but relying on the code generated by the Isabelle/HOL code generator.
+
+% 
+%   What implementations did we compare: Authors? Where do implementations come from? 
+%   [We must convince the reader that we did not intentionally chose bad implementations to compare us against!]
+% 
+%   Comparsion of algorithms (Modified data structure: Flow vs Flow-like vs ResGraph)
+%     Computing successors in BFS: Filter by visited set first.
+%       -> Technically more challenging, when computing on flow: 1 access first vs 2 accesses first. On resGraph: No advantage in memory accesses (However, plain array access vs. matrix access)
+%     Computing bottleneck during BFS. Saves one iteration over the path, at the cost of one extra memory access per discovered node.
+%     For our benchmarks, we observed that the BFS discovers considerably more nodes than the length of the ultimately returned path, 
+%     and the overall code was faster with the extra iteration for bottleneck computation.
 
 \section{Conclusion}
+  In this paper, we have presented a verification of the Ford-Fulkerson method, using a stepwise refinement approach.
+  Starting with a proof of the Ford-Fulkerson theorem, we have verified the generic Ford-Fulkerson method, then 
+  specialized it to the Edmonds-Karp algorithm and proved the upper bound $O(|E||V|)$ for the number of outer loop iterations.
+  We then used several refinement steps to derive an efficiently executable implementation of the algorithm, 
+  involving a verified breadth first search to obtain shortest augmenting paths. 
+  Finally, we added a verified algorithm to check whether the input is a valid network, and generated executable code in SML.
+  Our implementation compares well to a reference implementation in Java.
+
+  Our formalization has combined several techniques to achieve an elegant and accessible formalization: The Isar-proof language~\cite{}
+  allows us to write formal proofs that closely follow informal proofs found in textbooks. This way, we have created a completely rigorous, but 
+  still accessible, proof of the Ford-Fulkerson theorem. The Isabelle Refinement Framework allows us to first formalize an algorithm abstractly, 
+  and then refine it, in a modular fashion, towards an efficient implementation. This allowed us to present the Ford-Fulkerson method on a level 
+  of abstraction that closely resembles pseudocode presentations found in textbooks, and then formally link this presentation to an efficient
+  implementation. Moreover, modularity of refinement allowed us to develop the breadth first search algorithm independently, and later link it to the 
+  main algorithm. The BFS algorithm can be reused as building block for other algorithms. Also the data structures are re-usable, and in fact, the 
+  only new data structure we had to implement where the adjacency matrices. The other data structures where readily available in the growing library 
+  of the Sepref tool~\cite{La15} and the Isabelle Collection Framework~\cite{LaTu12}.
+
+  \subsection{Related Work}
+  We are only aware of one other formalization of the Ford-Fulkerson method conducted in Mizar~\cite{}. 
+  It formalizes a fixed instance of the Ford Fulkerson method on a rather abstract level. No code has been generated, 
+  nor does it contain a complexity analysis.
+
+  Apart from our own work~\cite{La14,NoLa12}, there are several other verifications of graph algorithms, using different techniques. 
+  Top-Down
+  Bottom-Up: Noschinski verifies a checker for (non-) planarity certificates using a bottom-up approach. Starting at a C implementation,
+  the AutoCorres tool generates a monadic representation of the program in Isabelle. 
+  XXX, ctd here
+
+  Unfortunately, it
+
   ... and related work

### 2016-03-02 paper/root.tex
draft of conclusion

@@
 We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
-Our formal proof closely follows a standard textbook proof, so it is easy to read even for those who 
-are not expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
+Our formal proof closely follows a standard textbook proof, and is easy to read even without being
+an expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
+% 
+% 
+% so it is easy to read even for those who 
+% are not expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
 
-We then use stepwise refinement to obtain the Edmonds-Karp algorithm and we formally prove its termination.
-Further refinement yields a verified implementation, which is almost as fast as an unverified reference implementation in Java.
+We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
+Further refinement yields a verified implementation, which compares well to an unverified reference implementation in Java.
 
@@
 
-One solution for the aforementioned problem is refinement~\cite{Morr87}.
+One solution for the aforementioned problem is refinement~\cite{Wirth71}.
 In order to generate executable code of an algorithm using refinement, we first formulate the algorithm on an abstract level.
@@
 There are several approaches to data refinement in Isabelle/HOL.
-We will be using the Autoref tool~\cite{La13}, which has been used for proving more complex results such as formalized implementation of Hopcroft’s DFA-minimization algorithm~\cite{LaTu12}.
+We will be using the Isabelle Refinement Framework~\cite{LaTu12,La12}, which has been used for proving more complex results such as formalized implementation of Hopcroft’s DFA-minimization algorithm~\cite{LaTu12}.
 Given an algorithm phrased over abstract concepts like sets and maps, it automatically synthesizes a concrete, executable algorithm and the corresponding refinement theorem.
@@
 
+  The formalization consists of roughly 8000 lines of proof text, where the graph theory up to the Ford-Fulkerson theorem required 2500 lines.
+  The abstract Edmonds-Karp algorithm and its complexity analysis contributes 900 lines, and its implementation (including BFS) another 1500 lines.
+  The remaining lines are contributed by the network checker and some auxiliary theories.
+The development of the theories required roughly 3 men month, a significant amount of this time going into a first, purely functional version of the implementation, which was later dropped in favor of the significantly faster imperative version.
+  TODO: Reza, is this estimate OK?
+
   \subsection{Related Work}
@@
   Bottom-Up: Noschinski verifies a checker for (non-) planarity certificates using a bottom-up approach. Starting at a C implementation,
-  the AutoCorres tool generates a monadic representation of the program in Isabelle. 
-  XXX, ctd here
+  the AutoCorres tool~\cite{AutoCorres. What's the standard paper on it?} generates a monadic representation of the program in Isabelle. Further abstractions are applied
+  to this representation to hide low-level details like pointer manipulations and fixed size integers~\cite{AutoCorres-Mind-the-Gap}. Finally, a verification condition
+  generator is used to prove the abstracted program correct.
+Note that approach is the opposite direction of ours: While they start at a concrete version of the algorithm and use abstraction steps to eliminate implementation details, we start at an abstract version, and use concretization steps to introduce implementation details.
 
-  Unfortunately, it
+  Chargu\'eraud also uses a bottom-up approach~\cite{char11} to verify imperative programs written in a subset of OCaml, amongst them a version of Dijkstra's algorithm:
+  A verification condition generator generates a \emph{characteristic formula}, which reflects the semantics of the program in the logic of the proof assistant~\cite{}.
 
-  ... and related work
-    Mizar-Formalization: What did they do.
+  \subsection{Future Work}
+  Future work includes the formalization of more advanced maximum flow algorithms, like Dinic's algorithm~\cite{Di06} or push-relabel algorithms~\cite{GoTa88}.
+  Both, formalizing the abstract theory, and developing efficient implementations will pose some challenges.
 
-    History of the formalization: 
-      Impl: From first impl that was hardly able to compute a 100 nodes graph, to current efficient one.
-
-    Future Work: Dinic. (Actually, our abstract scheme [almost] covers Dinic's algorithm!)
-
-Contributions
-
-  Formal proof of mincut maxflow
-  fofo-scheme
-  inst to edmonds karp
-  complexity analysis of edmonds karp
-  refinement down to executable code. Roughly 2.5 times slower than Java. (What about OCaml)
-    + NetCheck
-
-What shines (its a Pearl)
-  Min-Cut Max-Flow: Textboook like formal reasoning: Comprehensible proof, BUT machine checked
-    (Present one (carefully worked) example in paperI. We could use lemma \isai{augment_flow_presv_cap})
-
-  Refinement based approach: Fofu-Scheme, instantiation to EdsKa. 
-    +1: Abstract Algo looks almost like pseudo-code you would expect in textbook.
-    +2: Fofu-Scheme proved correct for all aug-path finders. EdsKa is instantiation of it.
-    +3: Modularity: Fofu-scheme and pathfinder developed+proved independently of each other.
-
-  Down to executable code, plugging in efficient data structures.
-
-Some minor contributions:
-  Reusable BFS algorithm
-  Imperative matrix data structure (really minor).
+%   
+%   can be performed in two directions. First of all, the abstract theory can be extended to also support more advanced 
+%   maximum flow algorithms like push-relabel algorithms~\cite{GoTa88}
+%   
+%   
+%   
+%   What do the Coq-Guys have: CFML? Others?
+%   
+%   
+%   XXX, ctd here
+%   
+%   Unfortunately, it
+% 
+%   ... and related work
+%     Mizar-Formalization: What did they do.
+%     
+%     History of the formalization: 
+%       Impl: From first impl that was hardly able to compute a 100 nodes graph, to current efficient one.
+%     
+%     Future Work: Dinic. (Actually, our abstract scheme [almost] covers Dinic's algorithm!)
+%     
+% 
+% 
+%   
+% Contributions
+% 
+%   Formal proof of mincut maxflow
+%   fofo-scheme
+%   inst to edmonds karp
+%   complexity analysis of edmonds karp
+%   refinement down to executable code. Roughly 2.5 times slower than Java. (What about OCaml)
+%     + NetCheck
+%     
+% What shines (its a Pearl)
+%   Min-Cut Max-Flow: Textboook like formal reasoning: Comprehensible proof, BUT machine checked
+%     (Present one (carefully worked) example in paperI. We could use lemma \isai{augment_flow_presv_cap})
+%     
+%   Refinement based approach: Fofu-Scheme, instantiation to EdsKa. 
+%     +1: Abstract Algo looks almost like pseudo-code you would expect in textbook.
+%     +2: Fofu-Scheme proved correct for all aug-path finders. EdsKa is instantiation of it.
+%     +3: Modularity: Fofu-scheme and pathfinder developed+proved independently of each other.
+%   
+%   Down to executable code, plugging in efficient data structures.
+%   
+% Some minor contributions:
+%   Reusable BFS algorithm
+%   Imperative matrix data structure (really minor).
+%   
+%   
+%   
+%     
+% 
+% 
+%
 

### 2016-03-02 paper/root.tex
Added TODOs to latex file

@@
 
-This paper present a formalization of the correctness proof of the Ford-Fulkerson method in Isabelle/HOL~\cite{NPW02}.
+(TODO: Here I have mentioned the Edmonds-Karp for the first time.) This paper present a formalization of the correctness proof of the Ford-Fulkerson method in Isabelle/HOL~\cite{NPW02}.
 Our formalization is based on the informal proof presented in the book \emph{Introduction to algorithms}~\cite{CLRS09}.
@@
 
-One solution for the aforementioned problem is refinement~\cite{Wirth71}.
+(TODO: I am not sure about the level of details that I describe refinement in introduction, co-relate this with the refinement section) One solution for the aforementioned problem is refinement~\cite{Wirth71}.
 In order to generate executable code of an algorithm using refinement, we first formulate the algorithm on an abstract level.
@@
 
-There are several approaches to data refinement in Isabelle/HOL.
+(TODO: check the citation order and description) There are several approaches to data refinement in Isabelle/HOL.
 We will be using the Isabelle Refinement Framework~\cite{LaTu12,La12}, which has been used for proving more complex results such as formalized implementation of Hopcroft’s DFA-minimization algorithm~\cite{LaTu12}.
@@
 
-There are two kinds of edges in the residual graph: 1) those of the residual graph that present the edges of the flow network with left-over capacities; and 2) those of the residual graph that do not occur in the flow network, and indicate the possibility for pushing the flow backward.
+(TODO: there is a shorter informal description available as comment.
+Adapt that one if required) There are two kinds of edges in the residual graph: 1) those of the residual graph that present the edges of the flow network with left-over capacities; and 2) those of the residual graph that do not occur in the flow network, and indicate the possibility for pushing the flow backward.
 Formally speaking, consider a flow $f$ on an $s$-$t$ network $G = (V, E)$ with the capacity function $c$.
@@
 
-The Ford-Fulkerson algorithm is sometimes called a method instead of an algorithm because it does not fully specify the procedure for finding the augmenting path in the residual graph.
+(TODO: Here I mention the Edmonds-Karp algorithm and I mention its running time, co-relate this with definition in later sections) The Ford-Fulkerson algorithm is sometimes called a method instead of an algorithm because it does not fully specify the procedure for finding the augmenting path in the residual graph.
 There are several implementations with different execution times.
@@
 
-Informal proofs focus on the relevant thoughts by leaving out technical details.
+(TODO: not sure about repeating citation of Cormen) Informal proofs focus on the relevant thoughts by leaving out technical details.
 However, in a formal proof each step has to be precisely specified as a valid application of some inference rule which makes formal proofs often lengthy and hard to follow.
@@
 
-  We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of 6, while for dense graphs, our implementation comes quite close 
+  (TODO: Table's caption is incorrect, I think we have enough space to print the whole table with SML, Java as rows and graph size as columns.
+Also, single lines for separating table looks nicer.
+I have seen it in your Hopcroft's paper) We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of 6, while for dense graphs, our implementation comes quite close
   to the Java implementation. We have no full explanation for this phenomenon. Note that the Java implementation operates on flows, while our implementation 

### 2016-03-02 paper/root.tex
Added TODOs to latex file

@@
 The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm and leaves out the implementation details.
-We present the Ford-Fulkerson method using the syntax of Autoref~\cite{La13} tool which is quite similar to pseudocode of the algorithm:
+We present the Ford-Fulkerson method using the syntax of (TODO: here I mention autoref again, I was not aware that we are not using it anymore.
+So please correct all the references that I made to autoref tool) Autoref~\cite{La13} tool which is quite similar to pseudocode of the algorithm:
 \Snippet{ford_fulkerson_algo}

### 2016-03-03 paper/root.tex
diagram for benchmark

@@
 
+\usepackage{etex}
 \pagestyle{plain} % turn on page numbers
@@
 \usepackage{wrapfig}
-
 \usepackage{multirow}
+\usepackage{pgfplots}
 
@@
   We have compared the running time of our algorithm against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}.
+
+\begin{figure}[!h]
+\centering    
+\begin{tikzpicture}[thick,scale=1, every node/.style={scale=1}] %change the scales if you like to reduce the size
+	\begin{axis}[
+		%title={Benchmark of the Edmonds-Karp algorithm},
+	    axis x line*= bottom,
+	    axis y line*= left,
+	    xmode = log,
+	    ymode = log,
+	    xlabel = Number of vertices in the network,
+	    ylabel = Execution time in milliseconds,
+    	smooth,
+	    cycle list name = black white,
+	    legend style = { 
+	    	at={(0.59,0.4)}, 
+    	    anchor=north west,
+	        draw=black, 
+	        fill= white,
+        	align=left
+    	}
+	]
+
+	%start data for plotting
+		\addplot table {
+			1000 204
+			2000 1430
+			3000 4928
+			4000 9299
+			5000 23625
+		}; \addlegendentry{SML, Sparse};
+
+		\addplot table {
+			1000 15
+			2000 231
+			3000 828
+			4000 2145
+			5000 4868
+		}; \addlegendentry{Java, Sparse};
+
+		\addplot table {
+			1000 6575
+			1100 8067
+			1200 10892
+			1300 14070
+			1400 17317
+		}; \addlegendentry{SML, Dense};
+
+		\addplot table {
+			1000 4011
+			1100 5487
+			1200 7011
+			1300 9815
+			1400 12156
+		}; \addlegendentry{Java, Dense};
+	%end data for plotting
+
+	\end{axis}
+\end{tikzpicture}
+%\caption{Edmonds-Karp benchmark}\label{fig:benchmark}
+\end{figure}
+
   We have done the comparison on randomly generated sparse and dense networks, the sparse networks having a \emph{density} ($\frac{|E|}{|V|(|V| - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.

### 2016-03-03 paper/root.tex
fixed listing fonts
modified diagram

@@
 \input{lstisabelle}
+\lstset{basicstyle=\small\ttfamily\slshape,breaklines=true}
 \lstset{captionpos=b}
@@
 	    ylabel = Execution time in milliseconds,
+		xticklabels={$0$, $1000$,$2000$,$3000$,$4000$},
     	smooth,

### 2016-03-03 paper/root.tex
Rephrasing + improvement up to and excluding benchmark section

--- 2016-03-03 paper/root.tex
@@
 \section{Introduction}
-The Ford-Fulkerson algorithm~\cite{FF56} is one of the important results of the graph theory and is used to find a solution to the maximum flow problem in flow networks.
-Many important mathematical problems like the maximum-bipartite-matching problem, the edge-disjoint-paths problem, the circulation-demand problem, and many other scheduling and resource allocating problems can be reduced to the maximum flow problem.
-Hence, the Ford-Fulkerson algorithm has major application in the field of mathematical optimization.
+The maximum flow problem is an important problem in graph theory, as many other problems like the maximum-bipartite-matching problem, the edge-disjoint-paths problem, the circulation-demand problem, and various scheduling and resource allocating problems can be reduced to it.
+The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
+which solves the maximum flow problem in polynomial time for the general case of real valued capacities.
 
-Despite its importance, no formalization of the Ford-Fulkerson algorithm has been developed in modern proof assistants.
+In this paper, we present a verified efficient implementation of the Edmonds-Karp algorithm and its polynomial complexity bound.
+The formalization is conducted entirely in the Isabelle/HOL proof assistant~\cite{NPW02}. 
+Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification into an abstract proof of the Ford-Fulkerson method,
+its instantiation to Edmonds-Karp algorithm, and finally its efficient implementation.
+The abstract parts of our verification closely follow the standard textbook presentation of Cormen et al.~\cite{CLRS09}.
+Being developed in the Isar~\cite{Wenzel99} proof language, our proofs are accessible even to non-Isabelle experts.
+
+The rest of this paper is structured as follows: In Section~\ref{sec:background} we give a short informal introduction to the Ford-Fulkerson method.
+In Section~\ref{sec:abs-formalization}, we report on our formalization of the abstract method. 
+Section~\ref{sec:refinement} gives a brief overview of the Isabelle Refinement Framework~\cite{LaTu12,La12}, which supports stepwise refinement based algorithm development in Isabelle/HOL.
+In Section~\ref{sec:edka}, we report on our instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm and the proof of its complexity.
+Section~\ref{sec:executable} reports on the further refinement steps required to yield an efficient implementation. Section~\ref{sec:benchmark} reports on 
+benchmarking our implementation against a reference implementation of the Edmonds-Karp algorithm from Sedgewick et al.~\cite{SeWa11}. 
+Finally, Section~\ref{sec:concl} gives a conclusion and discusses related and future work.
+
+% Despite its importance, no formalization of the Ford-Fulkerson algorithm has been developed in modern proof assistants.
 The only similar work that the authors are aware of is formalization of the Ford-Fulkerson algorithm in Mizar~\cite{Lee05}.
@@
 
-(TODO: Here I have mentioned the Edmonds-Karp for the first time.) This paper present a formalization of the correctness proof of the Ford-Fulkerson method in Isabelle/HOL~\cite{NPW02}.
-Our formalization is based on the informal proof presented in the book \emph{Introduction to algorithms}~\cite{CLRS09}.
-We then provide a verified implementation of the Edmons-Karp's algorithm~\cite{EK72}.
-This algorithm instantiates the Ford-Fulkerson method with breadth-first-search and provides a polynomial algorithm for solving the maximum flow problem in generic case--- maximum flow on networks with real edge capacities.
-
-Isabelle/HOL provides some automation for generating the code corresponding to a verified algorithm, however, in order to use the code generating features, one needs to do the formalization with executability in mind.
-Being limited only to executable concepts in the formalization has the disadvantage of cluttering the proofs with implementation details.
-Such approach makes the proofs more complicated, and may even render proofs of medium complex algorithms unmanageable.
-
-(TODO: I am not sure about the level of details that I describe refinement in introduction, co-relate this with the refinement section) One solution for the aforementioned problem is refinement~\cite{Wirth71}.
-In order to generate executable code of an algorithm using refinement, we first formulate the algorithm on an abstract level.
-The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm.
-Next, we refine the abstract definition of the algorithm towards an executable implementation in possibly multiple refinement steps.
-During each step, we only need to prove the correctness of the implementation of a particular abstract concept.
-Hence, we can be sure about the correctness of the resulting executable program as each refinement step preserves the correctness.
-
-(TODO: check the citation order and description) There are several approaches to data refinement in Isabelle/HOL.
-We will be using the Isabelle Refinement Framework~\cite{LaTu12,La12}, which has been used for proving more complex results such as formalized implementation of Hopcroft’s DFA-minimization algorithm~\cite{LaTu12}.
-Given an algorithm phrased over abstract concepts like sets and maps, it automatically synthesizes a concrete, executable algorithm and the corresponding refinement theorem.
-To make it applicable for the development of actual algorithms, Autoref is integrated with the Isabelle Refinement Framework~\cite{La12} and the Isabelle Collection Framework~\cite{LL10}.
-
-\section{A short background on the Ford-Fulkerson method}
+\section{A short background on the Ford-Fulkerson method}\label{sec:background}
 % \begin{comment}
@@
 %   \end{comment}
-A flow network is a weighted digraph where the set of vertices (nodes) $V$ is finite, and each edge $(u, v)$ has a real capacity $c (u, v) \geq 0$.
-There are two distinct vertices $s$ (source) and $t$ (sink) in the flow network.
-In this case the network is called an $s$-$t$ network.
-To make the formalization simple, some additional assumptions are added to the definition as following: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no edge $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$.
-Although these assumptions seems restrictive, any network may be transformed such that it satisfies all the aforementioned properties~\cite{CLRS09}.
 
-An $s$-$t$ flow on a weighted digraph is a function $\colon V \times V \rightarrow \mathbb{R}$ which satisfies the following conditions.
-1) \emph{Capacity constraint}: the flow of each edge is a non-negative value which is smaller or equal to the edge capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over incoming edges of the vertex is equal to the sum of flows over outgoing edges of the vertex.
+In this section, we give a short introduction to the Ford-Fulkerson method, closely following the presentation in~\cite{CLRS09}.
 
-The value of an $s$-$t$ flow f is denoted by $|f|$, and is computed by subtracting the sum of the flows over all incoming edges of $s$ from the sum of the flows over all outgoing edges of $s$.
-Given an $s$-$t$ network $G$, the maximum flow problem involves finding a flow $f$ in $G$ such that the value of $f$ is greater or equal to the value of any other $s$-$t$ flow in $G$~\footnote{Note that the value of the flow in our model of a flow network equals sum of flows over edges going out of $s$.
-Because the model prohibits edges coming into the source.}.
+A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real valued capacity $c(u,v)>0$.
+Moreover, their are two distinct vertices $s,t\in E$ which are called \emph{source} and \emph{sink}.
+To simplify reasoning about flow networks, it is common to make some additional assumptions: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no edge $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$.
+Although these assumptions seem restrictive, any network may be transformed such that it satisfies all the aforementioned properties~\cite{CLRS09}.
 
-One of the important results in mathematical optimization is the correlation between flows and cuts in flow networks.
-An $s$-$t$ cut in a flow network with source $s$ and sink $t$ is a partition of vertices that puts $s$ and $t$ in different subsets.
-Then the capacity of a cut is defined as sum of the edge capacities over all edges going from the source's side to the sink's side.
-Consider a valid flow in a flow network $G$.
-For any cut in $G$, the flow that crosses between the two subsets of the cut cannot exceed the capacity of the cut.
-So cuts define an upper bound for any flow in the flow network.
+An flow $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges.
+
+The value of flow $f$ is denoted by $|f|$, and defined to be the sum over the flows of all outgoing edges of the source\footnote{Note that we assumed that the source has no incoming edges.
+In a more general setting, the value of a flow is the sum over all flows leaving $s$ minus the sum over all incoming flows to $s$.}.
+Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.
+
+An important result is the relation between flows and cuts in a network. A \emph{cut} in a network is a partition of the vertices into two sets, such that one set contains 
+the source and the other set contains the sink. The capacity of a cut is the sum of the capacities of all edges going from the source's side to the sink's side of the cut.
+It is easy to see that the value of any flow cannot exceed the capacity of any cut, as all flow from the source must ultimately reach the sink, and thus go through the edges of the cut.
 The Ford-Fulkerson theorem tightens this bound and states that the value of the maximum flow is equal to the capacity of the minimum cut.
 
-The Ford-Fulkerson algorithm is the direct consequence of the Ford-Fulkerson theorem.
-It solves the maximum flow problem in a greedy approach.
-Assume an instance of the maximum flow problem in an $s$-$t$ network.
-The algorithm starts with a zero flow: $f(e) = 0$ for all the edges $e$.
-In each iteration of the algorithm the value of $f$ is increased by pushing flow along a path from $s$ to $t$ up to the limits imposed by the available edge capacities.
-The process would continue until no more paths can be found to push additional flow from $s$ to $t$.
-
-The total flow value is increased by manipulating the flow values along different edges.
-However, the flow along specific edges might decrease during this process.
-In order to perform these operations, the Ford-Fulkerson algorithm defines the residual graph of the flow network.
-Intuitively, the residual graph corresponding to a flow in a given flow network, has the same vertices as the flow network and consists of edges with capacities representing how flow can be changed along the edges of the flow network.
-
-%Assume a flow $f$ in an $s$-$t$ network with capacity function $c$.
-An edge $(u, v)$ of the network can admit an additional amount of flow equal to the edge’s capacity minus the flow along that edge.
-If this value is positive, then a $(u, v)$ edge is added to the residual graph with a residual capacity $c_f (u, v) = c (u, v) - f (u, v)$.
-So, the residual graph only contains those edges of the original flow network that have left-over capacities.
-The residual graph may also contain some edges that do not exist in the original flow network.
-For each edge $(u, v)$ in the flow network which has a positive flow value, a $(v, u)$ edge is added to the residual graph with the residual capacity $c_f (v, u) = f (u, v)$.
-The edge $(v, u)$ indicates the possibility for pushing the flow backward along the edge $(u, v)$ in the original flow network.
-
-(TODO: there is a shorter informal description available as comment.
-Adapt that one if required) There are two kinds of edges in the residual graph: 1) those of the residual graph that present the edges of the flow network with left-over capacities; and 2) those of the residual graph that do not occur in the flow network, and indicate the possibility for pushing the flow backward.
-Formally speaking, consider a flow $f$ on an $s$-$t$ network $G = (V, E)$ with the capacity function $c$.
-A residual capacity function $c_f \colon V \times V \rightarrow \mathbb{R}$ is defined as following:
+The Ford-Fulkerson method is a corollary of this theorem.
+It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached.
+In order to increase the flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges.
+For this purpose, the Ford-Fulkerson method defines the residual graph.
+It has edges in parallel and reverse to the edges of the network, each edge indicating how much flow can be moved between two nodes by either increasing or decreasing the flow on an edge in the network.
+Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the
+edges with positive labels according to the following labeling function $c_f$:
 \[ c_f (u, v) = 
@@
 
-Then the residual graph corresponding to $f$ and $G$ is indicated by $G_f = (V, E_f)$, where the residual edge set $E_f$ is defined as following:
-\[ E_f = \{(u,v) \in V \times V \colon c_f (u, v) > 0\}\]
-
-After each flow update, the algorithm considers the residual graph corresponding to the new flow.
-Then the algorithm looks for an augmenting path--- a simple path from source to the sink--- in the residual graph.
-If such a path could be found, it will be used to construct a flow in the residual graph.
-Let $p$ be such a path.
-The residual capacity of $p$ is defined as $c_f(p) = min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}$.
-Then an augmenting flow $f_p \colon V \times V \rightarrow \mathbb{R}$  in the residual graph can be defined using $p$ as following:
+In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph, and then 
+pushes as much flow as possible along this path to increase the value of the current flow. 
+Formally, for an augmenting path $p$, one defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
+\[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
+An augmenting path yields a residual flow $f_p$, which is the flow that can be passed along this path:
 \[ f_p (u, v) = 
@@
 \]
-
-This flow provides a road map for modifying the current flow in the flow network.
-The process of increasing the current flow using an augmenting flow is called augmentation.
-Let $f$ be the current flow in an $s$-$t$ network $G = (V, E)$ with capacity function $c$, and $f'$ be a flow in the corresponding residual graph.
-The augmentation of $f$ using $f'$ is a function from $V \times V$ to $\mathbb{R}$, defined as following:
-\[ (f \uparrow f') (u, v) = 
+\newcommand{\augment}{{\mathbin\uparrow}}
+In order to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
+by any \emph{augmenting flow} $f'$, i.e., any flow in the residual graph:
+\[ (f \augment f') (u, v) = 
   \begin{cases}
@@
 \]
+Note that, for any edge in the network, the augmenting flow parallel to that edge is added to the flow, while the augmenting flow reverse to this edge is subtracted. 
+This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.
 
-The intuition behind the above definition is based on the definition of the residual graph.
-The augment procedure increases the flow on $(u, v)$ by $f ( u, v )$ but decreases it by $f (v, u)$.
-This is due to the fact that pushing flow on the reverse edge on the residual graph signifies decreasing the flow in the original network.
+The correctness of the Ford-Fulkerson algorithm follows from the Ford-Fulkerson theorem, which is usually stated as the following three statements being equivalent:
+\begin{enumerate}
+\item $f$ is a maximum flow in a network $G$.
+\item there is no augmenting path in the residual graph $G_f$.
+\item there is a cut $C$ in $G$ such that the capacity of $C$ is equal to the value of $f$.
+\end{enumerate}
 
-In each iteration of the algorithm we replace the current flow with the result of the augment function--- another flow on the network but with greater value.
-The algorithm terminates if no more augmenting paths can be found in the residual graph.
-According to the Ford-Fulkerson theorem, the flow computed after termination is indeed the maximum flow in the flow network.
-According to the theorem, the following statements are equivalent:
+The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph, and there are several possible implementations with different execution times.
+The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}.
+When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$\footnote{Where unambiguous, we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges.}, even for the general case of real valued capacities.
+A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.
 
-\begin{itemize}
-\item $f$ is a maximum $s$-$t$ flow in flow network $G$.
-\item there is no augmenting path in the residual graph $G_f$.
-\item there is an $s$-$t$ cut $C$ in $G$ such that capacity of $C$ is equal to the value of $f$.
-\end{itemize}
-
-The termination of the Ford-Fulkerson algorithm depends on how the augmenting path is found.
-For the networks with irrational edge capacities the algorithm might even fail to terminate~\cite{Zwick95}.
-In practice, the maximum-flow problem often arises with integral capacities.
-Moreover, we may convert rational edge capacities to integral values by multiplying them with a big enough integer.
-If $f^*$ denotes a maximum-flow in the transformed network, then the algorithm will iterate at most $|f^*|$ times, since the augmentation procedure increases the flow value by at least one unit in each iteration.
-Assuming that we use breadth-first-search (BFS) or depth-first-search (DFS) for finding augmenting paths makes the total execution time of the Ford-Fulkerson algorithm $O ( E | f^* |)$, as BFS and DFS run in $O (E + V)$.
-
-(TODO: Here I mention the Edmonds-Karp algorithm and I mention its running time, co-relate this with definition in later sections) The Ford-Fulkerson algorithm is sometimes called a method instead of an algorithm because it does not fully specify the procedure for finding the augmenting path in the residual graph.
-There are several implementations with different execution times.
-For instance, the Edmond-Karp algorithm uses breadth-first-search (BFS) for finding augmenting paths and has $O (VE^2)$ running-time.
-The augmentings paths that are computed using BFS are also shortest paths connecting source to the sink in the residual graph.
-Edmond and Karp showed that in the Ford-Fulkerson algorithm, if each augmenting path is the shortest one, the length of the augmenting paths is non-decreasing and the algorithm always terminates.
-Using this statement they could finally present a polynomial algorithm for solving the maximum flow problem in generic case of real edge capacities.
-
-\section{Formalizing the Ford-Fulkerson method}
+\section{Formalizing the Ford-Fulkerson method}\label{sec:abs-formalization}
 %  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
@@
 
-(TODO: not sure about repeating citation of Cormen) Informal proofs focus on the relevant thoughts by leaving out technical details.
-However, in a formal proof each step has to be precisely specified as a valid application of some inference rule which makes formal proofs often lengthy and hard to follow.
-In the case that an interactive theorem prover is used, readers may even need to replay the proof to understand the idea behind it.
-In Isabelle proof assistant, the Isar proof language bridges the semantic gap between internal notions of proof given by interactive theorem proving systems and an appropriate level of abstraction for user-level work.
-Isar makes formal proofs readable even for those who are not familiar with Isabelle concepts.
-We have used such a functionality to present our formal proof of the Ford-Fulkerson method which resembels the informal proof described by Cormen et al.~\cite{CLRS09}.
+% (DONE: not sure about repeating citation of Cormen: I would definitely repeat the citation!) 
+Informal proofs focus on the relevant thoughts by leaving out technical details and obvious steps. 
+In contrast, a formal proof has to precisely specify each step as the application of some inference rules. 
+Although modern proof assistants provide high-level tactics to summarize some of these steps, formal proofs tend to be significantly more 
+verbose than informal proofs. Moreover, formal proofs are conducted in the tactic language of the proof assistant, which is often some dialect of ML. 
+Thus, many formal proofs are essentially programs that instruct the proof assistant how to conduct the proof, and are inaccessible without a deep knowledge of the used proof assistant, in many cases requiring to replay the proof in the proof assistant in order to understand the idea behind it.
 
-As an example consider the following proof obtained form Cormen's algorithm book.
-Let $cf$ be the residual graph corresponding to a flow $f$ in an $s$-$t$ network $c$, and let $f'$ be a flow in $cf$.
-Consider the definition of the augment function that was described in previous section.
-Then, the result of augmenting $f$ with $f'$ is another function which satisfies the capacity constraint of a $s$-$t$ flow in $c$.
+For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that strongly resemble standard mathematical textbook proofs, and are accessible even for those not familiar with Isabelle/HOL.
+We use Isar to present our proof of the Ford-Fulkerson method such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.
 
-\emph{\textbf{Proof.}} first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
+As an example consider the proof that augmenting a flow $f$ with a residual flow $f'$ yields a valid flow $f\augment f'$ again. In particular, one has to show 
+that the augmented flow satisfies the capacity constraint. Cormen et al.~give the following proof, which we copied literally, only replacing the references to 
+``Equation~26.4'' by ``definition of $\augment$'':
+
+For the capacity constraint, first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
 	\begin{align*}
-	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of augment)} \\
+	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of $\augment$)} \\
 	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
@@
 	\begin{align*}
-	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of augment)} \\
+	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of $\augment$)} \\
 	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
@@
 
-Following we present the equivalent Isar proof.
-Note that we have to additionally consider the case that $(u, v) \notin E$.
-Moreover in line 6, we have used the assumption that there are no parallel edges in a flow network.
-This important property is not mentioned in Cormen's proof.
+In the following we present the corresponding formal proof in Isar:
 \Snippet{augment_flow_presv_cap}
+Its structure is exactly the same as the textbook proof, except that we had to also consider the case $(u,v)\notin E$, which is not mentioned in the informal proof at all, and easily discharged in our formal proof by the \isai{auto}-tactic after the \isai{qed}.
+We also use exactly the same justifications as the original proof, except that we had to use the fact that there are no parallel edges to show $c_f(v,u)=f(u,v)$, which is not mentioned in the original proof.
 
-In reference books, it is a convention to use pseudocodes for presenting the algorithms.
-The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm and leaves out the implementation details.
-We present the Ford-Fulkerson method using the syntax of (TODO: here I mention autoref again, I was not aware that we are not using it anymore.
-So please correct all the references that I made to autoref tool) Autoref~\cite{La13} tool which is quite similar to pseudocode of the algorithm:
+In textbooks, it is common to present algorithms in pseudocode, which captures the essential ideas, but leaves open implementation details.
+As a formal equivalent to pseudocode, we use the monadic programming language provided by the Isabelle Refinement Framework~\cite{LaTu12,La12}:
 \Snippet{ford_fulkerson_algo}
 
-In the pseudocode of the Ford-Fulkerson method, the path found while evaluating the loop condition is later used for augmenting the flow inside the loop body.
-The best way for implementing such requirement using syntax of the refinement framework is moving the path lookup inside the loop body.
-To do so, we modified the while-loop by introducing a break flag.
-The flag is initially off, and is set if no augmenting path is found in the residual graph.
+TODO: Introduce obtain-syntax, and unfold find-augmenting-path spec!
+The code looks quite similar to pseudocode that one would expect in a textbook, but actually is a rigorous formal specification of the algorithm, using nondeterminism to leave open the implementation details (\cf Section~\ref{sec:refinement}).
+The only caveat is that our framework currently does not support breaking of loops, such that we had to use an explicit flag instead.
 
-\section{Refinement in Isabelle/HOL}
-When formalizing algorithms, there is often a trade off between verifiability 
-and efficiency: An efficient algorithm that uses elaborate data structures and optimizations is often harder to 
-verify than a simple but straightforward algorithm.
+Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Refinement Framework, it is straightforward to prove (partial) correctness of
+the Ford-Fulkerson method, which is stated in Isabelle/HOL by the following theorem:
+\Snippet{ford_fulkerson_partial_correct}
 
-The idea of stepwise refinement \cite{Wirth} is to first specify a simple, abstract version of an algorithm, which can be easily verified,
-and then refine this algorithm towards an efficient implementation. Here, each refinement step can focus on a single aspect of the algorithm or 
-its implementation, independently of the other refinement steps. This results in a greatly increased modularity of proofs. 
-Refinement calculi \cite{Back} are used to systematically perform refinement in a Hoare-logic setting, and are particularly well-suited for 
-implementation in theorem provers.
+\section{Refinement in Isabelle/HOL}\label{sec:refinement}
+After having stated and proved correct an algorithm on the abstract level, the next step is to provide an efficient implementation. 
+In our case, we first specialize the Ford-Fulkerson method to use shortest augmenting paths, then implement the search for shortest augmenting paths by BFS, and finally use efficient data structures to represent the abstract objects modified by the algorithm.
 
-Note that it is important to support nondeterminism in a refinement setting, as this allows to defer 
-implementation choices to the later refinement steps. For example, if we want to specify an algorithm 
-that returns a shortest path between two nodes of a graph, we do not want to fix a particular shortest path.
-Only later, when we decide to implement the specification by, e.g., breadth first search, we fix a particular path.
+A natural way to achieve this formally is \emph{stepwise refinement}~\cite{Wirth71}, and in particular \emph{refinement calculus}~\cite{Back78,BaWr98}, which allows us to systematically transform an abstract algorithm into a more concrete one, preserving its correctness.
 
@@
 It features a refinement calculus for programs phrased in a nondeterminism monad. 
-The monad's type is a set of results plus an additional value that indicates a failure:
+The monad's type is a set of possible results plus an additional value that indicates a failure:
 \begin{lstlisting}
@@
 \end{lstlisting}
-The return operation \isai{return x} of the monad describes the single result \isai{x}, and the bind 
+The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
 operation \isai{bind m f} nondeterministically picks a result from $m$ and executes $f$ on it. 
@@
 
-On nondeterministic results we define the \emph{refinement ordering} by lifting the subset ordering with \isai{fail} being the greatest element.
+We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
 Intuitively, $m\le m'$ means that $m$ is a refinement of $m'$, i.e., all possible results of $m$ are 
 also possible results of $m'$. 
-Note that this ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
+Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
 Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
@@
 
-Correctness of a program $m$ with precondition $P$ and postcondition $Q$ is expressed as \isai{P ==> m <= spec r. Q r}, which
+Correctness of a program $m$ with precondition $P$ and postcondition $Q$ is expressed as \isai{P ==> m <= spec r. Q r} (or, eta-contracted, just \isai{spec Q}), which
 means that, if $P$ holds, $m$ does not fail and all possible results of $m$ satisfy $Q$. Note that we provide different recursion constructs
@@
 The Isabelle Refinement Framework also supports data refinement, changing the representation of results according to a \emph{refinement relation}, 
-which relates concrete with abstract results: Given a relation $R$, \isai{\<Down> R m} is the largest set of concrete results that are related to an 
-abstract result in $m$ by $R$. If \isai{m=fail}, then also \isai{\<Down> R m = fail}.
+which relates concrete with abstract results: Given a relation $R$, \isai{\<Down>R m} is the largest set of concrete results that are related to an 
+abstract result in $m$ by $R$. If \isai{m=fail}, then also \isai{\<Down>R m = fail}.
 
-Finally, the Isabelle Refinement Framework provides a refinement calculus, which comes with a verification condition 
-generator to simplify its usage.
-
-In a typical program development using the Isabelle refinement framework, one first comes up with an initial version $m_0$ of
+In a typical program development, one first comes up with an initial version $m_0$ of
 the algorithm and its specification $P,Q$, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions $m_i$ of the algorithm,
@@
 gets \isai{P ==> $m_i$ <= \<Down>$R_i\ldots R_1$ spec Q}, showing the correctness of the refined algorithm.
+Various tools, including a verification condition generator, assist the use in conducting the refinement proofs by breaking 
+them down to statements that do not involve monad constructs any more, and in many cases precisely reflect the core idea of the proof.
 
-Monotonicity of the standard combinators also allows for modular refinement: Replacing part of a program by a refined version
+Monotonicity of the standard combinators also allows for modular refinement: Replacing a part of a program by a refined version
 results in a program that refines the original program.
+This gives us a natural formal model for statements like ``we implement shortest path finding by BFS'', or ``we use arrays to represent the edge labeling''.
 
-In Section~\ref{sec:fofu}, we presented the abstract Ford-Fulkerson algorithm as a program in the Isabelle Refinement Framework.
-In the next sections, we describe its stepwise refinement down to executable code.
-
-% Each refinement step is shown to preserve correctness
-% 
-% Refinement techniques \cite{Wirth,Back} structure the verification of an algorithm into
-% modular steps, such that one first verifies the abstract algorithmic ideas, and then 
-% 
 % 
@@
 
-\section{Edmonds-Karp Algorithm}
-  The general Ford-Fulkerson method can only be shown to terminate if the edge capacities are integer numbers. 
-  An improvement over that is the Edmonds-Karp algorithm~\cite{}, which is obtained when always choosing 
-  a shortest augmenting path. In this case, it can be shown that only $O(|V||E|)$ augmentations are performed until the
-  algorithm terminates, even for real-valued capacities. A shortest augmenting path can be obtained by breadth first search in time $O(|E|)$,
-  yielding an overall complexity of $O(|V||E|^2) $ algorithm.
+\section{Edmonds-Karp Algorithm}\label{sec:edka}
+  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm requires us to show that finding a shortest augmenting path is a correct refinement of finding any augmenting path, which is straightforward using basic graph theory.
 
-  In our formalization, we refine the specification of an augmenting path to a specification of a shortest augmenting path, which immediately yields 
-  an abstract version of the Edmonds-Karp algorithm which refines the Ford-Fulkerson method.
+  Considerable more effort is required on showing that the resulting algorithm terminates within $O(VE)$ iterations. 
+  The idea of the proof is as follows: Edges that are reverse to edges on a shortest path cannot lie 
+  on a shortest path itself.
+  On every augmentation, at least one edge of the residual graph that lies on a shortest augmenting path is flipped. Thus, either the length of the shortest path increases,
+  or the number of edges that lie on a shortest path decreases. As the length of a shortest path is at most $V$, there are no more than $O(VE)$ iterations.
 
-  The larger part of the formalization is spent on proving the complexity bound. Note that the refinement framework does not 
-  have a notion of computational complexity, so we cannot even define the runtime of an algorithm. However, we can instrument 
-  the while-loop of the algorithm with a counter, which is incremented on each iteration, and prove an upper bound on this counter.
-  Moreover, this also yields a termination argument for real-valued capacities.\footnote{As we restricted the algorithm 
-  to integer valued capacities at the very abstract level, this is actually not used.}
+  Note that Cormen et al.~present the same idea a bit differently: They define an edge of the residual graph being \emph{critical} if it lies on a shortest path such that it will be flipped by augmentation.
+Then, they establish an upper bound of how often an edge can get critical during the algorithm.
+Our presentation is more suited for a formal proof, as we can directly construct a measure function from it, \ie a function from flows to natural numbers, which decreases on every iteration and is bounded by $O(VE)$.
 
-  The idea of the complexity proof is as follows: Note that edges that are reverse to edges on a shortest augmenting path cannot lie 
-  on a shortest path itself.
-  On every augmentation, at least one edge that lies on a shortest path is flipped. Thus, either the length of the shortest path increases,
-  or the number of edges that lie on a shortest path decreases. As the length of a shortest path is at most $|V|$, there are no more than $O(|E||V|)$ iterations.
+  Formalizing the above intuitive argument was more tricky than it seemed on first glance:
+  While it easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
+  to a graph transformation which may add reverse edges and removes at least one original edge requires some generalization of the statement.
+  Note that a straightforward induction on the length of the augmenting path or on the number of flipped edges fails, as, after flipping the first 
+  edge, the path is no longer augmenting.
 
-  TODO: Relate to Cormen-proof: Critical edge, removed by augmentation, bound on how often an edge can get critical.  
-    Same idea, but yields a more precise bound. On the other hand, we found our proof to be simpler to formalize, and its enough to
-    get the $O(|V||E|)$ bound.
+%   We decided to generalize the statement as follows. Assume we have an original graph $cf$ with a shortest augmenting path $p$, and a transformed graph $cf'$ that has been created from
+%   $cf$ by adding some flipped edges from $p$ and removing some edges from $p$.
+%   Then, we consider an edge $(u,v) \in p$, a path $p_1$ from $s$ to $v$ in the original graph, and a path $p_2$ from $u$ to $t$ in the transformed graph.
+%   
+%   By induction on the number of reversed edges in $p_2$, we 
+%   show that $|p_1|+|p_2| > |p|$. In the induction step, we use the proof idea for the single flipped edge $(v,u)$, and,
+%   if $p_2$ contains more flipped edges, we split $p_2$ at the first flipped edge. Then, the initial segment of $p_2$ contains no flipped edges,
+%   and thus also is a path in the original graph, and we can apply the induction hypothesis.
+%   
+%   Proving that augmentation with a shortest augmenting path actually only adds flipped edges, and removes at least one original edge, required some more work,
+%   but was straightforward and yielded no surprises.
+%   
+%   Finally, we phrase the complexity argument as a measure function $m = l*2|E| + n$, where $l$ is the length of the shortest augmenting path,
+%   and $n$ is the number of edges that lie on any shortest path. We show that $m$ is decreased by the loop body.
+%   Adding a special case for loop termination (there are no augmenting paths), and observing that $m$ is bounded by $2|V||E|$, yields 
+%   the desired upper bound on loop iterations. Refinement of the algorithm to add an explicit loop counter, and asserting the upper bound, is then straightforward.
 
-  Formalizing the above intuitive argument is more tricky than it seems on first glance. 
-  While it easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
-  to a graph transformation which may add reverse edges and removes at least original edge, is tricky. Note that a straightforward induction
-  on the length of the augmenting path must fail, as after flipping the first edge, the path is no longer augmenting.
+  Having defined the measure function, and shown that it decreases on augmentation, it is straightforward to refine the partial correct while loop to a total correct one.
+Moreover, to make explicit the bound on the number of loop iterations, we instrument the loop with a counter, and assert the upper bound on the counter after the loop.
 
-  We decided to generalize the statement as follows. Assume we have an original graph $cf$ with a shortest augmenting path $p$, and a transformed graph $cf'$ that has been created from
-  $cf$ by adding some flipped edges from $p$ and removing some edges from $p$.
-  Then, we consider an edge $(u,v) \in p$, a path $p_1$ from $s$ to $v$ in the original graph, and a path $p_2$ from $u$ to $t$ in the transformed graph.
+\section{Refinement to Executable Code}\label{sec:executable}
+  In the previous sections, we have presented our abstract formalization of the Edmonds-Karp algorithm,
+  leaving open how to obtain a shortest augmenting path, and how to implement the algorithm. 
+  In this section, we outline the further refinement steps that were necessary to obtain an efficient implementation.
 
-  By induction on the number of reversed edges in $p_2$, we 
-  show that $|p_1|+|p_2| > |p|$. In the induction step, we use the proof idea for the single flipped edge $(v,u)$, and,
-  if $p_2$ contains more flipped edges, we split $p_2$ at the first flipped edge. Then, the initial segment of $p_2$ contains no flipped edges,
-  and thus also is a path in the original graph, and we can apply the induction hypothesis.
+  \subsection{Using Breadth First Search}
+  A standard way to find a shortest path in a graph is breadth first search (BFS).
+Luckily, we had already formalized a BFS algorithm as an example for the refinement framework.
+Unfortunately, the algorithm only computed the minimum distance between two nodes, without returning an actual path.
+For this project, we extended the formalization accordingly, and added an efficient imperative implementation, using the same stepwise refinement techniques as for the main algorithm.
+Note that the resulting BFS algorithm is independent, and can be reused to refine shortest path finding in other applications.
 
-  Proving that augmentation with a shortest augmenting path actually only adds flipped edges, and removes at least one original edge, required some more work,
-  but was straightforward and yielded no surprises.
+  Implementing shortest path finding by BFS in Edmonds-Karp algorithm yields a specification that algorithmically describes all major operations, but still leaves open the data structures used for implementation.
 
-  Finally, we phrase the complexity argument as a measure function $m = l*2|E| + n$, where $l$ is the length of the shortest augmenting path,
-  and $n$ is the number of edges that lie on any shortest path. We show that $m$ is decreased by the loop body.
-  Adding a special case for loop termination (there are no augmenting paths), and observing that $m$ is bounded by $2|V||E|$, yields 
-  the desired upper bound on loop iterations. Refinement of the algorithm to add an explicit loop counter, and asserting the upper bound, is then straightforward.
-
-\section{Refinement to Executable Code}
-  \subsection{Using Breadth First Search}
-
-  In the previous sections, we have described the Ford-Fulkerson and the Edmonds-Karp algorithm abstractly,  
-  leaving open how to obtain a (shortest) augmenting path. A standard way to find a shortest path in a graph 
-  is breadth first search (BFS). Luckily, we had already formalized a BFS algorithm as an example for the refinement framework.
-  However, the existing formalization could only compute the minimum distance between two nodes, without returning an actual path.
-  We briefly report on our adapted formalization here, which is displayed in Figure~\ref{TODO: Snippet}: 
-  TODO: Reference to figure (line numbers?)
-  The abstract algorithm keeps track of a set $C$ of current nodes, a set $N$ of new nodes, and a partial map $P$ from already visited nodes to predecessor nodes.
-  Initially, only the start node is in $C$ and $N$ is empty. In each iteration of the loop, a node $u\in C$ is picked, and its successors $v$ that have not yet been 
-  visited are added to $N$, and $P~v$ is set to $u$. If $C$ becomes empty, $C$ and $N$ are swapped. If the target node is encountered, the algorithm immediately terminates.
-  Note that this implementation is a generalized version of the usual queue implementation of BFS: While a queue enforces a complete order on the encountered nodes, our 
-  implementation only enforces an ordering between nodes on the current level and nodes on the next level.
-
-  For the actual algorithm, we wrap this algorithm by a procedure to handle the special case where source and target nodes are the same, and to extract a shortest path 
-  from $P$ upon termination. Moreover, we implement the abstract specification of adding the successor nodes of a node by a loop.
-
-  Finally, we prove the following theorem: %TODO: Not actually true, the impls are oly done in bfs2!
-%     theorem
-%       assumes "src∈V" 
-%       shows "bfs src dst ≤ (spec p. 
-%         case p of None ⇒ ¬conn src dst | Some p ⇒ shortestPath src p dst)"
-
-  Obviously, the BFS algorithm refines the specification for obtaining a shortest path.
-  Using this to refine the Edmonds-Karp formalization yields a program that algorithmically specifies all major operations.
+%   
+%   We briefly report on our adapted formalization here, which is displayed in Figure~\ref{TODO: Snippet}: 
+%   TODO: Reference to figure (line numbers?)
+%   The abstract algorithm keeps track of a set $C$ of current nodes, a set $N$ of new nodes, and a partial map $P$ from already visited nodes to predecessor nodes.
+%   Initially, only the start node is in $C$ and $N$ is empty. In each iteration of the loop, a node $u\in C$ is picked, and its successors $v$ that have not yet been 
+%   visited are added to $N$, and $P~v$ is set to $u$. If $C$ becomes empty, $C$ and $N$ are swapped. If the target node is encountered, the algorithm immediately terminates.
+%   Note that this implementation is a generalized version of the usual queue implementation of BFS: While a queue enforces a complete order on the encountered nodes, our 
+%   implementation only enforces an ordering between nodes on the current level and nodes on the next level. 
+% 
+%   For the actual algorithm, we wrap this algorithm by a procedure to handle the special case where source and target nodes are the same, and to extract a shortest path 
+%   from $P$ upon termination. Moreover, we implement the abstract specification of adding the successor nodes of a node by a loop.
+%   
+%   Finally, we prove the following theorem: %TODO: Not actually true, the impls are oly done in bfs2!
+% %     theorem
+% %       assumes "src∈V" 
+% %       shows "bfs src dst ≤ (spec p. 
+% %         case p of None ⇒ ¬conn src dst | Some p ⇒ shortestPath src p dst)"
+%   
+%   Obviously, the BFS algorithm refines the specification for obtaining a shortest path.
+%   Using this to refine the Edmonds-Karp formalization yields a program that algorithmically specifies all major operations.
 
@@
   in order to update the flow, the residual graph is used, which is a combination of the current flow and the network.
-  Obviously, computing the complete residual graph each time before searching for an augmenting path would be a bad idea. 
+  Obviously, computing the complete residual graph each time before searching for an augmenting path is a bad idea. 
   One solution to this problem is to compute the edges of the residual graph on demand from the network and the current flow.
@@
 
-  Thus, we chose to let the algorithm operate on a representation of the residual graph directly.
+  After our profiling indicated a hot spot on accessing the capacity matrices of the network and the flow, we switched to 
+  an algorithm that operates on a representation of the residual graph directly. This resulted in a speed-up of factor two.
 
-  The shortest path specification and bottleneck computation are already 
-  phrased on residual graphs. What remains is the computation of the initial residual graph (which equals the Network),
-  the augmentation of the residual graph, given an augmenting path and its bottleneck capacity, and the computation of the 
-  resulting maximum flow from the residual graph, at the end of the algorithm.
-
-  In order to refine these operations, we note that the residual graph uniquely determines the flow (and vice versa). 
-%         definition (in Network) "flow_of_cf cf e ≡ (if (e∈E) then c e - cf e else 0)"
-
-  For a valid flow $f$, we have that
-%         lemma (in NFlow)
-%           "flow_of_cf (residualGraph f) = f"
-%           by (rule fo_rg_inv)
-
-  Thus, augmentation can be expressed directly on the residual graph:
-%         "residualGraph (augment (augmentingFlow p)) (u,v) = (
-%           if (u,v)∈set p then (residualGraph f (u,v) - bottleNeck p)
-%           else if (v,u)∈set p then (residualGraph f (u,v) + bottleNeck p)
-%           else residualGraph f (u,v))"
-
-  Having these characterizations, a straightforward refinement yields an algorithm that operates on residual graphs.
-
-%   Note: Straightforward, so omitted here!
-%   In a next step, we have to specify iterative implementations for finding the bottleneck capacity and doing the augmentation.
-%   This is straightforward by folding over the path, exploiting that an augmenting path is simple, and thus never contains an edge 
-%   and its flipped edge at the same time.
+  The shortest path specification and bottleneck computation are already phrased on residual graphs. 
+  As the residual graph uniquely determines the flow (and vice versa), it is straightforward to phrase the other 
+  operations directly on the residual graph. Performing a data refinement of the flow \wrt the refinement 
+  relation $\{(c_f,f) \mid \text{$f$ is a flow}\}$ then yields the desired algorithm.
 
@@
   Although this can be implemented on the capacity matrix by iterating over all nodes, this implementation tends to be inefficient for sparse graphs,
-  where we would have to iterate over many nodes just to find that there is no edge.
+  where we would have to iterate over many possible successor nodes just to find that there is no edge.
 
-  A common optimization is to pre-compute a map from nodes to adjacent nodes in the network. As an edge in the residual graph is either parallel or reverse to 
-  a network edge, it is enough to iterate over the adjacent nodes in the network, and check whether they are actually successors in the residual graph.
-  It is straightforward to show that this implementation actually returns the successors in the residual graph:
-%         lemma (in RGraph) rg_succ_ref:
-%           assumes A: "is_pred_succ ps c"
-%           assumes B: "u∈V"
-%           shows "rg_succ2 ps cf u ≤ SPEC (λl. (l,cf.E``{u}) ∈ ⟨Id⟩list_set_rel)"
+  A common optimization is to pre-compute an \emph{adjacency map} from nodes to adjacent nodes in the network. 
+  As an edge in the residual graph is either parallel or reverse to 
+  a network edge, it is enough to iterate over the adjacent nodes in the network, and check whether they are actual successors in the residual graph.
+  It is straightforward to show that this implementation actually returns the successors in the residual graph.
 
@@
   The adjacency map of the network is implemented as an array of lists of nodes. 
+  An augmenting path is represented by a list of edges, \ie a list of pairs of nodes.
 
-  For the BFS algorithm, we implement the sets $C$ and $N$ by lists, and the predecessor map $P$ as an array of \isai{node option}. 
-  The successors of a node are represented as a list with no duplicate elements. The resulting path, which is extracted from $P$ upon termination of
-  the BFS, is represented as a list of edges.
+%   For the BFS algorithm, we implement the sets $C$ and $N$ by lists, and the predecessor map $P$ as an array of \isai{node option}. 
+%   The successors of a node are represented as a list with no duplicate elements. The resulting path, which is extracted from $P$ upon termination of
+%   the BFS, is represented as a list of edges.
 
@@
 
-  This allows flexibility in using the algorithm, as any capacity matrix representation can be converted into a function easily, and 
+  This gives us some flexibility in using the algorithm, as any capacity matrix representation can be converted into a function easily, and 
   without loosing efficiency for read-access\footnote{Due to technical limitations of our tools, this function 
   cannot depend on heap content, but it can use the efficient functional and pseudo-functional datastructures provided by the Isabelle Collection Framework.}. 
+  Similarly, our implementation expects an adjacency map as additional parameter, which is then tabulated into an array.
+This is convenient in our context, where a preprocessing step computes the adjacency map anyway.
 
-  The output flow of the algorithm is represented as the residual graph. The user can decide how he computes the maximum flow from it. For example,
+  The output flow of the algorithm is represented as the residual graph. The user can decide how to compute the maximum flow from it. For example,
   in order to compute the maximum flow value, only the outgoing edges of the source node have to be computed, which is typically less 
@@
 
-  Note that their is still some optimization potential left: For example, the BFS algorithm computes the predecessor map $P$.
-  Then, it iterates over $P$ to get the shortest path into a list. Then, this list is iterated again to compute the 
-  bottleneck, and, finally, another iteration over the list is done to do the augmentation. This calls for a deforestation optimization,
-  getting rid of the intermediate list altogether, and iterate two times over the predecessor map: A first time to compute the bottleneck, and a second time
-  to do the augmentation. Fortunately, iteration over the shortest path does not significantly contribute to the runtime of the algorithm, so that we leave this
+  Note that there is still some optimization potential left in the choice of our data structures: 
+  For example, the BFS algorithm computes a predecessor map $P$. It then iterates over $P$ to extract the shortest path as a list of edges.
+  A subsequent iteration over this list computes the residual capacity, and a final iteration performs the augmentation. 
+  This calls for a deforestation optimization,
+  getting rid of the intermediate list, and iterate only two times over the predecessor map directly.
+%   A first time to compute the bottleneck, and a second time to do the augmentation. 
+  Fortunately, iteration over the shortest path seems not to significantly contribute to the runtime of our implementation, such that we leave this
   optimization for future work.
 
-  Note that we performed the last refinement step using our Sepref tool~\cite{La15}, which provides tool support for data refinements from 
+  Note that we performed the last refinement step using our Sepref tool~\cite{La15,La16}, which provides tool support for data refinements from 
   purely functional programs of the Isabelle Refinement Framework into imperative programs expressed in Imperative/HOL~\cite{BKHEM08}.
   The formalization of this refinement step consists of setting up the mappings between the abstract and concrete data structures,
-  and then using the tools provided by Sepref to synthesize the Imperative/HOL programs and their refinement proofs.
+  and then using Sepref to synthesize the Imperative/HOL programs and their refinement proofs.
 
-  Finally, Imperative/HOL provides a heap-exception monad and comes with a setup for the Isabelle code generator to generate efficient imperative programs
+  Finally, Imperative/HOL comes with a setup for the Isabelle code generator to generate efficient imperative programs
   in OCaml, SML, Scala, and Haskell.
@@
   Combining the implementation of the Edmonds-Karp algorithm with the network checker yields our final algorithm,
-  for which we can export code, and have proved the theorem: [...Snippet! edmonds-karp-correct]. 
+  for which we can export code, and have proved the theorem: 
+  \Snippet{edmonds-karp-correct}
 %     theorem edmonds_karp_correct:
@@
   If the algorithm returns \isai{None}, then the edge list was malformed or described a graph that does not satisfy the network assumptions.
-  Here, \isai{ln_invar} describes well-formed edge lists, i.e., edge lists that have no duplicate edges and no edges with zero capacity,
+  Here, \isai{ln_invar} describes well-formed edge lists, i.e., edge lists that have no duplicate edges and only edges with positive capacity,
   and \isai{ln_\<alpha>} describes the mapping from (well-formed) edge lists to capacity matrices.
-  If the algorithm returns some number $N$ and residual graph $fi$, then the input was a well-formed edge list that describes a valid network.
+  If the algorithm returns some number $N$ and residual graph $fi$, then the input was a well-formed edge list that describes a valid network with at most $N$ nodes.
   Moreover, the returned residual graph describes a flow $f$ in the network, which is maximal. TODO: Swap in thm!
@@
 
-\section{Benchmarking}
+\section{Benchmarking}\label{sec:benchmark}
   We have compared the running time of our algorithm against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}.
@@
 
-\section{Conclusion}
+\section{Conclusion}\label{sec:concl}
   In this paper, we have presented a verification of the Ford-Fulkerson method, using a stepwise refinement approach.

### 2016-03-03 paper/root.tex
Rephrasing round 1 completed

@@
   We have compared the running time of our algorithm against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}.
+  We have done the comparison on randomly generated sparse and dense networks, the sparse networks having a \emph{density} ($\frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
+Note that the maximum density for networks that satisfy our assumptions is $0.5$, as we allow no parallel edges.
+  For sparse networks, we varied the number of nodes between $1000$ and $5000$, for dense networks between $1000$ and $1400$.
+  The results are shown in Figure~\ref{fig:benchmark}, in a double-logarithmic scale. 
+%   \begin{wrapfigure}{o}{.4\textwidth}
+%   \vspace{-20pt}
+%   \begin{center}
+%   \begin{tabular}{|c|c|c|}
+%   \hline
+%   Name        & SML       & Java   \\
+%   \hline\hline
+%   1000-sparse & 204       & 15     \\
+%   2000-sparse & 1430      & 231    \\
+%   3000-sparse & 4928      & 828    \\
+%   4000-sparse & 9299      & 2145   \\
+%   5000-sparse & 23625     & 4868   \\
+%   1000-dense  & 6575      & 4011   \\
+%   1100-dense  & 8067      & 5487   \\
+%   1200-dense  & 10892     & 7011   \\
+%   1300-dense  & 14070     & 9815   \\
+%   1400-dense  & 17317     & 12156  \\
+%   \hline
+%   \end{tabular}
+%   \end{center}
+%   \vspace{-15pt}
+%   \caption{Dijkstra benchmark}\label{tab:benchmark}
+%   \vspace{-20pt}
+%   \end{wrapfigure}
 
@@
   \end{tikzpicture}
-%\caption{Edmonds-Karp benchmark}\label{fig:benchmark}
+  \caption{Edmonds-Karp benchmark}\label{fig:benchmark}
   \end{figure}
 
-  We have done the comparison on randomly generated sparse and dense networks, the sparse networks having a \emph{density} ($\frac{|E|}{|V|(|V| - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
-For sparse networks, we varied the number of nodes between $1000$ and $5000$, for dense networks between $1000$ and $1400$.
-  The results are shown in Table~\ref{tab:benchmark}. 
-  \begin{wrapfigure}{o}{.4\textwidth}
-  \vspace{-20pt}
-  \begin{center}
-  \begin{tabular}{|c|c|c|}
-  \hline
-  Name        & SML       & Java   \\
-  \hline\hline
-  1000-sparse & 204       & 15     \\
-  2000-sparse & 1430      & 231    \\
-  3000-sparse & 4928      & 828    \\
-  4000-sparse & 9299      & 2145   \\
-  5000-sparse & 23625     & 4868   \\
-  1000-dense  & 6575      & 4011   \\
-  1100-dense  & 8067      & 5487   \\
-  1200-dense  & 10892     & 7011   \\
-  1300-dense  & 14070     & 9815   \\
-  1400-dense  & 17317     & 12156  \\
-  \hline
-  \end{tabular}
-  \end{center}
-  \vspace{-15pt}
-  \caption{Dijkstra benchmark}\label{tab:benchmark}
-  \vspace{-20pt}
-  \end{wrapfigure}
-
-  (TODO: Table's caption is incorrect, I think we have enough space to print the whole table with SML, Java as rows and graph size as columns.
-Also, single lines for separating table looks nicer.
-I have seen it in your Hopcroft's paper) We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of 6, while for dense graphs, our implementation comes quite close
+  We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of 6, while for dense graphs, our implementation comes quite close 
   to the Java implementation. We have no full explanation for this phenomenon. Note that the Java implementation operates on flows, while our implementation 
@@
   Starting with a proof of the Ford-Fulkerson theorem, we have verified the generic Ford-Fulkerson method, then 
-  specialized it to the Edmonds-Karp algorithm and proved the upper bound $O(|E||V|)$ for the number of outer loop iterations.
-  We then used several refinement steps to derive an efficiently executable implementation of the algorithm, 
-  involving a verified breadth first search to obtain shortest augmenting paths. 
+  specialized it to the Edmonds-Karp algorithm and proved the upper bound $O(VE)$ for the number of outer loop iterations.
+  We then conducted several refinement steps to derive an efficiently executable implementation of the algorithm, 
+  including a verified breadth first search algorithm to obtain shortest augmenting paths. 
   Finally, we added a verified algorithm to check whether the input is a valid network, and generated executable code in SML.
@@
 
-  Our formalization has combined several techniques to achieve an elegant and accessible formalization: The Isar-proof language~\cite{}
-  allows us to write formal proofs that closely follow informal proofs found in textbooks. This way, we have created a completely rigorous, but 
-  still accessible, proof of the Ford-Fulkerson theorem. The Isabelle Refinement Framework allows us to first formalize an algorithm abstractly, 
-  and then refine it, in a modular fashion, towards an efficient implementation. This allowed us to present the Ford-Fulkerson method on a level 
+  Our formalization has combined several techniques to achieve an elegant and accessible formalization: The Isar-proof language~\cite{Wenzel99}
+  supports writing formal proofs that closely resemble informal proofs found in textbooks. This way we have created a completely rigorous but 
+  still accessible proof of the Ford-Fulkerson theorem. The Isabelle Refinement Framework~\cite{LaTu12,La12} and the Sepref tool~\cite{La15,La16} 
+  provide stepwise refinement techniques to formalize an algorithm abstractly, and then refine it, in a modular fashion, towards an efficient implementation. 
+  This allowed us to present the Ford-Fulkerson method on a level 
   of abstraction that closely resembles pseudocode presentations found in textbooks, and then formally link this presentation to an efficient
   implementation. Moreover, modularity of refinement allowed us to develop the breadth first search algorithm independently, and later link it to the 
-  main algorithm. The BFS algorithm can be reused as building block for other algorithms. Also the data structures are re-usable, and in fact, the 
-  only new data structure we had to implement where the adjacency matrices. The other data structures where readily available in the growing library 
-  of the Sepref tool~\cite{La15} and the Isabelle Collection Framework~\cite{LaTu12}.
+  main algorithm. The BFS algorithm can be reused as building block for other algorithms. Also the data structures are re-usable. 
+  Although we had to implement the array representation of (capacity) matrices for this project, it will be added to the growing library of data structures 
+  supported by the Sepref tool, such that it can be re-used for future formalizations.
 
+  We conclude with some statistics:
   The formalization consists of roughly 8000 lines of proof text, where the graph theory up to the Ford-Fulkerson theorem required 2500 lines.
@@
   The remaining lines are contributed by the network checker and some auxiliary theories.
-The development of the theories required roughly 3 men month, a significant amount of this time going into a first, purely functional version of the implementation, which was later dropped in favor of the significantly faster imperative version.
+The development of the theories required roughly 3 man month, a significant amount of this time going into a first, purely functional version of the implementation, which was later dropped in favor of the faster imperative version.
   TODO: Reza, is this estimate OK?
@@
   \subsection{Related Work}
-  We are only aware of one other formalization of the Ford-Fulkerson method conducted in Mizar~\cite{}. 
-  It formalizes a fixed instance of the Ford Fulkerson method on a rather abstract level. No code has been generated, 
-  nor does it contain a complexity analysis.
+  We are only aware of one other formalization of the Ford-Fulkerson method conducted in Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication
+  on this formalization except~\cite{Lee05}, which provides a Mizar proof script without any additional comments except that it ``defines and proves correctness of Ford/Fulkerson's Maximum Network-Flow algorithm at the level of graph manipulations.''.
+Moreover, in Lee et al.~\cite{LeRu07}, which is about graph representation in Mizar, the formalization is shortly mentioned, and it is clarified that it does not provide any implementation or data structure formalization.
+  As far as we could understand the Mizar proof scripts, it provides an abstract formalization of the Edmonds/Karp algorithm, including breadth first search.
+Termination is only proved for the case of integer valued capacities, and no implementation is provided.
 
-  Apart from our own work~\cite{La14,NoLa12}, there are several other verifications of graph algorithms, using different techniques. 
-  Top-Down
-  Bottom-Up: Noschinski verifies a checker for (non-) planarity certificates using a bottom-up approach. Starting at a C implementation,
-  the AutoCorres tool~\cite{AutoCorres. What's the standard paper on it?} generates a monadic representation of the program in Isabelle. Further abstractions are applied
-  to this representation to hide low-level details like pointer manipulations and fixed size integers~\cite{AutoCorres-Mind-the-Gap}. Finally, a verification condition
+  Apart from our own work~\cite{La14,NoLa12}, there are several other verifications of graph algorithms and their implementations, using different techniques and proof assistants.
+Noschinski~\cite{Nosch15} verifies a checker for (non-)planarity certificates using a bottom-up approach.
+Starting at a C implementation,
+  the AutoCorres tool~\cite{Greenaway15,GAK12} generates a monadic representation of the program in Isabelle. Further abstractions are applied
+  to this representation to hide low-level details like pointer manipulations and fixed size integers. Finally, a verification condition
   generator is used to prove the abstracted program correct.
-Note that approach is the opposite direction of ours: While they start at a concrete version of the algorithm and use abstraction steps to eliminate implementation details, we start at an abstract version, and use concretization steps to introduce implementation details.
+Note that their approach takes the opposite direction than ours: While they start at a concrete version of the algorithm and use abstraction steps to eliminate implementation details, we start at an abstract version, and use concretization steps to introduce implementation details.
 
-  Chargu\'eraud also uses a bottom-up approach~\cite{char11} to verify imperative programs written in a subset of OCaml, amongst them a version of Dijkstra's algorithm:
-  A verification condition generator generates a \emph{characteristic formula}, which reflects the semantics of the program in the logic of the proof assistant~\cite{}.
+  Chargu\'eraud~\cite{char11} also uses a bottom-up approach to verify imperative programs written in a subset of OCaml, amongst them a version of Dijkstra's algorithm:
+  A verification condition generator generates a \emph{characteristic formula}, which reflects the semantics of the program in the logic of the proof assistant.
 
   \subsection{Future Work}
-  Future work includes the formalization of more advanced maximum flow algorithms, like Dinic's algorithm~\cite{Di06} or push-relabel algorithms~\cite{GoTa88}.
-  Both, formalizing the abstract theory, and developing efficient implementations will pose some challenges.
+  Future work includes the optimization of our implementation, and the formalization of more advanced maximum flow algorithms, like Dinic's algorithm~\cite{Di06} or push-relabel algorithms~\cite{GoTa88}.
+  We expect both, formalizing the abstract theory and developing efficient implementations, to be challenging but realistic tasks.
 

### 2016-03-04 paper/root.tex
modified diagram

@@
         xlabel = Number of vertices in the network,
-        ylabel = Execution time in milliseconds,
-      xticklabels={$0$, $1000$,$2000$,$3000$,$4000$},
+        ylabel = Execution time in seconds, % in milliseconds,
+        xtick = data,
+        ytick = data,
+        xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$},
+        yticklabels = {$0.2$, $1.4$, $4.8$, $9.3$, $23.4$},
         smooth,

### 2016-03-04 paper/root.tex
modified diagram

--- 2016-03-03 paper/root.tex
@@
         xlabel = Number of vertices in the network,
-        ylabel = Execution time in milliseconds,
-      xticklabels={$0$, $1000$,$2000$,$3000$,$4000$},
+        ylabel = Execution time in seconds, %in milliseconds, modify the scale for lables
+        xtick = data,
+        ytick = data,
+        xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$}, %corresponds to the first column in {SML, Sparse}
+        yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
         smooth,

### 2016-03-04 paper/root.tex
Reza's mods

@@
 an expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
-% 
-% 
-% so it is easy to read even for those who 
-% are not expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
-
 We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
@@
 \section{Introduction}
-The maximum flow problem is an important problem in graph theory, as many other problems like the maximum-bipartite-matching problem, the edge-disjoint-paths problem, the circulation-demand problem, and various scheduling and resource allocating problems can be reduced to it.
+Computing the maximum flow of a network is an important problem in graph theory.
+Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, and various scheduling and resource allocating problems can be reduced to it.
 The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
@@
 
-An flow $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges.
+An \emph{$s$-$t$ flow} $f$ on a graph is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges.
+In context of a network, a \emph{flow} is an $s$-$t$ flow, \ie a flow from the source to the sink.
 
-The value of flow $f$ is denoted by $|f|$, and defined to be the sum over the flows of all outgoing edges of the source\footnote{Note that we assumed that the source has no incoming edges.
-In a more general setting, the value of a flow is the sum over all flows leaving $s$ minus the sum over all incoming flows to $s$.}.
+The value of an $s$-$t$ flow $f$ is denoted by $|f|$, and defined to be the sum over all flows leaving $s$ minus the sum over all incoming flows to $s$\footnote{In the context of networks, there are no incoming edges to $s$, such the definition can be simplified.}.
 Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.
@@
 In order to increase the flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges.
-For this purpose, the Ford-Fulkerson method defines the residual graph.
-It has edges in parallel and reverse to the edges of the network, each edge indicating how much flow can be moved between two nodes by either increasing or decreasing the flow on an edge in the network.
+For this purpose the Ford-Fulkerson method defines the residual graph, which has edges in parallel and reverse to the edges of the network.
+Each edge indicates how much flow can be moved between two nodes by either increasing or decreasing the flow on an edge in the network.
 Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the
@@
 
-TODO: Introduce obtain-syntax, and unfold find-augmenting-path spec!
+TODO: Introduce obtain-syntax, and unfold find-augmenting-path spec! Or even while-exists loop as combinator?
 The code looks quite similar to pseudocode that one would expect in a textbook, but actually is a rigorous formal specification of the algorithm, using nondeterminism to leave open the implementation details (\cf Section~\ref{sec:refinement}).
-The only caveat is that our framework currently does not support breaking of loops, such that we had to use an explicit flag instead.
+Note that we had to use the available combinators of the refinement framework, which made the
+code slightly more verbose than we would have liked.
+
+% The only caveat is that our framework currently does not support breaking of loops, such that we had to use an explicit flag instead.
 
@@
 The development of the theories required roughly 3 man month, a significant amount of this time going into a first, purely functional version of the implementation, which was later dropped in favor of the faster imperative version.
-  TODO: Reza, is this estimate OK?
 
@@
 Moreover, in Lee et al.~\cite{LeRu07}, which is about graph representation in Mizar, the formalization is shortly mentioned, and it is clarified that it does not provide any implementation or data structure formalization.
-  As far as we could understand the Mizar proof scripts, it provides an abstract formalization of the Edmonds/Karp algorithm, including breadth first search.
+  As far as we could understand the Mizar proof scripts, it formalizes an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson method.
   Termination is only proved for the case of integer valued capacities, and no implementation is provided.

### 2016-03-05 paper/root.tex
Added missing snippets, latex fine-tuning, makefile-setup for snippets
The latex fine tuning is done via a sed-script that processes the generated snippet file. This is a hack, and can be achieved in a cleaner way on isabelle level, by defining output syntax, see Library/LaTeXSugar for examples.

The makefile intentionally does not contain a dependence of the snippets to the Isabelle-theories, to keep it decoupled. To force re-buld of snippets do a
  make clean_snippets
first.
Fixed size problems with listings, restored orig style
2nd iteration over paper
changed presentation of networks and f

@@
 
+% \isabellestyle{it}
+
+\def\isachardoublequote{}%
+\def\isachardoublequoteopen{}%
+\def\isachardoublequoteclose{}%
+
+\newcommand{\isainnerkeyword}[1]{{\tt #1}}
+\newcommand{\isasymexistsA}{\isamath{\exists_{\textsc A}\,}}
+
 \def\isadelimproof{}
@@
 
-\input{macros}
-
 \input{lstisabelle}
-\lstset{basicstyle=\small\ttfamily\slshape,breaklines=true}
+\lstset{basicstyle=\footnotesize\ttfamily\slshape}
 \lstset{captionpos=b}
 \lstset{numberbychapter=false}
+
+\newcommand{\isai}{\lstinline[language=isabelle,basicstyle=\normalsize\ttfamily\slshape]}
+
+\input{macros}
 
@@
      \end{quote}}}
-\newcommand{\Snippet}[1]{\csname snippet--#1\endcsname}
-\input{snippets/snippets}
+\newcommand{\Snippet}[1]{\ifcsname snippet--#1\endcsname\csname snippet--#1\endcsname\else\PackageError{}{No snippet '#1' defined.}{}\fi}
+\input{snippets/snippets.out}
 
@@
 We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
-Our formal proof closely follows a standard textbook proof, and is easy to read even without being
+Our formal proof closely follows a standard textbook proof, and is accessible even without being
 an expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
 We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
-Further refinement yields a verified implementation, which compares well to an unverified reference implementation in Java.
+Further refinement yields a verified implementation, whose execution time compares well to an unverified reference implementation in Java.
 
@@
 The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
-which solves the maximum flow problem in polynomial time for the general case of real valued capacities.
+which was one of the first algorithms to solve the maximum flow problem in polynomial time for the general case of real valued capacities.
 
-In this paper, we present a verified efficient implementation of the Edmonds-Karp algorithm and its polynomial complexity bound.
+In this paper, we present a formal verification of the Edmonds-Karp algorithm and its polynomial complexity bound.
 The formalization is conducted entirely in the Isabelle/HOL proof assistant~\cite{NPW02}. 
@@
 Being developed in the Isar~\cite{Wenzel99} proof language, our proofs are accessible even to non-Isabelle experts.
+
+While there exists another formalization of the Ford-Fulkerson method~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation.
 
@@
 benchmarking our implementation against a reference implementation of the Edmonds-Karp algorithm from Sedgewick et al.~\cite{SeWa11}. 
-Finally, Section~\ref{sec:concl} gives a conclusion and discusses related and future work.
+Finally, Section~\ref{sec:concl} gives a conclusion and discusses related and future work. The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/edmonds_karp/}.
 
@@
 %         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
-%         Can transform any Network to match these assumptions, preserving the flow \cite{???}.
+%         Can transform any Network to match these assumptions, preserving the flow cite{???}.
 %         
@@
 
+% TODO: Possible alternative presentation
+%   First define a relaxed network, without the constraints. Define the flow there. Then add network constraints.
+%
+
 A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real valued capacity $c(u,v)>0$.
-Moreover, their are two distinct vertices $s,t\in E$ which are called \emph{source} and \emph{sink}.
-To simplify reasoning about flow networks, it is common to make some additional assumptions: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no edge $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$.
-Although these assumptions seem restrictive, any network may be transformed such that it satisfies all the aforementioned properties~\cite{CLRS09}.
+Moreover, there are two distinct vertices $s,t\in V$ which are called \emph{source} and \emph{sink}.
 
-An \emph{$s$-$t$ flow} $f$ on a graph is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges.
-In context of a network, a \emph{flow} is an $s$-$t$ flow, \ie a flow from the source to the sink.
-
-The value of an $s$-$t$ flow $f$ is denoted by $|f|$, and defined to be the sum over all flows leaving $s$ minus the sum over all incoming flows to $s$\footnote{In the context of networks, there are no incoming edges to $s$, such the definition can be simplified.}.
+An \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges.
+The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
 Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.
 
-An important result is the relation between flows and cuts in a network. A \emph{cut} in a network is a partition of the vertices into two sets, such that one set contains 
+To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$.
+Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.
+
+An important result is the relation between flows and cuts in a network. A \emph{cut} is a partitioning of the vertices into two sets, such that one set contains 
 the source and the other set contains the sink. The capacity of a cut is the sum of the capacities of all edges going from the source's side to the sink's side of the cut.
@@
 It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached.
-In order to increase the flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges.
-For this purpose the Ford-Fulkerson method defines the residual graph, which has edges in parallel and reverse to the edges of the network.
-Each edge indicates how much flow can be moved between two nodes by either increasing or decreasing the flow on an edge in the network.
+In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges.
+For this purpose the Ford-Fulkerson method defines the residual graph, which has edges both in parallel and reverse to the edges of the network.
+Each edge is labeled by the amount of flow that can be effectively moved along this edge, by either increasing or decreasing the flow on an edge in the network.
 Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
@@
 In order to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
-by any \emph{augmenting flow} $f'$, i.e., any flow in the residual graph:
+by any \emph{augmenting flow} $f'$, \ie any flow in the residual graph:
 \[ (f \augment f') (u, v) = 
@@
 The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}.
-When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$\footnote{Where unambiguous, we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges.}, even for the general case of real valued capacities.
+When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real valued capacities.
+Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges.
 A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.
@@
 verbose than informal proofs. Moreover, formal proofs are conducted in the tactic language of the proof assistant, which is often some dialect of ML. 
-Thus, many formal proofs are essentially programs that instruct the proof assistant how to conduct the proof, and are inaccessible without a deep knowledge of the used proof assistant, in many cases requiring to replay the proof in the proof assistant in order to understand the idea behind it.
+Thus, many formal proofs are essentially programs that instruct the proof assistant how to conduct the proof.
+They tend to be inaccessible without a deep knowledge of the used proof assistant, in many cases requiring to replay the proof in the proof assistant in order to understand the idea behind it.
 
-For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that strongly resemble standard mathematical textbook proofs, and are accessible even for those not familiar with Isabelle/HOL.
+For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that resemble standard mathematical textbook proofs, and are accessible, to a certain extent, even for those not familiar with Isabelle/HOL.
 We use Isar to present our proof of the Ford-Fulkerson method such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.
@@
 As an example consider the proof that augmenting a flow $f$ with a residual flow $f'$ yields a valid flow $f\augment f'$ again. In particular, one has to show 
-that the augmented flow satisfies the capacity constraint. Cormen et al.~give the following proof, which we copied literally, only replacing the references to 
+that the augmented flow satisfies the capacity constraint. Cormen et al.~give the following proof, which we display literally here, only replacing the references to 
 ``Equation~26.4'' by ``definition of $\augment$'':
@@
 	\begin{align*}
-	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of $\augment$)} \\
+	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
 	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
@@
 	\begin{align*}
-	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of $\augment$)} \\
+	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
 	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
@@
 
-TODO: Introduce obtain-syntax, and unfold find-augmenting-path spec! Or even while-exists loop as combinator?
+% DONE: Introduce obtain-syntax, and unfold find-augmenting-path spec! Or even while-exists loop as combinator?
 The code looks quite similar to pseudocode that one would expect in a textbook, but actually is a rigorous formal specification of the algorithm, using nondeterminism to leave open the implementation details (\cf Section~\ref{sec:refinement}).
-Note that we had to use the available combinators of the refinement framework, which made the
-code slightly more verbose than we would have liked.
+Note that we had to use the available combinators of the Isabelle Refinement Framework, which made the code slightly more verbose than we would have liked.
+We leave it to future work to define a set of combinators and appropriate syntax that allow for more concise presentation of pseudocode.
 
@@
 the Ford-Fulkerson method, which is stated in Isabelle/HOL by the following theorem:
-\Snippet{ford_fulkerson_partial_correct}
+\Snippet{ford_fulkerson_correct}
 
 \section{Refinement in Isabelle/HOL}\label{sec:refinement}
-After having stated and proved correct an algorithm on the abstract level, the next step is to provide an efficient implementation. 
+After having stated and proved correct an algorithm on the abstract level, the next step is to provide an (efficient) implementation. 
 In our case, we first specialize the Ford-Fulkerson method to use shortest augmenting paths, then implement the search for shortest augmenting paths by BFS, and finally use efficient data structures to represent the abstract objects modified by the algorithm.
@@
 
-In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12}. 
+In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12,La12}. 
 It features a refinement calculus for programs phrased in a nondeterminism monad. 
@@
 We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
-Intuitively, $m\le m'$ means that $m$ is a refinement of $m'$, i.e., all possible results of $m$ are 
+Intuitively, $m\le m'$ means that $m$ is a refinement of $m'$, \ie all possible results of $m$ are 
 also possible results of $m'$. 
@@
 Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
-language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define constants for loop constructs (while, foreach), 
-assertions and specifications, and use a Haskell-like do-notation:
+language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define standard loop constructs (while, foreach), 
+a syntax for postcondition specifications, and use a Haskell-like do-notation:
 \begin{lstlisting}
-  assert \<Phi> = if \<Phi> then return () else fail
-  spec x. P x = RES {x. P x}
-  do {x <- m; f x} = bind m f
-  do {m; m'} = bind m (%_. m')
+  spec P == spec x. P x == res {x. P x}
+  do {x <- m; f x} == bind m f
+  do {m; m'} == bind m (%_. m')
 \end{lstlisting}
@@
 means that, if $P$ holds, $m$ does not fail and all possible results of $m$ satisfy $Q$. Note that we provide different recursion constructs
-for partial and total correctness: A nonterminating total correct recursion yields \isai{fail}, while a nonterminating partial correct 
-recursion yields \isai|res {}|, which is the least element of the refinement ordering.
+for partial and total correctness: A nonterminating total correct recursion yields \isai{fail}, which satisfies no specification, even if joined 
+with results from other possible runs. On the other hand, a nonterminating partial correct recursion yields \isai|res {}|, which refines any specification 
+and disappears when joined with other results.
 
@@
 gets \isai{P ==> $m_i$ <= \<Down>$R_i\ldots R_1$ spec Q}, showing the correctness of the refined algorithm.
-Various tools, including a verification condition generator, assist the use in conducting the refinement proofs by breaking 
+If no data refinement is performed, $R_i$ is set to the identity relation, in which case \isai{\<Down>$R_i$} becomes the identity function.
+
+Various tools, including a verification condition generator, assist the user in conducting the refinement proofs by breaking 
 them down to statements that do not involve monad constructs any more, and in many cases precisely reflect the core idea of the proof.
@@
 
-\section{Edmonds-Karp Algorithm}\label{sec:edka}
-  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm requires us to show that finding a shortest augmenting path is a correct refinement of finding any augmenting path, which is straightforward using basic graph theory.
+\section{The Edmonds-Karp Algorithm}\label{sec:edka}
+  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm requires us to show that finding a shortest augmenting path is a refinement of finding any augmenting path, which is straightforward using basic graph theory.
 
-  Considerable more effort is required on showing that the resulting algorithm terminates within $O(VE)$ iterations. 
+  Considerably more effort is required to show that the resulting algorithm terminates within $O(VE)$ iterations. 
   The idea of the proof is as follows: Edges that are reverse to edges on a shortest path cannot lie 
@@
   Formalizing the above intuitive argument was more tricky than it seemed on first glance:
-  While it easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
+  While it is easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
   to a graph transformation which may add reverse edges and removes at least one original edge requires some generalization of the statement.
@@
 \section{Refinement to Executable Code}\label{sec:executable}
-  In the previous sections, we have presented our abstract formalization of the Edmonds-Karp algorithm,
+  In the previous section, we have presented our abstract formalization of the Edmonds-Karp algorithm,
   leaving open how to obtain a shortest augmenting path, and how to implement the algorithm. 
@@
   A standard way to find a shortest path in a graph is breadth first search (BFS).
-Luckily, we had already formalized a BFS algorithm as an example for the refinement framework.
-Unfortunately, the algorithm only computed the minimum distance between two nodes, without returning an actual path.
+Luckily, we had already formalized a BFS algorithm as an example for the Refinement Framework.
+Unfortunately, this algorithm only computed the minimum distance between two nodes, without returning an actual path.
 For this project, we extended the formalization accordingly, and added an efficient imperative implementation, using the same stepwise refinement techniques as for the main algorithm.
-Note that the resulting BFS algorithm is independent, and can be reused to refine shortest path finding in other applications.
+Note that the resulting BFS algorithm is independent, and can be reused for shortest path finding in other applications.
 
@@
   Obviously, computing the complete residual graph each time before searching for an augmenting path is a bad idea. 
-  One solution to this problem is to compute the edges of the residual graph on demand from the network and the current flow.
+  One solution to this problem is to compute the edges of the residual graph on the fly from the network and the current flow.
   Although this solution seems to be common, it has the disadvantage that for each edge of the residual graph, two (or even three) 
@@
   an algorithm that operates on a representation of the residual graph directly. This resulted in a speed-up of factor two.
-
-  The shortest path specification and bottleneck computation are already phrased on residual graphs. 
-  As the residual graph uniquely determines the flow (and vice versa), it is straightforward to phrase the other 
+  As the residual graph uniquely determines the flow (and vice versa), it is straightforward to phrase the
   operations directly on the residual graph. Performing a data refinement of the flow \wrt the refinement 
@@
   In order to find an augmenting path, the BFS algorithm has to compute the successors of a node in the residual graph. 
-  Although this can be implemented on the capacity matrix by iterating over all nodes, this implementation tends to be inefficient for sparse graphs,
+  Although this can be implemented on the edge labeling function by iterating over all nodes, this implementation tends to be inefficient for sparse graphs,
   where we would have to iterate over many possible successor nodes just to find that there is no edge.
@@
 
-  \subsection{Using Efficient Data Structures}\label{sec:impl_datastructures}
-  In a final step, we have to chose efficient data structures for the algorithm. The objects that the algorithm works with are
+  \subsection{Using Efficient Data Structures}\label{sec:impl_data_structures}
+  In a final step, we have to choose efficient data structures for the algorithm. The objects that the algorithm works with are
   capacities, nodes, edges, the graph of the input network, residual graphs, the adjacency map, paths, and the resulting maximum flow.
 
-  We implement capacities as (arbitrary precision) integer numbers. Note that an implementation as fixed precision numbers would also be possible,
+  We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume real, rational, and integer values.}.
+Note that an implementation as fixed precision numbers would also be possible,
   but requires additional checks on the network to ensure that no overflows can happen. 
@@
 
-  The input network of the algorithm is represented as a function mapping network edges to capacities, which is tabulated into an array
-  to obtain the initial residual graph.
-
-  This gives us some flexibility in using the algorithm, as any capacity matrix representation can be converted into a function easily, and 
-  without loosing efficiency for read-access\footnote{Due to technical limitations of our tools, this function 
-  cannot depend on heap content, but it can use the efficient functional and pseudo-functional datastructures provided by the Isabelle Collection Framework.}. 
+  The input network of the algorithm is represented as a function from network edges to capacities, which is tabulated into an array
+  to obtain the initial residual graph. This gives us some flexibility in using the algorithm, as any capacity matrix representation can be converted into a 
+  function easily, and without loosing efficiency for read-access\footnote{Due to technical limitations of our tools, this function 
+  cannot depend on heap content, but it can use the efficient functional and pseudo-functional data structures provided by the Isabelle Collection Framework.}. 
   Similarly, our implementation expects an adjacency map as additional parameter, which is then tabulated into an array.
@@
   A subsequent iteration over this list computes the residual capacity, and a final iteration performs the augmentation. 
-  This calls for a deforestation optimization,
-  getting rid of the intermediate list, and iterate only two times over the predecessor map directly.
+  This calls for a deforestation optimization to get rid of the intermediate list, and iterate only two times over the predecessor map directly.
 %   A first time to compute the bottleneck, and a second time to do the augmentation. 
-  Fortunately, iteration over the shortest path seems not to significantly contribute to the runtime of our implementation, such that we leave this
-  optimization for future work.
+%   Fortunately, iteration over the shortest path seems not to significantly contribute to the runtime of our implementation, such that 
+  We leave this optimization for future work.
 
@@
   and then using Sepref to synthesize the Imperative/HOL programs and their refinement proofs.
-
-  Finally, Imperative/HOL comes with a setup for the Isabelle code generator to generate efficient imperative programs
-  in OCaml, SML, Scala, and Haskell.
+  Finally, Imperative/HOL comes with a setup for the Isabelle code generator~\cite{Haft09,HaNi10} to generate imperative programs in OCaml, SML, Scala, and Haskell.
 
@@
 
-  Combining the implementation of the Edmonds-Karp algorithm with the network checker yields our final algorithm,
-  for which we can export code, and have proved the theorem: 
-  \Snippet{edmonds-karp-correct}
-%     theorem edmonds_karp_correct:
-%     "<emp> edmonds_karp el s t <λ
-%         None ⇒ ↑(¬ln_invar el ∨ ¬Network (ln_α el) s t)
-%       | Some (N,fi) ⇒ ∃⇩Af. Network.is_rflow (ln_α el) N f fi * ↑(isMaxFlow (ln_α el) s t f)
-%           * ↑(ln_invar el ∧ Network (ln_α el) s t ∧ Graph.V (ln_α el) ⊆ {0..<N})
-%     >⇩t"
+  Combining the implementation of the Edmonds-Karp algorithm with the network checker yields our final implementation,
+  for which we can export code, and have proved the following theorem: 
+  \Snippet{edmonds_karp_correct}
   Note that this theorem is stated as a Hoare triple, using separation logic assertions. There are no preconditions on the input.
   If the algorithm returns \isai{None}, then the edge list was malformed or described a graph that does not satisfy the network assumptions.
-  Here, \isai{ln_invar} describes well-formed edge lists, i.e., edge lists that have no duplicate edges and only edges with positive capacity,
+  Here, \isai{ln_invar} describes well-formed edge lists, \ie edge lists that have no duplicate edges and only edges with positive capacity,
   and \isai{ln_\<alpha>} describes the mapping from (well-formed) edge lists to capacity matrices.
-  If the algorithm returns some number $N$ and residual graph $fi$, then the input was a well-formed edge list that describes a valid network with at most $N$ nodes.
-  Moreover, the returned residual graph describes a flow $f$ in the network, which is maximal. TODO: Swap in thm!
+  If the algorithm returns some number $N$ and residual graph $cf$, then the input was a well-formed edge list that describes a valid network with at most $N$ nodes.
+  Moreover, the returned residual graph describes a flow $f$ in the network, which is maximal. 
   As there are no other possible return values, this theorem completely describes the correctness of the algorithm. 
-  Note that Isabelle/HOL does not have a notion of execution, thus total correctness of the code finally produced by the code 
-  generator cannot be defined. However, the program is phrased in a heap-exception monad, thus introducing some (coarse grained) notion of
+  Note that Isabelle/HOL does not have a notion of execution, thus total correctness of the generated code cannot be expressed.
+However, the program is phrased in a heap-exception monad, thus introducing some (coarse grained) notion of
   computation. On this level, termination can be ensured, and, indeed, the above theorem implies that all the recursions stated by recursion 
-  combinators in the monad must terminate. However, it does not guarantee that we have not injected spurious code equations: E.g. $f~x = f~x$,
-  is provable by reflexivity, but causes the generated program to diverge.
+  combinators in the monad must terminate. However, it does not guarantee that we have not injected spurious code equations like $f~x = f~x$, 
+  which is provable by reflexivity, but causes the generated program to diverge.
 
@@
   We have compared the running time of our algorithm against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}.
-  We have done the comparison on randomly generated sparse and dense networks, the sparse networks having a \emph{density} ($\frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
+  We have done the comparison on randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
 Note that the maximum density for networks that satisfy our assumptions is $0.5$, as we allow no parallel edges.
@@
         5000 23625
-      }; \addlegendentry{SML, Sparse};
+      }; \addlegendentry{SML, sparse};
 
@@
         5000 4868
-      }; \addlegendentry{Java, Sparse};
+      }; \addlegendentry{Java, sparse};
 
@@
         1400 17317
-      }; \addlegendentry{SML, Dense};
+      }; \addlegendentry{SML, dense};
 
@@
         1400 12156
-      }; \addlegendentry{Java, Dense};
+      }; \addlegendentry{Java, dense};
     %end data for plotting
@@
   operates on residual graphs (\cf Section~\ref{sec:impl_res_graph}). Moreover, the Java implementation does not store the augmenting 
-  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_datastructures}).
+  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).
 
@@
   but roughly one order of magnitude faster for dense graphs. We leave it to future work to investigate this issue, and conclude that we were able to produce
-  a reasonably fast verified implementation, without doing any low-level optimizations, but relying on the code generated by the Isabelle/HOL code generator.
+  a reasonably fast verified implementation, without doing any low-level optimizations, but using the code as generated by the Isabelle/HOL code generator.
 
@@
 \section{Conclusion}\label{sec:concl}
-  In this paper, we have presented a verification of the Ford-Fulkerson method, using a stepwise refinement approach.
-  Starting with a proof of the Ford-Fulkerson theorem, we have verified the generic Ford-Fulkerson method, then 
-  specialized it to the Edmonds-Karp algorithm and proved the upper bound $O(VE)$ for the number of outer loop iterations.
+  We have presented a verification of the Edmonds-Karp algorithm, using a stepwise refinement approach.
+  Starting with a proof of the Ford-Fulkerson theorem, we have verified the generic Ford-Fulkerson method, 
+  specialized it to the Edmonds-Karp algorithm, and proved the upper bound $O(VE)$ for the number of outer loop iterations.
   We then conducted several refinement steps to derive an efficiently executable implementation of the algorithm, 
@@
   implementation. Moreover, modularity of refinement allowed us to develop the breadth first search algorithm independently, and later link it to the 
-  main algorithm. The BFS algorithm can be reused as building block for other algorithms. Also the data structures are re-usable. 
-  Although we had to implement the array representation of (capacity) matrices for this project, it will be added to the growing library of data structures 
+  main algorithm.
+The BFS algorithm can be reused as building block for other algorithms.
+The data structures are re-usable, too: although we had to implement the array representation of (capacity) matrices for this project, it will be added to the growing library of data structures
   supported by the Sepref tool, such that it can be re-used for future formalizations.
@@
   We conclude with some statistics:
-  The formalization consists of roughly 8000 lines of proof text, where the graph theory up to the Ford-Fulkerson theorem required 2500 lines.
+  The formalization consists of roughly 8000 lines of proof text, where the graph theory up to the Ford-Fulkerson algorithm required 2600 lines.
   The abstract Edmonds-Karp algorithm and its complexity analysis contributes 900 lines, and its implementation (including BFS) another 1500 lines.
@@
 
-  \subsection{Related Work}
+  \subsection{Related Work}\label{sec:related_work}
   We are only aware of one other formalization of the Ford-Fulkerson method conducted in Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication
-  on this formalization except~\cite{Lee05}, which provides a Mizar proof script without any additional comments except that it ``defines and proves correctness of Ford/Fulkerson's Maximum Network-Flow algorithm at the level of graph manipulations.''.
+  on this formalization except~\cite{Lee05}, which provides a Mizar proof script without any additional comments except that it ``defines and proves correctness of Ford/Fulkerson's Maximum Network-Flow algorithm at the level of graph manipulations''.
 Moreover, in Lee et al.~\cite{LeRu07}, which is about graph representation in Mizar, the formalization is shortly mentioned, and it is clarified that it does not provide any implementation or data structure formalization.
   As far as we could understand the Mizar proof scripts, it formalizes an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson method.
-  Termination is only proved for the case of integer valued capacities, and no implementation is provided.
+  Termination is only proved for integer valued capacities.
 
@@
   Future work includes the optimization of our implementation, and the formalization of more advanced maximum flow algorithms, like Dinic's algorithm~\cite{Di06} or push-relabel algorithms~\cite{GoTa88}.
-  We expect both, formalizing the abstract theory and developing efficient implementations, to be challenging but realistic tasks.
+  We expect both formalizing the abstract theory and developing efficient implementations, to be challenging but realistic tasks.
 

### 2016-03-05 paper/root.tex
Fixed layout problem with sample proof
Fixed matrix order

@@
   We implement nodes as natural numbers less than an upper bound $N$, and residual graphs are implemented by their capacity matrices, which, in turn,
-  are realized as arrays of size $N*N$ with row-major indexing.
+  are realized as arrays of size $N*N$ with row-major indexing, such that the successors of a node are close together in memory.
   The adjacency map of the network is implemented as an array of lists of nodes. 

### 2016-03-06 paper/root.tex
Not measuring initial tabulation on SML benchmark any more
3rd iteration over paper
Added introduction to "formalizing Fofu" section
Added "implementing augmentation" subsection
Updated benchmarks. Diagrams now next to each other. (I'm not yet completely satisfied with the result)

@@
 
-\author{S.~Reza Sefidgar, Peter Lammich}
+\author{Peter Lammich, S.~Reza Sefidgar}
 
-\institute{Technische Universit\"at M\"unchen, \email{\{sefidgar,lammich\}@in.tum.de}}
+\institute{Technische Universit\"at M\"unchen, \email{\{lammich,sefidgar\}@in.tum.de}}
 
@@
 Computing the maximum flow of a network is an important problem in graph theory.
-Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, and various scheduling and resource allocating problems can be reduced to it.
+Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, as well as various scheduling and resource allocating problems can be reduced to it.
 The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
@@
 
-While there exists another formalization of the Ford-Fulkerson method~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation.
+While there exists another formalization of the Ford-Fulkerson method in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation.
+Moreover, this paper is a case study on elegantly formalizing algorithms.
 
@@
 
-To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$.
+To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the flow network must be on a path from $s$ to $t$.
 Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.
@@
 For this purpose the Ford-Fulkerson method defines the residual graph, which has edges both in parallel and reverse to the edges of the network.
-Each edge is labeled by the amount of flow that can be effectively moved along this edge, by either increasing or decreasing the flow on an edge in the network.
+Each edge is labeled by the amount of flow that can be effectively moved along this edge, by either increasing or decreasing the flow on a network edge.
 Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
@@
 When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real valued capacities.
-Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges.
+Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges if the intended meaning is clear from the context.
 A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.
 
-\section{Formalizing the Ford-Fulkerson method}\label{sec:abs-formalization}
+\section{Formalizing the Ford-Fulkerson Method}\label{sec:abs-formalization}
 %  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
@@
 
+In this section, we provide a brief overview of our formalization of the Ford-Fulkerson method. 
+In order to develop theory in the context of a fixed graph or network, we use Isabelle's concept of \emph{locales}~\cite{Ballarin:2006:MKM}, which allows
+us to define named contexts that fix some parameters and assumptions.
+For example, the graph theory is developed in the locale \isai{Graph}, which 
+fixes the edge labeling function \isai{c}, and defines the set of edges and nodes based on $c$:
+\begin{lstlisting}
+  locale Graph = fixes c :: "edge => 'capacity" begin
+    definition "E == {(u, v). c (u, v) ~= 0}"
+    definition "V == {u. \<exists>v. (u, v) \<in> E \<or> (v, u) \<in> E}"
+    [...]
+\end{lstlisting}
+Moreover, we define basic concepts like (simple, shortest) paths, and provide lemmas to reason about them.
+
+Networks are based on graphs, and add the source and sink node, as well as the network assumptions:
+\begin{lstlisting}
+  locale Network = Graph + fixes s t :: node
+    assumes no_incoming_s: "\<forall>u. (u, s) \<notin> E"
+    [...]
+\end{lstlisting}
+Most theorems presented in this paper are in the context of the \isai{Network} locale.
+
+\subsection{Presentation of Proofs}
 % (DONE: not sure about repeating citation of Cormen: I would definitely repeat the citation!) 
@@
 
+\subsection{Presentation of Algorithms}
 In textbooks, it is common to present algorithms in pseudocode, which captures the essential ideas, but leaves open implementation details.
-As a formal equivalent to pseudocode, we use the monadic programming language provided by the Isabelle Refinement Framework~\cite{LaTu12,La12}:
+As a formal equivalent to pseudocode, we use the monadic programming language provided by the Isabelle Refinement Framework~\cite{LaTu12,La12}.
+For example, we define the Ford-Fulkerson method as follows:
 \Snippet{ford_fulkerson_algo}
@@
 Note that we had to use the available combinators of the Isabelle Refinement Framework, which made the code slightly more verbose than we would have liked.
-We leave it to future work to define a set of combinators and appropriate syntax that allow for more concise presentation of pseudocode.
+We leave it to future work to define a set of combinators and appropriate syntax that allows for more concise presentation of pseudocode.
 
@@
 The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
-operation \isai{bind m f} nondeterministically picks a result from $m$ and executes $f$ on it. 
-If either \isai{m = fail}, or \isai{f} may fail for a result in $m$, the bind operation fails.
+operation \isai{bind m f} nondeterministically picks a result from \isai{m} and executes \isai{f} on it. 
+If either \isai{m = fail}, or \isai{f} may fail for a result in \isai{m}, the bind operation fails.
 
 We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
-Intuitively, $m\le m'$ means that $m$ is a refinement of $m'$, \ie all possible results of $m$ are 
-also possible results of $m'$. 
+Intuitively, \isai{m<=m'} means that \isai{m} is a refinement of \isai{m'}, \ie all possible results of \isai{m} are 
+also possible results of \isai{m'}. 
 Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
@@
 
-Correctness of a program $m$ with precondition $P$ and postcondition $Q$ is expressed as \isai{P ==> m <= spec r. Q r} (or, eta-contracted, just \isai{spec Q}), which
-means that, if $P$ holds, $m$ does not fail and all possible results of $m$ satisfy $Q$. Note that we provide different recursion constructs
+Correctness of a program \isai{m} with precondition \isai{P} and postcondition \isai{Q} is expressed as \isai{P ==> m <= spec r. Q r} (or, eta-contracted, just \isai{spec Q}), which
+means that, if \isai{P} holds, \isai{m} does not fail and all possible results of \isai{m} satisfy \isai{Q}. Note that we provide different recursion constructs
 for partial and total correctness: A nonterminating total correct recursion yields \isai{fail}, which satisfies no specification, even if joined 
@@
 The Isabelle Refinement Framework also supports data refinement, changing the representation of results according to a \emph{refinement relation}, 
-which relates concrete with abstract results: Given a relation $R$, \isai{\<Down>R m} is the largest set of concrete results that are related to an 
-abstract result in $m$ by $R$. If \isai{m=fail}, then also \isai{\<Down>R m = fail}.
+which relates concrete with abstract results: Given a relation \isai{R}, \isai{\<Down>R m} is the largest set of concrete results that are related to an 
+abstract result in \isai{m} by \isai{R}. If \isai{m=fail}, then also \isai{\<Down>R m = fail}.
 
-In a typical program development, one first comes up with an initial version $m_0$ of
-the algorithm and its specification $P,Q$, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions $m_i$ of the algorithm,
-proving \isai{$m_i$ <= \<Down>$R_i$ $m_{i-1}$}. Using transitivity and composability of data refinement, one 
-gets \isai{P ==> $m_i$ <= \<Down>$R_i\ldots R_1$ spec Q}, showing the correctness of the refined algorithm.
-If no data refinement is performed, $R_i$ is set to the identity relation, in which case \isai{\<Down>$R_i$} becomes the identity function.
+In a typical program development, one first comes up with an initial version \isai{m_0} of
+the algorithm and its specification \isai{P,Q}, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions \isai{m$_i$} of the algorithm,
+proving \isai{m$_i$ <= \<Down>R$_i$ m$_{i-1}$}. Using transitivity and composability of data refinement, one 
+gets \isai{P ==> m$_i$ <= \<Down>R$_i$...R$_1$ spec Q}, showing the correctness of the refined algorithm.
+If no data refinement is performed, \isai{R$_i$} is set to the identity relation, in which case \isai{\<Down>R$_i$} becomes the identity function.
 
@@
 \section{The Edmonds-Karp Algorithm}\label{sec:edka}
-  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm requires us to show that finding a shortest augmenting path is a refinement of finding any augmenting path, which is straightforward using basic graph theory.
+  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm requires to show that finding a shortest augmenting path is a refinement of finding any augmenting path, which is straightforward using basic graph theory.
 
@@
   While it is easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
-  to a graph transformation which may add reverse edges and removes at least one original edge requires some generalization of the statement.
+  to a graph transformation which may add multiple reverse edges and removes at least one original edge requires some generalization of the statement.
   Note that a straightforward induction on the length of the augmenting path or on the number of flipped edges fails, as, after flipping the first 
@@
 
-  Having defined the measure function, and shown that it decreases on augmentation, it is straightforward to refine the partial correct while loop to a total correct one.
+  Having defined the measure function and shown that it decreases on augmentation, it is straightforward to refine the partial correct while loop to a total correct one.
 Moreover, to make explicit the bound on the number of loop iterations, we instrument the loop with a counter, and assert the upper bound on the counter after the loop.
@@
   In the previous section, we have presented our abstract formalization of the Edmonds-Karp algorithm,
-  leaving open how to obtain a shortest augmenting path, and how to implement the algorithm. 
+  leaving open how to obtain a shortest augmenting path and how to implement the algorithm. 
   In this section, we outline the further refinement steps that were necessary to obtain an efficient implementation.
@@
 
+  \subsection{Implementing Augmentation}    
+  In our abstract formalization, which matches the presentation in Section~\ref{sec:background}, we have formulated augmentation by first defining 
+  the residual capacity $c_p$ of the augmenting path. Using $c_p$, we have defined the residual flow $f_p$, which was finally added to the current flow.
+  In the refinement to operate on residual graphs, we have refined this to augment the residual graph.
+  For the implementation, we compute the residual capacity in a first iteration over the augmenting path, and modify the residual graph 
+  in a second iteration. Proving this implementation correct is straightforward by induction on the augmenting path.
+
   \subsection{Computing Successors}
@@
 
-  We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume real, rational, and integer values.}.
+  We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers.}.
 Note that an implementation as fixed precision numbers would also be possible,
@@
   to obtain the initial residual graph. This gives us some flexibility in using the algorithm, as any capacity matrix representation can be converted into a 
-  function easily, and without loosing efficiency for read-access\footnote{Due to technical limitations of our tools, this function 
-  cannot depend on heap content, but it can use the efficient functional and pseudo-functional data structures provided by the Isabelle Collection Framework.}. 
+  function easily, without loosing efficiency for read-access. 
   Similarly, our implementation expects an adjacency map as additional parameter, which is then tabulated into an array.
@@
 %   A first time to compute the bottleneck, and a second time to do the augmentation. 
-%   Fortunately, iteration over the shortest path seems not to significantly contribute to the runtime of our implementation, such that 
-  We leave this optimization for future work.
+  Fortunately, iteration over the shortest path seems not to significantly contribute to the runtime of our implementation, such that 
+  we leave this optimization for future work.
 
@@
   Here, \isai{ln_invar} describes well-formed edge lists, \ie edge lists that have no duplicate edges and only edges with positive capacity,
-  and \isai{ln_\<alpha>} describes the mapping from (well-formed) edge lists to capacity matrices.
-  If the algorithm returns some number $N$ and residual graph $cf$, then the input was a well-formed edge list that describes a valid network with at most $N$ nodes.
-  Moreover, the returned residual graph describes a flow $f$ in the network, which is maximal. 
-  As there are no other possible return values, this theorem completely describes the correctness of the algorithm. 
+  and \isai{ln_\<alpha>} describes the mapping from (well-formed) edge lists to capacity matrices (note that we set \isai{c==ln_\<alpha> el}).
+  If the algorithm returns some number \isai{N} and residual graph \isai{cf}, then the input was a well-formed edge list that describes a valid network with at most \isai{N} nodes.
+  Moreover, the returned residual graph describes a flow \isai{f} in the network, which is maximal. 
+  As the case distinction is exhaustive, this theorem states the correctness of the algorithm. 
   Note that Isabelle/HOL does not have a notion of execution, thus total correctness of the generated code cannot be expressed.
@@
   We have compared the running time of our algorithm against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}.
+We have used MLton 20100608~\cite{MLton} and OpenJDK Java 1.7.0\_95, running on a standard laptop machine with a 2.8GHz i7 quadcore processor and 16GiB of RAM.
+
   We have done the comparison on randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
 Note that the maximum density for networks that satisfy our assumptions is $0.5$, as we allow no parallel edges.
-  For sparse networks, we varied the number of nodes between $1000$ and $5000$, for dense networks between $1000$ and $1400$.
+  For sparse networks, we varied the number of nodes between $1000$ and $5500$, for dense networks between $1000$ and $1450$.
   The results are shown in Figure~\ref{fig:benchmark}, in a double-logarithmic scale.
-%   \begin{wrapfigure}{o}{.4\textwidth}
-%   \vspace{-20pt}
-%   \begin{center}
-%   \begin{tabular}{|c|c|c|}
-%   \hline
-%   Name        & SML       & Java   \\
-%   \hline\hline
-%   1000-sparse & 204       & 15     \\
-%   2000-sparse & 1430      & 231    \\
-%   3000-sparse & 4928      & 828    \\
-%   4000-sparse & 9299      & 2145   \\
-%   5000-sparse & 23625     & 4868   \\
-%   1000-dense  & 6575      & 4011   \\
-%   1100-dense  & 8067      & 5487   \\
-%   1200-dense  & 10892     & 7011   \\
-%   1300-dense  & 14070     & 9815   \\
-%   1400-dense  & 17317     & 12156  \\
-%   \hline
-%   \end{tabular}
-%   \end{center}
-%   \vspace{-15pt}
-%   \caption{Dijkstra benchmark}\label{tab:benchmark}
-%   \vspace{-20pt}
-%   \end{wrapfigure}
 
   \begin{figure}[!h]
-  \centering    
-  \begin{tikzpicture}[thick,scale=1, every node/.style={scale=1}] %change the scales if you like to reduce the size
+  \begin{tikzpicture}[thick,scale=.7, every node/.style={scale=1}] %change the scales if you like to reduce the size
     \begin{axis}[
@@
         xlabel = Number of vertices in the network,
-        ylabel = Execution time in seconds, %in milliseconds, modify the scale for lables
-        xtick = data,
-        ytick = data,
+        ylabel = Execution time in milliseconds, %in milliseconds, modify the scale for lables
+        xtick = {0,1000,...,5000},
+%         ytick = data,
         xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$}, %corresponds to the first column in {SML, Sparse}
-        yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
+%         yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
         smooth,
@@
       \addplot table {
-        1000 204
-        2000 1430
-        3000 4928
-        4000 9299
-        5000 23625
+        1000 27   
+        1500 124  
+        2000 299  
+        2500 665  
+        3000 1335 
+        3500 2526 
+        4000 3111 
+        4500 6917 
+        5000 7743 
+        5500 10452    
       }; \addlegendentry{SML, sparse};
@@
       \addplot table {
-        1000 15
-        2000 231
-        3000 828
-        4000 2145
-        5000 4868
+        1000 16   
+        1500 49   
+        2000 168  
+        2500 453  
+        3000 842  
+        3500 1524 
+        4000 2136 
+        4500 3438 
+        5000 4734 
+        5500 6369     
       }; \addlegendentry{Java, sparse};
+    %end data for plotting
+    \end{axis}
+  \end{tikzpicture}
+%
+  \begin{tikzpicture}[thick,scale=.7, every node/.style={scale=1}] %change the scales if you like to reduce the size
+    \begin{axis}[
+      %title={Benchmark of the Edmonds-Karp algorithm},
+        axis x line*= bottom,
+        axis y line*= left,
+        xmode = log,
+        ymode = log,
+        xlabel = Number of vertices in the network,
+        ylabel = Execution time in milliseconds, %in milliseconds, modify the scale for lables
+        xtick = {0,1000,1100,...,1400},
+        ytick = {0,100,1000,10000},
+        xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, %corresponds to the first column in {SML, Sparse}
+%         yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
+        smooth,
+        cycle list name = black white,
+        legend style = { 
+          at={(0.59,0.4)}, 
+            anchor=north west,
+            draw=black, 
+            fill= white,
+            align=left
+        }
+    ]
 
+    %start data for plotting
       \addplot table {
-        1000 6575
-        1100 8067
-        1200 10892
-        1300 14070
-        1400 17317
+        1000 3328 
+        1050 3345 
+        1100 4841 
+        1150 5036 
+        1200 5840 
+        1250 6858 
+        1300 7651 
+        1350 8950 
+        1400 8398 
+        1450 10769     
       }; \addlegendentry{SML, dense};
@@
       \addplot table {
-        1000 4011
-        1100 5487
-        1200 7011
-        1300 9815
-        1400 12156
+        1000 4141 
+        1050 4778 
+        1100 5465 
+        1150 6248 
+        1200 7128 
+        1250 8137 
+        1300 9864 
+        1350 11008
+        1400 11902
+        1450 13365     
       }; \addlegendentry{Java, dense};
     %end data for plotting
-
     \end{axis}
   \end{tikzpicture}
+
+%   \centering    
+%   \begin{tikzpicture}[thick,scale=1, every node/.style={scale=1}] %change the scales if you like to reduce the size
+%     \begin{axis}[
+%       %title={Benchmark of the Edmonds-Karp algorithm},
+%         axis x line*= bottom,
+%         axis y line*= left,
+%         xmode = log,
+%         ymode = log,
+%         xlabel = Number of vertices in the network,
+%         ylabel = Execution time in milliseconds, %in milliseconds, modify the scale for lables
+%         xtick = {0,1000,...,5000},
+% %         ytick = data,
+%         xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$}, %corresponds to the first column in {SML, Sparse}
+% %         yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
+%         smooth,
+%         cycle list name = black white,
+%         legend style = { 
+%           at={(0.59,0.4)}, 
+%             anchor=north west,
+%             draw=black, 
+%             fill= white,
+%             align=left
+%         }
+%     ]
+%     
+%     %start data for plotting
+%       \addplot table {
+%         1000 27   
+%         1500 124  
+%         2000 299  
+%         2500 665  
+%         3000 1335 
+%         3500 2526 
+%         4000 3111 
+%         4500 6917 
+%         5000 7743 
+%         5500 10452    
+%       }; \addlegendentry{SML, sparse};
+% 
+%       \addplot table {
+%         1000 16   
+%         1500 49   
+%         2000 168  
+%         2500 453  
+%         3000 842  
+%         3500 1524 
+%         4000 2136 
+%         4500 3438 
+%         5000 4734 
+%         5500 6369     
+%       }; \addlegendentry{Java, sparse};
+% 
+%       \addplot table {
+%         1000 3328 
+%         1050 3345 
+%         1100 4841 
+%         1150 5036 
+%         1200 5840 
+%         1250 6858 
+%         1300 7651 
+%         1350 8950 
+%         1400 8398 
+%         1450 10769     
+%       }; \addlegendentry{SML, dense};
+% 
+%       \addplot table {
+%         1000 4141 
+%         1050 4778 
+%         1100 5465 
+%         1150 6248 
+%         1200 7128 
+%         1250 8137 
+%         1300 9864 
+%         1350 11008
+%         1400 11902
+%         1450 13365     
+%       }; \addlegendentry{Java, dense};
+%     %end data for plotting
+%     \end{axis}
+%   \end{tikzpicture}
+
   \caption{Edmonds-Karp benchmark}\label{fig:benchmark}
@@
 
-  We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of 6, while for dense graphs, our implementation comes quite close 
-  to the Java implementation. We have no full explanation for this phenomenon. Note that the Java implementation operates on flows, while our implementation 
+  We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of $1.6$, while for dense graphs, our implementation is faster by a factor of $1.2$.
+Note that the Java implementation operates on flows, while our implementation
   operates on residual graphs (\cf Section~\ref{sec:impl_res_graph}). Moreover, the Java implementation does not store the augmenting 
@@
   but roughly one order of magnitude faster for dense graphs. We leave it to future work to investigate this issue, and conclude that we were able to produce
-  a reasonably fast verified implementation, without doing any low-level optimizations, but using the code as generated by the Isabelle/HOL code generator.
+  a reasonably fast verified implementation.
+
+%   , without doing any low-level optimizations, but using the code as generated by the Isabelle/HOL code generator.
 
@@
 
-  Our formalization has combined several techniques to achieve an elegant and accessible formalization: The Isar-proof language~\cite{Wenzel99}
-  supports writing formal proofs that closely resemble informal proofs found in textbooks. This way we have created a completely rigorous but 
+  Our formalization has combined several techniques to achieve an elegant and accessible formalization: 
+  Using the Isar proof language~\cite{Wenzel99}, we were able to provide a completely rigorous but 
   still accessible proof of the Ford-Fulkerson theorem. The Isabelle Refinement Framework~\cite{LaTu12,La12} and the Sepref tool~\cite{La15,La16}
-  provide stepwise refinement techniques to formalize an algorithm abstractly, and then refine it, in a modular fashion, towards an efficient implementation. 
-  This allowed us to present the Ford-Fulkerson method on a level 
+  allowed us to present the Ford-Fulkerson method on a level 
   of abstraction that closely resembles pseudocode presentations found in textbooks, and then formally link this presentation to an efficient
@@
 Moreover, in Lee et al.~\cite{LeRu07}, which is about graph representation in Mizar, the formalization is shortly mentioned, and it is clarified that it does not provide any implementation or data structure formalization.
-  As far as we could understand the Mizar proof scripts, it formalizes an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson method.
+  As far as we could understand the Mizar proof script, it formalizes an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson method.
   Termination is only proved for integer valued capacities.
@@
   Chargu\'eraud~\cite{char11} also uses a bottom-up approach to verify imperative programs written in a subset of OCaml, amongst them a version of Dijkstra's algorithm:
-  A verification condition generator generates a \emph{characteristic formula}, which reflects the semantics of the program in the logic of the proof assistant.
+  A verification condition generator generates a \emph{characteristic formula}, which reflects the semantics of the program in the logic of the Coq proof assistant~\cite{BeCa10}.
 
@@
   Future work includes the optimization of our implementation, and the formalization of more advanced maximum flow algorithms, like Dinic's algorithm~\cite{Di06} or push-relabel algorithms~\cite{GoTa88}.
-  We expect both formalizing the abstract theory and developing efficient implementations, to be challenging but realistic tasks.
+  We expect both formalizing the abstract theory and developing efficient implementations to be challenging but realistic tasks.
 

### 2016-03-07 paper/root.tex
Changes of meeting

@@
 Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, as well as various scheduling and resource allocating problems can be reduced to it.
-The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
-which was one of the first algorithms to solve the maximum flow problem in polynomial time for the general case of real valued capacities.
+The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. An important instance is the Edmonds-Karp algorithm~\cite{EK72},
+which was one of the first algorithms to solve the maximum flow problem in polynomial time for the general case of networks with real valued capacities.
 
@@
 Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification into an abstract proof of the Ford-Fulkerson method,
-its instantiation to Edmonds-Karp algorithm, and finally its efficient implementation.
+its instantiation to the Edmonds-Karp algorithm, and finally an efficient implementation.
 The abstract parts of our verification closely follow the standard textbook presentation of Cormen et al.~\cite{CLRS09}.
@@
 
-\section{A short background on the Ford-Fulkerson method}\label{sec:background}
+\section{The Ford-Fulkerson Method}\label{sec:background}
 % \begin{comment}
@@
 
-In this section, we give a short introduction to the Ford-Fulkerson method, closely following the presentation in~\cite{CLRS09}.
+In this section, we give a short introduction to the Ford-Fulkerson method, closely following the presentation by Cormen et al.~\cite{CLRS09}.
 
@@
 
-An \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges.
+A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges.
 The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
@@
 
-To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the flow network must be on a path from $s$ to $t$.
+To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the network must be on a path from $s$ to $t$.
 Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.
@@
 For this purpose the Ford-Fulkerson method defines the residual graph, which has edges both in parallel and reverse to the edges of the network.
-Each edge is labeled by the amount of flow that can be effectively moved along this edge, by either increasing or decreasing the flow on a network edge.
+Each edge is labeled by the amount of flow that can be effectively passed along this edge, by either increasing or decreasing the flow on a network edge.
 Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
@@
 pushes as much flow as possible along this path to increase the value of the current flow. 
-Formally, for an augmenting path $p$, one defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
+Formally, for an augmenting path $p$, one first defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
 \[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
-An augmenting path yields a residual flow $f_p$, which is the flow that can be passed along this path:
+An augmenting path then yields a residual flow $f_p$, which is the flow that can be passed along this path:
 \[ f_p (u, v) = 
@@
 \]
-\newcommand{\augment}{{\mathbin\uparrow}}
-In order to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
+\newcommand{\augment}{{\mathbin\uparrow}}%
+Finally, to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
 by any \emph{augmenting flow} $f'$, \ie any flow in the residual graph:
@@
 
-The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph, and there are several possible implementations with different execution times.
+The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph.
+There are several possible implementations with different execution times.
 The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}.
@@
 \Snippet{augment_flow_presv_cap}
-Its structure is exactly the same as the textbook proof, except that we had to also consider the case $(u,v)\notin E$, which is not mentioned in the informal proof at all, and easily discharged in our formal proof by the \isai{auto}-tactic after the \isai{qed}.
+The structure of the Isar proof is exactly the same as that of the textbook proof, except that we had to also consider the case $(u,v)\notin E$, which is not mentioned in the informal proof at all, and easily discharged in our formal proof by the \isai{auto}-tactic after the \isai{qed}.
 We also use exactly the same justifications as the original proof, except that we had to use the fact that there are no parallel edges to show $c_f(v,u)=f(u,v)$, which is not mentioned in the original proof.
@@
 which relates concrete with abstract results: Given a relation \isai{R}, \isai{\<Down>R m} is the largest set of concrete results that are related to an 
-abstract result in \isai{m} by \isai{R}. If \isai{m=fail}, then also \isai{\<Down>R m = fail}.
+abstract result in \isai{m} by \isai{R}. If \isai{m = fail}, then also \isai{\<Down>R m = fail}.
 
@@
   On every augmentation, at least one edge of the residual graph that lies on a shortest augmenting path is flipped. Thus, either the length of the shortest path increases,
-  or the number of edges that lie on a shortest path decreases. As the length of a shortest path is at most $V$, there are no more than $O(VE)$ iterations.
+  or the number of edges that lie on some shortest path decreases. As the length of a shortest path is at most $V$, there are no more than $O(VE)$ iterations.
+  % TODO: Add more explanation here, like: DIstinguish two cases: 1) ... 2) The distance s-t in cf is still the same, but then ...
 
@@
   \subsection{Network Checker}
-  We also implemented an algorithm that takes as input a list of edges, a source node, and a target node.
+  Additionally, we implemented an algorithm that takes as input a list of edges, a source node, and a target node.
   It converts these to a capacity matrix and an adjacency map, and checks whether the resulting graph satisfies our network assumptions.
@@
   \begin{figure}[!h]
-  \begin{tikzpicture}[thick,scale=.7, every node/.style={scale=1}] %change the scales if you like to reduce the size
+  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
     \begin{axis}[
@@
         ymode = log,
-        xlabel = Number of vertices in the network,
-        ylabel = Execution time in milliseconds, %in milliseconds, modify the scale for lables
+%         xlabel = Number of vertices in the network,
+%         ylabel = Execution time in seconds, 
         xtick = {0,1000,...,5000},
-%         ytick = data,
-        xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$}, %corresponds to the first column in {SML, Sparse}
-%         yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
+        ytick = {0.01,0.1,1,10},
+        xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$},
+        yticklabels = {$0.01$,$0.1$,$1$,$10$}, 
         smooth,
@@
       \addplot table {
-        1000 27   
-        1500 124  
-        2000 299  
-        2500 665  
-        3000 1335 
-        3500 2526 
-        4000 3111 
-        4500 6917 
-        5000 7743 
-        5500 10452    
+        1000 .027   
+        1500 .124  
+        2000 .299  
+        2500 .665  
+        3000 1.335 
+        3500 2.526 
+        4000 3.111 
+        4500 6.917 
+        5000 7.743 
+        5500 10.452    
       }; \addlegendentry{SML, sparse};
@@
       \addplot table {
-        1000 16   
-        1500 49   
-        2000 168  
-        2500 453  
-        3000 842  
-        3500 1524 
-        4000 2136 
-        4500 3438 
-        5000 4734 
-        5500 6369     
+        1000 .016   
+        1500 .049   
+        2000 .168  
+        2500 .453  
+        3000 .842  
+        3500 1.524 
+        4000 2.136 
+        4500 3.438 
+        5000 4.734 
+        5500 6.369     
       }; \addlegendentry{Java, sparse};
@@
 %
-  \begin{tikzpicture}[thick,scale=.7, every node/.style={scale=1}] %change the scales if you like to reduce the size
+  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
     \begin{axis}[
@@
         ymode = log,
-        xlabel = Number of vertices in the network,
-        ylabel = Execution time in milliseconds, %in milliseconds, modify the scale for lables
+%         xlabel = Number of vertices in the network,
+%         ylabel = Execution time in seconds, 
         xtick = {0,1000,1100,...,1400},
-        ytick = {0,100,1000,10000},
-        xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, %corresponds to the first column in {SML, Sparse}
-%         yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
+        ytick = {2,4,...,10},
+        xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, 
+        yticklabels = {$2$, $4$,$6$,$8$,$10$}, 
         smooth,
@@
       \addplot table {
-        1000 3328 
-        1050 3345 
-        1100 4841 
-        1150 5036 
-        1200 5840 
-        1250 6858 
-        1300 7651 
-        1350 8950 
-        1400 8398 
-        1450 10769     
+        1000 3.328 
+        1050 3.345 
+        1100 4.841 
+        1150 5.036 
+        1200 5.840 
+        1250 6.858 
+        1300 7.651 
+        1350 8.950 
+        1400 8.398 
+        1450 10.769     
       }; \addlegendentry{SML, dense};
@@
       \addplot table {
-        1000 4141 
-        1050 4778 
-        1100 5465 
-        1150 6248 
-        1200 7128 
-        1250 8137 
-        1300 9864 
-        1350 11008
-        1400 11902
-        1450 13365     
+        1000 4.141 
+        1050 4.778 
+        1100 5.465 
+        1150 6.248 
+        1200 7.128 
+        1250 8.137 
+        1300 9.864 
+        1350 11.008
+        1400 11.902
+        1450 13.365     
       }; \addlegendentry{Java, dense};
@@
 
-  \caption{Edmonds-Karp benchmark}\label{fig:benchmark}
+%   \center{\footnotesize The x-axis shows the number of nodes, the y axis the execution time in seconds.}
+  \caption{Benchmark of different implementations. The x-axis shows the number of nodes, the y axis the execution time in seconds.}\label{fig:benchmark}
   \end{figure}

### 2016-03-07 paper/root.tex
Added 'lessons learned' section. Minor tuning.

@@
 its instantiation to the Edmonds-Karp algorithm, and finally an efficient implementation.
-The abstract parts of our verification closely follow the standard textbook presentation of Cormen et al.~\cite{CLRS09}.
+The abstract parts of our verification closely follow the textbook presentation of Cormen et al.~\cite{CLRS09}.
 Being developed in the Isar~\cite{Wenzel99} proof language, our proofs are accessible even to non-Isabelle experts.
@@
 A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real valued capacity $c(u,v)>0$.
-Moreover, there are two distinct vertices $s,t\in V$ which are called \emph{source} and \emph{sink}.
+Moreover, there are two distinct vertices $s,t\in V$, which are called \emph{source} and \emph{sink}.
 
@@
 
-To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the network must be on a path from $s$ to $t$.
+To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) the source only has outgoing edges while the sink only has incoming edges; 2) if the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the network must be on a path from $s$ to $t$.
 Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.
@@
 
-In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph, and then 
-pushes as much flow as possible along this path to increase the value of the current flow. 
+In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph.
+It then pushes as much flow as possible along this path to increase the value of the current flow. 
 Formally, for an augmenting path $p$, one first defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
@@
 \begin{lstlisting}
-  locale Graph = fixes c :: "edge => 'capacity" begin
+  locale Graph = fixes c :: "edge => capacity" begin
     definition "E == {(u, v). c (u, v) ~= 0}"
@@
 
-Networks are based on graphs, and add the source and sink node, as well as the network assumptions:
+Networks are based on graphs, and add the source and sink nodes, as well as the network assumptions:
 \begin{lstlisting}
@@
 
-As an example consider the proof that augmenting a flow $f$ with a residual flow $f'$ yields a valid flow $f\augment f'$ again. In particular, one has to show 
+As an example consider the proof that for a flow $f$ and a residual flow $f'$, the augmented flow $f\augment f'$ is again a valid flow. In particular, one has to show 
 that the augmented flow satisfies the capacity constraint. Cormen et al.~give the following proof, which we display literally here, only replacing the references to 
@@
 	\end{align*}
-
+% 
 In the following we present the corresponding formal proof in Isar:
@@
 
-Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Refinement Framework, it is straightforward to prove (partial) correctness of
+Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Isabelle Refinement Framework, it is straightforward to prove (partial) correctness of
 the Ford-Fulkerson method, which is stated in Isabelle/HOL by the following theorem:
@@
 operation \isai{bind m f} nondeterministically picks a result from \isai{m} and executes \isai{f} on it. 
-If either \isai{m = fail}, or \isai{f} may fail for a result in \isai{m}, the bind operation fails.
+The bind operation fails iff either \isai{m = fail}, or \isai{f} may fail for a result in \isai{m},
 
 We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
-Intuitively, \isai{m<=m'} means that \isai{m} is a refinement of \isai{m'}, \ie all possible results of \isai{m} are 
+Intuitively, \isai{m <= m'} means that \isai{m} is a refinement of \isai{m'}, \ie all possible results of \isai{m} are 
 also possible results of \isai{m'}. 
@@
 
-The Isabelle Refinement Framework also supports data refinement, changing the representation of results according to a \emph{refinement relation}, 
-which relates concrete with abstract results: Given a relation \isai{R}, \isai{\<Down>R m} is the largest set of concrete results that are related to an 
+The Isabelle Refinement Framework also supports data refinement. The representation of results can be changed according to a \emph{refinement relation}, 
+which relates concrete with abstract results: Given a relation \isai{R}, \isai{\<Down>R m} is the set of concrete results that are related to an 
 abstract result in \isai{m} by \isai{R}. If \isai{m = fail}, then also \isai{\<Down>R m = fail}.
@@
 Various tools, including a verification condition generator, assist the user in conducting the refinement proofs by breaking 
-them down to statements that do not involve monad constructs any more, and in many cases precisely reflect the core idea of the proof.
+them down to statements that do not contain monad constructs any more. In many cases, these verification conditions reflect the core idea of the proof precisely.
 
@@
   Note that a straightforward induction on the length of the augmenting path or on the number of flipped edges fails, as, after flipping the first 
-  edge, the path is no longer augmenting.
+  edge, the path does no longer exist.
 
@@
   \subsection{Using Breadth First Search}
-  A standard way to find a shortest path in a graph is breadth first search (BFS).
-Luckily, we had already formalized a BFS algorithm as an example for the Refinement Framework.
+  A standard way to find a shortest path in a graph is breadth first search (BFS). Luckily, we had already formalized a BFS algorithm as an example for the 
+  Isabelle Refinement Framework.
 Unfortunately, this algorithm only computed the minimum distance between two nodes, without returning an actual path.
 For this project, we extended the formalization accordingly, and added an efficient imperative implementation, using the same stepwise refinement techniques as for the main algorithm.
-Note that the resulting BFS algorithm is independent, and can be reused for shortest path finding in other applications.
+Note that the resulting BFS algorithm is independent, and can be reused for finding shortest paths in other applications.
 
@@
   \subsection{Manipulating Residual Graphs Directly}\label{sec:impl_res_graph}
-  In a next step, we observe that the algorithm is phrased in terms of a flow which is updated until it is maximal. However,
-  in order to update the flow, the residual graph is used, which is a combination of the current flow and the network.
-  Obviously, computing the complete residual graph each time before searching for an augmenting path is a bad idea. 
+  In a next step, we observe that the algorithm is phrased in terms of a flow, which is updated until it is maximal. 
+  In each iteration, the augmenting path is searched on the residual graph induced by the current flow.
+  Obviously, computing the complete residual graph in each iteration is a bad idea. 
   One solution to this problem is to compute the edges of the residual graph on the fly from the network and the current flow.
@@
   After our profiling indicated a hot spot on accessing the capacity matrices of the network and the flow, we switched to 
-  an algorithm that operates on a representation of the residual graph directly. This resulted in a speed-up of factor two.
+  an algorithm that operates on a representation of the residual graph directly. This resulted in a speed-up of roughly factor two.
   As the residual graph uniquely determines the flow (and vice versa), it is straightforward to phrase the
@@
   a network edge, it is enough to iterate over the adjacent nodes in the network, and check whether they are actual successors in the residual graph.
-  It is straightforward to show that this implementation actually returns the successors in the residual graph.
+  It is straightforward to show that this implementation actually returns the successor nodes in the residual graph.
 
   \subsection{Using Efficient Data Structures}\label{sec:impl_data_structures}
-  In a final step, we have to choose efficient data structures for the algorithm. The objects that the algorithm works with are
-  capacities, nodes, edges, the graph of the input network, residual graphs, the adjacency map, paths, and the resulting maximum flow.
+  In a final step, we have to choose efficient data structures for the algorithm.
+
+  %   The objects that the algorithm works with are
+%   capacities, nodes, edges, the graph of the input network, residual graphs, the adjacency map, paths, and the resulting maximum flow.
 
@@
 
-  Note that we performed the last refinement step using our Sepref tool~\cite{La15,La16}, which provides tool support for data refinements from 
+  Note that we performed the last refinement step using our Sepref tool~\cite{La15,La16}, which provides tool support for refinement from the
   purely functional programs of the Isabelle Refinement Framework into imperative programs expressed in Imperative/HOL~\cite{BKHEM08}.
@@
   for which we can export code, and have proved the following theorem:
+
+  \vspace{\abovedisplayskip}{\raggedright\begin{minipage}{\textwidth}
   \Snippet{edmonds_karp_correct}
-  Note that this theorem is stated as a Hoare triple, using separation logic assertions. There are no preconditions on the input.
+  \end{minipage}}\vspace{\belowdisplayskip}
+
+  \noindent Note that this theorem is stated as a Hoare triple, using separation logic~\cite{Rey02} assertions. There are no preconditions on the input.
   If the algorithm returns \isai{None}, then the edge list was malformed or described a graph that does not satisfy the network assumptions.
   Here, \isai{ln_invar} describes well-formed edge lists, \ie edge lists that have no duplicate edges and only edges with positive capacity,
-  and \isai{ln_\<alpha>} describes the mapping from (well-formed) edge lists to capacity matrices (note that we set \isai{c==ln_\<alpha> el}).
+  and \isai{ln_\<alpha>} describes the mapping from (well-formed) edge lists to capacity matrices (note that we set \isai{c == ln_\<alpha> el}).
   If the algorithm returns some number \isai{N} and residual graph \isai{cf}, then the input was a well-formed edge list that describes a valid network with at most \isai{N} nodes.
@@
   path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).
-
   Finally note that a carefully optimized C++ implementation of the algorithm is only slightly faster than the Java implementation for sparse graphs,
@@
   Finally, we added a verified algorithm to check whether the input is a valid network, and generated executable code in SML.
-  Our implementation compares well to a reference implementation in Java.
+  The runtime of our verified implementation compares well to that of an unverified reference implementation in Java.
 
@@
 The BFS algorithm can be reused as building block for other algorithms.
-The data structures are re-usable, too: although we had to implement the array representation of (capacity) matrices for this project, it will be added to the growing library of data structures
+The data structures are re-usable, too: although we had to implement the array representation of (capacity) matrices for this project, it will be added to the growing library of verified imperative data structures
   supported by the Sepref tool, such that it can be re-used for future formalizations.
 
+  During this project, we have learned some lessons on verified algorithm development: 
+  \begin{itemize}
+  \item It is important to keep the levels of abstraction strictly separated.
+    For example, when implementing the capacity function with arrays, one needs to show that it is only applied to valid nodes.
+    However, proving that, e.g., augmenting paths only contain valid nodes is hard at this low level. 
+    Instead, one can protect the application of the capacity function by an assertion--- already on a high abstraction level where it can be easily discharged. 
+    On refinement, this assertion is passed down, and ultimately available for the implementation.
+    Optimally, one wraps the function together with an assertion of its precondition into a new constant, which is then refined independently.
+  \item Profiling has helped a lot in identifying candidates for optimization. For example, based on profiling data, we decided to delay a 
+    possible deforestation optimization on augmenting paths, and to first refine the algorithm to operate on residual graphs directly.
+  \item ``Efficiency bugs'' are as easy to introduce as for unverified software. For example, out of convenience, we implemented the successor list computation by
+    \emph{filter}. Profiling then indicated a hot-spot on this function. As the order of successors does not matter, we invested a bit more work to make the computation tail 
+    recursive and gained a significant speed-up. Moreover, we realized only lately that we had accidentally implemented and verified
+    matrices with column major ordering, which have a poor cache locality for our algorithm. Changing the order resulted in another significant speed-up.
+  \end{itemize}
+
   We conclude with some statistics:
-  The formalization consists of roughly 8000 lines of proof text, where the graph theory up to the Ford-Fulkerson algorithm required 2600 lines.
-  The abstract Edmonds-Karp algorithm and its complexity analysis contributes 900 lines, and its implementation (including BFS) another 1500 lines.
+  The formalization consists of roughly 8000 lines of proof text, where the graph theory up to the Ford-Fulkerson algorithm requires 3000 lines.
+  The abstract Edmonds-Karp algorithm and its complexity analysis contribute 800 lines, and its implementation (including BFS) another 1700 lines.
   The remaining lines are contributed by the network checker and some auxiliary theories.
@@
 Moreover, in Lee et al.~\cite{LeRu07}, which is about graph representation in Mizar, the formalization is shortly mentioned, and it is clarified that it does not provide any implementation or data structure formalization.
-  As far as we could understand the Mizar proof script, it formalizes an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson method.
+  As far as we understood the Mizar proof script, it formalizes an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson method.
   Termination is only proved for integer valued capacities.
@@
   the AutoCorres tool~\cite{Greenaway15,GAK12} generates a monadic representation of the program in Isabelle. Further abstractions are applied
-  to this representation to hide low-level details like pointer manipulations and fixed size integers. Finally, a verification condition
+  to hide low-level details like pointer manipulations and fixed size integers. Finally, a verification condition
   generator is used to prove the abstracted program correct.

### 2016-03-09 formalization/document/root.tex
Cleanup, building proof document

--- 2016-03-07 paper/root.tex
@@
-% Markierungen: TODO
-\documentclass{llncs}
+\documentclass[11pt,a4paper]{article}
+\usepackage{isabelle,isabellesym}
+\usepackage{amssymb}
+\usepackage[english]{babel}
+\usepackage[only,bigsqcap]{stmaryrd}
+\usepackage{wasysym}
 
-\usepackage{etex}
-\pagestyle{plain} % turn on page numbers
-\usepackage[utf8]{inputenc}
-\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
-\usepackage{amsmath}
-\usepackage{amssymb}
-%\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
-% \usepackage{mathpartir}
-%\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
-\usepackage[boxed]{algorithm}
-\usepackage{enumerate}
-\usepackage{listings}
-\usepackage{graphicx}
-\usepackage{tabularx}
-\usepackage{booktabs}
-\usepackage{color}
-\usepackage[noend]{algpseudocode}
-\usepackage{caption}
-\usepackage[font=scriptsize]{subcaption}
-\usepackage{hyperref}
-\usepackage{float}
-\usepackage{wrapfig}
-\usepackage{multirow}
-\usepackage{pgfplots}
+% this should be the last package used
+\usepackage{pdfsetup}
 
-\usepackage{packages/isabelle}
-\usepackage{packages/isabelletags}
-\usepackage{packages/isabellesym}
-\usepackage{packages/comment}
+% urls in roman style, theory text in math-similar italics
+\urlstyle{rm}
+\isabellestyle{it}
 
-% \isabellestyle{it}
+% Tweaks
+\newcounter{TTStweak_tag}
+\setcounter{TTStweak_tag}{0}
+\newcommand{\setTTS}{\setcounter{TTStweak_tag}{1}}
+\newcommand{\resetTTS}{\setcounter{TTStweak_tag}{0}}
+\newcommand{\insertTTS}{\ifnum\value{TTStweak_tag}=1 \ \ \ \fi}
 
-\def\isachardoublequote{}%
-\def\isachardoublequoteopen{}%
-\def\isachardoublequoteclose{}%
+\renewcommand{\isakeyword}[1]{\resetTTS\emph{\bf\def\isachardot{.}\def\isacharunderscore{\isacharunderscorekeyword}\def\isacharbraceleft{\{}\def\isacharbraceright{\}}#1}}
+\renewcommand{\isachardoublequoteopen}{\insertTTS}
+\renewcommand{\isachardoublequoteclose}{\setTTS}
+\renewcommand{\isanewline}{\mbox{}\par\mbox{}\resetTTS}
 
-\newcommand{\isainnerkeyword}[1]{{\tt #1}}
-\newcommand{\isasymexistsA}{\isamath{\exists_{\textsc A}\,}}
+\renewcommand{\isamarkupcmt}[1]{\hangindent5ex{\isastylecmt --- #1}}
 
-\def\isadelimproof{}
-\def\endisadelimproof{}
-\def\isatagproof{}
-\def\endisatagproof{}
-\def\isafoldproof{}
-\def\isadelimproof{}
-\def\endisadelimproof{}
+%\newcommand{\isaheader}[1]{\section{#1}}
+%\renewcommand{\isamarkupheader}[1]{#1}
 
-\input{lstisabelle}
-\lstset{basicstyle=\footnotesize\ttfamily\slshape}
-\lstset{captionpos=b}
-\lstset{numberbychapter=false}
-
-\newcommand{\isai}{\lstinline[language=isabelle,basicstyle=\normalsize\ttfamily\slshape]}
-
-\input{macros}
-
-% Include snippets
-\newcommand{\DefineSnippet}[2]{%
-   \expandafter\newcommand\csname snippet--#1\endcsname{%
-     \begin{quote}
-     \begin{isabelle}\footnotesize
-     #2
-     \end{isabelle}
-     \end{quote}}}
-\newcommand{\Snippet}[1]{\ifcsname snippet--#1\endcsname\csname snippet--#1\endcsname\else\PackageError{}{No snippet '#1' defined.}{}\fi}
-\input{snippets/snippets.out}
-
-% \overfullrule=8pt
+\newcommand{\DefineSnippet}[2]{#2}
 
@@
 \title{Formalizing the Edmonds-Karp Algorithm}
-% \subtitle{}
+\author{Peter Lammich and S.~Reza Sefidgar}
+\maketitle
 
-\author{Peter Lammich, S.~Reza Sefidgar}
-
-\institute{Technische Universit\"at M\"unchen, \email{\{lammich,sefidgar\}@in.tum.de}}
-
-\maketitle
 \begin{abstract}
@@
 Further refinement yields a verified implementation, whose execution time compares well to an unverified reference implementation in Java.
+\end{abstract}
 
-%We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network. 
-%The formal proofs are done in Isabelle/HOL. They closely follow a standard textbook proof and are relatively easy 
-%to read, even without being an Isabelle expert.
+\clearpage
+\tableofcontents
 
-%We then use stepwise refinement to obtain the Edmonds-Karp algorithm and prove its upper complexity bound of $O(|V||E|^2)$.
-%Further refinement yields a verified implementation, which is almost as fast as an unverified  
-%reference implementation in Java.
+\clearpage
 
-% 
-% and a verified imperative implementation of it. Our implementation is roughly 2.5 times slower than an unverified  
-% reference implementation in Java. 
-% 
-% Finally, we prove the upper complexity bound of $O(|V||E|^2)$ for the Edmonds-Karp algorithm.
-\end{abstract}
+% sane default for proof documents
+\parindent 0pt\parskip 0.5ex
 
@@
 Computing the maximum flow of a network is an important problem in graph theory.
-Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, as well as various scheduling and resource allocating problems can be reduced to it.
-The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. An important instance is the Edmonds-Karp algorithm~\cite{EK72},
-which was one of the first algorithms to solve the maximum flow problem in polynomial time for the general case of networks with real valued capacities.
+Many other problems, like maximum-bipartite-matching, edge-disjoint-paths,
+circulation-demand, as well as various scheduling and resource allocating
+problems can be reduced to it. The Ford-Fulkerson method~\cite{FF56} describes a
+class of algorithms to solve the maximum flow problem. An important instance is
+the Edmonds-Karp algorithm~\cite{EK72}, which was one of the first algorithms to
+solve the maximum flow problem in polynomial time for the general case of
+networks with real valued capacities.
 
-In this paper, we present a formal verification of the Edmonds-Karp algorithm and its polynomial complexity bound.
-The formalization is conducted entirely in the Isabelle/HOL proof assistant~\cite{NPW02}. 
-Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification into an abstract proof of the Ford-Fulkerson method,
-its instantiation to the Edmonds-Karp algorithm, and finally an efficient implementation.
-The abstract parts of our verification closely follow the standard textbook presentation of Cormen et al.~\cite{CLRS09}.
-Being developed in the Isar~\cite{Wenzel99} proof language, our proofs are accessible even to non-Isabelle experts.
+In this paper, we present a formal verification of the Edmonds-Karp algorithm
+and its polynomial complexity bound. The formalization is conducted entirely in
+the Isabelle/HOL proof assistant~\cite{NPW02}. Stepwise refinement
+techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our
+verification into an abstract proof of the Ford-Fulkerson method, its
+instantiation to the Edmonds-Karp algorithm, and finally an efficient
+implementation. The abstract parts of our verification closely follow the
+textbook presentation of Cormen et al.~\cite{CLRS09}. Being developed in the
+Isar~\cite{Wenzel99} proof language, our proofs are accessible even to
+non-Isabelle experts.
 
-While there exists another formalization of the Ford-Fulkerson method in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation.
-Moreover, this paper is a case study on elegantly formalizing algorithms.
-
-The rest of this paper is structured as follows: In Section~\ref{sec:background} we give a short informal introduction to the Ford-Fulkerson method.
-In Section~\ref{sec:abs-formalization}, we report on our formalization of the abstract method. 
-Section~\ref{sec:refinement} gives a brief overview of the Isabelle Refinement Framework~\cite{LaTu12,La12}, which supports stepwise refinement based algorithm development in Isabelle/HOL.
-In Section~\ref{sec:edka}, we report on our instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm and the proof of its complexity.
-Section~\ref{sec:executable} reports on the further refinement steps required to yield an efficient implementation. Section~\ref{sec:benchmark} reports on 
-benchmarking our implementation against a reference implementation of the Edmonds-Karp algorithm from Sedgewick et al.~\cite{SeWa11}. 
-Finally, Section~\ref{sec:concl} gives a conclusion and discusses related and future work. The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/edmonds_karp/}.
-
-% Despite its importance, no formalization of the Ford-Fulkerson algorithm has been developed in modern proof assistants.
-The only similar work that the authors are aware of is formalization of the Ford-Fulkerson algorithm in Mizar~\cite{Lee05}.
-This formalization defines and proves correctness of the algorithm at the level of graph manipulations without providing concrete implementation of the algorithm.
-Providing such an implementation is specially important, as it could provide the basis for many verfied programs with practical importance.
-
-\section{The Ford-Fulkerson Method}\label{sec:background}
-% \begin{comment}
-%   GOAL: 
-%     1) Remind the reader of FoFu. The educated reader should have the feeling:
-%       Yes, now I recall FoFu, and understand (again) what it does.
-%       
-%     2) Set the field for further, more detailed descriptions in next section.
-%   
-% 
-%   Short description of Network: Finite graph over nodes V, 
-%     edge (u,v) annotated with capacity c(u,v).
-%     Assuming distinct nodes s and t. 
-%     Note on additional assumptions: 
-%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
-%         Can transform any Network to match these assumptions, preserving the flow cite{???}.
-%         
-%   s-t Flow: Annotation of edges with values, such that: 
-%     1) Values smaller than capacities.
-%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
-%   
-%   Value of flow: Incoming flow - outgoing flow of s
-%   
-%   Max-Flow problem. 
-%     Rpt. why important?
-%   
-%   Cuts, intuition of min-cut >= max-flow
-%   
-%   Theorem: Min-Cut = Max-Flow
-%   Proven via augmenting flow in residual graph. 
-%     Present 1,2,3 of min-cut max-flow equivalences
-%   
-%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
-%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
-%       Termination? In General: Only for integer capacities.
-%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
-%   \end{comment}
-
-In this section, we give a short introduction to the Ford-Fulkerson method, closely following the presentation by Cormen et al.~\cite{CLRS09}.
-
-% TODO: Possible alternative presentation
-%   First define a relaxed network, without the constraints. Define the flow there. Then add network constraints.
-%
-
-A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real valued capacity $c(u,v)>0$.
-Moreover, there are two distinct vertices $s,t\in V$ which are called \emph{source} and \emph{sink}.
-
-A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges.
-The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
-Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.
-
-To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the network must be on a path from $s$ to $t$.
-Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.
-
-An important result is the relation between flows and cuts in a network. A \emph{cut} is a partitioning of the vertices into two sets, such that one set contains 
-the source and the other set contains the sink. The capacity of a cut is the sum of the capacities of all edges going from the source's side to the sink's side of the cut.
-It is easy to see that the value of any flow cannot exceed the capacity of any cut, as all flow from the source must ultimately reach the sink, and thus go through the edges of the cut.
-The Ford-Fulkerson theorem tightens this bound and states that the value of the maximum flow is equal to the capacity of the minimum cut.
-
-The Ford-Fulkerson method is a corollary of this theorem.
-It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached.
-In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges.
-For this purpose the Ford-Fulkerson method defines the residual graph, which has edges both in parallel and reverse to the edges of the network.
-Each edge is labeled by the amount of flow that can be effectively passed along this edge, by either increasing or decreasing the flow on a network edge.
-Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
-edges with positive labels according to the following labeling function $c_f$:
-\[ c_f (u, v) = 
-  \begin{cases}
-  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
-  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
-  0 \hfill & \text{ otherwise} \\
-  \end{cases} 
-\]
-
-In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph, and then 
-pushes as much flow as possible along this path to increase the value of the current flow. 
-Formally, for an augmenting path $p$, one first defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
-\[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
-An augmenting path then yields a residual flow $f_p$, which is the flow that can be passed along this path:
-\[ f_p (u, v) = 
-  \begin{cases}
-  c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\  
-  0 \hfill & \text{ otherwise} \\
-  \end{cases} 
-\]
-\newcommand{\augment}{{\mathbin\uparrow}}%
-Finally, to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
-by any \emph{augmenting flow} $f'$, \ie any flow in the residual graph:
-\[ (f \augment f') (u, v) = 
-  \begin{cases}
-  f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\  
-  0 \hfill & \text{ otherwise} \\
-  \end{cases} 
-\]
-Note that, for any edge in the network, the augmenting flow parallel to that edge is added to the flow, while the augmenting flow reverse to this edge is subtracted. 
-This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.
-
-The correctness of the Ford-Fulkerson algorithm follows from the Ford-Fulkerson theorem, which is usually stated as the following three statements being equivalent:
-\begin{enumerate}
-\item $f$ is a maximum flow in a network $G$.
-\item there is no augmenting path in the residual graph $G_f$.
-\item there is a cut $C$ in $G$ such that the capacity of $C$ is equal to the value of $f$.
-\end{enumerate}
-
-The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph.
-There are several possible implementations with different execution times.
-The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}.
-When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real valued capacities.
-Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges if the intended meaning is clear from the context.
-A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.
-
-\section{Formalizing the Ford-Fulkerson Method}\label{sec:abs-formalization}
-%  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
-%    Reader should think: Yeah, they did cool stuff!
-%
-%  [???* Thematize definition of flows: On Graphs vs. on Networks. You find both in the literature. 
-%      We decides for [...]. This is better suited, as is gives nice and elegant proofs, even formally, bla bla bla]
-%    
-%  Proof of MinCut-MaxFlow: Follows the textbook proof.
-%    Uses Isar to even look like a textbook proof, being comprehensible even without running
-%      Isabelle/HOL.
-%      
-%    For example, \isai{augment_flow_presv_cap}.
-%      Present formal proof text. Perhaps oppose it to textbook proof?.
-%    
-%  Abstract Algo looks like pseudocode presented in texbooks.
-%    Oppose our algo to textbook pseudocode.
-%
-
-%A formal proof contains many technical details that are not presented in an informal proof.
-While dealing with the technical details in a proof causes distraction, leaving out such details makes it hard to follow the proofs.
-.
-So, readers may use this informal proof as a guide to better understand the relevant thoughts in our formal proof.
-Moreover using the proof language Isar, we made the formal proof readable even for readers that are not familiar with Isabelle.
-
-In this section, we provide a brief overview of our formalization of the Ford-Fulkerson method. 
-In order to develop theory in the context of a fixed graph or network, we use Isabelle's concept of \emph{locales}~\cite{Ballarin:2006:MKM}, which allows
-us to define named contexts that fix some parameters and assumptions.
-For example, the graph theory is developed in the locale \isai{Graph}, which 
-fixes the edge labeling function \isai{c}, and defines the set of edges and nodes based on $c$:
-\begin{lstlisting}
-  locale Graph = fixes c :: "edge => 'capacity" begin
-    definition "E == {(u, v). c (u, v) ~= 0}"
-    definition "V == {u. \<exists>v. (u, v) \<in> E \<or> (v, u) \<in> E}"
-    [...]
-\end{lstlisting}
-Moreover, we define basic concepts like (simple, shortest) paths, and provide lemmas to reason about them.
-
-Networks are based on graphs, and add the source and sink node, as well as the network assumptions:
-\begin{lstlisting}
-  locale Network = Graph + fixes s t :: node
-    assumes no_incoming_s: "\<forall>u. (u, s) \<notin> E"
-    [...]
-\end{lstlisting}
-Most theorems presented in this paper are in the context of the \isai{Network} locale.
-
-\subsection{Presentation of Proofs}
-% (DONE: not sure about repeating citation of Cormen: I would definitely repeat the citation!) 
-Informal proofs focus on the relevant thoughts by leaving out technical details and obvious steps. 
-In contrast, a formal proof has to precisely specify each step as the application of some inference rules. 
-Although modern proof assistants provide high-level tactics to summarize some of these steps, formal proofs tend to be significantly more 
-verbose than informal proofs. Moreover, formal proofs are conducted in the tactic language of the proof assistant, which is often some dialect of ML. 
-Thus, many formal proofs are essentially programs that instruct the proof assistant how to conduct the proof.
-They tend to be inaccessible without a deep knowledge of the used proof assistant, in many cases requiring to replay the proof in the proof assistant in order to understand the idea behind it.
-
-For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that resemble standard mathematical textbook proofs, and are accessible, to a certain extent, even for those not familiar with Isabelle/HOL.
-We use Isar to present our proof of the Ford-Fulkerson method such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.
-
-As an example consider the proof that augmenting a flow $f$ with a residual flow $f'$ yields a valid flow $f\augment f'$ again. In particular, one has to show 
-that the augmented flow satisfies the capacity constraint. Cormen et al.~give the following proof, which we display literally here, only replacing the references to 
-``Equation~26.4'' by ``definition of $\augment$'':
-
-For the capacity constraint, first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
-	\begin{align*}
-	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
-	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
-	& = f'(u, v) \\
-	& \geq 0.
-	\end{align*}
-In addition,
-	\begin{align*}
-	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
-	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
-	& \leq f(u, v) + c_f(u, v) &&  \text{(capacity constraint)} \\
-	& = f(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
-	& = c (u, v).
-	\end{align*}
-
-In the following we present the corresponding formal proof in Isar:
-\Snippet{augment_flow_presv_cap}
-The structure of the Isar proof is exactly the same as that of the textbook proof, except that we had to also consider the case $(u,v)\notin E$, which is not mentioned in the informal proof at all, and easily discharged in our formal proof by the \isai{auto}-tactic after the \isai{qed}.
-We also use exactly the same justifications as the original proof, except that we had to use the fact that there are no parallel edges to show $c_f(v,u)=f(u,v)$, which is not mentioned in the original proof.
-
-\subsection{Presentation of Algorithms}
-In textbooks, it is common to present algorithms in pseudocode, which captures the essential ideas, but leaves open implementation details.
-As a formal equivalent to pseudocode, we use the monadic programming language provided by the Isabelle Refinement Framework~\cite{LaTu12,La12}.
-For example, we define the Ford-Fulkerson method as follows:
-\Snippet{ford_fulkerson_algo}
-
-% DONE: Introduce obtain-syntax, and unfold find-augmenting-path spec! Or even while-exists loop as combinator?
-The code looks quite similar to pseudocode that one would expect in a textbook, but actually is a rigorous formal specification of the algorithm, using nondeterminism to leave open the implementation details (\cf Section~\ref{sec:refinement}).
-Note that we had to use the available combinators of the Isabelle Refinement Framework, which made the code slightly more verbose than we would have liked.
-We leave it to future work to define a set of combinators and appropriate syntax that allows for more concise presentation of pseudocode.
-
-% The only caveat is that our framework currently does not support breaking of loops, such that we had to use an explicit flag instead.
-
-Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Refinement Framework, it is straightforward to prove (partial) correctness of
-the Ford-Fulkerson method, which is stated in Isabelle/HOL by the following theorem:
-\Snippet{ford_fulkerson_correct}
-
-\section{Refinement in Isabelle/HOL}\label{sec:refinement}
-After having stated and proved correct an algorithm on the abstract level, the next step is to provide an (efficient) implementation. 
-In our case, we first specialize the Ford-Fulkerson method to use shortest augmenting paths, then implement the search for shortest augmenting paths by BFS, and finally use efficient data structures to represent the abstract objects modified by the algorithm.
-
-A natural way to achieve this formally is \emph{stepwise refinement}~\cite{Wirth71}, and in particular \emph{refinement calculus}~\cite{Back78,BaWr98}, which allows us to systematically transform an abstract algorithm into a more concrete one, preserving its correctness.
-
-In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12,La12}. 
-It features a refinement calculus for programs phrased in a nondeterminism monad. 
-The monad's type is a set of possible results plus an additional value that indicates a failure:
-\begin{lstlisting}
-  datatype \<alpha> nres = res \<alpha> set | fail
-\end{lstlisting}
-The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
-operation \isai{bind m f} nondeterministically picks a result from \isai{m} and executes \isai{f} on it. 
-If either \isai{m = fail}, or \isai{f} may fail for a result in \isai{m}, the bind operation fails.
-
-We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
-Intuitively, \isai{m<=m'} means that \isai{m} is a refinement of \isai{m'}, \ie all possible results of \isai{m} are 
-also possible results of \isai{m'}. 
-Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
-Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
-language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define standard loop constructs (while, foreach), 
-a syntax for postcondition specifications, and use a Haskell-like do-notation:
-\begin{lstlisting}
-  spec P == spec x. P x == res {x. P x}
-  do {x <- m; f x} == bind m f
-  do {m; m'} == bind m (%_. m')
-\end{lstlisting}
-
-Correctness of a program \isai{m} with precondition \isai{P} and postcondition \isai{Q} is expressed as \isai{P ==> m <= spec r. Q r} (or, eta-contracted, just \isai{spec Q}), which
-means that, if \isai{P} holds, \isai{m} does not fail and all possible results of \isai{m} satisfy \isai{Q}. Note that we provide different recursion constructs
-for partial and total correctness: A nonterminating total correct recursion yields \isai{fail}, which satisfies no specification, even if joined 
-with results from other possible runs. On the other hand, a nonterminating partial correct recursion yields \isai|res {}|, which refines any specification 
-and disappears when joined with other results.
-
-The Isabelle Refinement Framework also supports data refinement, changing the representation of results according to a \emph{refinement relation}, 
-which relates concrete with abstract results: Given a relation \isai{R}, \isai{\<Down>R m} is the largest set of concrete results that are related to an 
-abstract result in \isai{m} by \isai{R}. If \isai{m = fail}, then also \isai{\<Down>R m = fail}.
-
-In a typical program development, one first comes up with an initial version \isai{m_0} of
-the algorithm and its specification \isai{P,Q}, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions \isai{m$_i$} of the algorithm,
-proving \isai{m$_i$ <= \<Down>R$_i$ m$_{i-1}$}. Using transitivity and composability of data refinement, one 
-gets \isai{P ==> m$_i$ <= \<Down>R$_i$...R$_1$ spec Q}, showing the correctness of the refined algorithm.
-If no data refinement is performed, \isai{R$_i$} is set to the identity relation, in which case \isai{\<Down>R$_i$} becomes the identity function.
-
-Various tools, including a verification condition generator, assist the user in conducting the refinement proofs by breaking 
-them down to statements that do not involve monad constructs any more, and in many cases precisely reflect the core idea of the proof.
-
-Monotonicity of the standard combinators also allows for modular refinement: Replacing a part of a program by a refined version
-results in a program that refines the original program.
-This gives us a natural formal model for statements like ``we implement shortest path finding by BFS'', or ``we use arrays to represent the edge labeling''.
-
-% 
-% 
-% Purpose: Separation of concerns.
-% 
-% Short intro on Refinement Framework:
-% 
-% Programs phrased as error-set monad. With order structure. 
-% Refinement: Less possible results (refinement ordering). Error is greatest element (refined by everything).
-% 
-% Building chains of refinement, at the end there should be a deterministic algorithm (yielding exactly one result), which is
-% contained (by transitivity) in the specification.
-% 
-% Standard combinators (bind, recursion, if, etc) are monotonic wrt. refinement ordering, which allows for modularity:
-% Replacing part of a program by a refined version, yields a refined program. 
-% 
-% Data refinement: Changing the representation of data. Relation between concrete and abstract representation. 
-% $\Uparrow m$ is biggest concrete program that refines $m$ wrt.\ relation $R$. 
-% 
-% The Isabelle Refinement Framework provides the nres-monad, and tools to reason about program refinement, 
-% the most important one being a VCG.
-
-\section{The Edmonds-Karp Algorithm}\label{sec:edka}
-  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm requires to show that finding a shortest augmenting path is a refinement of finding any augmenting path, which is straightforward using basic graph theory.
-
-  Considerably more effort is required to show that the resulting algorithm terminates within $O(VE)$ iterations. 
-  The idea of the proof is as follows: Edges that are reverse to edges on a shortest path cannot lie 
-  on a shortest path itself.
-  On every augmentation, at least one edge of the residual graph that lies on a shortest augmenting path is flipped. Thus, either the length of the shortest path increases,
-  or the number of edges that lie on some shortest path decreases. As the length of a shortest path is at most $V$, there are no more than $O(VE)$ iterations.
-  % TODO: Add more explanation here, like: DIstinguish two cases: 1) ... 2) The distance s-t in cf is still the same, but then ...
-
-  Note that Cormen et al.~present the same idea a bit differently: They define an edge of the residual graph being \emph{critical} if it lies on a shortest path such that it will be flipped by augmentation.
-Then, they establish an upper bound of how often an edge can get critical during the algorithm.
-Our presentation is more suited for a formal proof, as we can directly construct a measure function from it, \ie a function from flows to natural numbers, which decreases on every iteration and is bounded by $O(VE)$.
-
-  Formalizing the above intuitive argument was more tricky than it seemed on first glance:
-  While it is easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
-  to a graph transformation which may add multiple reverse edges and removes at least one original edge requires some generalization of the statement.
-  Note that a straightforward induction on the length of the augmenting path or on the number of flipped edges fails, as, after flipping the first 
-  edge, the path is no longer augmenting.
-
-%   We decided to generalize the statement as follows. Assume we have an original graph $cf$ with a shortest augmenting path $p$, and a transformed graph $cf'$ that has been created from
-%   $cf$ by adding some flipped edges from $p$ and removing some edges from $p$.
-%   Then, we consider an edge $(u,v) \in p$, a path $p_1$ from $s$ to $v$ in the original graph, and a path $p_2$ from $u$ to $t$ in the transformed graph.
-%   
-%   By induction on the number of reversed edges in $p_2$, we 
-%   show that $|p_1|+|p_2| > |p|$. In the induction step, we use the proof idea for the single flipped edge $(v,u)$, and,
-%   if $p_2$ contains more flipped edges, we split $p_2$ at the first flipped edge. Then, the initial segment of $p_2$ contains no flipped edges,
-%   and thus also is a path in the original graph, and we can apply the induction hypothesis.
-%   
-%   Proving that augmentation with a shortest augmenting path actually only adds flipped edges, and removes at least one original edge, required some more work,
-%   but was straightforward and yielded no surprises.
-%   
-%   Finally, we phrase the complexity argument as a measure function $m = l*2|E| + n$, where $l$ is the length of the shortest augmenting path,
-%   and $n$ is the number of edges that lie on any shortest path. We show that $m$ is decreased by the loop body.
-%   Adding a special case for loop termination (there are no augmenting paths), and observing that $m$ is bounded by $2|V||E|$, yields 
-%   the desired upper bound on loop iterations. Refinement of the algorithm to add an explicit loop counter, and asserting the upper bound, is then straightforward.
-
-  Having defined the measure function and shown that it decreases on augmentation, it is straightforward to refine the partial correct while loop to a total correct one.
-Moreover, to make explicit the bound on the number of loop iterations, we instrument the loop with a counter, and assert the upper bound on the counter after the loop.
-
-\section{Refinement to Executable Code}\label{sec:executable}
-  In the previous section, we have presented our abstract formalization of the Edmonds-Karp algorithm,
-  leaving open how to obtain a shortest augmenting path and how to implement the algorithm. 
-  In this section, we outline the further refinement steps that were necessary to obtain an efficient implementation.
-
-  \subsection{Using Breadth First Search}
-  A standard way to find a shortest path in a graph is breadth first search (BFS).
-Luckily, we had already formalized a BFS algorithm as an example for the Refinement Framework.
-Unfortunately, this algorithm only computed the minimum distance between two nodes, without returning an actual path.
-For this project, we extended the formalization accordingly, and added an efficient imperative implementation, using the same stepwise refinement techniques as for the main algorithm.
-Note that the resulting BFS algorithm is independent, and can be reused for shortest path finding in other applications.
-
-  Implementing shortest path finding by BFS in Edmonds-Karp algorithm yields a specification that algorithmically describes all major operations, but still leaves open the data structures used for implementation.
-
-%   
-%   We briefly report on our adapted formalization here, which is displayed in Figure~\ref{TODO: Snippet}: 
-%   TODO: Reference to figure (line numbers?)
-%   The abstract algorithm keeps track of a set $C$ of current nodes, a set $N$ of new nodes, and a partial map $P$ from already visited nodes to predecessor nodes.
-%   Initially, only the start node is in $C$ and $N$ is empty. In each iteration of the loop, a node $u\in C$ is picked, and its successors $v$ that have not yet been 
-%   visited are added to $N$, and $P~v$ is set to $u$. If $C$ becomes empty, $C$ and $N$ are swapped. If the target node is encountered, the algorithm immediately terminates.
-%   Note that this implementation is a generalized version of the usual queue implementation of BFS: While a queue enforces a complete order on the encountered nodes, our 
-%   implementation only enforces an ordering between nodes on the current level and nodes on the next level. 
-% 
-%   For the actual algorithm, we wrap this algorithm by a procedure to handle the special case where source and target nodes are the same, and to extract a shortest path 
-%   from $P$ upon termination. Moreover, we implement the abstract specification of adding the successor nodes of a node by a loop.
-%   
-%   Finally, we prove the following theorem: %TODO: Not actually true, the impls are oly done in bfs2!
-% %     theorem
-% %       assumes "src∈V" 
-% %       shows "bfs src dst ≤ (spec p. 
-% %         case p of None ⇒ ¬conn src dst | Some p ⇒ shortestPath src p dst)"
-%   
-%   Obviously, the BFS algorithm refines the specification for obtaining a shortest path.
-%   Using this to refine the Edmonds-Karp formalization yields a program that algorithmically specifies all major operations.
-
-  \subsection{Manipulating Residual Graphs Directly}\label{sec:impl_res_graph}
-  In a next step, we observe that the algorithm is phrased in terms of a flow which is updated until it is maximal. However,
-  in order to update the flow, the residual graph is used, which is a combination of the current flow and the network.
-  Obviously, computing the complete residual graph each time before searching for an augmenting path is a bad idea. 
-  One solution to this problem is to compute the edges of the residual graph on the fly from the network and the current flow.
-  Although this solution seems to be common, it has the disadvantage that for each edge of the residual graph, two (or even three) 
-  edges of the network and the flow have to be accessed. As edges of the residual graph are accessed in the inner loop, during 
-  the BFS, these operations are time critical.
-
-  After our profiling indicated a hot spot on accessing the capacity matrices of the network and the flow, we switched to 
-  an algorithm that operates on a representation of the residual graph directly. This resulted in a speed-up of factor two.
-  As the residual graph uniquely determines the flow (and vice versa), it is straightforward to phrase the
-  operations directly on the residual graph. Performing a data refinement of the flow \wrt the refinement 
-  relation $\{(c_f,f) \mid \text{$f$ is a flow}\}$ then yields the desired algorithm.
-
-  \subsection{Implementing Augmentation}    
-  In our abstract formalization, which matches the presentation in Section~\ref{sec:background}, we have formulated augmentation by first defining 
-  the residual capacity $c_p$ of the augmenting path. Using $c_p$, we have defined the residual flow $f_p$, which was finally added to the current flow.
-  In the refinement to operate on residual graphs, we have refined this to augment the residual graph.
-  For the implementation, we compute the residual capacity in a first iteration over the augmenting path, and modify the residual graph 
-  in a second iteration. Proving this implementation correct is straightforward by induction on the augmenting path.
-
-  \subsection{Computing Successors}
-  In order to find an augmenting path, the BFS algorithm has to compute the successors of a node in the residual graph. 
-  Although this can be implemented on the edge labeling function by iterating over all nodes, this implementation tends to be inefficient for sparse graphs,
-  where we would have to iterate over many possible successor nodes just to find that there is no edge.
-
-  A common optimization is to pre-compute an \emph{adjacency map} from nodes to adjacent nodes in the network. 
-  As an edge in the residual graph is either parallel or reverse to 
-  a network edge, it is enough to iterate over the adjacent nodes in the network, and check whether they are actual successors in the residual graph.
-  It is straightforward to show that this implementation actually returns the successors in the residual graph.
-
-  \subsection{Using Efficient Data Structures}\label{sec:impl_data_structures}
-  In a final step, we have to choose efficient data structures for the algorithm. The objects that the algorithm works with are
-  capacities, nodes, edges, the graph of the input network, residual graphs, the adjacency map, paths, and the resulting maximum flow.
-
-  We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers.}.
-Note that an implementation as fixed precision numbers would also be possible,
-  but requires additional checks on the network to ensure that no overflows can happen. 
-  % TODO: Can we trick the Java reference implementation on integer overflow?
-
-  We implement nodes as natural numbers less than an upper bound $N$, and residual graphs are implemented by their capacity matrices, which, in turn,
-  are realized as arrays of size $N*N$ with row-major indexing, such that the successors of a node are close together in memory.
-  The adjacency map of the network is implemented as an array of lists of nodes. 
-  An augmenting path is represented by a list of edges, \ie a list of pairs of nodes.
-
-%   For the BFS algorithm, we implement the sets $C$ and $N$ by lists, and the predecessor map $P$ as an array of \isai{node option}. 
-%   The successors of a node are represented as a list with no duplicate elements. The resulting path, which is extracted from $P$ upon termination of
-%   the BFS, is represented as a list of edges.
-
-  The input network of the algorithm is represented as a function from network edges to capacities, which is tabulated into an array
-  to obtain the initial residual graph. This gives us some flexibility in using the algorithm, as any capacity matrix representation can be converted into a 
-  function easily, without loosing efficiency for read-access. 
-  Similarly, our implementation expects an adjacency map as additional parameter, which is then tabulated into an array.
-This is convenient in our context, where a preprocessing step computes the adjacency map anyway.
-
-  The output flow of the algorithm is represented as the residual graph. The user can decide how to compute the maximum flow from it. For example,
-  in order to compute the maximum flow value, only the outgoing edges of the source node have to be computed, which is typically less 
-  expensive than computing the complete flow matrix. The correctness theorem of the algorithm abstractly states how to obtain the maximum flow from the output.
-
-  Note that there is still some optimization potential left in the choice of our data structures: 
-  For example, the BFS algorithm computes a predecessor map $P$. It then iterates over $P$ to extract the shortest path as a list of edges.
-  A subsequent iteration over this list computes the residual capacity, and a final iteration performs the augmentation. 
-  This calls for a deforestation optimization to get rid of the intermediate list, and iterate only two times over the predecessor map directly.
-%   A first time to compute the bottleneck, and a second time to do the augmentation. 
-  Fortunately, iteration over the shortest path seems not to significantly contribute to the runtime of our implementation, such that 
-  we leave this optimization for future work.
-
-  Note that we performed the last refinement step using our Sepref tool~\cite{La15,La16}, which provides tool support for data refinements from 
-  purely functional programs of the Isabelle Refinement Framework into imperative programs expressed in Imperative/HOL~\cite{BKHEM08}.
-  The formalization of this refinement step consists of setting up the mappings between the abstract and concrete data structures,
-  and then using Sepref to synthesize the Imperative/HOL programs and their refinement proofs.
-  Finally, Imperative/HOL comes with a setup for the Isabelle code generator~\cite{Haft09,HaNi10} to generate imperative programs in OCaml, SML, Scala, and Haskell.
-
-  \subsection{Network Checker}
-  Additionally, we implemented an algorithm that takes as input a list of edges, a source node, and a target node.
-  It converts these to a capacity matrix and an adjacency map, and checks whether the resulting graph satisfies our network assumptions.
-  We proved that this algorithm returns the correct capacity matrix and adjacency map iff the input describes a valid network,
-  and returns a failure value otherwise.
-
-  Combining the implementation of the Edmonds-Karp algorithm with the network checker yields our final implementation,
-  for which we can export code, and have proved the following theorem: 
-  \Snippet{edmonds_karp_correct}
-  Note that this theorem is stated as a Hoare triple, using separation logic assertions. There are no preconditions on the input.
-  If the algorithm returns \isai{None}, then the edge list was malformed or described a graph that does not satisfy the network assumptions.
-  Here, \isai{ln_invar} describes well-formed edge lists, \ie edge lists that have no duplicate edges and only edges with positive capacity,
-  and \isai{ln_\<alpha>} describes the mapping from (well-formed) edge lists to capacity matrices (note that we set \isai{c==ln_\<alpha> el}).
-  If the algorithm returns some number \isai{N} and residual graph \isai{cf}, then the input was a well-formed edge list that describes a valid network with at most \isai{N} nodes.
-  Moreover, the returned residual graph describes a flow \isai{f} in the network, which is maximal. 
-  As the case distinction is exhaustive, this theorem states the correctness of the algorithm. 
-  Note that Isabelle/HOL does not have a notion of execution, thus total correctness of the generated code cannot be expressed.
-However, the program is phrased in a heap-exception monad, thus introducing some (coarse grained) notion of
-  computation. On this level, termination can be ensured, and, indeed, the above theorem implies that all the recursions stated by recursion 
-  combinators in the monad must terminate. However, it does not guarantee that we have not injected spurious code equations like $f~x = f~x$, 
-  which is provable by reflexivity, but causes the generated program to diverge.
-
-\section{Benchmarking}\label{sec:benchmark}
-  We have compared the running time of our algorithm against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}.
-We have used MLton 20100608~\cite{MLton} and OpenJDK Java 1.7.0\_95, running on a standard laptop machine with a 2.8GHz i7 quadcore processor and 16GiB of RAM.
-
-  We have done the comparison on randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
-Note that the maximum density for networks that satisfy our assumptions is $0.5$, as we allow no parallel edges.
-  For sparse networks, we varied the number of nodes between $1000$ and $5500$, for dense networks between $1000$ and $1450$.
-  The results are shown in Figure~\ref{fig:benchmark}, in a double-logarithmic scale.
-
-  \begin{figure}[!h]
-  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
-    \begin{axis}[
-      %title={Benchmark of the Edmonds-Karp algorithm},
-        axis x line*= bottom,
-        axis y line*= left,
-        xmode = log,
-        ymode = log,
-%         xlabel = Number of vertices in the network,
-%         ylabel = Execution time in seconds, 
-        xtick = {0,1000,...,5000},
-        ytick = {0.01,0.1,1,10},
-        xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$},
-        yticklabels = {$0.01$,$0.1$,$1$,$10$}, 
-        smooth,
-        cycle list name = black white,
-        legend style = { 
-          at={(0.59,0.4)}, 
-            anchor=north west,
-            draw=black, 
-            fill= white,
-            align=left
-        }
-    ]
-
-    %start data for plotting
-      \addplot table {
-        1000 .027   
-        1500 .124  
-        2000 .299  
-        2500 .665  
-        3000 1.335 
-        3500 2.526 
-        4000 3.111 
-        4500 6.917 
-        5000 7.743 
-        5500 10.452    
-      }; \addlegendentry{SML, sparse};
-
-      \addplot table {
-        1000 .016   
-        1500 .049   
-        2000 .168  
-        2500 .453  
-        3000 .842  
-        3500 1.524 
-        4000 2.136 
-        4500 3.438 
-        5000 4.734 
-        5500 6.369     
-      }; \addlegendentry{Java, sparse};
-    %end data for plotting
-    \end{axis}
-  \end{tikzpicture}
-%
-  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
-    \begin{axis}[
-      %title={Benchmark of the Edmonds-Karp algorithm},
-        axis x line*= bottom,
-        axis y line*= left,
-        xmode = log,
-        ymode = log,
-%         xlabel = Number of vertices in the network,
-%         ylabel = Execution time in seconds, 
-        xtick = {0,1000,1100,...,1400},
-        ytick = {2,4,...,10},
-        xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, 
-        yticklabels = {$2$, $4$,$6$,$8$,$10$}, 
-        smooth,
-        cycle list name = black white,
-        legend style = { 
-          at={(0.59,0.4)}, 
-            anchor=north west,
-            draw=black, 
-            fill= white,
-            align=left
-        }
-    ]
-
-    %start data for plotting
-      \addplot table {
-        1000 3.328 
-        1050 3.345 
-        1100 4.841 
-        1150 5.036 
-        1200 5.840 
-        1250 6.858 
-        1300 7.651 
-        1350 8.950 
-        1400 8.398 
-        1450 10.769     
-      }; \addlegendentry{SML, dense};
-
-      \addplot table {
-        1000 4.141 
-        1050 4.778 
-        1100 5.465 
-        1150 6.248 
-        1200 7.128 
-        1250 8.137 
-        1300 9.864 
-        1350 11.008
-        1400 11.902
-        1450 13.365     
-      }; \addlegendentry{Java, dense};
-    %end data for plotting
-    \end{axis}
-  \end{tikzpicture}
-
-%   \centering    
-%   \begin{tikzpicture}[thick,scale=1, every node/.style={scale=1}] %change the scales if you like to reduce the size
-%     \begin{axis}[
-%       %title={Benchmark of the Edmonds-Karp algorithm},
-%         axis x line*= bottom,
-%         axis y line*= left,
-%         xmode = log,
-%         ymode = log,
-%         xlabel = Number of vertices in the network,
-%         ylabel = Execution time in milliseconds, %in milliseconds, modify the scale for lables
-%         xtick = {0,1000,...,5000},
-% %         ytick = data,
-%         xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$}, %corresponds to the first column in {SML, Sparse}
-% %         yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
-%         smooth,
-%         cycle list name = black white,
-%         legend style = { 
-%           at={(0.59,0.4)}, 
-%             anchor=north west,
-%             draw=black, 
-%             fill= white,
-%             align=left
-%         }
-%     ]
-%     
-%     %start data for plotting
-%       \addplot table {
-%         1000 27   
-%         1500 124  
-%         2000 299  
-%         2500 665  
-%         3000 1335 
-%         3500 2526 
-%         4000 3111 
-%         4500 6917 
-%         5000 7743 
-%         5500 10452    
-%       }; \addlegendentry{SML, sparse};
-% 
-%       \addplot table {
-%         1000 16   
-%         1500 49   
-%         2000 168  
-%         2500 453  
-%         3000 842  
-%         3500 1524 
-%         4000 2136 
-%         4500 3438 
-%         5000 4734 
-%         5500 6369     
-%       }; \addlegendentry{Java, sparse};
-% 
-%       \addplot table {
-%         1000 3328 
-%         1050 3345 
-%         1100 4841 
-%         1150 5036 
-%         1200 5840 
-%         1250 6858 
-%         1300 7651 
-%         1350 8950 
-%         1400 8398 
-%         1450 10769     
-%       }; \addlegendentry{SML, dense};
-% 
-%       \addplot table {
-%         1000 4141 
-%         1050 4778 
-%         1100 5465 
-%         1150 6248 
-%         1200 7128 
-%         1250 8137 
-%         1300 9864 
-%         1350 11008
-%         1400 11902
-%         1450 13365     
-%       }; \addlegendentry{Java, dense};
-%     %end data for plotting
-%     \end{axis}
-%   \end{tikzpicture}
-
-%   \center{\footnotesize The x-axis shows the number of nodes, the y axis the execution time in seconds.}
-  \caption{Benchmark of different implementations. The x-axis shows the number of nodes, the y axis the execution time in seconds.}\label{fig:benchmark}
-  \end{figure}
-
-  We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of $1.6$, while for dense graphs, our implementation is faster by a factor of $1.2$.
-Note that the Java implementation operates on flows, while our implementation
-  operates on residual graphs (\cf Section~\ref{sec:impl_res_graph}). Moreover, the Java implementation does not store the augmenting 
-  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).
-
-  Finally note that a carefully optimized C++ implementation of the algorithm is only slightly faster than the Java implementation for sparse graphs,
-  but roughly one order of magnitude faster for dense graphs. We leave it to future work to investigate this issue, and conclude that we were able to produce
-  a reasonably fast verified implementation.
-
-%   , without doing any low-level optimizations, but using the code as generated by the Isabelle/HOL code generator.
-
-% 
-%   What implementations did we compare: Authors? Where do implementations come from? 
-%   [We must convince the reader that we did not intentionally chose bad implementations to compare us against!]
-% 
-%   Comparsion of algorithms (Modified data structure: Flow vs Flow-like vs ResGraph)
-%     Computing successors in BFS: Filter by visited set first.
-%       -> Technically more challenging, when computing on flow: 1 access first vs 2 accesses first. On resGraph: No advantage in memory accesses (However, plain array access vs. matrix access)
-%     Computing bottleneck during BFS. Saves one iteration over the path, at the cost of one extra memory access per discovered node.
-%     For our benchmarks, we observed that the BFS discovers considerably more nodes than the length of the ultimately returned path, 
-%     and the overall code was faster with the extra iteration for bottleneck computation.
+While there exists another formalization of the Ford-Fulkerson method in
+Mizar~\cite{Lee05}, we are, to the best of our knowledge, the first that verify a
+polynomial maximum flow algorithm, prove the polynomial complexity bound, or
+provide a verified executable implementation. Moreover, this paper is a case
+study on elegantly formalizing algorithms.
+% generated text of all theories
+\input{session}
 
 \section{Conclusion}\label{sec:concl}
-  We have presented a verification of the Edmonds-Karp algorithm, using a stepwise refinement approach.
-  Starting with a proof of the Ford-Fulkerson theorem, we have verified the generic Ford-Fulkerson method, 
-  specialized it to the Edmonds-Karp algorithm, and proved the upper bound $O(VE)$ for the number of outer loop iterations.
-  We then conducted several refinement steps to derive an efficiently executable implementation of the algorithm, 
-  including a verified breadth first search algorithm to obtain shortest augmenting paths. 
-  Finally, we added a verified algorithm to check whether the input is a valid network, and generated executable code in SML.
-  Our implementation compares well to a reference implementation in Java.
-
-  Our formalization has combined several techniques to achieve an elegant and accessible formalization: 
-  Using the Isar proof language~\cite{Wenzel99}, we were able to provide a completely rigorous but 
-  still accessible proof of the Ford-Fulkerson theorem. The Isabelle Refinement Framework~\cite{LaTu12,La12} and the Sepref tool~\cite{La15,La16}
-  allowed us to present the Ford-Fulkerson method on a level 
-  of abstraction that closely resembles pseudocode presentations found in textbooks, and then formally link this presentation to an efficient
-  implementation. Moreover, modularity of refinement allowed us to develop the breadth first search algorithm independently, and later link it to the 
-  main algorithm.
-The BFS algorithm can be reused as building block for other algorithms.
-The data structures are re-usable, too: although we had to implement the array representation of (capacity) matrices for this project, it will be added to the growing library of data structures
-  supported by the Sepref tool, such that it can be re-used for future formalizations.
-
-  We conclude with some statistics:
-  The formalization consists of roughly 8000 lines of proof text, where the graph theory up to the Ford-Fulkerson algorithm required 2600 lines.
-  The abstract Edmonds-Karp algorithm and its complexity analysis contributes 900 lines, and its implementation (including BFS) another 1500 lines.
-  The remaining lines are contributed by the network checker and some auxiliary theories.
-The development of the theories required roughly 3 man month, a significant amount of this time going into a first, purely functional version of the implementation, which was later dropped in favor of the faster imperative version.
+  We have presented a verification of the Edmonds-Karp algorithm, using a
+  stepwise refinement approach. Starting with a proof of the Ford-Fulkerson
+  theorem, we have verified the generic Ford-Fulkerson method, specialized it to
+  the Edmonds-Karp algorithm, and proved the upper bound $O(VE)$ for the number
+  of outer loop iterations. We then conducted several refinement steps to derive
+  an efficiently executable implementation of the algorithm, including a
+  verified breadth first search algorithm to obtain shortest augmenting paths.
+  Finally, we added a verified algorithm to check whether the input is a valid
+  network, and generated executable code in SML. The runtime of our verified
+  implementation compares well to that of an unverified reference implementation
+  in Java. Our formalization has combined several techniques to achieve an
+  elegant and accessible formalization: Using the Isar proof
+  language~\cite{Wenzel99}, we were able to provide a completely rigorous but
+  still accessible proof of the Ford-Fulkerson theorem. The Isabelle Refinement
+  Framework~\cite{LaTu12,La12} and the Sepref tool~\cite{La15,La16} allowed us
+  to present the Ford-Fulkerson method on a level of abstraction that closely
+  resembles pseudocode presentations found in textbooks, and then formally link
+  this presentation to an efficient implementation. Moreover, modularity of
+  refinement allowed us to develop the breadth first search algorithm
+  independently, and later link it to the main algorithm. The BFS algorithm can
+  be reused as building block for other algorithms. The data structures are
+  re-usable, too: although we had to implement the array representation of
+  (capacity) matrices for this project, it will be added to the growing library
+  of verified imperative data structures supported by the Sepref tool, such that
+  it can be re-used for future formalizations. During this project, we have
+  learned some lessons on verified algorithm development: 
+  \begin{itemize} 
+  \item
+  It is important to keep the levels of abstraction strictly separated. For
+  example, when implementing the capacity function with arrays, one needs to
+  show that it is only applied to valid nodes. However, proving that, e.g.,
+  augmenting paths only contain valid nodes is hard at this low level. Instead,
+  one can protect the application of the capacity function by an assertion---
+  already on a high abstraction level where it can be easily discharged. On
+  refinement, this assertion is passed down, and ultimately available for the
+  implementation. Optimally, one wraps the function together with an assertion
+  of its precondition into a new constant, which is then refined independently.
+  \item Profiling has helped a lot in identifying candidates for optimization.
+  For example, based on profiling data, we decided to delay a possible
+  deforestation optimization on augmenting paths, and to first refine the
+  algorithm to operate on residual graphs directly. 
+  \item ``Efficiency bugs''
+  are as easy to introduce as for unverified software. For example, out of
+  convenience, we implemented the successor list computation by \emph{filter}.
+  Profiling then indicated a hot-spot on this function. As the order of
+  successors does not matter, we invested a bit more work to make the
+  computation tail recursive and gained a significant speed-up. Moreover, we
+  realized only lately that we had accidentally implemented and verified
+  matrices with column major ordering, which have a poor cache locality for our
+  algorithm. Changing the order resulted in another significant speed-up.
+  \end{itemize} 
+  We conclude with some statistics: The formalization consists of
+  roughly 8000 lines of proof text, where the graph theory up to the
+  Ford-Fulkerson algorithm requires 3000 lines. The abstract Edmonds-Karp
+  algorithm and its complexity analysis contribute 800 lines, and its
+  implementation (including BFS) another 1700 lines. The remaining lines are
+  contributed by the network checker and some auxiliary theories. The
+  development of the theories required roughly 3 man month, a significant amount
+  of this time going into a first, purely functional version of the
+  implementation, which was later dropped in favor of the faster imperative
+  version.
 
   \subsection{Related Work}\label{sec:related_work} 
-  We are only aware of one other formalization of the Ford-Fulkerson method conducted in Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication
-  on this formalization except~\cite{Lee05}, which provides a Mizar proof script without any additional comments except that it ``defines and proves correctness of Ford/Fulkerson's Maximum Network-Flow algorithm at the level of graph manipulations''.
-Moreover, in Lee et al.~\cite{LeRu07}, which is about graph representation in Mizar, the formalization is shortly mentioned, and it is clarified that it does not provide any implementation or data structure formalization.
-  As far as we could understand the Mizar proof script, it formalizes an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson method.
-  Termination is only proved for integer valued capacities.
+  We are only aware
+  of one other formalization of the Ford-Fulkerson method conducted in
+  Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication on
+  this formalization except~\cite{Lee05}, which provides a Mizar proof script
+  without any additional comments except that it ``defines and proves
+  correctness of Ford/Fulkerson's Maximum Network-Flow algorithm at the level of
+  graph manipulations''. Moreover, in Lee et al.~\cite{LeRu07}, which is about
+  graph representation in Mizar, the formalization is shortly mentioned, and it
+  is clarified that it does not provide any implementation or data structure
+  formalization. As far as we understood the Mizar proof script, it formalizes
+  an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson
+  method. Termination is only proved for integer valued capacities. Apart from
+  our own work~\cite{La14,NoLa12}, there are several other verifications of
+  graph algorithms and their implementations, using different techniques and
+  proof assistants. Noschinski~\cite{Nosch15} verifies a checker for
+  (non-)planarity certificates using a bottom-up approach. Starting at a C
+  implementation, the AutoCorres tool~\cite{Greenaway15,GAK12} generates a
+  monadic representation of the program in Isabelle. Further abstractions are
+  applied to hide low-level details like pointer manipulations and fixed size
+  integers. Finally, a verification condition generator is used to prove the
+  abstracted program correct. Note that their approach takes the opposite
+  direction than ours: While they start at a concrete version of the algorithm
+  and use abstraction steps to eliminate implementation details, we start at an
+  abstract version, and use concretization steps to introduce implementation
+  details.
 
-  Apart from our own work~\cite{La14,NoLa12}, there are several other verifications of graph algorithms and their implementations, using different techniques and proof assistants.
-Noschinski~\cite{Nosch15} verifies a checker for (non-)planarity certificates using a bottom-up approach.
-Starting at a C implementation,
-  the AutoCorres tool~\cite{Greenaway15,GAK12} generates a monadic representation of the program in Isabelle. Further abstractions are applied
-  to this representation to hide low-level details like pointer manipulations and fixed size integers. Finally, a verification condition
-  generator is used to prove the abstracted program correct.
-Note that their approach takes the opposite direction than ours: While they start at a concrete version of the algorithm and use abstraction steps to eliminate implementation details, we start at an abstract version, and use concretization steps to introduce implementation details.
-
-  Chargu\'eraud~\cite{char11} also uses a bottom-up approach to verify imperative programs written in a subset of OCaml, amongst them a version of Dijkstra's algorithm:
-  A verification condition generator generates a \emph{characteristic formula}, which reflects the semantics of the program in the logic of the Coq proof assistant~\cite{BeCa10}.
+  Chargu\'eraud~\cite{char11} also uses a bottom-up approach to verify
+  imperative programs written in a subset of OCaml, amongst them a version of
+  Dijkstra's algorithm: A verification condition generator generates a
+  \emph{characteristic formula}, which reflects the semantics of the program in
+  the logic of the Coq proof assistant~\cite{BeCa10}.
 
   \subsection{Future Work}
-  Future work includes the optimization of our implementation, and the formalization of more advanced maximum flow algorithms, like Dinic's algorithm~\cite{Di06} or push-relabel algorithms~\cite{GoTa88}.
-  We expect both formalizing the abstract theory and developing efficient implementations to be challenging but realistic tasks.
-
-%   
-%   can be performed in two directions. First of all, the abstract theory can be extended to also support more advanced 
-%   maximum flow algorithms like push-relabel algorithms~\cite{GoTa88}
-%   
-%   
-%   
-%   What do the Coq-Guys have: CFML? Others?
-%   
-%   
-%   XXX, ctd here
-%   
-%   Unfortunately, it
-% 
-%   ... and related work
-%     Mizar-Formalization: What did they do.
-%     
-%     History of the formalization: 
-%       Impl: From first impl that was hardly able to compute a 100 nodes graph, to current efficient one.
-%     
-%     Future Work: Dinic. (Actually, our abstract scheme [almost] covers Dinic's algorithm!)
-%     
-% 
-% 
-%   
-% Contributions
-% 
-%   Formal proof of mincut maxflow
-%   fofo-scheme
-%   inst to edmonds karp
-%   complexity analysis of edmonds karp
-%   refinement down to executable code. Roughly 2.5 times slower than Java. (What about OCaml)
-%     + NetCheck
-%     
-% What shines (its a Pearl)
-%   Min-Cut Max-Flow: Textboook like formal reasoning: Comprehensible proof, BUT machine checked
-%     (Present one (carefully worked) example in paperI. We could use lemma \isai{augment_flow_presv_cap})
-%     
-%   Refinement based approach: Fofu-Scheme, instantiation to EdsKa. 
-%     +1: Abstract Algo looks almost like pseudo-code you would expect in textbook.
-%     +2: Fofu-Scheme proved correct for all aug-path finders. EdsKa is instantiation of it.
-%     +3: Modularity: Fofu-scheme and pathfinder developed+proved independently of each other.
-%   
-%   Down to executable code, plugging in efficient data structures.
-%   
-% Some minor contributions:
-%   Reusable BFS algorithm
-%   Imperative matrix data structure (really minor).
-%   
-%   
-%   
-%     
-% 
-% 
-%
-
+  Future work includes the optimization of our implementation, and the
+  formalization of more advanced maximum flow algorithms, like Dinic's
+  algorithm~\cite{Di06} or push-relabel algorithms~\cite{GoTa88}. We expect both
+  formalizing the abstract theory and developing efficient implementations to be
+  challenging but realistic tasks.
+% optional bibliography
 \bibliographystyle{abbrv}
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2016-03-09 formalization/fix_urls.sh
HTML page + scripts to generate it

@@ -1,0 +1,16 @@
+AFP_URL="http://afp.sourceforge.net/browser_info/current/AFP"
+HOL_URL="https://isabelle.in.tum.de/dist/library/HOL"
+  
+if test $# -eq 0; then
+  echo "Usage: $0 directories"
+  echo "Fix URLs in html-files generated by Isabelle/HOL such that AFP and HOL links point to the AFP and Isabelle/HOL homepages"
+  exit 1
+fi
+  
+  
+# Fixup URLs in browser_info
+
+for i in `find $@ -name '*.html'`; do
+  sed -i -re "s|../../AFP/|$AFP_URL/|g; s|../../HOL/|$HOL_URL/|g" $i
+  
+done

### 2016-03-09 formalization/make_html.sh

@@ -1,0 +1,7 @@
+#!/bin/bash
+isabelle build -d. -v Edka_Document
+rm -rf html/browser_info/*
+
+cp -a ~/.isabelle/Isabelle2015/browser_info/Unsorted/Edka_Document/* html/browser_info/
+
+./fix_urls.sh html/browser_info

### 2016-03-09 formalization/makedistr.sh

@@ -1,0 +1,41 @@
+#!/bin/bash
+
+echo 'Collecting'
+# Collect theory files
+THYS=`isabelle build -d. -nl Edka | egrep '^ +[^/ ]'`
+
+# Collect root files
+ROOTS=""
+function follow_roots {
+  local DIR="$1"
+  
+  if test -f "$DIR/ROOT"; then
+    ROOTS="$ROOTS $DIR/ROOT"
+  fi
+  if test -f "$DIR/ROOTS"; then
+    ROOTS="$ROOTS $DIR/ROOTS"
+    for r in `cat $DIR/ROOTS`; do
+      local ND="$DIR/$r"
+      [[ "$r" != '.' ]] && follow_roots "$ND"
+    done  
+  fi
+}
+
+follow_roots "."
+
+
+# Collect extra files
+
+EXTRA=""
+EXTRA="$EXTRA `find evaluation -type f -not -name '*~' -not -path '*/samples/*' -not -name '*.log' -not -name '*.class'`"
+EXTRA="$EXTRA `find evaluation -iname "README*"`"
+EXTRA="$EXTRA README.md"
+
+FILES="$THYS $ROOTS $EXTRA"
+
+OUTNAME="edmonds_karp.tgz"
+
+echo 'Making archive'
+tar --transform 's,^,edmonds_karp/,' -czf $OUTNAME $FILES
+
+cp $OUTNAME html

### 2016-03-09 formalization/makedistr.sh
Cleanup of line overflows in proof document, up to edka-algo

@@ -28,7 +28,6 @@ follow_roots "."
 
 EXTRA=""
 EXTRA="$EXTRA `find evaluation -type f -not -name '*~' -not -path '*/samples/*' -not -name '*.log' -not -name '*.class'`"
-EXTRA="$EXTRA `find evaluation -iname "README*"`"
 EXTRA="$EXTRA README.md"
 
 FILES="$THYS $ROOTS $EXTRA"
@@ -36,6 +35,6 @@ FILES="$THYS $ROOTS $EXTRA"
 OUTNAME="edmonds_karp.tgz"
 
 echo 'Making archive'
-tar --transform 's,^,edmonds_karp/,' -czf $OUTNAME $FILES
+tar --transform 's,^\./,,;s,^,edmonds_karp/,' -czf $OUTNAME $FILES
 
 cp $OUTNAME html

### 2016-05-23 paper/root.tex
Reviewer comments on paper

@@
 Our formal proof closely follows a standard textbook proof, and is accessible even without being
-an expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
+an expert in Isabelle/HOL --- the interactive theorem prover used for the formalization.
 We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
@@
 The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. An important instance is the Edmonds-Karp algorithm~\cite{EK72},
-which was one of the first algorithms to solve the maximum flow problem in polynomial time for the general case of networks with real valued capacities.
+which was one of the first algorithms to solve the maximum flow problem in polynomial time for the general case of networks with real-valued capacities.
 
 In this paper, we present a formal verification of the Edmonds-Karp algorithm and its polynomial complexity bound.
-The formalization is conducted entirely in the Isabelle/HOL proof assistant~\cite{NPW02}. 
+The formalization is conducted in the Isabelle/HOL proof assistant~\cite{NPW02}. 
 Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification into an abstract proof of the Ford-Fulkerson method,
@@
 The abstract parts of our verification closely follow the textbook presentation of Cormen et al.~\cite{CLRS09}.
-Being developed in the Isar~\cite{Wenzel99} proof language, our proofs are accessible even to non-Isabelle experts.
+Using the Isar~\cite{Wenzel99} proof language, we were able to produce proofs that are accessible even to non-Isabelle experts.
 
@@
 
-A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real valued capacity $c(u,v)>0$.
+A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real-valued capacity $c(u,v)>0$.
 Moreover, there are two distinct vertices $s,t\in V$, which are called \emph{source} and \emph{sink}.
@@
 In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges.
-For this purpose the Ford-Fulkerson method defines the residual graph, which has edges both in parallel and reverse to the edges of the network.
+For this purpose the Ford-Fulkerson method defines the residual graph, which has edges in the same and opposite direction as the network edges.
+% , which has edges both in parallel and reverse to the edges of the network. 
 Each edge is labeled by the amount of flow that can be effectively passed along this edge, by either increasing or decreasing the flow on a network edge.
@@
 \]
-Note that, for any edge in the network, the augmenting flow parallel to that edge is added to the flow, while the augmenting flow reverse to this edge is subtracted. 
+Note that, for any edge in the network, the augmenting flow in the same direction is added to the flow, while the augmenting flow in the opposite direction is subtracted. 
 This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.
@@
 The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}.
-When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real valued capacities.
+When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real-valued capacities.
 Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges if the intended meaning is clear from the context.
@@
 
-As an example consider the proof that for a flow $f$ and a residual flow $f'$, the augmented flow $f\augment f'$ is again a valid flow. In particular, one has to show 
+As an example, consider the proof that for a flow $f$ and a residual flow $f'$, the augmented flow $f\augment f'$ is again a valid flow. In particular, one has to show 
 that the augmented flow satisfies the capacity constraint. Cormen et al.~give the following proof, which we display literally here, only replacing the references to 
@@
 also possible results of \isai{m'}. 
-Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
+Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed-point construction~\cite{Kr10}.
 Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
@@
 \section{The Edmonds-Karp Algorithm}\label{sec:edka}
-  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm requires to show that finding a shortest augmenting path is a refinement of finding any augmenting path, which is straightforward using basic graph theory.
+  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm is straightforward, as finding a shortest augmenting path is a refinement of finding any augmenting path.
 
   Considerably more effort is required to show that the resulting algorithm terminates within $O(VE)$ iterations. 
-  The idea of the proof is as follows: Edges that are reverse to edges on a shortest path cannot lie 
+  The idea of the proof is as follows: Edges in the opposite direction to an edge on a shortest path cannot lie 
   on a shortest path itself.
+%   that are reverse to edges on a shortest path cannot lie 
+%   on a shortest path itself.
   On every augmentation, at least one edge of the residual graph that lies on a shortest augmenting path is flipped. Thus, either the length of the shortest path increases,
@@
   Formalizing the above intuitive argument was more tricky than it seemed on first glance:
-  While it is easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
-  to a graph transformation which may add multiple reverse edges and removes at least one original edge requires some generalization of the statement.
+  While it is easy to prove that, in a \emph{fixed graph}, an edge and its opposite cannot both lie on shortest paths, generalizing the argument
+  to a graph transformation which may add multiple flipped edges and removes at least one original edge requires some generalization of the statement.
   Note that a straightforward induction on the length of the augmenting path or on the number of flipped edges fails, as, after flipping the first 
-  edge, the path does no longer exist.
+  edge, the path no longer exists.
 
@@
   Having defined the measure function and shown that it decreases on augmentation, it is straightforward to refine the partial correct while loop to a total correct one.
-Moreover, to make explicit the bound on the number of loop iterations, we instrument the loop with a counter, and assert the upper bound on the counter after the loop.
+Moreover, to make explicit the bound on the number of loop iterations, we instrument the loop to count its iterations, and assert the upper bound after the loop.
 
@@
   \subsection{Manipulating Residual Graphs Directly}\label{sec:impl_res_graph}
-  In a next step, we observe that the algorithm is phrased in terms of a flow, which is updated until it is maximal. 
+  Next, we observe that the algorithm is phrased in terms of a flow, which is updated until it is maximal. 
   In each iteration, the augmenting path is searched on the residual graph induced by the current flow.
@@
   After our profiling indicated a hot spot on accessing the capacity matrices of the network and the flow, we switched to 
-  an algorithm that operates on a representation of the residual graph directly. This resulted in a speed-up of roughly factor two.
+  an algorithm that operates on a representation of the residual graph directly. This resulted in a speed-up of roughly a factor of two.
   As the residual graph uniquely determines the flow (and vice versa), it is straightforward to phrase the
@@
   A common optimization is to pre-compute an \emph{adjacency map} from nodes to adjacent nodes in the network. 
-  As an edge in the residual graph is either parallel or reverse to 
+  As an edge in the residual graph is either in the same or opposite direction of 
   a network edge, it is enough to iterate over the adjacent nodes in the network, and check whether they are actual successors in the residual graph.
@@
 
-  We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers.}.
+  We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers.
+Thus, we could chose any executable number representation here.}.
 Note that an implementation as fixed precision numbers would also be possible,
@@
   We implement nodes as natural numbers less than an upper bound $N$, and residual graphs are implemented by their capacity matrices, which, in turn,
-  are realized as arrays of size $N*N$ with row-major indexing, such that the successors of a node are close together in memory.
+  are realized as arrays of size $N\times N$ with row-major indexing, such that the successors of a node are close together in memory.
   The adjacency map of the network is implemented as an array of lists of nodes. 
@@
   to obtain the initial residual graph. This gives us some flexibility in using the algorithm, as any capacity matrix representation can be converted into a 
-  function easily, without loosing efficiency for read-access. 
+  function easily, without losing efficiency for read-access. 
   Similarly, our implementation expects an adjacency map as additional parameter, which is then tabulated into an array.
@@
 
-  \noindent Note that this theorem is stated as a Hoare triple, using separation logic~\cite{Rey02} assertions. There are no preconditions on the input.
+  \noindent Note that this theorem is stated as a Hoare triple, using separation logic~\cite{Rey02,LaMe12} assertions. There are no preconditions on the input.
   If the algorithm returns \isai{None}, then the edge list was malformed or described a graph that does not satisfy the network assumptions.
@@
 \section{Benchmarking}\label{sec:benchmark}
-  We have compared the running time of our algorithm against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}.
+  We have compared the running time of our algorithm in SML against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}.
 We have used MLton 20100608~\cite{MLton} and OpenJDK Java 1.7.0\_95, running on a standard laptop machine with a 2.8GHz i7 quadcore processor and 16GiB of RAM.

### 2016-05-23 paper/root.tex
---

@@
     However, proving that, e.g., augmenting paths only contain valid nodes is hard at this low level. 
-    Instead, one can protect the application of the capacity function by an assertion--- already on a high abstraction level where it can be easily discharged. 
+    Instead, one can protect the application of the capacity function by an assertion --- already on a high abstraction level where it can be easily discharged. 
     On refinement, this assertion is passed down, and ultimately available for the implementation.

### 2016-08-08 formalization/EdmondsKarp_Maxflow/make_submission_tgz.sh
AFP submission

@@ -1,0 +1,3 @@
+#!/bin/sh
+
+tar -czf EdmondsKarp_Maxflow.tgz --exclude="EdmondsKarp_Maxflow.tgz" --exclude-from=TGZ_EXCLUDE --exclude-tag-all=NOTGZ --exclude-backups --exclude-vcs --transform 's|^|EdmondsKarp_Maxflow/|' *

### 2017-02-14 paper/root.tex
paper final version
switched to condensed format
fixed affiliations
introduced the section about push-relabel algorithm

@@
-% Markierungen: TODO
-\documentclass{llncs}
+%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This is a general template file for the LaTeX package SVJour3
+% for Springer journals.          Springer Heidelberg 2010/09/16
+%
+% Copy it to a new file with a new name and use it as the basis
+% for your article. Delete % signs as needed.
+%
+% This template includes a few options for different layouts and
+% content for various journals. Please consult a previous issue of
+% your journal as needed.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\RequirePackage{fix-cm}
+%
+%\documentclass{svjour3}                     % onecolumn (standard format)
+\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
+%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+%\documentclass[twocolumn]{svjour3}          % twocolumn
+%
 
@@
 
+%
 \begin{document}
 
-\title{Formalizing the Edmonds-Karp Algorithm}
-% \subtitle{}
-
-\author{Peter Lammich, S.~Reza Sefidgar}
-
-\institute{Technische Universit\"at M\"unchen, \email{\{lammich,sefidgar\}@in.tum.de}}
+\title{Formalization of Network Flow Algorithms}
+\subtitle{a refinement approach in Isabelle/HOL}
+
+%\titlerunning{Short form of title}        % if too long for running head
+
+\author{Peter Lammich \and S. Reza Sefidgar}
+
+%\authorrunning{Short form of author list} % if too long for running head
+
+\institute{
+  Peter Lammich \at
+  Institut f\"ur Informatik,
+  Technische Universit\"at M\"unchen \\
+  \email{lammich@in.tum.de} %
+\and
+  S. Reza Sefidgar \at
+  Institute of Information Security,
+  Department of Computer Science,
+  ETH Zurich, Switzerland \\
+  \email{reza.sefidgar@inf.ethz.ch} %
+}
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
 
 \maketitle
+
 \begin{abstract}
@@
 Further refinement yields a verified implementation, whose execution time compares well to an unverified reference implementation in Java.
-
-%We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network. 
-%The formal proofs are done in Isabelle/HOL. They closely follow a standard textbook proof and are relatively easy 
-%to read, even without being an Isabelle expert.
-
-%We then use stepwise refinement to obtain the Edmonds-Karp algorithm and prove its upper complexity bound of $O(|V||E|^2)$.
-%Further refinement yields a verified implementation, which is almost as fast as an unverified  
-%reference implementation in Java.
-
-% 
-% and a verified imperative implementation of it. Our implementation is roughly 2.5 times slower than an unverified  
-% reference implementation in Java. 
-% 
-% Finally, we prove the upper complexity bound of $O(|V||E|^2)$ for the Edmonds-Karp algorithm.
+\keywords{First keyword \and Second keyword \and More}
+% \PACS{PACS code1 \and PACS code2 \and more}
+% \subclass{MSC code1 \and MSC code2 \and more}
 \end{abstract}
@@
 	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
-	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
+	& \geq f(u, v) + f'(u, v) - f(u, v) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
 	& = f'(u, v) \\
@@
   which is provable by reflexivity, but causes the generated program to diverge.
+
+  \section{Push-Relabel algorithms}\label{sec:push_relabel}
+  In this section we present the formalization of Push-relabel algorithms by
+  Goldberg(XXX) that build up on Karzanov's idea of a preflow ---a relaxation
+  of the flow definition by allowing the sum of flows over incoming edges to be
+  greater equal than the sum of flows over outgoing edges of each vertex of the
+  network. Karzanov was the first to leave the augmenting path idea in research
+  on finding the maximum-flow in networks. He came up with an $O(V^3)$
+  maximum-flow algorithm by relaxing some of the invariants in Dinitz's
+  algorithm (XXX Dinitz paper). Years later, Goldberg proposed the generic
+  Push-relabel algorithm which shares some ideas with Karzanov's algorithm, yet
+  it provides freedom for choosing the order of ``push'' and ``relabel''
+  operations. The push–relabel algorithm is considered one of the most efficient
+  maximum flow algorithms. The generic algorithm has a strongly polynomial
+  $O(V^2 E)$ time complexity, while specific variants of the algorithms achieve
+  even lower time complexities up to $O(V^2 \sqrt{E})$. In what follows we first
+  overview the transition from Ford-Fulkerson method to Push-relabel algorithm,
+  and then provide the formal explanation of Goldberg's generic algorithm.
+
+  One year after publication of Edmonds-Karp algorithm and without being aware
+  of their result, Dinitz proposed another strongly polynomial algorithm based
+  on Ford-Fulkerson method with $O(V^2 E)$ complexity. Original intention of
+  Dinitz was just to accelerate iterations of Ford-Fulkerson method by means of
+  a smart data structure. As finding the augmenting path is the bottleneck in each
+  iteration of the Ford-Fulkerson method, he decided to keep track of all shortest
+  paths from source to sink in a data-structure called the layered-network.
+  Therefore, unless the length of the shortest path from source to sink has
+  increased, we could use the layered-network for finding augmenting paths.
+  Finding an augmenting path in the layered-network would require O(V)
+  operations which is considerably more efficient than O(E) complexity of
+  finding such path by searching the residual graph. Note that initializing the
+  layered-network would has O(E) complexity, however this is only required
+  whenever the length of the shortest path from source to sink has increased.
+
+  Karzanov observed that at each phase of the Dinitz's algorithm it is
+  sufficient to use a blocking-flow for increasing the current flow, and
+  hence determining the next layered-network. He uses the term to refere to an
+  augmenting flow in the current layared-network that saturates at least one edge
+  on each path from source to the sink. During each phase, Karzanov's algorithm
+  maintains a preflow in a layered-network. The algorithm ``pushes'' flow
+  through the network to find a blocking-flow, and then ``balances'' the
+  resulting preflow to obtain a valid flow at the end of each phase. Cherkarsky
+  tired to eliminate balancing operation by canceling the previous goal of
+  Karzanov's algorithm for constructing a feasible flow at the end of each
+  phase. He proved that balancing the flow at some vertex is equivalent to
+  pushing from it in future when it will appear in some layared-network at a
+  greater distance from the sink. However, Cherkarsky could not come up with the
+  generic push-relabel algorithm. The leftovers of phase structures of Dinitz's
+  and Karsanov's algorithms prevented Cherkarsky from canceling the devision of
+  his algorithm into phases.
+
+  Goldberg's generic Push-relabel algorithm takes Karzanov's idea of using preflows
+  to the next level. This algorithm abandons the idea of finding a flow in each
+  phase, and also abandons the idea of global phases. Instead, Goldberg idea
+  maintains a preflow in the original network and pushes flow toward the sink
+  along what it estimates to be shortest paths. When the algorithm terminates,
+  the preflow becomes a valid flow which is also proved to be a maximum flow.
+
+  The Push-relabels algorithms maintain a preflow as well as a height-labeling
+  during the execution. A preflow in a network with source $s$ and sink $t$ is a
+  function $f: V \times V \mapsto \mathbb{R}$ that satisfies the capacity
+  constraint as well as a relaxed conservation constraint\footnote{every flow
+    related definition, like flow's value or flow's corresponding residual
+    graph can be generalized for preflows.}. That is, for all vertices except
+  the source $s$ the sum of flows over all incoming edges is greater equal than
+  the sum of flows over all outgoing edges. The excess flow $e(v)$ of a vertex
+  $v$ in the network is the amount by which in flow exceeds the out flow, and a
+  $v$ is overflowing if $e(v) > 0$. Moreover, a height-labeling in a $s$-$t$
+  network is a funtion $l: V \mapsto \mathbb{N}$ where $l(s) =
+  \left\vert{V}\right\vert$, $l(t) = 0$, and for every edge $(u, v)$ of the
+  residual graph we have that $l(u) \le l(v) + 1$.
+
+  We summerize Cormen's intutition behind Push-relabel algorithms to clarify why
+  such a labeling function is required. Consider vertices of the network to be
+  liquid containers and residual network's edges to be pipes that are connecting
+  those containers. The amount of flow that can be passed through each pipe
+  corresponds to its residual capacity. Moreover, each of the containers is
+  lying on a platform which elevates it to higher altitudes as the algorithm
+  makes progress. The height-labeling function keeps track of the heights of
+  each container, and the flow can only go from higher altitude containers to
+  the containers at lower heights.
+
+  At the beginning of the algorithm the source is at height
+  $\left\vert{V}\right\vert$, and all other containers are on the ground (height
+  $0$). The algorithm sends as much as flow downhill from the source to the
+  sink, hence all the edges outgoing from source are filled to the capacity.
+  This flow is stored in the intermediate containers (resembeling the idea of a
+  preflow), and is eventually ``pushed'' to the sink. During the execution, it
+  may happen that a non-empty container(resembeling the idea of an overflowing
+  vertex) is only attached to pipes with left-over capacities that are
+  connecting it to a containers at higher altitude. In such a case, the
+  container is elevated (``relabeled'') to a higher altitude so it can push its
+  excess flow. Eventually, all the flows that could get to the sink has reached
+  there, and the intermediate containers has sent their liquids back to the
+  source (resembeling a valid flow which is later proved to be also a maximum
+  flow).
+
+  \newcommand{\push}{{\mathbin\uparrow_P}}%
+  As it was presented in the above intuition, Push-relabel algorithms perform
+  two basic operations. The push operation $\push$ increases or decreases the preflow
+  value of any edge $(u, v)$ of the network by pushing flow along its
+  corresponding residual edges $(u, v)$ or $(v, u)$ respectively
+  \footnote{applying push operation to an edge$(u, v)$ decreases $e(u)$ while
+    increases $e(v)$ at the same time.}. However, it can be applied only if $u$
+  is an overflowing vertex and $l(u) = l(v) + 1$. This restriction is due to the
+  fact that no residual edges exists between two vertices with hight-labeling
+  difference of more that 1. Hence, we would gain nothing by allowing such a
+  push operation and the resulting preflow will remains unchanged. In the
+  following definition, let $\Delta = min (e(u), c_f(u, v))$.
+\[ (f \push (u,v)) (u', v') = 
+  \begin{cases}
+    f (u', v') \hfill & \text{ if $(u', v') \neq (u, v) \wedge (u', v') \neq (v, u)$}\\
+    f (u', v') + \Delta & \text{ if $(u, v) \in E$}\\  
+    f (v', u') - \Delta & \text{ if $(v, u) \in E$}\\
+    0 & \text{ otherwise}
+  \end{cases}
+\]
+
+Goldberg distinguishes two types of pushes which are latered used for proving
+the correctness and complexity of his algorithm. A push is called saturating if
+the edge $(u, v)$ is removed from the residual graph after the push operation,
+i.e. $e(u) \ge c_f(u, v)$. Otherwise, the push is called non-saturating, and the
+edge $(u, v)$ remains in the residual graph after the push operation.
+
+  \newcommand{\relabel}{{\mathbin\uparrow_L}}%
+Next, the relabel operation $\relabel$ which increases the height-labeling of an
+overflowing vertex $u$, when for all edges $(u, v)$ that are going out of $u$ in
+the residual graph we have $l(u) \le l(v)$. The overflowing precondition is
+necessary to make sure that in the following definition minimization is not
+applied to an empty set. Note that, according to the definition of the push
+operation $\push$ flow can only be pushed downhill. So, with relabel operation
+we enable an overflowing vertex to push its excess flow into one of its
+neighbours. By definition, neither the source $s$ nor the sink $t$ can change
+their height-labeling, so, the relabel operation does not apply to them.
+\[ (l \relabel u) (u') = 
+  \begin{cases}
+    l(u') \hfill & \text{ if $u' \neq u$}\\
+    1 + min \{l(v) \vert \text{$(u, v)$ is an edge of the residual graph}\}  & \text{ otherwise}
+  \end{cases}
+\]
+
+The generic Push-relabel algorithm is then defined by iterative application of the aformentioned basic
+operations. The algorithm starts by setting up an initial height-lableling and
+preflow function. \isai{init-labeling} operation outputs a function $l$ such that
+$l(s) = \vert V \vert$ and $l(u) = 0$ for any $u \neq s$. Also,
+\isai{init-preflow} procedure produces a preflow function $f$ such that $f(s, u)
+= c(s, u)$ for all edges going out of the source and $f(u, v) = 0$ for every
+other edge of the network. The algorithm, terminates when no more push or
+relabel operation is applicable, and according to correctness theorem of the
+algorithm the resulting preflow upon termination is indeed a maximum-flow of the network.
+
+\texttt{---Present either the refinement code or pseudocode---}
+
+\texttt{---Discuss correctness and termination proofs if required}
+
+\texttt{---Corelate to the FIFO and gap heuristic implementations}
+
+\texttt{---Discuss our time complexity proof, if required}
 
@@
 
-%   
-%   can be performed in two directions. First of all, the abstract theory can be extended to also support more advanced 
-%   maximum flow algorithms like push-relabel algorithms~\cite{GoTa88}
-%   
-%   
-%   
-%   What do the Coq-Guys have: CFML? Others?
-%   
-%   
-%   XXX, ctd here
-%   
-%   Unfortunately, it
-% 
-%   ... and related work
-%     Mizar-Formalization: What did they do.
-%     
-%     History of the formalization: 
-%       Impl: From first impl that was hardly able to compute a 100 nodes graph, to current efficient one.
-%     
-%     Future Work: Dinic. (Actually, our abstract scheme [almost] covers Dinic's algorithm!)
-%     
-% 
-% 
-%   
-% Contributions
-% 
-%   Formal proof of mincut maxflow
-%   fofo-scheme
-%   inst to edmonds karp
-%   complexity analysis of edmonds karp
-%   refinement down to executable code. Roughly 2.5 times slower than Java. (What about OCaml)
-%     + NetCheck
-%     
-% What shines (its a Pearl)
-%   Min-Cut Max-Flow: Textboook like formal reasoning: Comprehensible proof, BUT machine checked
-%     (Present one (carefully worked) example in paperI. We could use lemma \isai{augment_flow_presv_cap})
-%     
-%   Refinement based approach: Fofu-Scheme, instantiation to EdsKa. 
-%     +1: Abstract Algo looks almost like pseudo-code you would expect in textbook.
-%     +2: Fofu-Scheme proved correct for all aug-path finders. EdsKa is instantiation of it.
-%     +3: Modularity: Fofu-scheme and pathfinder developed+proved independently of each other.
-%   
-%   Down to executable code, plugging in efficient data structures.
-%   
-% Some minor contributions:
-%   Reusable BFS algorithm
-%   Imperative matrix data structure (really minor).
-%   
-%   
-%   
-%     
-% 
-% 
-%
+%\section{Introduction}
+%\label{intro}
+%Your text comes here. Separate text sections with
+%\section{Section title}
+%\label{sec:1}
+%\subsection{Subsection title}
+%\label{sec:2}
+%as required. Don't forget to give each section
+%and subsection a unique label (see Sect.~\ref{sec:1}).
+%\paragraph{Paragraph headings} Use paragraph headings as needed.
+%\begin{equation}
+%a^2+b^2=c^2
+%\end{equation}
+%
+%% For one-column wide figures use
+%\begin{figure}
+%% Use the relevant command to insert your figure file.
+%% For example, with the graphicx package use
+%  %\includegraphics{example.eps}
+%% figure caption is below the figure
+%\caption{Please write your figure caption here}
+%\label{fig:1}       % Give a unique label
+%\end{figure}
+%%
+%% For two-column wide figures use
+%\begin{figure*}
+%% Use the relevant command to insert your figure file.
+%% For example, with the graphicx package use
+% % \includegraphics[width=0.75\textwidth]{example.eps}
+%% figure caption is below the figure
+%\caption{Please write your figure caption here}
+%\label{fig:2}       % Give a unique label
+%\end{figure*}
+%%
+%% For tables use
+%\begin{table}
+%% table caption is above the table
+%\caption{Please write your table caption here}
+%\label{tab:1}       % Give a unique label
+%% For LaTeX tables use
+%\begin{tabular}{lll}
+%\hline\noalign{\smallskip}
+%first & second & third  \\
+%\noalign{\smallskip}\hline\noalign{\smallskip}
+%number & number & number \\
+%number & number & number \\
+%\noalign{\smallskip}\hline
+%\end{tabular}
+%\end{table}
+%
+%
+%%\begin{acknowledgements}
+%%If you'd like to thank anyone, place your comments here
+%%and remove the percent signs.
+%%\end{acknowledgements}
+%
+%% BibTeX users please use one of
+%%\bibliographystyle{spbasic}      % basic style, author-year citations
+%%\bibliographystyle{spmpsci}      % mathematics and physical sciences
+%%\bibliographystyle{spphys}       % APS-like style for physics
+%%\bibliography{}   % name your BibTeX data base
+%
+%% Non-BibTeX users please use
+%%\begin{thebibliography}{}
+%%
+%% and use \bibitem to create references. Consult the Instructions
+%% for authors for reference list style.
+%%
+%%\bibitem{RefJ}
+%% Format for Journal Reference
+%%Author, Article title, Journal, Volume, page numbers (year)
+%% Format for books
+%%\bibitem{RefB}
+%%Author, Book title, page numbers. Publisher, place (year)
+%% etc
+%% \end{thebibliography}
 

### 2017-02-22 formalization/Push_Relabel_Maxflow/evaluation/bench_exec.sh
Performance tuning + not measuring cf and am init time

@@ -1,8 +1,34 @@
 #!/bin/bash
 
-tests=`cat tests`
-platforms_fofu=`cat fofu-platforms`
-platforms_prpu=`cat prpu-platforms`
+for i in "$@"
+do
+case $i in
+    --tests=*)
+    tests="${i#*=}"
+    shift
+    ;;
+    --fofu=*)
+    platforms_fofu="${i#*=}"
+    shift
+    ;;
+    --prpu=*)
+    platforms_prpu="${i#*=}"
+    shift
+    ;;
+    *)
+      echo "Unknown command line argument: $i"
+      exit 1
+    ;;
+esac
+done
+
+
+test -n "$tests" || tests=`cat tests`
+test -n "$platforms_fofu" || platforms_fofu=`cat fofu-platforms`
+test -n "$platforms_prpu" || platforms_prpu=`cat prpu-platforms`
+
+test "$platforms_fofu" = "-" &&  platforms_fofu=""
+test "$platforms_prpu" = "-" &&  platforms_prpu=""
 
 for t in $tests; do
   echo "*********************"

### 2017-02-22 formalization/Push_Relabel_Maxflow/evaluation/bench_exec.sh
Command line arguments for bench-scripts

@@ -1,34 +1,6 @@
 #!/bin/bash
 
-for i in "$@"
-do
-case $i in
-    --tests=*)
-    tests="${i#*=}"
-    shift
-    ;;
-    --fofu=*)
-    platforms_fofu="${i#*=}"
-    shift
-    ;;
-    --prpu=*)
-    platforms_prpu="${i#*=}"
-    shift
-    ;;
-    *)
-      echo "Unknown command line argument: $i"
-      exit 1
-    ;;
-esac
-done
-
-
-test -n "$tests" || tests=`cat tests`
-test -n "$platforms_fofu" || platforms_fofu=`cat fofu-platforms`
-test -n "$platforms_prpu" || platforms_prpu=`cat prpu-platforms`
-
-test "$platforms_fofu" = "-" &&  platforms_fofu=""
-test "$platforms_prpu" = "-" &&  platforms_prpu=""
+. bench_cmdline.inc
 
 for t in $tests; do
   echo "*********************"

### 2017-02-22 formalization/Push_Relabel_Maxflow/evaluation/bench_make.sh

@@ -1,11 +1,10 @@
 #!/bin/bash
 
+. bench_cmdline.inc
+
 cd data
 ./build
 cd ..
-
-platforms_fofu=`cat fofu-platforms`
-platforms_prpu=`cat prpu-platforms`
 
 for p in $platforms_fofu; do
   echo "Compiling fofu-$p"

### 2017-02-22 formalization/Push_Relabel_Maxflow/evaluation/bench_print.sh

@@ -1,8 +1,6 @@
 #!/bin/bash
 
-tests=`cat tests`
-platforms_fofu=`cat fofu-platforms`
-platforms_prpu=`cat prpu-platforms`
+. bench_cmdline.inc
 
 tabs -9
 echo -ne "Name \t& "

### 2017-02-22 formalization/Push_Relabel_Maxflow/evaluation/prpu-Jadj/Prpu.java
Right placement of push-counter, added gap counter, added counters to SML

@@ -35,7 +35,7 @@ public class Prpu {
 
         CounterPR stat = G.getCounter();
 
-        return new MaxFlowResult(endTime - startTime, maxflow, stat.getDischargeCount(), stat.getPushCount(), stat.getRelabelCount());
+        return new MaxFlowResult(endTime - startTime, maxflow, stat.getDischargeCount(), stat.getPushCount(), stat.getRelabelCount(), stat.getGapCount());
     }
 
     // test client that reads input network, solves max flow, and prints results
@@ -64,6 +64,7 @@ public class Prpu {
             System.out.format(">>>disc: %d\n", result.getDischargeCount());
             System.out.format(">>>push: %d\n", result.getPushCount());
             System.out.format(">>>rlbl: %d\n", result.getRelabelCount());
+            System.out.format(">>>gap : %d\n", result.getGapCount());
 
         }
     }

### 2017-02-22 formalization/Push_Relabel_Maxflow/evaluation/prpu-Jadj/lib/CounterPR.java

@@ -1,7 +1,7 @@
 package lib;
 
 public class CounterPR {
-    private int dc, pc, rc;
+    private int dc, pc, rc, gc;
 
     public CounterPR() {
     }
@@ -18,6 +18,10 @@ public class CounterPR {
         return rc;
     }
 
+    public int getGapCount() {
+        return gc;
+    }
+    
     public void incDischarge(){
         this.dc++;
     }
@@ -29,4 +33,9 @@ public class CounterPR {
     public void incRelabel(){
         this.rc++;
     }
+    
+    public void incGap(){
+        this.gc++;
+    }
+    
 }

### 2017-02-22 formalization/Push_Relabel_Maxflow/evaluation/prpu-Jadj/lib/MaxFlowResult.java

@@ -2,16 +2,17 @@ package lib;
 
 public class MaxFlowResult {
     private long time, flowValue;
-    private int dc, pc, rc;
+    private int dc, pc, rc,gc;
 
 
-    public MaxFlowResult(long time, long flowValue, int dischargeCount, int pushCount, int relabelCount) {
+    public MaxFlowResult(long time, long flowValue, int dischargeCount, int pushCount, int relabelCount, int gapCount) {
         this.time = time;
         this.flowValue = flowValue;
 
         this.dc = dischargeCount;
         this.pc = pushCount;
         this.rc = relabelCount;
+        this.gc = gapCount;
     }
 
     public long getTime() {
@@ -32,6 +33,10 @@ public class MaxFlowResult {
 
     public int getRelabelCount() {
         return rc;
+    }
+
+    public int getGapCount() {
+        return gc;
     }
     
 }

### 2017-02-22 formalization/Push_Relabel_Maxflow/evaluation/prpu-Jadj/lib/PushRelabel.java

@@ -39,13 +39,13 @@ public class PushRelabel {
     }
 
     private void push(int from, int to) {
-        sc.incPush();
-
         int amt = (int) (min(excess[from], G[from][to]));
 
         if (dist[from] <= dist[to] || amt == 0)
             return;
 
+        sc.incPush();
+          
         G[from][to] -= amt;
         G[to][from] += amt;
 
@@ -56,6 +56,7 @@ public class PushRelabel {
     }
 
     private void gap(int k) {
+        sc.incGap();
         for (int v = 0; v < N; v++) {
             if (dist[v] < k) continue;
             count[dist[v]]--;

### 2017-02-23 paper/root.tex
draft for preflow-push section

@@
 
-  \section{Push-Relabel algorithms}\label{sec:push_relabel}
-  In this section we present the formalization of Push-relabel algorithms by
-  Goldberg(XXX) that build up on Karzanov's idea of a preflow ---a relaxation
-  of the flow definition by allowing the sum of flows over incoming edges to be
-  greater equal than the sum of flows over outgoing edges of each vertex of the
-  network. Karzanov was the first to leave the augmenting path idea in research
-  on finding the maximum-flow in networks. He came up with an $O(V^3)$
-  maximum-flow algorithm by relaxing some of the invariants in Dinitz's
-  algorithm (XXX Dinitz paper). Years later, Goldberg proposed the generic
-  Push-relabel algorithm which shares some ideas with Karzanov's algorithm, yet
-  it provides freedom for choosing the order of ``push'' and ``relabel''
-  operations. The push–relabel algorithm is considered one of the most efficient
-  maximum flow algorithms. The generic algorithm has a strongly polynomial
-  $O(V^2 E)$ time complexity, while specific variants of the algorithms achieve
-  even lower time complexities up to $O(V^2 \sqrt{E})$. In what follows we first
-  overview the transition from Ford-Fulkerson method to Push-relabel algorithm,
-  and then provide the formal explanation of Goldberg's generic algorithm.
-
-  One year after publication of Edmonds-Karp algorithm and without being aware
-  of their result, Dinitz proposed another strongly polynomial algorithm based
-  on Ford-Fulkerson method with $O(V^2 E)$ complexity. Original intention of
-  Dinitz was just to accelerate iterations of Ford-Fulkerson method by means of
-  a smart data structure. As finding the augmenting path is the bottleneck in each
-  iteration of the Ford-Fulkerson method, he decided to keep track of all shortest
-  paths from source to sink in a data-structure called the layered-network.
-  Therefore, unless the length of the shortest path from source to sink has
-  increased, we could use the layered-network for finding augmenting paths.
-  Finding an augmenting path in the layered-network would require O(V)
-  operations which is considerably more efficient than O(E) complexity of
-  finding such path by searching the residual graph. Note that initializing the
-  layered-network would has O(E) complexity, however this is only required
-  whenever the length of the shortest path from source to sink has increased.
-
-  Karzanov observed that at each phase of the Dinitz's algorithm it is
-  sufficient to use a blocking-flow for increasing the current flow, and
-  hence determining the next layered-network. He uses the term to refere to an
-  augmenting flow in the current layared-network that saturates at least one edge
-  on each path from source to the sink. During each phase, Karzanov's algorithm
-  maintains a preflow in a layered-network. The algorithm ``pushes'' flow
-  through the network to find a blocking-flow, and then ``balances'' the
-  resulting preflow to obtain a valid flow at the end of each phase. Cherkarsky
-  tired to eliminate balancing operation by canceling the previous goal of
-  Karzanov's algorithm for constructing a feasible flow at the end of each
-  phase. He proved that balancing the flow at some vertex is equivalent to
-  pushing from it in future when it will appear in some layared-network at a
-  greater distance from the sink. However, Cherkarsky could not come up with the
-  generic push-relabel algorithm. The leftovers of phase structures of Dinitz's
-  and Karsanov's algorithms prevented Cherkarsky from canceling the devision of
-  his algorithm into phases.
-
-  Goldberg's generic Push-relabel algorithm takes Karzanov's idea of using preflows
-  to the next level. This algorithm abandons the idea of finding a flow in each
-  phase, and also abandons the idea of global phases. Instead, Goldberg idea
-  maintains a preflow in the original network and pushes flow toward the sink
-  along what it estimates to be shortest paths. When the algorithm terminates,
-  the preflow becomes a valid flow which is also proved to be a maximum flow.
-
-  The Push-relabels algorithms maintain a preflow as well as a height-labeling
+  \section{Preflow-Push algorithms}\label{sec:push_relabel}
+  In this section we present the formalization of Preflow-push algorithms (a.k.a
+  Push-relabel algorithms) by Goldberg et al.~\cite{GoTa88}  that build up on
+  Karzanov's idea of a preflow~\cite{karzanov1974determination}. Karzanov was
+  the first to leave the augmenting path idea in research on finding the
+  maximum-flow in networks. He came up with an $O(V^3)$ maximum-flow algorithm
+  by relaxing some of the invariants in Dinitz's algorithm~\cite{Di06}. Years
+  later, Goldberg proposed the generic Preflow-push algorithm which shares some
+  ideas with Karzanov's algorithm, yet it provides freedom for choosing the
+  order of ``push'' and ``relabel'' operations. The Preflow-push algorithm is
+  considered one of the most efficient maximum flow algorithms. The generic
+  algorithm has a strongly polynomial $O(V^2 E)$ time complexity, while specific
+  variants of the algorithms achieve even lower time complexities up to $O(V^2
+  \sqrt{E})$.
+
+  % TODO
+  % @Peter: I think this first paragraph is enough for history, if you think
+  % more is required, you can look into the following paragraphs (for intro mabybe)
+  % TODO
+
+  %One year after publication of Edmonds-Karp algorithm and without being aware
+  %of their result, Dinitz proposed another strongly polynomial algorithm based
+  %on Ford-Fulkerson method with $O(V^2 E)$ complexity. Original intention of
+  %Dinitz was just to accelerate iterations of Ford-Fulkerson method by means of
+  %a smart data structure. As finding the augmenting path is the bottleneck in each
+  %iteration of the Ford-Fulkerson method, he decided to keep track of all shortest
+  %paths from source to sink in a data-structure called the layered-network.
+  %Therefore, unless the length of the shortest path from source to sink has
+  %increased, we could use the layered-network for finding augmenting paths.
+  %Finding an augmenting path in the layered-network would require O(V)
+  %operations which is considerably more efficient than O(E) complexity of
+  %finding such path by searching the residual graph. Note that initializing the
+  %layered-network would has O(E) complexity, however this is only required
+  %whenever the length of the shortest path from source to sink has increased.
+
+  %Karzanov observed that at each phase of the Dinitz's algorithm it is
+  %sufficient to use a blocking-flow for increasing the current flow, and
+  %hence determining the next layered-network. He uses the term to refere to an
+  %augmenting flow in the current layared-network that saturates at least one edge
+  %on each path from source to the sink. During each phase, Karzanov's algorithm
+  %maintains a preflow in a layered-network. The algorithm ``pushes'' flow
+  %through the network to find a blocking-flow, and then ``balances'' the
+  %resulting preflow to obtain a valid flow at the end of each phase. Cherkarsky
+  %tired to eliminate balancing operation by canceling the previous goal of
+  %Karzanov's algorithm for constructing a feasible flow at the end of each
+  %phase. He proved that balancing the flow at some vertex is equivalent to
+  %pushing from it in future when it will appear in some layared-network at a
+  %greater distance from the sink. However, Cherkarsky could not come up with the
+  %generic push-relabel algorithm. The leftovers of phase structures of Dinitz's
+  %and Karsanov's algorithms prevented Cherkarsky from canceling the devision of
+  %his algorithm into phases.
+
+  %Goldberg's generic Push-relabel algorithm takes Karzanov's idea of using preflows
+  %to the next level. This algorithm abandons the idea of finding a flow in each
+  %phase, and also abandons the idea of global phases. Instead, Goldberg idea
+  %maintains a preflow in the original network and pushes flow toward the sink
+  %along what it estimates to be shortest paths. When the algorithm terminates,
+  %the preflow becomes a valid flow which is also proved to be a maximum flow.
+
+  The Push-relabel algorithms maintain a preflow as well as a height-labeling
   during the execution. A preflow in a network with source $s$ and sink $t$ is a
@@
 
-Goldberg distinguishes two types of pushes which are latered used for proving
+Goldberg distinguishes two types of pushes which are later used for proving
 the correctness and complexity of his algorithm. A push is called saturating if
@@
 
-\texttt{---Present either the refinement code or pseudocode---}
-
-\texttt{---Discuss correctness and termination proofs if required}
-
-\texttt{---Corelate to the FIFO and gap heuristic implementations}
-
-\texttt{---Discuss our time complexity proof, if required}
+\texttt{\\\\---Present either the refinement code or pseudocode---}
+
+Correctness of the generic Push-relabel algorithm is proved in two steps. First,
+we show that algorithm returns a maximum flow if it terminates. Second, we show
+that the algorithm always terminates, hence, it is totaly correct.
+
+We start with the first step, and try to prove computation of a maximum flow
+upon termination of the algorithm. Maintaining helight-labeling function as well
+as a preflow during the execution of the algorithm is crucial for its
+correctness. Goldberg's idea is to use these functions to estimate a shortest
+path for pushing flow toward the sink.
+
+We proceed by proving that generic Push-relabel maintains a height-labeling
+function. Using the definition of relabel operation $\relabel$, we observe that
+the height-labeling $l(u)$ of any vertex $u \in V$ never decreases during the
+execution of the algorithm, and $(l \relabel u)$ increases $u$'s height by at
+least one. These two basic results prove that $l$ remains a height-labeling
+whenever it is modified by applying the relabel operation. Note taht, we also
+need to consider push operations. Pushing more flow along an edge $(u, v)$ in
+the residual graph, not only may remove $(u, v)$ from the residual graph, but
+also it can add $(v, u)$ to it. However, both of these modifications still
+maintain the height-labeling function because of the preconditions that push
+operator has.
+
+The maintanance of the preflow during the execution of the algorithm can be
+similiarly proved using the definition of the push operation $\push$. Next, we
+proceed to prove the correctness of the algorithm, by showing that it computes a
+maximum flow upon its termination. We use the fact about maintanance of
+height-labeling function to build an isabelle context \isa{Labeling}. We embed
+the rest of the correctness theorems inside this context, as it simplifies the
+proofs and gives us a cleaner formalization. That is, as we can access the
+height-labeling assumptions whenever they are required in some proof step.
+
+As an intermediate step, we should fist prove that for any preflow $f$ and
+height-labeling $l$, the residual graph corresponding to $f$ does not contain a
+path from the source $s$ to the sink $t$. In order to see this, let's by
+contradiction assume there is such a path. Using this path, we could obtain a
+simple path $p$ conntecting $s$ to $t$. As $l$ is a valid labeling, for each
+edge $(u, v)$ on $p$ we must have $l(u) \le l(v) + 1$, hence, $l(s) \le l(t) +
+\vert p \vert$. This yeilds the result that $l(s) \le \vert V \vert$, since
+$l(t) = 0$ and $p$ is assumed to be a simple path in the residual graph.
+However, this contradicts the assumption that $l$ is a height-labeling function,
+and completes the proof.
+
+We use the previous results to complete the proof of the first step. It is easy
+to see that \isai{init-preflow} and \isa{init-labeling} produce a valid preflow
+as well as a valid height-labeling function. We have also shown the maintanance
+of these properties earlier. When the algorithm terminates, the excess value
+$e(v)$ of each vertex $v \in V$ is equal to $0$. So, the preflow that was
+maintained during the execution becomes a valid flow. Given the fact that the
+height-labeling assumptions are also maintained and using the latest auxillary
+lemma, we can conclude that there is no path from source to sink in the residual
+graph. Hence, the result of the algorithm is a maximum flow using the
+Ford-fulkerson theorem.
+
+\texttt{\\\\---Discuss termination proofs if required}
+
+\texttt{\\\\---Discuss our time complexity proof, if required}
+
+\texttt{\\\\---talk about relabel-to-front and fifo instances}
 

### 2017-02-26 formalization/Push_Relabel_Maxflow/evaluation/data/generators/washington/washington.c
copied DIMACS generators to repository

@@ -1,0 +1,1420 @@
+  /* makegraph.c */
+
+/* graph.h */
+
+#include <sys/time.h>
+#include <stdio.h>
+/*#include <time.h>
+*/
+#include <ctype.h>
+#include <strings.h>
+
+/* ghead.h */
+
+#define DATA_DIRECTORY "/local/users/anderson/netflow/data/"
+
+#define FAILURE    0
+#define SUCCESS    1
+#define FALSE      0
+#define TRUE       1
+
+
+#define MAX_N     20000  
+/* #define MAX_N 60 */
+
+#define MAX_CAP   100000000
+
+/* Dimacs problem types */
+#define UNDEFINED        0
+#define MINCOSTFLOW      1
+#define MAXFLOW          2
+#define ASSIGNMENT       3
+
+typedef struct enode {
+  struct enode *next;
+  struct enode *mate;
+  int c;
+  int f;
+  int h;
+  int t;
+  int flag;
+} Edge;
+
+
+typedef struct {
+  Edge *A[MAX_N];
+  int V[MAX_N];
+  int size;
+  int max_v;
+} Graph;
+
+typedef struct {
+  int head, tail, size;
+  int *data;
+} Queue;
+
+
+
+
+
+#define MAX_RANGE 1000000
+#define MAX_DEGREE 20
+#define VERY_BIG 1000000
+
+int Range[] = {1000000, 500000, 250000, 125000, 62500, 31250,
+		 15625, 7812, 3906, 1953, 976, 488, 244, 122,
+		 61, 31, 15, 7, 4, 2};
+
+
+main(argc, argv)
+int argc;
+char *argv[];
+{
+  Graph *G, *Mesh(), *RLevel(), *R2Level(), *Match(), *SquareMesh(), 
+        *BasicLine(), *ExponentialLine(), *DExponentialLine(), *DinicBadCase(),
+        *GoldBadCase(), *Cheryian();
+
+  FILE *f;
+  int dim1, dim2, range, fct, s, t;
+
+  InitRandom(0);
+
+  if (argc != 6)
+    Barf("Usage: makegraph fct dim1 dim2 range file");
+
+  fct = atoi(argv[1]);
+  dim1 = atoi(argv[2]);
+  dim2 = atoi(argv[3]);
+  range = atoi(argv[4]);
+
[…]

### 2017-02-27 formalization/Push_Relabel_Maxflow/evaluation/data/data-cnv.c
parser for DIMACS format

@@ -3,16 +3,89 @@
 #include "time.h"
 #include "stdio.h"
 #include "stdlib.h"
+#include "limits.h"
 
 /* This function creates a square matrix to store our graph,
- * It also initializes all the capacities to zero */
-int** graph_init (int size) {
-	int** graph = malloc(size * sizeof(int*));
+ * then parses the DIMACS input until either EOF is entered
+ * or total edges are read. We add the source at index 0 
+ * and sink at index size - 1. Then we connect DIMACS s t
+ * to these new srouce sink. (Compatible with cleanup) */
+int** graph_init (int* v_count, int* e_count) {
+	int** graph;
+    
+    char dummy;
+    char dummies[10]; 
+    int in_edges = -1;
+    bool graph_init = false;
+    bool read_complete = false;
+    while(!read_complete){
+        
+        char* line;
+        size_t len = 0;
+        if(getline(&line, &len, stdin)!= -1) {
+            char command = line[0];
 
-	int i;
-	for (i = 0; i < size; i++)
-		graph[i] = calloc(size, sizeof(int));
+            switch (command) {
+                case 'p':
+                    if (!graph_init){
+                        int vs, es;
+                        sscanf(line, "%c %s %d %d", &dummy, dummies, &vs, &es);
 
+                        *v_count = vs + 2;
+                        *e_count = es + 2;
+
+                        in_edges = es;
+
+                        graph = malloc(*v_count * sizeof(int*));
+                        
+                        int i;
+                        for (i = 0; i < *v_count; i++)
+		                    graph[i] = calloc(*v_count, sizeof(int));
+
+                        graph_init = true;
+                    }
+                    break;
+                case 'n':
+                    {
+                        char ts;
+                        int id;
+                        sscanf(line, "%c %d %c", &dummy, &id, &ts);
+
+                        if (graph_init){
+                            if(ts == 's')
+                                graph[0][id] = INT_MAX;
+                            else
+                                graph[id][*v_count - 1] = INT_MAX;                                
+                        }
+                    }
+                    break;
+                case 'a':
+                    {
[…]

### 2017-02-27 formalization/Push_Relabel_Maxflow/evaluation/bench_clean.sh
enhanced clean up script

@@ -3,6 +3,8 @@
 find . -name "g-*.txt" -type f -delete
 find . -name "*.log" -type f -delete
 find . -name "data-gen" -type f -delete
+find . -name "data-cnv" -type f -delete
+find . -name "gen" -type f -delete
 find . -name "*.class" -type f -delete
 find . -name "Fofu" -type f -delete
 find . -name "Prpu" -type f -delete

### 2017-02-27 formalization/Push_Relabel_Maxflow/evaluation/data/data-cnv.c
modified converter to ignore self-loops and parallel edges

@@ -65,7 +65,8 @@ int** graph_init (int* v_count, int* e_count) {
                         sscanf(line, "%c %d %d %d", &dummy, &from, &to, &cap);
 
                         if (graph_init) {
-                            if (graph[to][from] == 0)
+							/* no self-loop or parallel edge otw. ignore */
+                            if (graph[to][from] == 0 && from != to)
                                 graph[from][to] = cap;
 
                             in_edges--;

### 2017-02-27 formalization/Push_Relabel_Maxflow/evaluation/data/data-cnv.c
integrated external generators into the build system --- modify according to the required graph's size now

@@ -26,14 +26,19 @@ int** graph_init (int* v_count, int* e_count) {
             char command = line[0];
 
             switch (command) {
+                /* Each Dimacs file must have only one p line, and it must occure before everything else. 
+                 * Hence, no requirement to check for recreation of the graph, or its existance in orther 
+                   cases either. (i assume all inputs are correct...) */
                 case 'p':
-                    if (!graph_init){
+                    {
                         int vs, es;
                         sscanf(line, "%c %s %d %d", &dummy, dummies, &vs, &es);
 
+                        /* index 0 and index *v_count - 1, are used for s an t. no matter what the input is */
                         *v_count = vs + 2;
                         *e_count = es + 2;
 
+                        /* set the counter for the loop. When all edges are read (in_edges = 0), exit */
                         in_edges = es;
 
                         graph = malloc(*v_count * sizeof(int*));
@@ -41,40 +46,42 @@ int** graph_init (int* v_count, int* e_count) {
                         int i;
                         for (i = 0; i < *v_count; i++)
 		                    graph[i] = calloc(*v_count, sizeof(int));
-
-                        graph_init = true;
                     }
                     break;
+
+                /* Read souce and sink nodes*/
                 case 'n':
                     {
                         char ts;
                         int id;
                         sscanf(line, "%c %d %c", &dummy, &id, &ts);
 
-                        if (graph_init){
-                            if(ts == 's')
-                                graph[0][id] = INT_MAX;
-                            else
-                                graph[id][*v_count - 1] = INT_MAX;                                
-                        }
+                        /* if it is s, connect our s (index 0) to it with infinit capacity 
+                         * if it is t, connect out t (index *v_count - 1) to it with infinite capacity */
+                        if(ts == 's')
+                            graph[0][id] = INT_MAX;
+                        else
+                            graph[id][*v_count - 1] = INT_MAX;                                
                     }
                     break;
+
+                /* Add next edge to the graph and decrease in_edge. (termination check...)*/
                 case 'a':
                     {
                         int from, to, cap;
                         sscanf(line, "%c %d %d %d", &dummy, &from, &to, &cap);
 
-                        if (graph_init) {
-							/* no self-loop or parallel edge otw. ignore */
-                            if (graph[to][from] == 0 && from != to)
-                                graph[from][to] = cap;
+                        /* no self-loop or parallel edge otw. ignore (according to DIMACS spec. this should never happer) */
[…]

### 2017-02-27 formalization/Push_Relabel_Maxflow/evaluation/data/data-cnv.c
modified test cases according to gen/conv/exec limitations

@@ -10,12 +10,12 @@
  * or total edges are read. We add the source at index 0 
  * and sink at index size - 1. Then we connect DIMACS s t
  * to these new srouce sink. (Compatible with cleanup) */
-int** graph_init (int* v_count, int* e_count) {
-	int** graph;
+long** graph_init (long* v_count, long* e_count) {
+	long** graph;
     
     char dummy;
     char dummies[10]; 
-    int in_edges = -1;
+    long in_edges = -1;
     bool graph_init = false;
     bool read_complete = false;
     while(!read_complete){
@@ -31,8 +31,8 @@ int** graph_init (int* v_count, int* e_count) {
                    cases either. (i assume all inputs are correct...) */
                 case 'p':
                     {
-                        int vs, es;
-                        sscanf(line, "%c %s %d %d", &dummy, dummies, &vs, &es);
+                        long vs, es;
+                        sscanf(line, "%c %s %ld %ld", &dummy, dummies, &vs, &es);
 
                         /* index 0 and index *v_count - 1, are used for s an t. no matter what the input is */
                         *v_count = vs + 2;
@@ -41,11 +41,11 @@ int** graph_init (int* v_count, int* e_count) {
                         /* set the counter for the loop. When all edges are read (in_edges = 0), exit */
                         in_edges = es;
 
-                        graph = malloc(*v_count * sizeof(int*));
+                        graph = malloc(*v_count * sizeof(long*));
                         
-                        int i;
+                        long i;
                         for (i = 0; i < *v_count; i++)
-		                    graph[i] = calloc(*v_count, sizeof(int));
+		                    graph[i] = calloc(*v_count, sizeof(long));
                     }
                     break;
 
@@ -53,23 +53,23 @@ int** graph_init (int* v_count, int* e_count) {
                 case 'n':
                     {
                         char ts;
-                        int id;
-                        sscanf(line, "%c %d %c", &dummy, &id, &ts);
+                        long id;
+                        sscanf(line, "%c %ld %c", &dummy, &id, &ts);
 
                         /* if it is s, connect our s (index 0) to it with infinit capacity 
                          * if it is t, connect out t (index *v_count - 1) to it with infinite capacity */
[…]

### 2017-02-27 formalization/Push_Relabel_Maxflow/evaluation/data/data-gen.c

@@ -182,7 +182,7 @@ int main (int argc, char** argv) {
 
       float density = (float)edge_count / (v_count * (v_count - 1));
       
-			printf("Successfully generated a graph with %d nodes and %d edges. Density is %.2f\n", v_count, edge_count, density);
+			printf("Generated a graph with %d nodes and %d edges. Density is %.2f\n", v_count, edge_count, density);
 		}
 		else {
 			printf("Number of vertices must be positive integer, target density must be in ]0..1].\n");

### 2017-02-27 formalization/Push_Relabel_Maxflow/evaluation/data/data-cnv.c
simple error with edges with infinite capacity

@@ -59,9 +59,9 @@ long** graph_init (long* v_count, long* e_count) {
                         /* if it is s, connect our s (index 0) to it with infinit capacity 
                          * if it is t, connect out t (index *v_count - 1) to it with infinite capacity */
                         if(ts == 's')
-                            graph[0][id] = LONG_MAX;
+                            graph[0][id] = INT_MAX;
                         else
-                            graph[id][*v_count - 1] = LONG_MAX;                                
+                            graph[id][*v_count - 1] = INT_MAX;                                
                     }
                     break;
 

### 2017-02-27 paper/root.tex
First draft of integrated fofu+prpu paper

@@
 \begin{abstract}
-We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
+We present a formalization of two classical algorithms for computing the maximum flow in a network:
+The Edmonds-Karp algorithm and Goldberg's Generic Push-Relabel algorithm. We prove the correctness and complexity of the algorithms.
 Our formal proof closely follows a standard textbook proof, and is accessible even without being
 an expert in Isabelle/HOL --- the interactive theorem prover used for the formalization.
-We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
-Further refinement yields a verified implementation, whose execution time compares well to an unverified reference implementation in Java.
-\keywords{First keyword \and Second keyword \and More}
+
+Using stepwise refinement, we obtain the Fifo-Push-Relabel algorithm with gap heuristics. 
+Further refinement yields verified efficient implementations of both, Edmonds-Karp and Fifo-Push-Relabel, 
+which compare well to unverified reference implementations in Java.
+
+\keywords{Maxflow Algorithm \and Edmonds-Karp Algorithm \and Push-Relabel Algorithm \and Formal Verification \and Isabelle/HOL \and Stepwise Refinement}
 % \PACS{PACS code1 \and PACS code2 \and more}
@@
 Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, as well as various scheduling and resource allocating problems can be reduced to it.
-The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. An important instance is the Edmonds-Karp algorithm~\cite{EK72},
-which was one of the first algorithms to solve the maximum flow problem in polynomial time for the general case of networks with real-valued capacities.
-
-In this paper, we present a formal verification of the Edmonds-Karp algorithm and its polynomial complexity bound.
+The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. 
+The basic idea is to repeatedly search for \emph{augmenting path} along which more flow can be moved from the source to the sink.
+An important instance is the Edmonds-Karp algorithm~\cite{EK72},
+which was one of the first algorithms to solve the maximum flow problem in polynomial time ($O(VE^2)$) for the general case of networks with real-valued capacities.
+Dinitz~\cite{Di06} generalized the idea of augmenting paths to blocking flows, obtaining an $O(V^2E)$ algorithm. 
+Karzanov~\cite{Ka74} was the first to propose an $O(V^3)$ algorithm based on the idea of \emph{preflows}. Based on Karzanov's ideas, Goldberg et al.\ proposed
+the generic push-relabel algorithm, which consists of repeated application of two local operations, push and relabel, in any order.
+Implementations of the push-relabel algorithm are among the most efficient maximum flow algorithms. While the generic algorithm has a time complexity of $O(V^2E)$,
+specific variants of the algorithms achieve even lower complexities up to $O(V^2\sqrt{E})$.
+
+In this paper, we present a formal verification of the Edmonds-Karp algorithm and of the generic push-relabel algorithm.
+We prove correctness of the algorithms as well as the time complexity bounds of $O(VE^2)$ and $O(V^2E)$, respectively. 
 The formalization is conducted in the Isabelle/HOL proof assistant~\cite{NPW02}. 
-Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification into an abstract proof of the Ford-Fulkerson method,
-its instantiation to the Edmonds-Karp algorithm, and finally an efficient implementation.
+Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification to separate the abstract algorithmic ideas from
+the implementation details: After proving the min-cut-max-flow theorem, it is straightforward to verify a generic version of the Ford-Fulkerson algorithm,
+which we then refine to Edmonds-Karp algorithm and finally to an efficient implementation. Similar, the generic push-relabel algorithm is refined to 
+the fifo-push-relabel algorithm, which, in turn, is refined to an efficient implementation.
 The abstract parts of our verification closely follow the textbook presentation of Cormen et al.~\cite{CLRS09}. 
@@
 
-While there exists another formalization of the Ford-Fulkerson method in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation.
+This paper is an extended version of our conference paper~\cite{LaSe16}, which only covered the Edmonds-Karp algorithm.
+
+While there exists another formalization of the Ford-Fulkerson method in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial time maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation.
 Moreover, this paper is a case study on elegantly formalizing algorithms.
@@
 
-A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges.
+A \emph{preflow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Non-generation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is not smaller than the sum of flows over all outgoing edges.
+We define the \emph{excess} flow $x_f(u)$ of a node $u$ to be the difference of the incoming and outgoing flow of $u$.
+Intuitively, a \emph{preflow} must not send more flow over an edge than its capacity allows, and only the source node may generate flow. 
+The excess flow is the amount of flow that is stuck at a node, \ie which flows in but does not flow out.
+A \emph{flow} is a \emph{preflow} that satisfies the \emph{Conservation constraint}: For all vertices except $s$ and $t$, the excess flow is zero.
+Intuitively, only the source may generate flow, and only the sink may consume flow, and for all other nodes, the incoming flow must be equal to the outgoing flow.
 The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
@@
 Each edge is labeled by the amount of flow that can be effectively passed along this edge, by either increasing or decreasing the flow on a network edge.
-Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
+Formally, the residual graph $G_f$ of a (pre)flow $f$ is the graph induced by the 
 edges with positive labels according to the following labeling function $c_f$:
@@
 
+\section{The Generic Push-Relabel Algorithm}  
+  We present our formalization of the generic push relabel algorithm of Goldberg et al.~\cite{GoTa88}. 
+  Again, our formalization closely follows the textbook presentation of Cormen et al.~\cite{CLRS09}.
+  The push-relabel algorithm maintains a preflow $f$ and a height-labeling $l: V \to \mathbb {N}$ of the nodes, 
+  with the invariant that for every edge $(u, v)$ of the residual graph we have that $l(u) \le l(v) + 1$. 
+  Moreover, the height of the source is fixed to $\left\vert{V}\right\vert$, and the height of the sink is fixed to $0$.
+  By generalizing the labeling invariant to paths, we get that for each path $p$ from $u$ to $v$, we have $l(u) \le l(v) + |p|$.
+  This implies that there cannot be a path from $s$ to $t$: The length of any (simple) path is less than $|V|$, however, a valid labeling 
+  implies a length greater than or equal to $|V|$.
+
+  A node that has positive excess is called \emph{active}. Upon termination of the algorithm, there will be no active nodes. 
+  Thus, the preflow is actually a flow, and as we have no $s$--$t$ paths, the min-cut-max-flow theorem implies that it is maximal.
+
+  Initially, the height of all nodes but the source is $0$, and the outgoing edges of the source get the maximum possible flow, 
+  and all other edges get no flow at all. Then, the algorithm repeats the following operations:
+  \begin{description}
+    \item[push] Send as much flow as possible over an edge $(u,v)$ with $l(u) = l(v)+1$. The maximum possible flow is constrained
+      by the residual capacity of the edge, and the excess flow available at $u$.
+    \item[relabel] Increase the height of an active (non-sink) node from which no push operation is possible. 
+      The height is increased to the minimum value such that a push becomes possible.
+  \end{description}
+  The algorithm is generic in the sense that it leaves open the order in which the push and relabel operations on the different nodes and edges are performed.
+  The algorithm terminates if no more operations apply. Then, there are no active nodes left, and the algorithm has constructed a maximum flow.
+
+\subsection{Formalization of the Generic Algorithm}
+For the complexity reasoning, we will count the different types of operations. Thus, it is convenient to formalize the algorithm as a labeled transition system
+over flows and heights, the labels indicating the performed operation:
+\begin{lstlisting}
+  inductive_set pr_lts :: "((preflow\<times>labeling) \<times> op_ty \<times> (preflow\<times>labeling)) set" 
+  where
+    "push_pre f l e ==> ((f,l),PUSH,(push_eff f e,l)) \<in> pr_lts"
+  | "relabel_pre f l u ==> ((f,l),RELABEL,(f,relabel_eff f l u)) \<in> pr_lts"
+\end{lstlisting}  
+Here, the \isai{inductive_set} construction defines the least relation that can be derived by the specified rules. 
+A push operation can be implemented in time $O(1)$, (only the flow on a single edge has to be modified) and a relabel operation can be implemented in time $O(V)$.
+(to find the new height, we have to consider the heights of all adjacent nodes in the residual graph) To this end, we define 
+\isai{cost PUSH == 1 | cost RELABEL == card(V)}. We prove the following theorem:
+\begin{lstlisting}
+theorem (in Network) generic_preflow_push_OV2E_and_correct:
+  assumes "((f_0, l_0), p, (f, l)) \<in> pr_lts^*" 
+  shows "(\<sum>x<-p. cost x) \<le> 38 * (card V)$^2$ * card E"
+    and "(f,l)\<notin>Domain pr_lts --> isMaxFlow f"
+\end{lstlisting}  
+where \isai$(f_0,l_0)$ is the initial flow and height-labeling. Intuitively, the cost of any (partial) run \isai$p$ is bounded by $O(V^2E)$, and if the 
+run is complete, \ie no more operations apply (\isai$(f,l)\<notin>Domain pr_lts$), the resulting flow is maximal.
+
+For the proof of this theorem, we follow the presentation of Cormen~\cite{CLRS09}. 
+However, at several places, we trade precision of the constant for simplicity of the proof. 
+Thus, our constant of $38$ is somewhat higher than what would result from Cormen's proof.
+
+For the correctness argument, we show that the operations of the algorithm maintain a valid labeling, and that existence of an active node implies applicability of 
+a push or relabel operation.
+The complexity argument is more tricky: First, we show that the labels of each node remain smaller than $2|V|$. This bounds the number of relabel operations to $O(V^2)$.
+For the push operations, we first define a push to be \emph{saturating}, if the excess at the source node of the push is not less than the capacity of the edge.
+In this case, the edge is completely saturated by the push and disappears from the residual graph. However, the source node of the push remains active.
+On the other hand, a non-saturating push deactivates the source node of the push.
+
+We count the number of saturating and non-saturating pushes separately.
+For saturating pushes, we regard the number of saturating pushes for each edge $(u,v)$ during execution of the algorithm: After a saturating push, the edge has disappeared from the residual graph.
+Thus, before the next saturating push on the same edge,
+there must be an operation that restores the edge. This can only be a push on the reverse edge $(v,e)$. 
+However, this push can only occur after the labeling of $v$ has increased. This can only happen $O(V)$ times, which yields a bound of $O(VE)$ on the number 
+of saturating push operations. Finally, the number of non-saturating pushes is bounded by defining a potential function that is decreased by non-saturating pushes,
+and increased by at most $O(V)$ by the other operations. The initial potential is $O(V)$, and with the already established bounds on the other operations we get a bound of
+$O(V^3 + V^2E) = O(V^2E)$ on the number of non-saturating push operations. Combining the bounds for the different operation types with the costs for the operations 
+then yields the desired time complexity bound of $O(V^2E)$.
+
+The simple structure of the generic algorithm allowed us to conveniently define it as labeled transition system, such that the above intuitive arguments could be formalized
+as counting different operation types on execution paths of the algorithm. Moreover, the powerful Isabelle/HOL standard library, in particular the lemmas for summations
+over finite sets, were very helpful. Anyway, the whole complexity proof is roughly 1000 lines of Isabelle text, and took a few men weeks to complete.
+
+\section{Implementing the Push-Relabel Algorithm}
+The generic algorithm does not specify the order in which the operations should be applied. 
+Moreover, choosing the right ordering may further decrease the worst-case time complexity, and, more important, the practical efficiency of the algorithm.
+
+We refine the generic push-relabel algorithm to different implementations. However, these have a more complicated algorithmic structure that
+prevents a convenient specification as labeled transition system, and thus a complexity proof in the style of the generic algorithm.
+Thus, we specify the algorithms using the Monadic Refinement Framework, and only prove correctness and termination 
+(which we get for free, as the complexity bound for the generic algorithm is also valid for all implementations).
+We leave it to future work to integrate complexity reasoning tools into the Refinement Framework and prove tighter complexity bounds for the specific implementations.
+
+\subsection{Relabel-to-Front Algorithm}
+A general idea of the implementation is to \emph{discharge} a node by repeatedly applying push and relabel operations to the same node, until it gets inactive.
+Cormen et al.\ present the relabel to front algorithm, which scans through a list of all nodes, discharging every node as it is visited.
+If the node gets relabeled during discharging, it is moved to the front of the list, and the scan of the list is restarted at the beginning of the list.
+The algorithm terminates if it reaches the end of the list. 
+During discharging a node $u$, the algorithm iterates over the list of all adjacent nodes $v$ in the network, and checks whether there is an edge $(u,v)$ in 
+the residual graph. When the discharge operation stops without relabeling $u$, the current position in the list of adjacent nodes is stored, and 
+the next discharge operations starts at exactly this position.
+
+The main idea of the correctness proof is to show that all nodes left of 
+the current position in the list are inactive. Thus, upon reaching the end of the list, all nodes are inactive, and the current flow is maximal.
+In order to show maintenance of this invariant, one has to strengthen it by adding that the list is a topological ordering of the 
+\emph{admissible network}, which consists of all edges $(u,v)$ of the residual graph with $l(u) = l(v) + 1$.
+
+Following the outline by Cormen et al., and using the Isabelle Refinement Framework to conveniently describe the abstract algorithm,
+the correctness proof offered no surprises. It is about 700 lines of Isabelle text plus 150 lines of general lemmas on topological orderings, 
+and took only a few days to develop.
+
+\subsection{FIFO Push-Relabel Algorithm}
+While the relabel-to-front algorithm has a theoretical worst case complexity of $O(V^3)$, it seems to be rather inefficient in practice.\footnote{We have not found any implementation based on relabel-to-front, and our own experiments indicate a rather poor performance.}
+An alternative implementation is the FIFO push-relabel algorithm, which maintains a first-in-first-out queue of the active nodes. 
+In each iteration, it pulls a node from the FIFO, and repeatedly pushes flow out of it.
+If no push operations are possible any more, but the node is still active, it is relabeled and appended to the FIFO.
+
+The FIFO push relabel algorithm has several advantages: It's correctness is straightforward to prove, 
+the only required additional invariant is that the queue contains exactly the active nodes. 
+Moreover, it performs well in practice, and we have found several implementations of it~\cite{ChGo97,ICPC-notebook}.
+
+Our algorithm follows the implementation that we found in the Stanford ACM-ICPC notebook~\cite{ICPC-notebook}, 
+which uses an additional gap heuristics: If there is a value $0<k<|V|$ such that there is no node that has height $k$, 
+all nodes with $k<l(v)<|V|$ may be relabeled to $|V|+1$, preserving a valid labeling. In practice, this may drastically accelerate convergence of the algorithm,
+as it prevents unnecessary pushes of flow towards the sink when the path to the sink is already blocked.
+The correctness proof of this heuristics is straightforward from the definitions, and also the overall correctness proof 
+is straightforward: It's size is about 400 lines of code, and it took roughly a day to complete.
+
 \section{Refinement to Executable Code}\label{sec:executable}
-  In the previous section, we have presented our abstract formalization of the Edmonds-Karp algorithm,
-  leaving open how to obtain a shortest augmenting path and how to implement the algorithm. 
+  In the previous section, we have presented our abstract formalization of the Edmonds-Karp and push-relabel algorithms.
+  However, we left open several algorithmic aspects. For example, we did not specify how to find a shortest augmenting path.
+  Moreover, some of the used operations require additional auxiliary data structures to be maintained. For example, 
+  we maintain the current excess of each node explicitly, instead of summing up the flows of incoming and outgoing edges each time 
+  we need to check whether a node is active.
+  Finally, we have to specify the concrete data structures that we want to use for implementation. 
+  For example, the residual graph is modeled as a function from edges to capacities on the abstract level. In the implementation, we want to use a two-dimensional array.
+
   In this section, we outline the further refinement steps that were necessary to obtain an efficient implementation.
@@
   \subsection{Manipulating Residual Graphs Directly}\label{sec:impl_res_graph}
-  Next, we observe that the algorithm is phrased in terms of a flow, which is updated until it is maximal. 
+  Next, we observe that the Edmonds-Karp algorithm is phrased in terms of a flow, which is updated until it is maximal. 
   In each iteration, the augmenting path is searched on the residual graph induced by the current flow.
@@
 
+  For the push-relabel algorithms, we combine the transition from preflows to residual graphs with the introduction of an 
+  explicit map from nodes to their current excess values: The refinement relation $\{((x_f,c_f),f) \mid \text{$f$ is a flow}\}$
+  relates pairs of excess values and residual graphs to the corresponding flow.
+
+  \subsection{Height Frequencies}
+  For an efficient implementation of the gap heuristics, we introduce an explicit map from height-values to then number of nodes with this height.
+  Using this map, it is easy to detect when we relabel the last node of a specific height and the gap heuristics applies.
+
   \subsection{Implementing Augmentation}    
@@
 
+  \subsection{Algorithms for Relabel and Discharge}    
+  For the discharge operation, it is straightforward to find the new label by iterating over the adjacent edges in the residual graph.
+  The discharge operation is a bit more tricky: Abstractly, we nondeterministically select a new edge to push, until we find no more edges.
+  For the push-to-front algorithm, we select the edges from a set of unprocessed edges which is maintained globally.
+  In the concrete implementation, however, we fold over a list of edges. The structure of the resulting concrete algorithms is somewhat different from the structure
+  of the abstract algorithm --- too different for the heuristics of our verification condition generator. Thus, we had to manually prove some rules specifically 
+  tailored to the required structural changes.
+
   \subsection{Computing Successors}
   In order to find an augmenting path, the BFS algorithm has to compute the successors of a node in the residual graph. 
-  Although this can be implemented on the edge labeling function by iterating over all nodes, this implementation tends to be inefficient for sparse graphs,
+  Similar, the relabel operation has to iterate over all successors to find the new minimal labeling, and the discharge operations iterate over all outgoing edges of a node.
+  Although these iterations can be implemented on the edge labeling function by iterating over all nodes, this implementation tends to be inefficient for sparse graphs,
   where we would have to iterate over many possible successor nodes just to find that there is no edge.
@@
   function easily, without losing efficiency for read-access. 
-  Similarly, our implementation expects an adjacency map as additional parameter, which is then tabulated into an array.
-This is convenient in our context, where a preprocessing step computes the adjacency map anyway.
-
-  The output flow of the algorithm is represented as the residual graph. The user can decide how to compute the maximum flow from it. For example,
+  Similarly, our implementations expect an adjacency map as additional parameter, which is then tabulated into an array. This is convenient in our context, 
+  where a preprocessing step computes the adjacency map anyway.
+
+  The output flow of the algorithms is represented as the residual graph. The user can decide how to compute the maximum flow from it. For example,
   in order to compute the maximum flow value, only the outgoing edges of the source node have to be computed, which is typically less 
   expensive than computing the complete flow matrix. The correctness theorem of the algorithm abstractly states how to obtain the maximum flow from the output.
+
+  For the push-relabel algorithms, the map from nodes to the excess, and the height frequency map are implemented by arrays. 
+  The queue of the FIFO push-relabel algorithm is implemented by two stacks, yielding amortized constant time enqueue and dequeue operations.
+  Similar, the list of the relabel to front algorithm is implemented as two lists, \emph{left} and \emph{right}, that contain the nodes left and right of the current position,
+  the left list being stored in reverse order. Thus, advancing the position is implemented in constant time by popping the first element of the right list, and pushing 
+  it to the left list. Resetting the position to the front of the list is implemented by pushing all elements from the left list to the right list. 
+  By accounting these push operations to the operations that created the left list, resetting the position can be estimated to be amortized constant time.
 
@@
   This calls for a deforestation optimization to get rid of the intermediate list, and iterate only two times over the predecessor map directly.
-%   A first time to compute the bottleneck, and a second time to do the augmentation. 
-  Fortunately, iteration over the shortest path seems not to significantly contribute to the runtime of our implementation, such that 
-  we leave this optimization for future work.
+  Also the FIFO queue could be implemented by a ring buffer based on an array, yielding true constant-time operations with less overhead.
+  Fortunately, our profiling data shows that these operations do not significantly contribute to the runtime of our implementations, such that 
+  we leave these optimizations for future work.
 
@@
 
-  Combining the implementation of the Edmonds-Karp algorithm with the network checker yields our final implementation,
-  for which we can export code, and have proved the following theorem:
+  Combining the implementations of our algorithms with the network checker yields our final implementations for which we export code.
+  On any input, the implementation checks whether the input is well-formed and returns a maximal flow for well-formed inputs.
+  We display the theorem for the Edmonds-Karp algorithm here, the theorems for relabel-to-front and FIFO push-relabel are analogous.
 
@@
 
+  \begin{comment}
   \section{Preflow-Push algorithms}\label{sec:push_relabel}
@@
 \texttt{\\\\---talk about relabel-to-front and fifo instances}
+
+\end{comment}
 
@@
   We conclude with some statistics:
-  The formalization consists of roughly 8000 lines of proof text, where the graph theory up to the Ford-Fulkerson algorithm requires 3000 lines.
+  The formalization consists of roughly 13000 lines of proof text, where the graph theory up to the Ford-Fulkerson algorithm requires 3000 lines.
   The abstract Edmonds-Karp algorithm and its complexity analysis contribute 800 lines, and its implementation (including BFS) another 1700 lines.
+  The generic push relabel algorithm and its complexity analysis is 1700 lines, the abstract implementations of relabel-to-front and FIFO push-relabel is another 1000 lines,
+  and their refinement to an efficient implementation takes 2000 lines.
+
   The remaining lines are contributed by the network checker and some auxiliary theories.
-The development of the theories required roughly 3 man month, a significant amount of this time going into a first, purely functional version of the implementation, which was later dropped in favor of the faster imperative version.
+The development of the theories required roughly 5 man month, a significant amount of this time going into a first, purely functional implementation of the Edmonds-Karp algorithm, which was later dropped in favor of the faster imperative version.
 
@@
   \subsection{Future Work}
-  Future work includes the optimization of our implementation, and the formalization of more advanced maximum flow algorithms, like Dinic's algorithm~\cite{Di06} or push-relabel algorithms~\cite{GoTa88}.
-  We expect both formalizing the abstract theory and developing efficient implementations to be challenging but realistic tasks.
+  While we implemented a simple version of the push-relabel algorithm, there are more advanced implementations, like the one of Cherkassky et el.~\cite{ChGo97} that 
+  we also use in the benchmarks. The same authors also provide an implementation based on the highest label selection rule, which is reported to be even more efficient 
+  in practice.
+Our formalized algorithm misses two important heuristics of these advanced algorithms: First, the global relabeling heuristics periodically performs a breadth first search on the reversed residual graph, to determine exact node labels based on their distance to the sink.
+Second, the algorithm is split in two phases:
+  The first phase ignores nodes with labels greater $|V|$, and already determines the exact value of the flow. The second phase only pushes excess flow back to the source. 
+  Using different heuristics for both phases is reported to improve practical performance~\cite{ChGo97}. 
+  Thus, the main focus of future work will be to verify these heuristics in order to obtain more efficient verified algorithms.
 

### 2017-02-28 formalization/Push_Relabel_Maxflow/evaluation/prpu-C++/prpu.cpp
Suitable benchmarks for Prpu

@@ -13,7 +13,7 @@
 
 using namespace std;
 
-typedef long LL;
+typedef long long LL;
 
 struct Edge {
   int from, to, cap, flow, index;
@@ -23,22 +23,12 @@ struct Edge {
 
 struct PushRelabel {
   int N;
-  vector<Edge>*  G;
-  LL* excess;
-  int *dist, *count;
-  bool * active;
+  vector<vector<Edge> > G;
+  vector<LL> excess;
+  vector<int> dist, active, count;
   queue<int> Q;
 
-  PushRelabel(int N) : N(N) {
-    this->G = new vector<Edge>[N];
-    for(int i = 0; i < N; i++)
-      G[i].reserve(10);
-
-    this->excess = new LL[N];
-    this->dist = new int[N];
-    this->active = new bool[N];
-    this->count = new int[2 * N];
-  }
+  PushRelabel(int N) : N(N), G(N), excess(N), dist(N), active(N), count(2*N) {}
 
   void AddEdge(int from, int to, int cap) {
     G[from].push_back(Edge(from, to, cap, 0, G[to].size()));
@@ -113,7 +103,6 @@ struct PushRelabel {
   }
 };
 
-
 int main(int argc, char** argv) {
 	if (argc < 2) {
 		cout << "Usage: <GRAPH-PATH>\n";
@@ -137,7 +126,7 @@ int main(int argc, char** argv) {
 
 		printf("@@@time: %.0f ms\n", ((double)(clock() - tStart)/CLOCKS_PER_SEC) * 1000);
 		printf("[Input (V E): %d %d]\n", V_size, E_size);
-		printf("@@@max-flow: %ld\n", maxFlow);
+		printf("@@@max-flow: %lld\n", maxFlow);
 	}
 
 	return 0;

### 2017-02-28 paper/root.tex
diagram of all benchmarks
final diagrams

@@
         legend style = { 
-          at={(0.59,0.4)}, 
+          at={(0.515,0.24)}, 
             anchor=north west,
@@
       \addplot table {
-        1000 .027   
-        1500 .124  
-        2000 .299  
-        2500 .665  
-        3000 1.335 
-        3500 2.526 
-        4000 3.111 
-        4500 6.917 
-        5000 7.743 
-        5500 10.452    
-      }; \addlegendentry{SML, sparse};
+        1000 .024   
+        1500 .098  
+        2000 .211  
+        2500 .693  
+        3000 2.014 
+        3500 3.153 
+        4000 3.679 
+        4500 8.364 
+        5000 10.328 
+        5500 11.749    
+      }; \addlegendentry{edka-SML, sparse};
 
       \addplot table {
-        1000 .016   
-        1500 .049   
-        2000 .168  
-        2500 .453  
-        3000 .842  
-        3500 1.524 
-        4000 2.136 
-        4500 3.438 
-        5000 4.734 
-        5500 6.369     
-      }; \addlegendentry{Java, sparse};
+        1000 .024   
+        1500 .054   
+        2000 .103  
+        2500 .399  
+        3000 .719  
+        3500 1.362 
+        4000 2.036 
+        4500 3.334 
+        5000 4.785 
+        5500 6.079     
+      }; \addlegendentry{edka-Java, sparse};
+
+      \addplot table {
+        1000 .005   
+        1500 .013  
+        2000 .024  
+        2500 .037  
+        3000 .066 
+        3500 .077 
+        4000 .096 
+        4500 .146 
+        5000 .194 
+        5500 .192    
+      }; \addlegendentry{prpu-SML, sparse};
     %end data for plotting
@@
         xtick = {0,1000,1100,...,1400},
-        ytick = {2,4,...,10},
+        ytick = {1,4,...,16},
         xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, 
-        yticklabels = {$2$, $4$,$6$,$8$,$10$}, 
+        yticklabels = {$1$,$4$,$8$,$12$,$16$}, 
         smooth,
@@
         legend style = { 
-          at={(0.59,0.4)}, 
+          at={(0.53,0.6)}, 
             anchor=north west,
@@
       \addplot table {
-        1000 3.328 
-        1050 3.345 
-        1100 4.841 
-        1150 5.036 
-        1200 5.840 
-        1250 6.858 
-        1300 7.651 
-        1350 8.950 
-        1400 8.398 
-        1450 10.769     
-      }; \addlegendentry{SML, dense};
+        1000	2.794
+        1050	2.9
+        1100	4.405
+        1150	4.435
+        1200	5.232
+        1250	6.029
+        1300	7.699
+        1350	8.91
+        1400	7.225
+        1450	10.042 
+      }; \addlegendentry{edka-SML, dense};
 
       \addplot table {
-        1000 4.141 
-        1050 4.778 
-        1100 5.465 
-        1150 6.248 
-        1200 7.128 
-        1250 8.137 
-        1300 9.864 
-        1350 11.008
-        1400 11.902
-        1450 13.365     
-      }; \addlegendentry{Java, dense};
+        1000	3.982
+        1050	4.463
+        1100	5.278
+        1150	6.27
+        1200	6.939
+        1250	7.953
+        1300	9.392
+        1350	10.317
+        1400	14.71
+        1450	13.11  
+      }; \addlegendentry{edka-Java, dense};
+
+      \addplot table {
+        1000	0.074
+        1050	0.059
+        1100	0.068
+        1150	0.075
+        1200	0.102
+        1250	0.085
+        1300	0.096
+        1350	0.103
+        1400	0.088
+        1450	0.114  
+      }; \addlegendentry{prpu-SML, dense};
+    \end{axis}
+  \end{tikzpicture}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
+    \begin{axis}[
+      %title={Benchmark of the Edmonds-Karp algorithm},
+        axis x line*= bottom,
+        axis y line*= left,
+        xmode = log,
+        ymode = log,
+%         xlabel = Number of vertices in the network,
+%         ylabel = Execution time in seconds, 
+        xtick = {3000, 6000, 10500, 13500, 16500},
+        ytick = {0.1,1,10,50,100},
+        xticklabels = {$3k$, $6k$, $10k$, $13k$, $16k$},
+        yticklabels = {$\text{\ \,}0.1$,$1$,$10$,$50$,$100$}, 
+        smooth,
+        cycle list name = black white,
+        legend style = { 
+          at={(0.53,0.24)}, 
+            anchor=north west,
+            draw=black, 
+            fill= white,
+            align=left
+        }
+    ]
+
+    %start data for plotting
+
+      \addplot table {
+        3000	1.185
+        4500	3.524
+        6000	6.457
+        7500	11.509
+        9000	17.529
+        10500	26.185
+        12000	36.828
+        13500	48.614
+        15000	86.449
+        16500	89.661
+      }; \addlegendentry{prpu-SML, wash};
+
+      \addplot table {
+        3000	0.906
+        4500	2.708
+        6000	5.083
+        7500	8.792
+        9000	12.951
+        10500	19.075
+        12000	27.466
+        13500	35.964
+        15000	61.839
+        16500	67.039
+      }; \addlegendentry{prpu-C++, wash};
+
+      \addplot table {
+        3000	0.006
+        4500	0.015
+        6000	0.029
+        7500	0.042
+        9000	0.063
+        10500	0.084
+        12000	0.115
+        13500	0.144
+        15000	0.183
+        16500	0.223
+      }; \addlegendentry{prpu-C, wash};
     %end data for plotting
@@
   \end{tikzpicture}
-
-%   \centering    
-%   \begin{tikzpicture}[thick,scale=1, every node/.style={scale=1}] %change the scales if you like to reduce the size
-%     \begin{axis}[
-%       %title={Benchmark of the Edmonds-Karp algorithm},
-%         axis x line*= bottom,
-%         axis y line*= left,
-%         xmode = log,
-%         ymode = log,
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
+    \begin{axis}[
+      %title={Benchmark of the Edmonds-Karp algorithm},
+        axis x line*= bottom,
+        axis y line*= left,
+        xmode = log,
+        ymode = log,
 %         xlabel = Number of vertices in the network,
-%         ylabel = Execution time in milliseconds, %in milliseconds, modify the scale for lables
-%         xtick = {0,1000,...,5000},
-% %         ytick = data,
-%         xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$}, %corresponds to the first column in {SML, Sparse}
-% %         yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
-%         smooth,
-%         cycle list name = black white,
-%         legend style = { 
-%           at={(0.59,0.4)}, 
-%             anchor=north west,
-%             draw=black, 
-%             fill= white,
-%             align=left
-%         }
-%     ]
-%     
-%     %start data for plotting
-%       \addplot table {
-%         1000 27   
-%         1500 124  
-%         2000 299  
-%         2500 665  
-%         3000 1335 
-%         3500 2526 
-%         4000 3111 
-%         4500 6917 
-%         5000 7743 
-%         5500 10452    
-%       }; \addlegendentry{SML, sparse};
-% 
-%       \addplot table {
-%         1000 16   
-%         1500 49   
-%         2000 168  
-%         2500 453  
-%         3000 842  
-%         3500 1524 
-%         4000 2136 
-%         4500 3438 
-%         5000 4734 
-%         5500 6369     
-%       }; \addlegendentry{Java, sparse};
-% 
-%       \addplot table {
-%         1000 3328 
-%         1050 3345 
-%         1100 4841 
-%         1150 5036 
-%         1200 5840 
-%         1250 6858 
-%         1300 7651 
-%         1350 8950 
-%         1400 8398 
-%         1450 10769     
-%       }; \addlegendentry{SML, dense};
-% 
-%       \addplot table {
-%         1000 4141 
-%         1050 4778 
-%         1100 5465 
-%         1150 6248 
-%         1200 7128 
-%         1250 8137 
-%         1300 9864 
-%         1350 11008
-%         1400 11902
-%         1450 13365     
-%       }; \addlegendentry{Java, dense};
-%     %end data for plotting
-%     \end{axis}
-%   \end{tikzpicture}
-
-%   \center{\footnotesize The x-axis shows the number of nodes, the y axis the execution time in seconds.}
+%         ylabel = Execution time in seconds, 
+        xtick = {0,1000,1100,...,1400},
+        ytick = {0.01, 0.1, 0.2, 0.5, 1},
+        xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, 
+        yticklabels = {$0.01$, $0.1$,$0.2$,$0.5$,$1$}, 
+        smooth,
+        cycle list name = black white,
+        legend style = { 
+          at={(0.52,0.24)}, 
+            anchor=north west,
+            draw=black, 
+            fill= white,
+            align=left
+        }
+    ]
+
+    %start data for plotting
+      \addplot table {
+        1000	0.141
+        1050	0.202
+        1100	0.157
+        1150	0.273
+        1200	0.24
+        1250	0.408
+        1300	0.465
+        1350	1.416
+        1400	0.657
+        1450	0.806
+      }; \addlegendentry{prpu-SML, acnet};
+
+      \addplot table {
+        1000	0.121
+        1050	0.17
+        1100	0.123
+        1150	0.216
+        1200	0.197
+        1250	0.237
+        1300	0.298
+        1350	0.621
+        1400	0.399
+        1450	0.486
+      }; \addlegendentry{prpu-C++, acnet};
+
+      \addplot table {
+        1000	0.056
+        1050	0.09
+        1100	0.069
+        1150	0.106
+        1200	0.09
+        1250	0.152
+        1300	0.163
+        1350	0.176
+        1400	0.223
+        1450	0.209
+      }; \addlegendentry{prpu-C, acnet};
+    \end{axis}
+  \end{tikzpicture}
+  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
+    \begin{axis}[
+      %title={Benchmark of the Edmonds-Karp algorithm},
+        axis x line*= bottom,
+        axis y line*= left,
+        xmode = log,
+        ymode = log,
+%         xlabel = Number of vertices in the network,
+%         ylabel = Execution time in seconds, 
+        xtick = {16000,19000,23000,27000,30000},
+        ytick = {0.1,1,5,10,25},
+        xticklabels = {$16k$, $19k$, $23k$, $27k$, $30k$}, 
+        yticklabels = {$\text{\ \,}0.1$, $1$,$5$,$10$,$25$}, 
+        smooth,
+        cycle list name = black white,
+        legend style = { 
+          at={(0.49,0.6)}, 
+            anchor=north west,
+            draw=black, 
+            fill= white,
+            align=left
+        }
+    ]
+
+    %start data for plotting
+      \addplot table {
+        16000	3.602
+        17600	4.223
+        19200	4.839
+        20800	8.201
+        22400	8.273
+        24000	13.426
+        25600	10.089
+        27200	15.889
+        28800	16.431
+        30400	21.437
+      }; \addlegendentry{prpu-SML, genrmf};
+
+      \addplot table {
+        16000	4.51
+        17600	4.804
+        19200	5.899
+        20800	8.052
+        22400	8.448
+        24000	11.555
+        25600	8.878
+        27200	13.999
+        28800	12.99
+        30400	13.791
+      }; \addlegendentry{prpu-C++, genrmf};
+
+      \addplot table {
+        16000	0.009
+        17600	0.012
+        19200	0.015
+        20800	0.021
+        22400	0.021
+        24000	0.019
+        25600	0.023
+        27200	0.028
+        28800	0.026
+        30400	0.026
+      }; \addlegendentry{prpu-C, genrmf};
+    \end{axis}
+  \end{tikzpicture}
   \caption{Benchmark of different implementations. The x-axis shows the number of nodes, the y axis the execution time in seconds.}\label{fig:benchmark}

### 2017-02-28 paper/root.tex
Iteration #2 over paper

@@
 \usepackage{listings}
+\usepackage{lstautogobble}
 \usepackage{graphicx}
@@
 \lstset{numberbychapter=false}
+\lstset{autogobble=true}
 
@@
   Institut f\"ur Informatik,
-  Technische Universit\"at M\"unchen \\
+  Technische Universit\"at M\"unchen, Germany \\
   \email{lammich@in.tum.de} %
@@
 \begin{abstract}
-We present a formalization of two classical algorithms for computing the maximum flow in a network:
-The Edmonds-Karp algorithm and Goldberg's Generic Push-Relabel algorithm. We prove the correctness and complexity of the algorithms.
+We present a formalization of classical algorithms for computing the maximum flow in a network:
+The Edmonds-Karp algorithm and the push-relabel algorithm. 
+We prove correctness and time complexity of these algorithms.
 Our formal proof closely follows a standard textbook proof, and is accessible even without being
@@
 
-Using stepwise refinement, we obtain the Fifo-Push-Relabel algorithm with gap heuristics. 
-Further refinement yields verified efficient implementations of both, Edmonds-Karp and Fifo-Push-Relabel, 
-which compare well to unverified reference implementations in Java.
-
-\keywords{Maxflow Algorithm \and Edmonds-Karp Algorithm \and Push-Relabel Algorithm \and Formal Verification \and Isabelle/HOL \and Stepwise Refinement}
+Using stepwise refinement techniques, we instantiate the generic Ford-Fulkerson method to Edmonds-Karp algorithm, 
+and the generic push-relabel algorithm of Goldberg and Tarjan to both, the relabel-to-front and the FIFO push-relabel algorithm.
+Further refinement then yields verified efficient implementations of the algorithms, which compare well to unverified reference implementations.
+
+\keywords{Maximum Flow Problem \and Edmonds-Karp Algorithm \and Push-Relabel Algorithm \and Formal Verification \and Isabelle/HOL \and Stepwise Refinement}
 % \PACS{PACS code1 \and PACS code2 \and more}
@@
 Dinitz~\cite{Di06} generalized the idea of augmenting paths to blocking flows, obtaining an $O(V^2E)$ algorithm. 
-Karzanov~\cite{Ka74} was the first to propose an $O(V^3)$ algorithm based on the idea of \emph{preflows}. Based on Karzanov's ideas, Goldberg et al.\ proposed
-the generic push-relabel algorithm, which consists of repeated application of two local operations, push and relabel, in any order.
+Karzanov~\cite{Ka74} was the first to propose an $O(V^3)$ algorithm based on the idea of \emph{preflows}. Based on Karzanov's ideas, Goldberg and Tarjan proposed
+the generic push-relabel algorithm~\cite{GoTa88}.
 Implementations of the push-relabel algorithm are among the most efficient maximum flow algorithms. While the generic algorithm has a time complexity of $O(V^2E)$,
@@
 the implementation details: After proving the min-cut-max-flow theorem, it is straightforward to verify a generic version of the Ford-Fulkerson algorithm,
-which we then refine to Edmonds-Karp algorithm and finally to an efficient implementation. Similar, the generic push-relabel algorithm is refined to 
-the fifo-push-relabel algorithm, which, in turn, is refined to an efficient implementation.
+which we then refine to Edmonds-Karp algorithm and finally to an efficient implementation. 
+Similarly, the generic push-relabel algorithm is refined to both, the relabel-to-front algorithm~\cite{CLRS09} and the FIFO push-relabel algorithm~\cite{GoTa88},
+which, in turn, are refined to efficient implementations. The efficiency of our verified implementations compare well to unverified reference implementations of 
+the algorithms.
+
 The abstract parts of our verification closely follow the textbook presentation of Cormen et al.~\cite{CLRS09}. 
@@
 This paper is an extended version of our conference paper~\cite{LaSe16}, which only covered the Edmonds-Karp algorithm.
-
 While there exists another formalization of the Ford-Fulkerson method in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial time maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation.
@@
 
-The rest of this paper is structured as follows: In Section~\ref{sec:background} we give a short informal introduction to the Ford-Fulkerson method.
-In Section~\ref{sec:abs-formalization}, we report on our formalization of the abstract method. 
-Section~\ref{sec:refinement} gives a brief overview of the Isabelle Refinement Framework~\cite{LaTu12,La12}, which supports stepwise refinement based algorithm development in Isabelle/HOL.
-In Section~\ref{sec:edka}, we report on our instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm and the proof of its complexity.
-Section~\ref{sec:executable} reports on the further refinement steps required to yield an efficient implementation. Section~\ref{sec:benchmark} reports on 
-benchmarking our implementation against a reference implementation of the Edmonds-Karp algorithm from Sedgewick et al.~\cite{SeWa11}. 
-Finally, Section~\ref{sec:concl} gives a conclusion and discusses related and future work. The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/edmonds_karp/}.
+In the rest of this paper, we give a short informal introduction to flow networks and the Ford-Fulkerson method (\S\ref{sec:background}).
+We then present our formalization of these material (\S\ref{sec:abs-formalization}).
+After introducing the Isabelle Refinement Framework~\cite{LaTu12,La12} (\S\ref{sec:refinement}), we report on the instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm, and on the complexity proof (\S\ref{sec:edka}).
+Next, we describe the correctness and complexity proof of the generic push-relabel algorithm (\S\ref{sec:gen-prpu}), and its instantiations 
+to the relabel-to-front and FIFO push-relabel algorithms (\S\ref{sec:prpu-inst}).
+We then report on the refinement of all three algorithms to efficient implementations (\S\ref{sec:executable}), and their comparison to unverified reference 
+implementations (\S\ref{sec:benchmark}).
+
+The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/edmonds_karp/}. 
+The part on Edmonds-Karp algorithm is also available in the Archive of Formal Proofs~\cite{LaSe16_afp}.
 
@@
 
-A \emph{preflow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Non-generation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is not smaller than the sum of flows over all outgoing edges.
-We define the \emph{excess} flow $x_f(u)$ of a node $u$ to be the difference of the incoming and outgoing flow of $u$.
-Intuitively, a \emph{preflow} must not send more flow over an edge than its capacity allows, and only the source node may generate flow. 
-The excess flow is the amount of flow that is stuck at a node, \ie which flows in but does not flow out.
-A \emph{flow} is a \emph{preflow} that satisfies the \emph{Conservation constraint}: For all vertices except $s$ and $t$, the excess flow is zero.
-Intuitively, only the source may generate flow, and only the sink may consume flow, and for all other nodes, the incoming flow must be equal to the outgoing flow.
+A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of incoming flows equals the
+sum of outgoing flows.
+% 
+% \emph{Non-generation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is not smaller than the sum of flows over all outgoing edges.
+% We define the \emph{excess} flow $x_f(u)$ of a node $u$ to be the difference of the incoming and outgoing flow of $u$.
+% Intuitively, a \emph{preflow} must not send more flow over an edge than its capacity allows, and only the source node may generate flow. 
+% The excess flow is the amount of flow that is stuck at a node, \ie which flows in but does not flow out.
+% A \emph{flow} is a \emph{preflow} that satisfies the \emph{Conservation constraint}: For all vertices except $s$ and $t$, the excess flow is zero.
+% Intuitively, only the source may generate flow, and only the sink may consume flow, and for all other nodes, the incoming flow must be equal to the outgoing flow.
 The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
@@
 
-The correctness of the Ford-Fulkerson algorithm follows from the Ford-Fulkerson theorem, which is usually stated as the following three statements being equivalent:
+The correctness of the Ford-Fulkerson algorithm follows from the Ford-Fulkerson theorem, which is also known as min-cut-max-flow theorem.
+It states that the following three propositions are equivalent:
 \begin{enumerate}
@@
 
-In a typical program development, one first comes up with an initial version \isai{m_0} of
+In a typical program development, one first designs an initial version \isai{m_0} of
 the algorithm and its specification \isai{P,Q}, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions \isai{m$_i$} of the algorithm,
-proving \isai{m$_i$ <= \<Down>R$_i$ m$_{i-1}$}. Using transitivity and composability of data refinement, one 
+proving \isai{m$_i$ <= \<Down>R$_i$ m$_{i-1}$}. Using transitivity and composability of refinement, one 
 gets \isai{P ==> m$_i$ <= \<Down>R$_i$...R$_1$ spec Q}, showing the correctness of the refined algorithm.
@@
 
-\section{The Generic Push-Relabel Algorithm}  
-  We present our formalization of the generic push relabel algorithm of Goldberg et al.~\cite{GoTa88}. 
+\section{The Generic Push-Relabel Algorithm}\label{sec:gen-prpu}
+  We present our formalization of the generic push relabel algorithm of Goldberg and Tarjan~\cite{GoTa88}. 
   Again, our formalization closely follows the textbook presentation of Cormen et al.~\cite{CLRS09}.
+
+  Push-relabel algorithms are based on a generalization of flows called \emph{preflows}: 
+  The conservation constraint is relaxed to allow more incoming flow than outgoing flow on a vertex.
+  For a vertex $u\in V$, the \emph{excess} $x_f(u)$ is the difference between incoming and outgoing flow. 
+  For a preflow, the excess of any node but the source must not be negative.
+  A non-sink vertex with positive excess is called \emph{active}. Thus, a preflow without any active vertices is actually a flow.
+
   The push-relabel algorithm maintains a preflow $f$ and a height-labeling $l: V \to \mathbb {N}$ of the nodes, 
@@
   Moreover, the height of the source is fixed to $\left\vert{V}\right\vert$, and the height of the sink is fixed to $0$.
-  By generalizing the labeling invariant to paths, we get that for each path $p$ from $u$ to $v$, we have $l(u) \le l(v) + |p|$.
+  By generalizing the labeling invariant to paths, we get that for each path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$.
   This implies that there cannot be a path from $s$ to $t$: The length of any (simple) path is less than $|V|$, however, a valid labeling 
-  implies a length greater than or equal to $|V|$.
-
-  A node that has positive excess is called \emph{active}. Upon termination of the algorithm, there will be no active nodes. 
-  Thus, the preflow is actually a flow, and as we have no $s$--$t$ paths, the min-cut-max-flow theorem implies that it is maximal.
-
-  Initially, the height of all nodes but the source is $0$, and the outgoing edges of the source get the maximum possible flow, 
-  and all other edges get no flow at all. Then, the algorithm repeats the following operations:
+  implies a length greater than or equal to $|V|$. Upon termination of the algorithm, there will be no active nodes. 
+  Thus, the preflow is actually a flow, and as we have no $s$--$t$ paths, the Ford-Fulkerson theorem implies that it is maximal.
+
+  The push relabel algorithm initializes the heights of all nodes but the source to zero. Moreover, the outgoing edges of the source are 
+  set to carry maximum flow, all other edges get zero flow. Then, the algorithm repeats the following operation as long as they apply:
   \begin{description}
-    \item[push] Send as much flow as possible over an edge $(u,v)$ with $l(u) = l(v)+1$. The maximum possible flow is constrained
-      by the residual capacity of the edge, and the excess flow available at $u$.
-    \item[relabel] Increase the height of an active (non-sink) node from which no push operation is possible. 
-      The height is increased to the minimum value such that a push becomes possible.
+    \item[$\textrm{push}(u,v)$] for an edge $(u,v)$ of the residual graph with $l(u) = l(v)+1$ and $u$ active: Move as much flow as possible from $u$ to $v$.
+    The maximum possible flow is constrained by the residual capacity of the edge, and the excess flow available at $u$.
+    \item[$\textrm{relabel}(u)$] for an active node $u$ from which no push operation is possible: Increase the height of $u$ to the minimum value such that a push
+      becomes possible.
   \end{description}
-  The algorithm is generic in the sense that it leaves open the order in which the push and relabel operations on the different nodes and edges are performed.
-  The algorithm terminates if no more operations apply. Then, there are no active nodes left, and the algorithm has constructed a maximum flow.
+  Obviously, if none of the operations apply, all nodes are inactive and the resulting preflow is a maximum flow.
+  The algorithm is generic in the sense that it leaves open the order in which the operations are applied to the different nodes and edges.
 
@@
 A push operation can be implemented in time $O(1)$, (only the flow on a single edge has to be modified) and a relabel operation can be implemented in time $O(V)$.
-(to find the new height, we have to consider the heights of all adjacent nodes in the residual graph) To this end, we define 
-\isai{cost PUSH == 1 | cost RELABEL == card(V)}. We prove the following theorem:
+(to find the new height, we have to consider the heights of all successor nodes in the residual graph) We define 
+\isai{cost PUSH == 1 | cost RELABEL == card(V)} and prove the following theorem:
 \begin{lstlisting}
@@
 \end{lstlisting}  
-where \isai$(f_0,l_0)$ is the initial flow and height-labeling. Intuitively, the cost of any (partial) run \isai$p$ is bounded by $O(V^2E)$, and if the 
+where \isai$(f_0,l_0)$ is the initial flow and height-labeling and \isai{pr_lts^*} denotes the reflexive transitive closure of \isai{pr_lts}. 
+Intuitively, the cost of any (partial) run \isai$p$ from the initial state to $(f,l)$ is bounded by $O(V^2E)$, and if the 
 run is complete, \ie no more operations apply (\isai$(f,l)\<notin>Domain pr_lts$), the resulting flow is maximal.
 
-For the proof of this theorem, we follow the presentation of Cormen~\cite{CLRS09}. 
-However, at several places, we trade precision of the constant for simplicity of the proof. 
-Thus, our constant of $38$ is somewhat higher than what would result from Cormen's proof.
-
-For the correctness argument, we show that the operations of the algorithm maintain a valid labeling, and that existence of an active node implies applicability of 
-a push or relabel operation.
-The complexity argument is more tricky: First, we show that the labels of each node remain smaller than $2|V|$. This bounds the number of relabel operations to $O(V^2)$.
-For the push operations, we first define a push to be \emph{saturating}, if the excess at the source node of the push is not less than the capacity of the edge.
+For the proof of this theorem, we follow the presentation of Cormen et al.~\cite{CLRS09}.
+The correctness proof shows that the algorithm maintains a valid labeling, and its formalization poses no further difficulties. 
+It consists of less than 500 lines, and took us only a few days to complete.
+
+The complexity argument is more tricky. Instead of defining a single measure function,
+Cormen counts the different types of operations separately, using quite involved arguments.
+For our formalization, we have adapted some Cormen's proofs, trading precision of the constant factors for simplicity of the formalization. 
+Thus, our constant of $38$ is somewhat higher than the $17$ that would result from Cormen's proof.
+
+First, we show that the labels of each node remain smaller than $2|V|$. This bounds the number of relabel operations to $O(V^2)$.
+For the push operations, we define a push to be \emph{saturating}, if the excess at the source node of the push is not less than the capacity of the edge.
 In this case, the edge is completely saturated by the push and disappears from the residual graph. However, the source node of the push remains active.
@@
 However, this push can only occur after the labeling of $v$ has increased. This can only happen $O(V)$ times, which yields a bound of $O(VE)$ on the number 
-of saturating push operations. Finally, the number of non-saturating pushes is bounded by defining a potential function that is decreased by non-saturating pushes,
-and increased by at most $O(V)$ by the other operations. The initial potential is $O(V)$, and with the already established bounds on the other operations we get a bound of
+of saturating push operations. 
+Finally, the number of non-saturating pushes is bounded by defining a potential function that is decreased by a non-saturating push.
+A saturating push or relabel operation increases the potential by $O(V)$.
+The initial potential is $O(V)$, and with the already established bounds on the other operations we get a bound of
 $O(V^3 + V^2E) = O(V^2E)$ on the number of non-saturating push operations. Combining the bounds for the different operation types with the costs for the operations 
@@
 
-The simple structure of the generic algorithm allowed us to conveniently define it as labeled transition system, such that the above intuitive arguments could be formalized
-as counting different operation types on execution paths of the algorithm. Moreover, the powerful Isabelle/HOL standard library, in particular the lemmas for summations
-over finite sets, were very helpful. Anyway, the whole complexity proof is roughly 1000 lines of Isabelle text, and took a few men weeks to complete.
-
-\section{Implementing the Push-Relabel Algorithm}
-The generic algorithm does not specify the order in which the operations should be applied. 
-Moreover, choosing the right ordering may further decrease the worst-case time complexity, and, more important, the practical efficiency of the algorithm.
-
-We refine the generic push-relabel algorithm to different implementations. However, these have a more complicated algorithmic structure that
-prevents a convenient specification as labeled transition system, and thus a complexity proof in the style of the generic algorithm.
-Thus, we specify the algorithms using the Monadic Refinement Framework, and only prove correctness and termination 
-(which we get for free, as the complexity bound for the generic algorithm is also valid for all implementations).
-We leave it to future work to integrate complexity reasoning tools into the Refinement Framework and prove tighter complexity bounds for the specific implementations.
+The simple structure of the generic algorithm allowed us to conveniently define it as labeled transition system, such that the above informal proof could be formalized
+by inductive arguments over the runs of the algorithm.
+Moreover, the powerful Isabelle/HOL standard library, in particular the lemmas for summations
+over finite sets, were very helpful to keep the formalization effort manageable. 
+Anyway, the whole complexity proof is roughly 1000 lines of Isabelle text, and took a few weeks to complete.
+This includes the time that we spent on an initial termination proof that used simpler arguments, but did not derive an exact complexity bound.
+
+\section{Instantiating the Generic Push-Relabel Algorithm}\label{sec:prpu-inst}
+The generic algorithm does not specify the order in which the operations are applied. 
+Choosing the right order decreases the worst-case time complexity, and, more important, the practical efficiency of the algorithm.
+
+We consider two common instantiations of the generic algorithm: The relabel-to-front algorithm~\cite{CLRS09}, and tzhe FIFO push-relabel algorithm~\cite{GoTa88}.
+Both have a theoretical time complexity of $O(V^3)$, which is better than the $O(V^2E)$ bound of the generic algorithm.
+However, the more complicated structure of these algorithms prevent a convenient specification as labeled transition system, and thus 
+a complexity proof in the style of the generic algorithm.
+Instead, we use the Monadic Refinement Framework to specify the algorithms and only prove correctness and 
+termination.\footnote{Which we get for free from the bound of the generic algorithm.} We leave it to future work to integrate complexity reasoning tools into 
+the Refinement Framework and prove the tighter complexity bounds for the specific implementations.
+
+A general idea of most efficient implementations is to \emph{discharge} a node by repeatedly applying push operations until the node gets inactive or needs to be relabeled.
 
 \subsection{Relabel-to-Front Algorithm}
-A general idea of the implementation is to \emph{discharge} a node by repeatedly applying push and relabel operations to the same node, until it gets inactive.
-Cormen et al.\ present the relabel to front algorithm, which scans through a list of all nodes, discharging every node as it is visited.
-If the node gets relabeled during discharging, it is moved to the front of the list, and the scan of the list is restarted at the beginning of the list.
+Cormen et al.\ present the relabel to front algorithm.
+It discharges a node until it gets inactive, applying relabel operations as necessary. 
+Moreover, if a node was discharged without being relabeled, the already saturated edges during this discharge are stored, and not considered 
+during the next discharge. This reduces the work required to find the next edge to push.
+The algorithm scans through a list of all nodes, discharging every node as it is visited.
+If the node gets relabeled during discharging, it is moved to the front of the list, and the scan of the list begins anew.
 The algorithm terminates if it reaches the end of the list.
-During discharging a node $u$, the algorithm iterates over the list of all adjacent nodes $v$ in the network, and checks whether there is an edge $(u,v)$ in 
-the residual graph. When the discharge operation stops without relabeling $u$, the current position in the list of adjacent nodes is stored, and 
-the next discharge operations starts at exactly this position.
 
@@
 
-Following the outline by Cormen et al., and using the Isabelle Refinement Framework to conveniently describe the abstract algorithm,
+Following the proof by Cormen et al., and using the Isabelle Refinement Framework to conveniently describe the abstract algorithm,
 the correctness proof offered no surprises. It is about 700 lines of Isabelle text plus 150 lines of general lemmas on topological orderings, 
@@
 An alternative implementation is the FIFO push-relabel algorithm, which maintains a first-in-first-out queue of the active nodes. 
-In each iteration, it pulls a node from the FIFO, and repeatedly pushes flow out of it.
-If no push operations are possible any more, but the node is still active, it is relabeled and appended to the FIFO.
-
-The FIFO push relabel algorithm has several advantages: It's correctness is straightforward to prove, 
-the only required additional invariant is that the queue contains exactly the active nodes. 
+In each iteration, it dequeues a node from the FIFO, and tries to discharge it by only using push operations.
+If the node is still active, it is relabeled and enqueued.
+
+The FIFO push-relabel algorithm has several advantages: It's correctness is straightforward to prove, 
+as the only required additional invariant is that the queue contains exactly the active nodes. 
 Moreover, it performs well in practice, and we have found several implementations of it~\cite{ChGo97,ICPC-notebook}.
@@
 which uses an additional gap heuristics: If there is a value $0<k<|V|$ such that there is no node that has height $k$, 
-all nodes with $k<l(v)<|V|$ may be relabeled to $|V|+1$, preserving a valid labeling. In practice, this may drastically accelerate convergence of the algorithm,
-as it prevents unnecessary pushes of flow towards the sink when the path to the sink is already blocked.
-The correctness proof of this heuristics is straightforward from the definitions, and also the overall correctness proof 
-is straightforward: It's size is about 400 lines of code, and it took roughly a day to complete.
+all nodes with $k<l(v)<|V|$ may be relabeled to $|V|+1$, preserving a valid labeling. 
+This heuristics prevents unnecessary pushes of flow towards the sink when the path to the sink is already blocked.
+The correctness of the heuristics follows easily from the definitions of a valid height-labeling.
+In total, the correctness proof of the FIFO push-relabel algorithm is about 400 lines of code and, building on the experience from relabel-to-front, it 
+took only a day to complete.
 
@@
   However, we left open several algorithmic aspects. For example, we did not specify how to find a shortest augmenting path.
-  Moreover, some of the used operations require additional auxiliary data structures to be maintained. For example, 
-  we maintain the current excess of each node explicitly, instead of summing up the flows of incoming and outgoing edges each time 
+  Moreover, some of the operations require additional auxiliary data structures to be maintained. For example, 
+  we maintain the current excess of each node explicitly, instead of summing up the incoming and outgoing flows each time 
   we need to check whether a node is active.
   Finally, we have to specify the concrete data structures that we want to use for implementation. 
-  For example, the residual graph is modeled as a function from edges to capacities on the abstract level. In the implementation, we want to use a two-dimensional array.
+  For example, the residual graph is modeled as a function from edges to capacities on the abstract level. In the implementation, we use a two-dimensional array.
 
@@
 
-  Implementing shortest path finding by BFS in Edmonds-Karp algorithm yields a specification that algorithmically describes all major operations, but still leaves open the data structures used for implementation.
+%   Implementing shortest path finding by BFS in the Edmonds-Karp algorithm yields a specification that algorithmically describes all major operations, 
+%   but still leaves open the data structures used for implementation.
 
@@
   \subsection{Manipulating Residual Graphs Directly}\label{sec:impl_res_graph}
-  Next, we observe that the Edmonds-Karp algorithm is phrased in terms of a flow, which is updated until it is maximal. 
-  In each iteration, the augmenting path is searched on the residual graph induced by the current flow.
+  The Edmonds-Karp algorithm is phrased in terms of a flow, which is updated until it is maximal. 
+  In each iteration, an augmenting path is searched on the residual graph induced by the current flow.
   Obviously, computing the complete residual graph in each iteration is a bad idea. 
@@
   \subsection{Height Frequencies}
-  For an efficient implementation of the gap heuristics, we introduce an explicit map from height-values to then number of nodes with this height.
-  Using this map, it is easy to detect when we relabel the last node of a specific height and the gap heuristics applies.
+  For an efficient implementation of the gap heuristics, we introduce an explicit map from height-values to the number of nodes with this height.
+  Using this map, it is easy to detect when we relabel the last node of a specific height such that the gap heuristics applies.
 
   \subsection{Implementing Augmentation}    
-  In our abstract formalization, which matches the presentation in Section~\ref{sec:background}, we have formulated augmentation by first defining 
+  In our abstract formalization of the Ford-Fulkerson method, which matches the presentation in Section~\ref{sec:background}, we have formulated augmentation by first defining 
   the residual capacity $c_p$ of the augmenting path. Using $c_p$, we have defined the residual flow $f_p$, which was finally added to the current flow.
@@
   \subsection{Algorithms for Relabel and Discharge}    
-  For the discharge operation, it is straightforward to find the new label by iterating over the adjacent edges in the residual graph.
+  For the relabel operation, it is straightforward to find the new label by iterating over the outgoing edges in the residual graph.
   The discharge operation is a bit more tricky: Abstractly, we nondeterministically select a new edge to push, until we find no more edges.
-  For the push-to-front algorithm, we select the edges from a set of unprocessed edges which is maintained globally.
-  In the concrete implementation, however, we fold over a list of edges. The structure of the resulting concrete algorithms is somewhat different from the structure
-  of the abstract algorithm --- too different for the heuristics of our verification condition generator. Thus, we had to manually prove some rules specifically 
-  tailored to the required structural changes.
+  In the concrete implementation, however, we fold over the list of successor nodes.
+  This structural difference between abstract and concrete algorithm cannot be handled by the heuristics of the Refinement Framework's verification condition generator.
+  Thus, we had to manually prove some refinement rules that are specifically tailored to our case.
 
@@
   In order to find an augmenting path, the BFS algorithm has to compute the successors of a node in the residual graph. 
-  Similar, the relabel operation has to iterate over all successors to find the new minimal labeling, and the discharge operations iterate over all outgoing edges of a node.
-  Although these iterations can be implemented on the edge labeling function by iterating over all nodes, this implementation tends to be inefficient for sparse graphs,
-  where we would have to iterate over many possible successor nodes just to find that there is no edge.
+  Similar, the relabel operation has to iterate over all successors to find the new minimal labeling, and the discharge operation iterates over all outgoing edges of a node.
+  Although these iterations can be implemented by iterating over all nodes, this would be inefficient for sparse graphs,
+  where we would iterate over many possible successor nodes just to find that there is no edge.
 
@@
   \subsection{Using Efficient Data Structures}\label{sec:impl_data_structures}
-  In a final step, we have to choose efficient data structures for the algorithm.
-
-  %   The objects that the algorithm works with are
-%   capacities, nodes, edges, the graph of the input network, residual graphs, the adjacency map, paths, and the resulting maximum flow.
-
+  In a final step, we choose efficient data structures. 
   We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers.
@@
 
-  For the push-relabel algorithms, the map from nodes to the excess, and the height frequency map are implemented by arrays. 
+  For the push-relabel algorithms, the maps from a node to its height and excess, as well as the height frequency map are implemented by arrays. 
   The queue of the FIFO push-relabel algorithm is implemented by two stacks, yielding amortized constant time enqueue and dequeue operations.
-  Similar, the list of the relabel to front algorithm is implemented as two lists, \emph{left} and \emph{right}, that contain the nodes left and right of the current position,
-  the left list being stored in reverse order. Thus, advancing the position is implemented in constant time by popping the first element of the right list, and pushing 
+  Similarly, the list of the relabel-to-front algorithm is implemented by two lists, \emph{left} and \emph{right}, that contain the nodes left and right of the current position,
+  the left list being stored in reverse order. Advancing the position is implemented in constant time by popping the first element of the right list, and pushing 
   it to the left list. Resetting the position to the front of the list is implemented by pushing all elements from the left list to the right list. 
@@
   This calls for a deforestation optimization to get rid of the intermediate list, and iterate only two times over the predecessor map directly.
-  Also the FIFO queue could be implemented by a ring buffer based on an array, yielding true constant-time operations with less overhead.
+  Also the FIFO queue could be implemented by a ring buffer based on a fixed size array, yielding true constant-time operations with less overhead.
   Fortunately, our profiling data shows that these operations do not significantly contribute to the runtime of our implementations, such that 
@@
 
-  Note that we performed the last refinement step using our Sepref tool~\cite{La15,La16}, which provides tool support for refinement from the
-  purely functional programs of the Isabelle Refinement Framework into imperative programs expressed in Imperative/HOL~\cite{BKHEM08}.
-  The formalization of this refinement step consists of setting up the mappings between the abstract and concrete data structures,
+  Note that we performed the last refinement step using our Sepref tool~\cite{La15,La16}, which automates the refinement from 
+  purely functional programs into imperative programs expressed in Imperative/HOL~\cite{BKHEM08}.
+  This refinement step consists of setting up the mappings between the abstract and concrete data structures,
   and then using Sepref to synthesize the Imperative/HOL programs and their refinement proofs.
@@
 
-  \subsection{Network Checker}
+  \subsection{Network Checker and Main Correctness Theorem}
   Additionally, we implemented an algorithm that takes as input a list of edges, a source node, and a target node.
@@
         legend style = { 
-          at={(0.515,0.24)}, 
+          at={(0.59,0.4)}, 
             anchor=north west,
@@
       }; \addlegendentry{prpu-SML, sparse};
+
+      \addplot table {
+        1000 .053   
+        1500 .089   
+        2000 .091   
+        2500 .120 
+        3000 .256 
+        3500 .210 
+        4000 .265 
+        4500 .499 
+        5000 .457 
+        5500 .522     
+      }; \addlegendentry{prpu-Java2, sparse};
+
+      \addplot table {
+        1000 .014 
+        1500 .029 
+        2000 .027 
+        2500 .037 
+        3000 .043 
+        3500 .051 
+        4000 .053 
+        4500 .080 
+        5000 .084 
+        5500 .085 
+      }; \addlegendentry{prpu-Java1, sparse};
+
+      \addplot table {
+        1000 .006 
+        1500 .013
+        2000 .014
+        2500 .019
+        3000 .041
+        3500 .035
+        4000 .042
+        4500 .076
+        5000 .079
+        5500 .082    
+      }; \addlegendentry{prpu-C++, sparse};
+
+      \addplot table {
+        1000 0  
+        1500 .001  
+        2000 .003  
+        2500 .002  
+        3000 .003  
+        3500 .004  
+        4000 .005  
+        4500 .007  
+        5000 .010 
+        5500 .009      
+      }; \addlegendentry{prpu-C, sparse};
     %end data for plotting
@@
         xtick = {0,1000,1100,...,1400},
-        ytick = {1,4,...,16},
+        ytick = {2,4,...,10},
         xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, 
-        yticklabels = {$1$,$4$,$8$,$12$,$16$}, 
+        yticklabels = {$2$, $4$,$6$,$8$,$10$}, 
         smooth,
@@
         legend style = { 
-          at={(0.53,0.6)}, 
+          at={(0.59,0.4)}, 
             anchor=north west,
@@
       }; \addlegendentry{prpu-SML, dense};
+
+      \addplot table {
+        1000	0.035
+        1050	0.037
+        1100	0.043
+        1150	0.044
+        1200	0.051
+        1250	0.065
+        1300	0.058
+        1350	0.068
+        1400	0.059
+        1450	0.069  
+      }; \addlegendentry{prpu-Java2, dense};
+
+      \addplot table {
+        1000	0.046
+        1050	0.05
+        1100	0.048
+        1150	0.059
+        1200	0.056
+        1250	0.056
+        1300	0.058
+        1350	0.062
+        1400	0.062
+        1450	0.066
+      }; \addlegendentry{prpu-Java1, dense};
+
+      \addplot table {
+        1000	0.034
+        1050	0.035
+        1100	0.039
+        1150	0.042
+        1200	0.052
+        1250	0.052
+        1300	0.054
+        1350	0.057
+        1400	0.064
+        1450	0.066  
+      }; \addlegendentry{prpu-C++, dense};
+
+      \addplot table {
+        1000	0.006
+        1050	0.004
+        1100	0.006
+        1150	0.005
+        1200	0.007
+        1250	0.007
+        1300	0.007
+        1350	0.007
+        1400	0.007
+        1450	0.008
+      }; \addlegendentry{prpu-C, dense};
     \end{axis}
@@
 %         ylabel = Execution time in seconds, 
-        xtick = {3000, 6000, 10500, 13500, 16500},
-        ytick = {0.1,1,10,50,100},
-        xticklabels = {$3k$, $6k$, $10k$, $13k$, $16k$},
-        yticklabels = {$\text{\ \,}0.1$,$1$,$10$,$50$,$100$}, 
+        xtick = {0,1000,...,5000},
+        ytick = {0.01,0.1,1,10},
+        xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$},
+        yticklabels = {$0.01$,$0.1$,$1$,$10$}, 
         smooth,
@@
         legend style = { 
-          at={(0.53,0.24)}, 
+          at={(0.59,0.4)}, 
             anchor=north west,
@@
     %start data for plotting
-
-      \addplot table {
-        3000	1.185
-        4500	3.524
-        6000	6.457
-        7500	11.509
-        9000	17.529
-        10500	26.185
-        12000	36.828
-        13500	48.614
-        15000	86.449
-        16500	89.661
+      \addplot table {
+        1000	0.356
+        1500	0.805
+        2000	1.539
+        2500	2.43
+        3000	3.28
+        3500	4.639
+        4000	5.61
+        4500	7.198
+        5000	8.904
+        5500	11.302
+      }; \addlegendentry{edka-SML, wash};
+
+      \addplot table {
+        1000	0.105
+        1500	0.228
+        2000	0.385
+        2500	0.466
+        3000	0.7
+        3500	0.992
+        4000	1.194
+        4500	1.503
+        5000	1.735
+        5500	2.243
+      }; \addlegendentry{edka-Java, wash};
+
+      \addplot table {
+        1000	1.185
+        1500	3.524
+        2000	6.457
+        2500	11.509
+        3000	17.529
+        3500	26.185
+        4000	36.828
+        4500	48.614
+        5000	86.449
+        5500	89.661
       }; \addlegendentry{prpu-SML, wash};
@@
       \addplot table {
-        3000	0.906
-        4500	2.708
-        6000	5.083
-        7500	8.792
-        9000	12.951
-        10500	19.075
-        12000	27.466
-        13500	35.964
-        15000	61.839
-        16500	67.039
+        1000	11.515
+        1500	37.224
+        2000	60.226
+        2500	96.657
+        3000	134.77
+        3500	199.812
+        4000	328.824
+        4500	383.112
+        5000	745.695
+        5500	661.721
+      }; \addlegendentry{prpu-Java2, wash};
+
+      \addplot table {
+        1000	0.654
+        1500	1.626
+        2000	3.096
+        2500	5.313
+        3000	8.572
+        3500	12.552
+        4000	18.103
+        4500	18.646
+        5000	40.456
+        5500	39.012
+      }; \addlegendentry{prpu-Java1, wash};
+
+      \addplot table {
+        1000	0.906
+        1500	2.708
+        2000	5.083
+        2500	8.792
+        3000	12.951
+        3500	19.075
+        4000	27.466
+        4500	35.964
+        5000	61.839
+        5500	67.039
       }; \addlegendentry{prpu-C++, wash};
@@
       \addplot table {
-        3000	0.006
-        4500	0.015
-        6000	0.029
-        7500	0.042
-        9000	0.063
-        10500	0.084
-        12000	0.115
-        13500	0.144
-        15000	0.183
-        16500	0.223
+        1000	0.006
+        1500	0.015
+        2000	0.029
+        2500	0.042
+        3000	0.063
+        3500	0.084
+        4000	0.115
+        4500	0.144
+        5000	0.183
+        5500	0.223
       }; \addlegendentry{prpu-C, wash};
@@
         xtick = {0,1000,1100,...,1400},
-        ytick = {0.01, 0.1, 0.2, 0.5, 1},
+        ytick = {2,4,...,10},
         xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, 
-        yticklabels = {$0.01$, $0.1$,$0.2$,$0.5$,$1$}, 
+        yticklabels = {$2$, $4$,$6$,$8$,$10$}, 
         smooth,
@@
         legend style = { 
-          at={(0.52,0.24)}, 
+          at={(0.59,0.4)}, 
             anchor=north west,
@@
     %start data for plotting
+      \addplot table {
+        1000	28.6
+        1050	42.725
+        1100	45.527
+        1150	56.67
+        1200	57.768
+        1250	78.733
+        1300	89.811
+        1350	103.254
+        1400	80.91
+        1450	87.823
+      }; \addlegendentry{edka-SML, acnet};
+
+      \addplot table {
+        1000	21.355
+        1050	23.84
+        1100	27.647
+        1150	36.513
+        1200	35.793
+        1250	40.851
+        1300	45.18
+        1350	49.515
+        1400	60.096
+        1450	65.875
+      }; \addlegendentry{edka-Java, acnet};
+
       \addplot table {
@@
       }; \addlegendentry{prpu-SML, acnet};
+
+      \addplot table {
+        1000	0.069
+        1050	0.08
+        1100	0.066
+        1150	0.094
+        1200	0.091
+        1250	0.094
+        1300	0.101
+        1350	0.209
+        1400	0.16
+        1450	0.185
+      }; \addlegendentry{prpu-Java2, acnet};
+
+      \addplot table {
+        1000	0.081
+        1050	0.112
+        1100	0.091
+        1150	0.151
+        1200	0.141
+        1250	0.135
+        1300	0.16
+        1350	0.29
+        1400	0.21
+        1450	0.269
+      }; \addlegendentry{prpu-Java1, acnet};
 
@@
   \end{tikzpicture}
-  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
-    \begin{axis}[
-      %title={Benchmark of the Edmonds-Karp algorithm},
-        axis x line*= bottom,
-        axis y line*= left,
-        xmode = log,
-        ymode = log,
-%         xlabel = Number of vertices in the network,
-%         ylabel = Execution time in seconds, 
-        xtick = {16000,19000,23000,27000,30000},
-        ytick = {0.1,1,5,10,25},
-        xticklabels = {$16k$, $19k$, $23k$, $27k$, $30k$}, 
-        yticklabels = {$\text{\ \,}0.1$, $1$,$5$,$10$,$25$}, 
-        smooth,
-        cycle list name = black white,
-        legend style = { 
-          at={(0.49,0.6)}, 
-            anchor=north west,
-            draw=black, 
-            fill= white,
-            align=left
-        }
-    ]
-
-    %start data for plotting
-      \addplot table {
-        16000	3.602
-        17600	4.223
-        19200	4.839
-        20800	8.201
-        22400	8.273
-        24000	13.426
-        25600	10.089
-        27200	15.889
-        28800	16.431
-        30400	21.437
-      }; \addlegendentry{prpu-SML, genrmf};
-
-      \addplot table {
-        16000	4.51
-        17600	4.804
-        19200	5.899
-        20800	8.052
-        22400	8.448
-        24000	11.555
-        25600	8.878
-        27200	13.999
-        28800	12.99
-        30400	13.791
-      }; \addlegendentry{prpu-C++, genrmf};
-
-      \addplot table {
-        16000	0.009
-        17600	0.012
-        19200	0.015
-        20800	0.021
-        22400	0.021
-        24000	0.019
-        25600	0.023
-        27200	0.028
-        28800	0.026
-        30400	0.026
-      }; \addlegendentry{prpu-C, genrmf};
-    \end{axis}
-  \end{tikzpicture}
+
   \caption{Benchmark of different implementations. The x-axis shows the number of nodes, the y axis the execution time in seconds.}\label{fig:benchmark}

### 2017-02-28 paper/root.tex

@@
       }; \addlegendentry{prpu-SML, sparse};
-
-      \addplot table {
-        1000 .053   
-        1500 .089   
-        2000 .091   
-        2500 .120 
-        3000 .256 
-        3500 .210 
-        4000 .265 
-        4500 .499 
-        5000 .457 
-        5500 .522     
-      }; \addlegendentry{prpu-Java2, sparse};
-
-      \addplot table {
-        1000 .014 
-        1500 .029 
-        2000 .027 
-        2500 .037 
-        3000 .043 
-        3500 .051 
-        4000 .053 
-        4500 .080 
-        5000 .084 
-        5500 .085 
-      }; \addlegendentry{prpu-Java1, sparse};
-
-      \addplot table {
-        1000 .006 
-        1500 .013
-        2000 .014
-        2500 .019
-        3000 .041
-        3500 .035
-        4000 .042
-        4500 .076
-        5000 .079
-        5500 .082    
-      }; \addlegendentry{prpu-C++, sparse};
-
-      \addplot table {
-        1000 0  
-        1500 .001  
-        2000 .003  
-        2500 .002  
-        3000 .003  
-        3500 .004  
-        4000 .005  
-        4500 .007  
-        5000 .010 
-        5500 .009      
-      }; \addlegendentry{prpu-C, sparse};
     %end data for plotting
@@
       }; \addlegendentry{prpu-SML, dense};
-
-      \addplot table {
-        1000	0.035
-        1050	0.037
-        1100	0.043
-        1150	0.044
-        1200	0.051
-        1250	0.065
-        1300	0.058
-        1350	0.068
-        1400	0.059
-        1450	0.069  
-      }; \addlegendentry{prpu-Java2, dense};
-
-      \addplot table {
-        1000	0.046
-        1050	0.05
-        1100	0.048
-        1150	0.059
-        1200	0.056
-        1250	0.056
-        1300	0.058
-        1350	0.062
-        1400	0.062
-        1450	0.066
-      }; \addlegendentry{prpu-Java1, dense};
-
-      \addplot table {
-        1000	0.034
-        1050	0.035
-        1100	0.039
-        1150	0.042
-        1200	0.052
-        1250	0.052
-        1300	0.054
-        1350	0.057
-        1400	0.064
-        1450	0.066  
-      }; \addlegendentry{prpu-C++, dense};
-
-      \addplot table {
-        1000	0.006
-        1050	0.004
-        1100	0.006
-        1150	0.005
-        1200	0.007
-        1250	0.007
-        1300	0.007
-        1350	0.007
-        1400	0.007
-        1450	0.008
-      }; \addlegendentry{prpu-C, dense};
     \end{axis}
@@
 %         ylabel = Execution time in seconds, 
-        xtick = {0,1000,...,5000},
-        ytick = {0.01,0.1,1,10},
-        xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$},
-        yticklabels = {$0.01$,$0.1$,$1$,$10$}, 
+        xtick = {3000, 6000, 10500, 13500, 16500},
+        ytick = {0.1,1,10,50,100},
+        xticklabels = {$3k$, $6k$, $10k$, $13k$, $16k$},
+        yticklabels = {$\text{\ \,}0.1$,$1$,$10$,$50$,$100$}, 
         smooth,
@@
         legend style = { 
-          at={(0.59,0.4)}, 
+          at={(0.53,0.24)}, 
             anchor=north west,
@@
     %start data for plotting
+
       \addplot table {
-        1000	0.356
-        1500	0.805
-        2000	1.539
-        2500	2.43
-        3000	3.28
-        3500	4.639
-        4000	5.61
-        4500	7.198
-        5000	8.904
-        5500	11.302
-      }; \addlegendentry{edka-SML, wash};
+        3000	1.185
+        4500	3.524
+        6000	6.457
+        7500	11.509
+        9000	17.529
+        10500	26.185
+        12000	36.828
+        13500	48.614
+        15000	86.449
+        16500	89.661
+      }; \addlegendentry{prpu-SML, wash};
 
       \addplot table {
-        1000	0.105
-        1500	0.228
-        2000	0.385
-        2500	0.466
-        3000	0.7
-        3500	0.992
-        4000	1.194
-        4500	1.503
-        5000	1.735
-        5500	2.243
-      }; \addlegendentry{edka-Java, wash};
+        3000	0.906
+        4500	2.708
+        6000	5.083
+        7500	8.792
+        9000	12.951
+        10500	19.075
+        12000	27.466
+        13500	35.964
+        15000	61.839
+        16500	67.039
+      }; \addlegendentry{prpu-C++, wash};
 
       \addplot table {
-        1000	1.185
-        1500	3.524
-        2000	6.457
-        2500	11.509
-        3000	17.529
-        3500	26.185
-        4000	36.828
-        4500	48.614
-        5000	86.449
-        5500	89.661
-      }; \addlegendentry{prpu-SML, wash};
-
-      \addplot table {
-        1000	11.515
-        1500	37.224
-        2000	60.226
-        2500	96.657
-        3000	134.77
-        3500	199.812
-        4000	328.824
-        4500	383.112
-        5000	745.695
-        5500	661.721
-      }; \addlegendentry{prpu-Java2, wash};
-
-      \addplot table {
-        1000	0.654
-        1500	1.626
-        2000	3.096
-        2500	5.313
-        3000	8.572
-        3500	12.552
-        4000	18.103
-        4500	18.646
-        5000	40.456
-        5500	39.012
-      }; \addlegendentry{prpu-Java1, wash};
-
-      \addplot table {
-        1000	0.906
-        1500	2.708
-        2000	5.083
-        2500	8.792
-        3000	12.951
-        3500	19.075
-        4000	27.466
-        4500	35.964
-        5000	61.839
-        5500	67.039
-      }; \addlegendentry{prpu-C++, wash};
-
-      \addplot table {
-        1000	0.006
-        1500	0.015
-        2000	0.029
-        2500	0.042
-        3000	0.063
-        3500	0.084
-        4000	0.115
-        4500	0.144
-        5000	0.183
-        5500	0.223
+        3000	0.006
+        4500	0.015
+        6000	0.029
+        7500	0.042
+        9000	0.063
+        10500	0.084
+        12000	0.115
+        13500	0.144
+        15000	0.183
+        16500	0.223
       }; \addlegendentry{prpu-C, wash};
@@
         xtick = {0,1000,1100,...,1400},
-        ytick = {2,4,...,10},
+        ytick = {0.01, 0.1, 0.2, 0.5, 1},
         xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, 
-        yticklabels = {$2$, $4$,$6$,$8$,$10$}, 
+        yticklabels = {$0.01$, $0.1$,$0.2$,$0.5$,$1$}, 
         smooth,
@@
         legend style = { 
-          at={(0.59,0.4)}, 
+          at={(0.52,0.24)}, 
             anchor=north west,
@@
     %start data for plotting
-      \addplot table {
-        1000	28.6
-        1050	42.725
-        1100	45.527
-        1150	56.67
-        1200	57.768
-        1250	78.733
-        1300	89.811
-        1350	103.254
-        1400	80.91
-        1450	87.823
-      }; \addlegendentry{edka-SML, acnet};
-
-      \addplot table {
-        1000	21.355
-        1050	23.84
-        1100	27.647
-        1150	36.513
-        1200	35.793
-        1250	40.851
-        1300	45.18
-        1350	49.515
-        1400	60.096
-        1450	65.875
-      }; \addlegendentry{edka-Java, acnet};
-
       \addplot table {
@@
       }; \addlegendentry{prpu-SML, acnet};
-
-      \addplot table {
-        1000	0.069
-        1050	0.08
-        1100	0.066
-        1150	0.094
-        1200	0.091
-        1250	0.094
-        1300	0.101
-        1350	0.209
-        1400	0.16
-        1450	0.185
-      }; \addlegendentry{prpu-Java2, acnet};
-
-      \addplot table {
-        1000	0.081
-        1050	0.112
-        1100	0.091
-        1150	0.151
-        1200	0.141
-        1250	0.135
-        1300	0.16
-        1350	0.29
-        1400	0.21
-        1450	0.269
-      }; \addlegendentry{prpu-Java1, acnet};
 
@@
   \end{tikzpicture}
-
+  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
+    \begin{axis}[
+      %title={Benchmark of the Edmonds-Karp algorithm},
+        axis x line*= bottom,
+        axis y line*= left,
+        xmode = log,
+        ymode = log,
+%         xlabel = Number of vertices in the network,
+%         ylabel = Execution time in seconds, 
+        xtick = {16000,19000,23000,27000,30000},
+        ytick = {0.1,1,5,10,25},
+        xticklabels = {$16k$, $19k$, $23k$, $27k$, $30k$}, 
+        yticklabels = {$\text{\ \,}0.1$, $1$,$5$,$10$,$25$}, 
+        smooth,
+        cycle list name = black white,
+        legend style = { 
+          at={(0.49,0.6)}, 
+            anchor=north west,
+            draw=black, 
+            fill= white,
+            align=left
+        }
+    ]
+
+    %start data for plotting
+      \addplot table {
+        16000	3.602
+        17600	4.223
+        19200	4.839
+        20800	8.201
+        22400	8.273
+        24000	13.426
+        25600	10.089
+        27200	15.889
+        28800	16.431
+        30400	21.437
+      }; \addlegendentry{prpu-SML, genrmf};
+
+      \addplot table {
+        16000	4.51
+        17600	4.804
+        19200	5.899
+        20800	8.052
+        22400	8.448
+        24000	11.555
+        25600	8.878
+        27200	13.999
+        28800	12.99
+        30400	13.791
+      }; \addlegendentry{prpu-C++, genrmf};
+
+      \addplot table {
+        16000	0.009
+        17600	0.012
+        19200	0.015
+        20800	0.021
+        22400	0.021
+        24000	0.019
+        25600	0.023
+        27200	0.028
+        28800	0.026
+        30400	0.026
+      }; \addlegendentry{prpu-C, genrmf};
+    \end{axis}
+  \end{tikzpicture}
   \caption{Benchmark of different implementations. The x-axis shows the number of nodes, the y axis the execution time in seconds.}\label{fig:benchmark}

### 2017-02-28 paper/root.tex
better diagrams

--- 2017-02-28 paper/root.tex
@@
     %start data for plotting
-      \addplot table {
+      \addplot+[mark=square] table {
         1000 .024   
@@
 
-      \addplot table {
+      \addplot+[mark=*] table {
         1000 .024   
@@
 
-      \addplot table {
+      \addplot+[mark=square*] table {
         1000 .005   
@@
     %start data for plotting
-      \addplot table {
+      \addplot+[mark=square] table {
         1000	2.794
@@
 
-      \addplot table {
+      \addplot+[mark=*] table {
         1000	3.982
@@
 
-      \addplot table {
+      \addplot+[mark=square*] table {
         1000	0.074
@@
 
-      \addplot table {
+      \addplot+[mark=square*] table {
         3000	1.185
@@
 
-      \addplot table {
+      \addplot+[mark=o] table {
         3000	0.906
@@
 
-      \addplot table {
+      \addplot+[mark=otimes*] table {
         3000	0.006
@@
     %start data for plotting
-      \addplot table {
+      \addplot+[mark=square*] table {
         1000	0.141
@@
 
-      \addplot table {
+      \addplot+[mark=o] table {
         1000	0.121
@@
 
-      \addplot table {
+      \addplot+[mark=otimes*] table {
         1000	0.056
@@
     %start data for plotting
-      \addplot table {
+      \addplot+[mark=square*] table {
         16000	3.602
@@
 
-      \addplot table {
+      \addplot+[mark=o] table {
         16000	4.51
@@
 
-      \addplot table {
+      \addplot+[mark=otimes*] table {
         16000	0.009

### 2017-02-28 paper/root.tex
Upd concludion

--- 2017-02-28 paper/root.tex
@@
 \section{Conclusion}\label{sec:concl}
-  We have presented a verification of the Edmonds-Karp algorithm, using a stepwise refinement approach.
-  Starting with a proof of the Ford-Fulkerson theorem, we have verified the generic Ford-Fulkerson method, 
-  specialized it to the Edmonds-Karp algorithm, and proved the upper bound $O(VE)$ for the number of outer loop iterations.
-  We then conducted several refinement steps to derive an efficiently executable implementation of the algorithm, 
-  including a verified breadth first search algorithm to obtain shortest augmenting paths. 
-  Finally, we added a verified algorithm to check whether the input is a valid network, and generated executable code in SML.
-  The runtime of our verified implementation compares well to that of an unverified reference implementation in Java.
+  We have presented a mechanical verification of the correctness and time complexity of maximum flow algorithms based on both, 
+  the augmenting path and the push-relabel approaches, as well as the Ford-Fulkerson (min-cut-max-flow) theorem, which underlies all these algorithms.
+
+  For the augmenting path algorithms, we start with a verification of the generic Ford-Fulkerson method, 
+  specialize it to the Edmonds-Karp algorithm, and prove its time complexity of $O(VE^2)$.
+  For the push-relabel approaches, we start with a verification of the generic push-relabel algorithm of Goldberg and Tarjan, show its time complexity of $O(V^2E)$, and
+  instantiate it to both, the relabel-to-front and the FIFO push-relabel algorithm.
+  We then conduct several refinement steps to derive efficiently executable implementations of all three algorithm, from which we extract SML code.
+  The runtime of our verified implementations compares well to that of unverified reference implementations.
 
@@
   still accessible proof of the Ford-Fulkerson theorem. The Isabelle Refinement Framework~\cite{LaTu12,La12} and the Sepref tool~\cite{La15,La16}
-  allowed us to present the Ford-Fulkerson method on a level 
-  of abstraction that closely resembles pseudocode presentations found in textbooks, and then formally link this presentation to an efficient
-  implementation. Moreover, modularity of refinement allowed us to develop the breadth first search algorithm independently, and later link it to the 
-  main algorithm.
-The BFS algorithm can be reused as building block for other algorithms.
+  allowed us to present the algorithms on a level 
+  of abstraction that closely resembles pseudocode presentations found in textbooks, and then formally link this presentation to efficient
+  implementations. Moreover, modularity of refinement allowed us to independently develop subroutines like the breadth first search algorithm, 
+  which can be reused as building block for other algorithms.
   The data structures are re-usable, too: although we had to implement the array representation of (capacity) matrices for this project, it will be added to the growing library of verified imperative data structures
@@
   We conclude with some statistics:
-  The formalization consists of roughly 13000 lines of proof text, where the graph theory up to the Ford-Fulkerson algorithm requires 3000 lines.
+  The formalization consists of roughly 14000 lines of proof text, where the graph theory up to the Ford-Fulkerson algorithm requires 2500 lines.
   The abstract Edmonds-Karp algorithm and its complexity analysis contribute 800 lines, and its implementation (including BFS) another 1700 lines.
-  The generic push relabel algorithm and its complexity analysis is 1700 lines, the abstract implementations of relabel-to-front and FIFO push-relabel is another 1000 lines,
-  and their refinement to an efficient implementation takes 2000 lines.
+  The generic push relabel algorithm and its complexity analysis is 1700 lines, the abstract implementations of relabel-to-front and FIFO push-relabel is another 1100 lines,
+  and their refinement to an efficient implementation takes 3000 lines.
 
@@
   While we implemented a simple version of the push-relabel algorithm, there are more advanced implementations, like the one of Cherkassky et el.~\cite{ChGo97} that 
-  we also use in the benchmarks. The same authors also provide an implementation based on the highest label selection rule, which is reported to be even more efficient 
+  we use in the benchmarks. The same authors also provide an implementation based on the highest label selection rule, which is reported to be even more efficient 
   in practice. 
-Our formalized algorithm misses two important heuristics of these advanced algorithms: First, the global relabeling heuristics periodically performs a breadth first search on the reversed residual graph, to determine exact node labels based on their distance to the sink.
+  Compared to these advanced algorithms, our formalization misses two important heuristics: First, the global relabeling heuristics periodically performs a breadth first search on the reversed residual graph, to determine exact node labels based on their distance to the sink.
 Second, the algorithm is split in two phases:
   The first phase ignores nodes with labels greater $|V|$, and already determines the exact value of the flow. The second phase only pushes excess flow back to the source. 
-  Using different heuristics for both phases is reported to improve practical performance~\cite{ChGo97}. 
+  Using different implementations for the two phases is reported to improve practical performance~\cite{ChGo97}. 
   Thus, the main focus of future work will be to verify these heuristics in order to obtain more efficient verified algorithms.

### 2017-02-28 paper/root.tex
Merged all diagrams into one Legend, added titles

--- 2017-02-28 paper/root.tex
@@
   \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
-    \begin{axis}[
-      %title={Benchmark of the Edmonds-Karp algorithm},
+    \begin{groupplot}[
+        group style={
+            group size=2 by 3,
+            vertical sep=1.5cm,
+        }
+    ]
+
+    \nextgroupplot[
+        title={Random Sparse Graphs},
+        title style={yshift=-2ex},
         axis x line*= bottom,
@@
         ymode = log,
-%         xlabel = Number of vertices in the network,
-%         ylabel = Execution time in seconds, 
         xtick = {0,1000,...,5000},
@@
         cycle list name = black white,
-        legend style = { 
-          at={(0.515,0.24)}, 
-            anchor=north west,
-            draw=black, 
-            fill= white,
-            align=left
-        }
+        legend cell align={left},
+        legend to name=grouplegend
     ]
+    % legend information
+      \addlegendimage{fill=gray, mark=square}
+      \addlegendimage{fill=gray, mark=square*}
+      \addlegendimage{fill=gray, mark=*}
+      \addlegendimage{fill=gray, mark=o}
+      \addlegendimage{fill=gray, mark=otimes*}
+      \addlegendentry{Edmonds-Karp, SML}
+      \addlegendentry{FIFO push-relabel, SML}
+      \addlegendentry{Edmonds-Karp, Java}
+      \addlegendentry{FIFO push-relabel, C++}
+      \addlegendentry{FIFO push-relabel, C}
+    % end legend information
 
 		%start data for plotting
-      \addplot table {
+		  \addplot+[mark=square] table {
 		    1000 .024   
@@
 		    5500 11.749    
-      }; \addlegendentry{edka-SML, sparse};
-
-      \addplot table {
+		  }; %Sparse, Edka, SML
+
+		  \addplot+[mark=*] table {
 		    1000 .024   
@@
 		    5500 6.079     
-      }; \addlegendentry{edka-Java, sparse};
-
-      \addplot table {
+		  }; %Sparse, Edka, Java
+
+		  \addplot+[mark=square*] table {
 		    1000 .005   
@@
 		    5500 .192    
-      }; \addlegendentry{prpu-SML, sparse};
+		  }; %Sparse, Prpu, SML 
 		%end data for plotting
-    \end{axis}
-  \end{tikzpicture}
-%
-  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
-    \begin{axis}[
-      %title={Benchmark of the Edmonds-Karp algorithm},
+
+	\nextgroupplot[
+    	title={Random Dense Graphs},
+    	title style={yshift=-2ex},
       axis x line*= bottom,
@@
       ymode = log,
-%         xlabel = Number of vertices in the network,
-%         ylabel = Execution time in seconds, 
       xtick = {0,1000,1100,...,1400},
@@
       smooth,
-        cycle list name = black white,
-        legend style = { 
-          at={(0.53,0.6)}, 
-            anchor=north west,
-            draw=black, 
-            fill= white,
-            align=left
-        }
+      cycle list name = black white
 	]
@@
 		%start data for plotting 
-      \addplot table {
+		  \addplot+[mark=square] table {
 		    1000	2.794
@@
 		    1450	10.042 
-      }; \addlegendentry{edka-SML, dense};
-
-      \addplot table {
+		  }; %Dense, Edka, SML
+
+		  \addplot+[mark=*] table {
 		    1000	3.982
@@
 		    1450	13.11  
-      }; \addlegendentry{edka-Java, dense};
-
-      \addplot table {
+		  }; %Dense, Edka, Java
+
+		  \addplot+[mark=square*] table {
 		    1000	0.074
@@
 		    1450	0.114  
-      }; \addlegendentry{prpu-SML, dense};
-    \end{axis}
-  \end{tikzpicture}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
-    \begin{axis}[
-      %title={Benchmark of the Edmonds-Karp algorithm},
+		  }; %Dense, Prpu, SML  
+		%end data for plotting
+
+	\nextgroupplot[
+      title={Goldberg's Bad Case (Washington)},
+      title style={yshift=-2ex},
       axis x line*= bottom,
@@
       ymode = log,
-%         xlabel = Number of vertices in the network,
-%         ylabel = Execution time in seconds, 
       xtick = {3000, 6000, 10500, 13500, 16500},
@@
       xticklabels = {$3k$, $6k$, $10k$, $13k$, $16k$},
-        yticklabels = {$\text{\ \,}0.1$,$1$,$10$,$50$,$100$}, 
+      yticklabels = {$0.1$,$1$,$10$,$50$,$100$}, 
       smooth,
-        cycle list name = black white,
-        legend style = { 
-          at={(0.53,0.24)}, 
-            anchor=north west,
-            draw=black, 
-            fill= white,
-            align=left
-        }
+      cycle list name = black white
 	]
@@
     %start data for plotting
-
-      \addplot table {
+      \addplot+[mark=square*] table {
         3000	1.185
@@
         16500	89.661
-      }; \addlegendentry{prpu-SML, wash};
-
-      \addplot table {
+      }; %WASH, Prpu, SML
+
+      \addplot+[mark=o] table {
         3000	0.906
@@
         16500	67.039
-      }; \addlegendentry{prpu-C++, wash};
-
-      \addplot table {
+      }; %WASH, Prpu, C++
+
+      \addplot+[mark=otimes*] table {
         3000	0.006
@@
         16500	0.223
-      }; \addlegendentry{prpu-C, wash};
-    %end data for plotting
-    \end{axis}
-  \end{tikzpicture}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
-    \begin{axis}[
-      %title={Benchmark of the Edmonds-Karp algorithm},
+      }; %WASH, Prpu, C
+
+	\nextgroupplot[
+      title={Fully Dense Acyclic Networks (Waissi)},
+      title style={yshift=-2ex},
       axis x line*= bottom,
@@
       smooth,
-        cycle list name = black white,
-        legend style = { 
-          at={(0.52,0.24)}, 
-            anchor=north west,
-            draw=black, 
-            fill= white,
-            align=left
-        }
+      cycle list name = black white
 	]
@@
     %start data for plotting 
-      \addplot table {
+      \addplot+[mark=square*] table {
         1000	0.141
@@
         1450	0.806
-      }; \addlegendentry{prpu-SML, acnet};
-
-      \addplot table {
+      }; %ACNET, Prpu, SML
+
+      \addplot+[mark=o] table {
         1000	0.121
@@
         1450	0.486
-      }; \addlegendentry{prpu-C++, acnet};
-
-      \addplot table {
+      }; %ACNET, Prpu, C++
+
+      \addplot+[mark=otimes*] table {
         1000	0.056
@@
         1450	0.209
-      }; \addlegendentry{prpu-C, acnet};
-    \end{axis}
-  \end{tikzpicture}
-  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
-    \begin{axis}[
-      %title={Benchmark of the Edmonds-Karp algorithm},
+      }; %ACNET, Prpu, C
+
+	\nextgroupplot[
+      title={RMFGEN Networks (Goldfarb et. al)},
+      title style={yshift=-2ex},
       axis x line*= bottom,
@@
       ymode = log,
-%         xlabel = Number of vertices in the network,
-%         ylabel = Execution time in seconds, 
       xtick = {16000,19000,23000,27000,30000},
@@
       xticklabels = {$16k$, $19k$, $23k$, $27k$, $30k$}, 
-        yticklabels = {$\text{\ \,}0.1$, $1$,$5$,$10$,$25$}, 
+      yticklabels = {$0.1$, $1$,$5$,$10$,$25$}, 
       smooth,
-        cycle list name = black white,
-        legend style = { 
-          at={(0.49,0.6)}, 
-            anchor=north west,
-            draw=black, 
-            fill= white,
-            align=left
-        }
+      cycle list name = black white
 	]
@@
         30400	21.437
-      }; \addlegendentry{prpu-SML, genrmf};
-
-      \addplot table {
+      }; %RPFGEN, Prpu, SML
+
+      \addplot+[mark=o] table {
         16000	4.51
@@
         30400	13.791
-      }; \addlegendentry{prpu-C++, genrmf};
-
-      \addplot table {
+      }; %RPFGEN, Prpu, C++
+
+      \addplot+[mark=otimes*] table {
         16000	0.009
@@
         30400	0.026
-      }; \addlegendentry{prpu-C, genrmf};
-    \end{axis}
+      }; %RPFGEN, Prpu, C  
+    \end{groupplot}
+
+    \node at (group c1r3.east) [anchor=west, xshift= 1cm] {\ref{grouplegend}};
   \end{tikzpicture}

### 2017-02-28 paper/root.tex
better node points for diagrams

@@
         ytick = {0.01,0.1,1,10},
-        xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$},
+        xticklabels = {$1k$, $2k$, $3k$, $4k$, $5k$},
         yticklabels = {$0.01$,$0.1$,$1$,$10$},
@@
       xtick = {0,1000,1100,...,1400},
-      ytick = {1,4,...,16},
-      xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, 
-      yticklabels = {$1$,$4$,$8$,$12$,$16$}, 
+      ytick = {2,4,6,10,16},
+      xticklabels = {$1k$, $1.1k$, $1.2k$, $1.3k$, $1.4k$}, 
+      yticklabels = {$2$,$4$,$6$,$10$,$16$}, 
       smooth,
@@
       ytick = {0.01, 0.1, 0.2, 0.5, 1},
-      xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, 
+      xticklabels = {$1k$, $1.1k$, $1.2k$, $1.3k$, $1.4k$}, 
       yticklabels = {$0.01$, $0.1$,$0.2$,$0.5$,$1$}, 
@@
       xtick = {16000,19000,23000,27000,30000},
-      ytick = {0.1,1,5,10,25},
+      ytick = {0.1,3,6,12,24},
       xticklabels = {$16k$, $19k$, $23k$, $27k$, $30k$}, 
-      yticklabels = {$0.1$, $1$,$5$,$10$,$25$}, 
+      yticklabels = {$0.1$, $3$,$6$,$12$,$24$}, 
       smooth,

### 2017-03-01 paper/root.tex

@@
 
+\usepgfplotslibrary{groupplots}
+
 \usepackage{packages/isabelle}
@@
 \begin{figure}[!h]
-  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
+\begin{center}    
+  \begin{tikzpicture}[thick,scale=.75, every node/.style={scale=1}] %change the scales if you like to reduce the size
     \begin{groupplot}[
@@
             group size=2 by 3,
-            vertical sep=1.5cm,
+            vertical sep=1.4cm,
         }
@@
       \addlegendimage{fill=gray, mark=square*}
-      \addlegendimage{fill=gray, mark=*}
+      \addlegendimage{dashed, every mark/.append style={solid, fill=gray}, mark=*}
       \addlegendimage{fill=gray, mark=o}
@@
 
-		  \addplot+[mark=*] table {
+		  \addplot+[dashed, every mark/.append style={solid, fill=gray},mark=*] table {
 		    1000 .024   
@@
 
-		  \addplot+[mark=*] table {
+		  \addplot+[dashed, every mark/.append style={solid, fill=gray},mark=*] table {
 		    1000	3.982
@@
 
-    \node at (group c1r3.east) [anchor=west, xshift= 1cm] {\ref{grouplegend}};
+    \node at (group c1r3.east) [anchor=west, xshift= 1.1cm] {\ref{grouplegend}};
   \end{tikzpicture}
+\end{center}
 \caption{Benchmark of different implementations. The x-axis shows the number of nodes, the y axis the execution time in seconds.}\label{fig:benchmark}

### 2017-03-01 paper/root.tex
Rephrased impl-section

@@
 
-  \subsection{Using Breadth First Search}
-  A standard way to find a shortest path in a graph is breadth first search (BFS). Luckily, we had already formalized a BFS algorithm as an example for the 
-  Isabelle Refinement Framework.
-Unfortunately, this algorithm only computed the minimum distance between two nodes, without returning an actual path.
-For this project, we extended the formalization accordingly, and added an efficient imperative implementation, using the same stepwise refinement techniques as for the main algorithm.
-Note that the resulting BFS algorithm is independent, and can be reused for finding shortest paths in other applications.
-
-%   Implementing shortest path finding by BFS in the Edmonds-Karp algorithm yields a specification that algorithmically describes all major operations, 
-%   but still leaves open the data structures used for implementation.
-
-%   
-%   We briefly report on our adapted formalization here, which is displayed in Figure~\ref{TODO: Snippet}: 
-%   TODO: Reference to figure (line numbers?)
-%   The abstract algorithm keeps track of a set $C$ of current nodes, a set $N$ of new nodes, and a partial map $P$ from already visited nodes to predecessor nodes.
-%   Initially, only the start node is in $C$ and $N$ is empty. In each iteration of the loop, a node $u\in C$ is picked, and its successors $v$ that have not yet been 
-%   visited are added to $N$, and $P~v$ is set to $u$. If $C$ becomes empty, $C$ and $N$ are swapped. If the target node is encountered, the algorithm immediately terminates.
-%   Note that this implementation is a generalized version of the usual queue implementation of BFS: While a queue enforces a complete order on the encountered nodes, our 
-%   implementation only enforces an ordering between nodes on the current level and nodes on the next level. 
-% 
-%   For the actual algorithm, we wrap this algorithm by a procedure to handle the special case where source and target nodes are the same, and to extract a shortest path 
-%   from $P$ upon termination. Moreover, we implement the abstract specification of adding the successor nodes of a node by a loop.
-%   
-%   Finally, we prove the following theorem: %TODO: Not actually true, the impls are oly done in bfs2!
-% %     theorem
-% %       assumes "src∈V" 
-% %       shows "bfs src dst ≤ (spec p. 
-% %         case p of None ⇒ ¬conn src dst | Some p ⇒ shortestPath src p dst)"
-%   
-%   Obviously, the BFS algorithm refines the specification for obtaining a shortest path.
-%   Using this to refine the Edmonds-Karp formalization yields a program that algorithmically specifies all major operations.
-
-  \subsection{Manipulating Residual Graphs Directly}\label{sec:impl_res_graph}
-  The Edmonds-Karp algorithm is phrased in terms of a flow, which is updated until it is maximal. 
-  In each iteration, an augmenting path is searched on the residual graph induced by the current flow.
-  Obviously, computing the complete residual graph in each iteration is a bad idea. 
-  One solution to this problem is to compute the edges of the residual graph on the fly from the network and the current flow.
-  Although this solution seems to be common, it has the disadvantage that for each edge of the residual graph, two (or even three) 
-  edges of the network and the flow have to be accessed. As edges of the residual graph are accessed in the inner loop, during 
-  the BFS, these operations are time critical.
-
-  After our profiling indicated a hot spot on accessing the capacity matrices of the network and the flow, we switched to 
-  an algorithm that operates on a representation of the residual graph directly. This resulted in a speed-up of roughly a factor of two.
-  As the residual graph uniquely determines the flow (and vice versa), it is straightforward to phrase the
-  operations directly on the residual graph. Performing a data refinement of the flow \wrt the refinement 
-  relation $\{(c_f,f) \mid \text{$f$ is a flow}\}$ then yields the desired algorithm.
-
-  For the push-relabel algorithms, we combine the transition from preflows to residual graphs with the introduction of an 
-  explicit map from nodes to their current excess values: The refinement relation $\{((x_f,c_f),f) \mid \text{$f$ is a flow}\}$
-  relates pairs of excess values and residual graphs to the corresponding flow.
-
-  \subsection{Height Frequencies}
-  For an efficient implementation of the gap heuristics, we introduce an explicit map from height-values to the number of nodes with this height.
-  Using this map, it is easy to detect when we relabel the last node of a specific height such that the gap heuristics applies.
-
-  \subsection{Implementing Augmentation}    
-  In our abstract formalization of the Ford-Fulkerson method, which matches the presentation in Section~\ref{sec:background}, we have formulated augmentation by first defining 
-  the residual capacity $c_p$ of the augmenting path. Using $c_p$, we have defined the residual flow $f_p$, which was finally added to the current flow.
-  In the refinement to operate on residual graphs, we have refined this to augment the residual graph.
-  For the implementation, we compute the residual capacity in a first iteration over the augmenting path, and modify the residual graph 
-  in a second iteration. Proving this implementation correct is straightforward by induction on the augmenting path.
-
-  \subsection{Algorithms for Relabel and Discharge}    
-  For the relabel operation, it is straightforward to find the new label by iterating over the outgoing edges in the residual graph.
-  The discharge operation is a bit more tricky: Abstractly, we nondeterministically select a new edge to push, until we find no more edges.
+  \subsection{Algorithms}
+  The most complex problem for which we specified no algorithm is finding a shortest augmenting path.
+  We use breadth first search (BFS) for this. We adapted a BFS formalization from the examples collection of the Isabelle Refinement Framework to
+  our needs and added an efficient imperative implementation. The resulting algorithm 
+  is independent of flow networks, and can be re-used to find shortest path in directed graphs also in other applications.
+
+  Minor problems for which we had to develop algorithms are augmentation of the flow along the shortest path, and the relabel and discharge operations.
+  While it is straightforward to find the height to relabel a node with by iterating over the outgoing edges, the discharge operation is a bit more tricky:
+  Abstractly, we nondeterministically select a new edge to push, until we find no more edges.
   In the concrete implementation, however, we fold over the list of successor nodes.
@@
 
-  \subsection{Computing Successors}
-  In order to find an augmenting path, the BFS algorithm has to compute the successors of a node in the residual graph. 
-  Similar, the relabel operation has to iterate over all successors to find the new minimal labeling, and the discharge operation iterates over all outgoing edges of a node.
-  Although these iterations can be implemented by iterating over all nodes, this would be inefficient for sparse graphs,
-  where we would iterate over many possible successor nodes just to find that there is no edge.
-
-  A common optimization is to pre-compute an \emph{adjacency map} from nodes to adjacent nodes in the network. 
-  As an edge in the residual graph is either in the same or opposite direction of 
-  a network edge, it is enough to iterate over the adjacent nodes in the network, and check whether they are actual successors in the residual graph.
-  It is straightforward to show that this implementation actually returns the successor nodes in the residual graph.
+  \subsection{Additional Data}\label{sec:impl_add_data}
+  An important optimization is to let the algorithms manipulate residual graphs instead of flows: 
+  Most operations can be phrased to only access one edge of the residual graph, while they would 
+  need to access multiple edges if phrased on flows. At the end of the algorithm, the flow can be 
+  computed from the residual graph. 
+  An initial version of our Edmonds-Karp algorithm worked on flows, and switching to residual graphs yielded a speedup of roughly factor 2.
+
+  Another important optimization is to pre-compute an adjacency map of the network:
+  To find the successors of a node in the residual graph, it is enough to search among the adjacent nodes in the network. 
+  As iterations over successor nodes occur in the inner loops of the algorithms, this optimization is quite effective, in particular for sparse graphs.
+
+  Moreover, for the push-relabel algorithms, we introduce a map from nodes to their current excess value: 
+  A single lookup in this map is sufficient to compute the flow for a push operation --- instead of summing up all 
+  the incoming and outgoing flows.
+
+  For the gap-heuristics, we introduce a height frequency map, which maps each height value to the number of nodes labeled with this value.
+  A single lookup in this map suffices to detect when we relabel the last node of a specific height such that the gap heuristics applies.
 
   \subsection{Using Efficient Data Structures}\label{sec:impl_data_structures}
-  In a final step, we choose efficient data structures. 
+  In a final step, we refine our algorithms to use efficient data structures.
+  This step is partially automated by the Sepref tool~\cite{La15,La16}, which synthesizes imperative programs and 
+  the corresponding refinement proofs, based on a user defined mapping from abstract types to data structures.
+  The imperative programs are specified in Imperative HOL~\cite{BKHEM08}, which defines a heap monad for Isabelle/HOL.
+
   We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers.
@@
   % TODO: Can we trick the Java reference implementation on integer overflow?
-
-  We implement nodes as natural numbers less than an upper bound $N$, and residual graphs are implemented by their capacity matrices, which, in turn,
+  Nodes are represented by natural numbers less than an upper bound $N$, and residual graphs are implemented by their capacity matrices, which, in turn,
   are realized as arrays of size $N\times N$ with row-major indexing, such that the successors of a node are close together in memory.
@@
   An augmenting path is represented by a list of edges, \ie a list of pairs of nodes. 
-
-%   For the BFS algorithm, we implement the sets $C$ and $N$ by lists, and the predecessor map $P$ as an array of \isai{node option}. 
-%   The successors of a node are represented as a list with no duplicate elements. The resulting path, which is extracted from $P$ upon termination of
-%   the BFS, is represented as a list of edges.
-
-  The input network of the algorithm is represented as a function from network edges to capacities, which is tabulated into an array
-  to obtain the initial residual graph. This gives us some flexibility in using the algorithm, as any capacity matrix representation can be converted into a 
-  function easily, without losing efficiency for read-access. 
-  Similarly, our implementations expect an adjacency map as additional parameter, which is then tabulated into an array. This is convenient in our context, 
-  where a preprocessing step computes the adjacency map anyway.
-
-  The output flow of the algorithms is represented as the residual graph. The user can decide how to compute the maximum flow from it. For example,
-  in order to compute the maximum flow value, only the outgoing edges of the source node have to be computed, which is typically less 
-  expensive than computing the complete flow matrix. The correctness theorem of the algorithm abstractly states how to obtain the maximum flow from the output.
-
-  For the push-relabel algorithms, the maps from a node to its height and excess, as well as the height frequency map are implemented by arrays. 
-  The queue of the FIFO push-relabel algorithm is implemented by two stacks, yielding amortized constant time enqueue and dequeue operations.
-  Similarly, the list of the relabel-to-front algorithm is implemented by two lists, \emph{left} and \emph{right}, that contain the nodes left and right of the current position,
-  the left list being stored in reverse order. Advancing the position is implemented in constant time by popping the first element of the right list, and pushing 
-  it to the left list. Resetting the position to the front of the list is implemented by pushing all elements from the left list to the right list. 
-  By accounting these push operations to the operations that created the left list, resetting the position can be estimated to be amortized constant time.
+  The height frequency map and the maps from nodes to heights and excess values are implemented as arrays.
+  The queue of the FIFO push-relabel algorithm and the list of the relabel-to-front algorithm are implemented by two stacks, allowing for amortized constant time operations.
 
@@
 
-  Note that we performed the last refinement step using our Sepref tool~\cite{La15,La16}, which automates the refinement from 
-  purely functional programs into imperative programs expressed in Imperative/HOL~\cite{BKHEM08}.
-  This refinement step consists of setting up the mappings between the abstract and concrete data structures,
-  and then using Sepref to synthesize the Imperative/HOL programs and their refinement proofs.
-  Finally, Imperative/HOL comes with a setup for the Isabelle code generator~\cite{Haft09,HaNi10} to generate imperative programs in OCaml, SML, Scala, and Haskell.
+  Finally, we use the Isabelle code generator~\cite{Haft09,HaNi10} to extract programs in OCaml, SML, Scala, and Haskell.
+
+%   
+%   
+%   automates the refinement from 
+%   purely functional programs into imperative programs expressed in Imperative/HOL~\cite{BKHEM08}.
+%   
+%   choose efficient data structures. 
+%   
+%   
+%   We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers.
+Thus, we could chose any executable number representation here.}.
+Note that an implementation as fixed precision numbers would also be possible,
+%   but requires additional checks on the network to ensure that no overflows can happen. 
+%   % TODO: Can we trick the Java reference implementation on integer overflow?
+%   
+%   We implement nodes as natural numbers less than an upper bound $N$, and residual graphs are implemented by their capacity matrices, which, in turn,
+%   are realized as arrays of size $N\times N$ with row-major indexing, such that the successors of a node are close together in memory.
+%   The adjacency map of the network is implemented as an array of lists of nodes. 
+%   An augmenting path is represented by a list of edges, \ie a list of pairs of nodes. 
+%   
+%   
+% %   For the BFS algorithm, we implement the sets $C$ and $N$ by lists, and the predecessor map $P$ as an array of \isai{node option}. 
+% %   The successors of a node are represented as a list with no duplicate elements. The resulting path, which is extracted from $P$ upon termination of
+% %   the BFS, is represented as a list of edges. 
+%   
+%   
+%   The input network of the algorithm is represented as a function from network edges to capacities, which is tabulated into an array
+%   to obtain the initial residual graph. This gives us some flexibility in using the algorithm, as any capacity matrix representation can be converted into a 
+%   function easily, without losing efficiency for read-access. 
+%   Similarly, our implementations expect an adjacency map as additional parameter, which is then tabulated into an array. This is convenient in our context, 
+%   where a preprocessing step computes the adjacency map anyway.
+%       
+%   The output flow of the algorithms is represented as the residual graph. The user can decide how to compute the maximum flow from it. For example,
+%   in order to compute the maximum flow value, only the outgoing edges of the source node have to be computed, which is typically less 
+%   expensive than computing the complete flow matrix. The correctness theorem of the algorithm abstractly states how to obtain the maximum flow from the output.
+%   
+%   For the push-relabel algorithms, the maps from a node to its height and excess, as well as the height frequency map are implemented by arrays. 
+%   The queue of the FIFO push-relabel algorithm is implemented by two stacks, yielding amortized constant time enqueue and dequeue operations.
+%   Similarly, the list of the relabel-to-front algorithm is implemented by two lists, \emph{left} and \emph{right}, that contain the nodes left and right of the current position,
+%   the left list being stored in reverse order. Advancing the position is implemented in constant time by popping the first element of the right list, and pushing 
+%   it to the left list. Resetting the position to the front of the list is implemented by pushing all elements from the left list to the right list. 
+%   By accounting these push operations to the operations that created the left list, resetting the position can be estimated to be amortized constant time.
+%   
+%   
+%   
+%   
+%   
+%   
+%   
+%   
+%   
+%   
+%   \subsection{Using Breadth First Search}
+%   A standard way to find a shortest path in a graph is breadth first search (BFS). Luckily, we had already formalized a BFS algorithm as an example for the 
+%   Isabelle Refinement Framework.
+Unfortunately, this algorithm only computed the minimum distance between two nodes, without returning an actual path.
+For this project, we extended the formalization accordingly, and added an efficient imperative implementation, using the same stepwise refinement techniques as for the main algorithm.
+Note that the resulting BFS algorithm is independent, and can be reused for finding shortest paths in other applications.
+%   
+% %   Implementing shortest path finding by BFS in the Edmonds-Karp algorithm yields a specification that algorithmically describes all major operations, 
+% %   but still leaves open the data structures used for implementation.
+%   
+% %   
+% %   We briefly report on our adapted formalization here, which is displayed in Figure~\ref{TODO: Snippet}: 
+% %   TODO: Reference to figure (line numbers?)
+% %   The abstract algorithm keeps track of a set $C$ of current nodes, a set $N$ of new nodes, and a partial map $P$ from already visited nodes to predecessor nodes.
+% %   Initially, only the start node is in $C$ and $N$ is empty. In each iteration of the loop, a node $u\in C$ is picked, and its successors $v$ that have not yet been 
+% %   visited are added to $N$, and $P~v$ is set to $u$. If $C$ becomes empty, $C$ and $N$ are swapped. If the target node is encountered, the algorithm immediately terminates.
+% %   Note that this implementation is a generalized version of the usual queue implementation of BFS: While a queue enforces a complete order on the encountered nodes, our 
+% %   implementation only enforces an ordering between nodes on the current level and nodes on the next level. 
+% % 
+% %   For the actual algorithm, we wrap this algorithm by a procedure to handle the special case where source and target nodes are the same, and to extract a shortest path 
+% %   from $P$ upon termination. Moreover, we implement the abstract specification of adding the successor nodes of a node by a loop.
+% %   
+% %   Finally, we prove the following theorem: %TODO: Not actually true, the impls are oly done in bfs2!
+% % %     theorem
+% % %       assumes "src∈V" 
+% % %       shows "bfs src dst ≤ (spec p. 
+% % %         case p of None ⇒ ¬conn src dst | Some p ⇒ shortestPath src p dst)"
+% %   
+% %   Obviously, the BFS algorithm refines the specification for obtaining a shortest path.
+% %   Using this to refine the Edmonds-Karp formalization yields a program that algorithmically specifies all major operations.
+%      
+%   \subsection{Manipulating Residual Graphs Directly}\label{sec:impl_res_graph}
+%   The Edmonds-Karp algorithm is phrased in terms of a flow, which is updated until it is maximal. 
+%   In each iteration, an augmenting path is searched on the residual graph induced by the current flow.
+%   Obviously, computing the complete residual graph in each iteration is a bad idea. 
+%   One solution to this problem is to compute the edges of the residual graph on the fly from the network and the current flow.
+%   Although this solution seems to be common, it has the disadvantage that for each edge of the residual graph, two (or even three) 
+%   edges of the network and the flow have to be accessed. As edges of the residual graph are accessed in the inner loop, during 
+%   the BFS, these operations are time critical.
+%   
+%   After our profiling indicated a hot spot on accessing the capacity matrices of the network and the flow, we switched to 
+%   an algorithm that operates on a representation of the residual graph directly. This resulted in a speed-up of roughly a factor of two.
+%   As the residual graph uniquely determines the flow (and vice versa), it is straightforward to phrase the
+%   operations directly on the residual graph. Performing a data refinement of the flow \wrt the refinement 
+%   relation $\{(c_f,f) \mid \text{$f$ is a flow}\}$ then yields the desired algorithm. 
+% 
+%   For the push-relabel algorithms, we combine the transition from preflows to residual graphs with the introduction of an 
+%   explicit map from nodes to their current excess values: The refinement relation $\{((x_f,c_f),f) \mid \text{$f$ is a flow}\}$
+%   relates pairs of excess values and residual graphs to the corresponding flow.
+%   
+%   \subsection{Height Frequencies}
+%   For an efficient implementation of the gap heuristics, we introduce an explicit map from height-values to the number of nodes with this height.
+%   Using this map, it is easy to detect when we relabel the last node of a specific height such that the gap heuristics applies.
+%   
+%   \subsection{Implementing Augmentation}    
+%   In our abstract formalization of the Ford-Fulkerson method, which matches the presentation in Section~\ref{sec:background}, we have formulated augmentation by first defining 
+%   the residual capacity $c_p$ of the augmenting path. Using $c_p$, we have defined the residual flow $f_p$, which was finally added to the current flow.
+%   In the refinement to operate on residual graphs, we have refined this to augment the residual graph.
+%   For the implementation, we compute the residual capacity in a first iteration over the augmenting path, and modify the residual graph 
+%   in a second iteration. Proving this implementation correct is straightforward by induction on the augmenting path.
+%       
+%   \subsection{Algorithms for Relabel and Discharge}    
+%   For the relabel operation, it is straightforward to find the new label by iterating over the outgoing edges in the residual graph.
+%   The discharge operation is a bit more tricky: Abstractly, we nondeterministically select a new edge to push, until we find no more edges.
+%   In the concrete implementation, however, we fold over the list of successor nodes.
+%   This structural difference between abstract and concrete algorithm cannot be handled by the heuristics of the Refinement Framework's verification condition generator.
+%   Thus, we had to manually prove some refinement rules that are specifically tailored to our case.
+%       
+%   \subsection{Computing Successors}
+%   In order to find an augmenting path, the BFS algorithm has to compute the successors of a node in the residual graph. 
+%   Similar, the relabel operation has to iterate over all successors to find the new minimal labeling, and the discharge operation iterates over all outgoing edges of a node.
+%   Although these iterations can be implemented by iterating over all nodes, this would be inefficient for sparse graphs,
+%   where we would iterate over many possible successor nodes just to find that there is no edge. 
+%   
+%   A common optimization is to pre-compute an \emph{adjacency map} from nodes to adjacent nodes in the network. 
+%   As an edge in the residual graph is either in the same or opposite direction of 
+%   a network edge, it is enough to iterate over the adjacent nodes in the network, and check whether they are actual successors in the residual graph.
+%   It is straightforward to show that this implementation actually returns the successor nodes in the residual graph.
+%   
+% 
+%   \subsection{Using Efficient Data Structures}\label{sec:impl_data_structures}
+%   In a final step, we choose efficient data structures. 
+%   We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers.
+Thus, we could chose any executable number representation here.}.
+Note that an implementation as fixed precision numbers would also be possible,
+%   but requires additional checks on the network to ensure that no overflows can happen. 
+%   % TODO: Can we trick the Java reference implementation on integer overflow?
+%   
+%   We implement nodes as natural numbers less than an upper bound $N$, and residual graphs are implemented by their capacity matrices, which, in turn,
+%   are realized as arrays of size $N\times N$ with row-major indexing, such that the successors of a node are close together in memory.
+%   The adjacency map of the network is implemented as an array of lists of nodes. 
+%   An augmenting path is represented by a list of edges, \ie a list of pairs of nodes. 
+%   
+%   
+% %   For the BFS algorithm, we implement the sets $C$ and $N$ by lists, and the predecessor map $P$ as an array of \isai{node option}. 
+% %   The successors of a node are represented as a list with no duplicate elements. The resulting path, which is extracted from $P$ upon termination of
+% %   the BFS, is represented as a list of edges. 
+%   
+%   
+%   The input network of the algorithm is represented as a function from network edges to capacities, which is tabulated into an array
+%   to obtain the initial residual graph. This gives us some flexibility in using the algorithm, as any capacity matrix representation can be converted into a 
+%   function easily, without losing efficiency for read-access. 
+%   Similarly, our implementations expect an adjacency map as additional parameter, which is then tabulated into an array. This is convenient in our context, 
+%   where a preprocessing step computes the adjacency map anyway.
+%       
+%   The output flow of the algorithms is represented as the residual graph. The user can decide how to compute the maximum flow from it. For example,
+%   in order to compute the maximum flow value, only the outgoing edges of the source node have to be computed, which is typically less 
+%   expensive than computing the complete flow matrix. The correctness theorem of the algorithm abstractly states how to obtain the maximum flow from the output.
+%   
+%   For the push-relabel algorithms, the maps from a node to its height and excess, as well as the height frequency map are implemented by arrays. 
+%   The queue of the FIFO push-relabel algorithm is implemented by two stacks, yielding amortized constant time enqueue and dequeue operations.
+%   Similarly, the list of the relabel-to-front algorithm is implemented by two lists, \emph{left} and \emph{right}, that contain the nodes left and right of the current position,
+%   the left list being stored in reverse order. Advancing the position is implemented in constant time by popping the first element of the right list, and pushing 
+%   it to the left list. Resetting the position to the front of the list is implemented by pushing all elements from the left list to the right list. 
+%   By accounting these push operations to the operations that created the left list, resetting the position can be estimated to be amortized constant time.
+%   
+%   
+% 
+%   Note that there is still some optimization potential left in the choice of our data structures: 
+%   For example, the BFS algorithm computes a predecessor map $P$. It then iterates over $P$ to extract the shortest path as a list of edges.
+%   A subsequent iteration over this list computes the residual capacity, and a final iteration performs the augmentation. 
+%   This calls for a deforestation optimization to get rid of the intermediate list, and iterate only two times over the predecessor map directly.
+%   Also the FIFO queue could be implemented by a ring buffer based on a fixed size array, yielding true constant-time operations with less overhead.
+%   Fortunately, our profiling data shows that these operations do not significantly contribute to the runtime of our implementations, such that 
+%   we leave these optimizations for future work.
+%   
+%   Note that we performed the last refinement step using our Sepref tool~\cite{La15,La16}, which automates the refinement from 
+%   purely functional programs into imperative programs expressed in Imperative/HOL~\cite{BKHEM08}.
+%   This refinement step consists of setting up the mappings between the abstract and concrete data structures,
+%   and then using Sepref to synthesize the Imperative/HOL programs and their refinement proofs.
+%   Finally, Imperative/HOL comes with a setup for the Isabelle code generator~\cite{Haft09,HaNi10} to generate imperative programs in OCaml, SML, Scala, and Haskell.
 
@@
 \section{Benchmarking}\label{sec:benchmark}
-  We have compared the running time of our algorithm in SML against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}.
-We have used MLton 20100608~\cite{MLton} and OpenJDK Java 1.7.0\_95, running on a standard laptop machine with a 2.8GHz i7 quadcore processor and 16GiB of RAM.
-
-  We have done the comparison on randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
+  We have benchmarked our verified algorithms against unverified reference implementations.
+  We extract SML code for our verified Edmonds-Karp and FIFO push-relabel algorithms, and compile with MLton 20100608~\cite{MLton}.
+
+  For the Edmonds-Karp algorithm, we use a reference implementation in Java from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}, which 
+  is compiled and ran with (TODO, Reza's machine?) OpenJDK Java 1.8.0.
+  As data set, we use randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
 Note that the maximum density for networks that satisfy our assumptions is $0.5$, as we allow no parallel edges.
-  For sparse networks, we varied the number of nodes between $1000$ and $5500$, for dense networks between $1000$ and $1450$.
-  The results are shown in Figure~\ref{fig:benchmark}, in a double-logarithmic scale.
+
+  For the push-relabel algorithm, we use the C++ implementation from the Stanford ACM-ICPC notebook~\cite{ICPC-notebook}, which served as 
+  a template for our verified FIFO push-relabel algorithm. Moreover, we use a highly optimized C implementation of a FIFO push-relabel 
+  algorithm with various additional heuristics by Cherkassky et al.~\cite{ChGo97}. These were compiled using (TODO: Reza's machine) gcc 5.4 at optimization level 4.
+  The data set consists of networks from the DIMACS implementation challenge~\cite{JGC93}: 
+  The ``Goldberg bad case networks'' from the ``Washington'' generator, which are tailored to be hard for push-relabel algorithms, 
+  the ``fully dense acyclic networks'' of Waissi, and the RMFGEN networks of Goldfarb and Grigoriadis.
+
+  The tests were run on a standard laptop machine with a (TODO: Reza's machine) 2.8GHz i7 quadcore processor and 16GiB of RAM.
+
+  The results are displayed in Figure~\ref{fig:benchmark}, using a double-logarithmic scale.
+  Note that we only measure the time required to run the algorithms themselves, excluding the time required for parsing the input or the network checker.
+
+  For the Edmonds-Karp algorithm, we observe that, for sparse graphs, the Java implementation is roughly faster by a factor of $1.6$, while for dense graphs, 
+  our implementation is faster by a factor of $1.2$. Note that the Java implementation operates on flows, while our implementation 
+  operates on residual graphs (\cf Section~\ref{sec:impl_add_data}). Moreover, the Java implementation does not store the augmenting 
+  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).
+  Our push-relabel algorithm is more than one order of magnitude faster than Edmonds-Karp algorithm.
+
+  For the push-relabel algorithms, our verified algorithm is roughly a factor of 8 (TODO) slower than the unverified C++ implementation.
+  The highly optimized C implementation is, depending on the test case, one to two orders of magnitudes faster. 
+  On the ``Goldberg Bad Case'' networks, which are designed as worst-case problems for push-relabel algorithms, 
+  even Edmonds-Karp algorithm is competitive to the simple FIFO push-relabel algorithms, while the highly optimized one shows no weakness on these networks.
 
@@
 	\nextgroupplot[
-      title={Goldberg's Bad Case (Washington)},
+      title={Goldberg Bad Case (Washington)},
       title style={yshift=-2ex},
@@
     %start data for plotting
+
+		  \addplot+[mark=square] table {
+        3000	0.356
+        4500	0.805
+        6000	1.539
+        7500	2.43
+        9000	3.28
+        10500	4.639
+        12000	5.61
+        13500	7.198
+        15000	8.904
+        16500	11.302
+		  }; %Wash, Edka, SML
+
       \addplot+[mark=square*] table {
@@
 
-  We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of $1.6$, while for dense graphs, our implementation is faster by a factor of $1.2$.
+% 
+%   We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of $1.6$, while for dense graphs, our implementation is faster by a factor of $1.2$.
 Note that the Java implementation operates on flows, while our implementation
-  operates on residual graphs (\cf Section~\ref{sec:impl_res_graph}). Moreover, the Java implementation does not store the augmenting 
-  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).
-  Finally note that a carefully optimized C++ implementation of the algorithm is only slightly faster than the Java implementation for sparse graphs,
-  but roughly one order of magnitude faster for dense graphs. We leave it to future work to investigate this issue, and conclude that we were able to produce
-  a reasonably fast verified implementation.
-
+%   operates on residual graphs (\cf Section~\ref{sec:impl_add_data}). Moreover, the Java implementation does not store the augmenting 
+%   path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).
+%   Finally note that a carefully optimized C++ implementation of the algorithm is only slightly faster than the Java implementation for sparse graphs,
+%   but roughly one order of magnitude faster for dense graphs. We leave it to future work to investigate this issue, and conclude that we were able to produce
+%   a reasonably fast verified implementation.
+%   
 %   , without doing any low-level optimizations, but using the code as generated by the Isabelle/HOL code generator.

### 2017-03-02 formalization/Push_Relabel_Maxflow/evaluation/bench_print.sh
new Benchmark results

@@ -13,13 +13,23 @@ for t in $tests; do
   echo -ne "$t\t& "
   for p in $platforms_fofu; do
     log="fofu-$p/$t.$p.log"
-    time=`grep "@@@time:" $log | sed -re 's/.*@@@time: //g;s/ms//g'`
+    time="-1"
+    if [ -e $log ]; then
+      if [ -n "$(grep "@@@time:" $log)" ]; then
+        time=`grep "@@@time:" $log | sed -re 's/.*@@@time: //g;s/ms//g'`
+      fi
+    fi    
     echo -ne "$time\t& "
   done
   
   for p in $platforms_prpu; do
     log="prpu-$p/$t.$p.log"
-    time=`grep "@@@time:" $log | sed -re 's/.*@@@time: //g;s/ms//g'`
+    time="-1"
+    if [ -e $log ]; then
+      if [ -n "$(grep "@@@time:" $log)" ]; then
+        time=`grep "@@@time:" $log | sed -re 's/.*@@@time: //g;s/ms//g'`
+      fi
+    fi 
     echo -ne "$time\t& "
   done
 

### 2017-03-02 formalization/Prpu_Maxflow/document/root.tex

--- 2016-03-09 formalization/document/root.tex
@@
 
+\newcommand{\cormen}[1]{[Cormen~$#1$]}
+
 \begin{document}
 
-\title{Formalizing the Edmonds-Karp Algorithm}
+\title{Formalization of Push-Relabel Maxflow Algorithms}
 \author{Peter Lammich and S.~Reza Sefidgar}
@@
 \begin{abstract}
-We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
-Our formal proof closely follows a standard textbook proof, and is accessible even without being
-an expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
-We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
-Further refinement yields a verified implementation, whose execution time compares well to an unverified reference implementation in Java.
+We present a formalization of push-relabel algorithms for computing the 
+maximum flow in a network. We start with Goldberg's et al.~generic push-relabel
+algorithm, for which we show correctness and the time complexity bound of 
+$O(V^2E)$. We then derive the relabel-to-front and FIFO implementation.
+Using stepwise refinement techniques, we derive an efficient verified 
+implementation.
+
+Our formal proof of the abstract algorithms closely follows a standard textbook 
+proof, and is accessible even without being an expert 
+in Isabelle/HOL--- the interactive theorem prover used for the formalization.
 \end{abstract}
@@
 circulation-demand, as well as various scheduling and resource allocating
-problems can be reduced to it. The Ford-Fulkerson method~\cite{FF56} describes a
-class of algorithms to solve the maximum flow problem. An important instance is
-the Edmonds-Karp algorithm~\cite{EK72}, which was one of the first algorithms to
-solve the maximum flow problem in polynomial time for the general case of
-networks with real valued capacities.
+problems can be reduced to it.
 
-In this paper, we present a formal verification of the Edmonds-Karp algorithm
-and its polynomial complexity bound. The formalization is conducted entirely in
-the Isabelle/HOL proof assistant~\cite{NPW02}. Stepwise refinement
-techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our
-verification into an abstract proof of the Ford-Fulkerson method, its
-instantiation to the Edmonds-Karp algorithm, and finally an efficient
-implementation. The abstract parts of our verification closely follow the
-textbook presentation of Cormen et al.~\cite{CLRS09}. Being developed in the
-Isar~\cite{Wenzel99} proof language, our proofs are accessible even to
-non-Isabelle experts.
+The practically most efficient algorithms to solve the maximum flow problem
+are push-relabel algorithms~\cite{ChGo97}. In this entry, we present 
+a formalization of Goldberg's et al.\ generic push-relabel algorithm~\cite{GoTa88}, 
+and two instances: The relabel-to-front algorithm~\cite{CLRS09} and the 
+FIFO push-relabel algorithm~\cite{GoTa88}. 
+Using stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98}, we derive 
+efficient verified implementations. Moreover, we show that the generic 
+push-relabel algorithm has a time complexity of $O(V^2E)$.
+
+This entry re-uses and extends theory developed for our formalization of
+the Edmonds-Karp maximum flow algorithm~\cite{LaSe16,LaSe16_afp}.
 
 While there exists another formalization of the Ford-Fulkerson method in
-Mizar~\cite{Lee05}, we are, to the best of our knowledge, the first that verify a
-polynomial maximum flow algorithm, prove the polynomial complexity bound, or
-provide a verified executable implementation. Moreover, this paper is a case
-study on elegantly formalizing algorithms.
+Mizar~\cite{Lee05}, we are, to the best of our knowledge, the first that verify
+a polynomial maximum flow algorithm, prove a polynomial complexity bound, or
+provide a verified executable implementation.
+
 % generated text of all theories
@@
 \section{Conclusion}\label{sec:concl}
-  We have presented a verification of the Edmonds-Karp algorithm, using a
-  stepwise refinement approach. Starting with a proof of the Ford-Fulkerson
-  theorem, we have verified the generic Ford-Fulkerson method, specialized it to
-  the Edmonds-Karp algorithm, and proved the upper bound $O(VE)$ for the number
-  of outer loop iterations. We then conducted several refinement steps to derive
-  an efficiently executable implementation of the algorithm, including a
-  verified breadth first search algorithm to obtain shortest augmenting paths.
-  Finally, we added a verified algorithm to check whether the input is a valid
-  network, and generated executable code in SML. The runtime of our verified
-  implementation compares well to that of an unverified reference implementation
-  in Java. Our formalization has combined several techniques to achieve an
-  elegant and accessible formalization: Using the Isar proof
-  language~\cite{Wenzel99}, we were able to provide a completely rigorous but
-  still accessible proof of the Ford-Fulkerson theorem. The Isabelle Refinement
-  Framework~\cite{LaTu12,La12} and the Sepref tool~\cite{La15,La16} allowed us
-  to present the Ford-Fulkerson method on a level of abstraction that closely
-  resembles pseudocode presentations found in textbooks, and then formally link
-  this presentation to an efficient implementation. Moreover, modularity of
-  refinement allowed us to develop the breadth first search algorithm
-  independently, and later link it to the main algorithm. The BFS algorithm can
-  be reused as building block for other algorithms. The data structures are
-  re-usable, too: although we had to implement the array representation of
-  (capacity) matrices for this project, it will be added to the growing library
-  of verified imperative data structures supported by the Sepref tool, such that
-  it can be re-used for future formalizations. During this project, we have
-  learned some lessons on verified algorithm development: 
-  \begin{itemize} 
-  \item
-  It is important to keep the levels of abstraction strictly separated. For
-  example, when implementing the capacity function with arrays, one needs to
-  show that it is only applied to valid nodes. However, proving that, e.g.,
-  augmenting paths only contain valid nodes is hard at this low level. Instead,
-  one can protect the application of the capacity function by an assertion---
-  already on a high abstraction level where it can be easily discharged. On
-  refinement, this assertion is passed down, and ultimately available for the
-  implementation. Optimally, one wraps the function together with an assertion
-  of its precondition into a new constant, which is then refined independently.
-  \item Profiling has helped a lot in identifying candidates for optimization.
-  For example, based on profiling data, we decided to delay a possible
-  deforestation optimization on augmenting paths, and to first refine the
-  algorithm to operate on residual graphs directly. 
-  \item ``Efficiency bugs''
-  are as easy to introduce as for unverified software. For example, out of
-  convenience, we implemented the successor list computation by \emph{filter}.
-  Profiling then indicated a hot-spot on this function. As the order of
-  successors does not matter, we invested a bit more work to make the
-  computation tail recursive and gained a significant speed-up. Moreover, we
-  realized only lately that we had accidentally implemented and verified
-  matrices with column major ordering, which have a poor cache locality for our
-  algorithm. Changing the order resulted in another significant speed-up.
-  \end{itemize} 
-  We conclude with some statistics: The formalization consists of
-  roughly 8000 lines of proof text, where the graph theory up to the
-  Ford-Fulkerson algorithm requires 3000 lines. The abstract Edmonds-Karp
-  algorithm and its complexity analysis contribute 800 lines, and its
-  implementation (including BFS) another 1700 lines. The remaining lines are
-  contributed by the network checker and some auxiliary theories. The
-  development of the theories required roughly 3 man month, a significant amount
-  of this time going into a first, purely functional version of the
-  implementation, which was later dropped in favor of the faster imperative
-  version.
+  We have presented a verification of two push-relabel algorithms for solving
+  the maximum flow problem. Starting with a generic push-relabel algorithm,
+  we have used stepwise refinement techniques to derive the relabel-to-front
+  and FIFO push-relabel algorithms. Further refinement yields 
+  verified efficient imperative implementations of the algorithms.
 
-  \subsection{Related Work}\label{sec:related_work} 
-  We are only aware
-  of one other formalization of the Ford-Fulkerson method conducted in
-  Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication on
-  this formalization except~\cite{Lee05}, which provides a Mizar proof script
-  without any additional comments except that it ``defines and proves
-  correctness of Ford/Fulkerson's Maximum Network-Flow algorithm at the level of
-  graph manipulations''. Moreover, in Lee et al.~\cite{LeRu07}, which is about
-  graph representation in Mizar, the formalization is shortly mentioned, and it
-  is clarified that it does not provide any implementation or data structure
-  formalization. As far as we understood the Mizar proof script, it formalizes
-  an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson
-  method. Termination is only proved for integer valued capacities. Apart from
-  our own work~\cite{La14,NoLa12}, there are several other verifications of
-  graph algorithms and their implementations, using different techniques and
-  proof assistants. Noschinski~\cite{Nosch15} verifies a checker for
-  (non-)planarity certificates using a bottom-up approach. Starting at a C
-  implementation, the AutoCorres tool~\cite{Greenaway15,GAK12} generates a
-  monadic representation of the program in Isabelle. Further abstractions are
-  applied to hide low-level details like pointer manipulations and fixed size
-  integers. Finally, a verification condition generator is used to prove the
-  abstracted program correct. Note that their approach takes the opposite
-  direction than ours: While they start at a concrete version of the algorithm
-  and use abstraction steps to eliminate implementation details, we start at an
-  abstract version, and use concretization steps to introduce implementation
-  details.
-
-  Chargu\'eraud~\cite{char11} also uses a bottom-up approach to verify
-  imperative programs written in a subset of OCaml, amongst them a version of
-  Dijkstra's algorithm: A verification condition generator generates a
-  \emph{characteristic formula}, which reflects the semantics of the program in
-  the logic of the Coq proof assistant~\cite{BeCa10}.
-
-  \subsection{Future Work}
-  Future work includes the optimization of our implementation, and the
-  formalization of more advanced maximum flow algorithms, like Dinic's
-  algorithm~\cite{Di06} or push-relabel algorithms~\cite{GoTa88}. We expect both
-  formalizing the abstract theory and developing efficient implementations to be
-  challenging but realistic tasks.
 % optional bibliography

### 2017-03-02 paper/root.tex

@@
 \title{Formalization of Network Flow Algorithms}
-\subtitle{a refinement approach in Isabelle/HOL}
+\subtitle{A Refinement Approach in Isabelle/HOL}
 
@@
 Using stepwise refinement techniques, we instantiate the generic Ford-Fulkerson method to Edmonds-Karp algorithm, 
-and the generic push-relabel algorithm of Goldberg and Tarjan to both, the relabel-to-front and the FIFO push-relabel algorithm.
+and the generic push-relabel algorithm of Goldberg and Tarjan to both the relabel-to-front and the FIFO push-relabel algorithm.
 Further refinement then yields verified efficient implementations of the algorithms, which compare well to unverified reference implementations.
@@
 Computing the maximum flow of a network is an important problem in graph theory.
-Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, as well as various scheduling and resource allocating problems can be reduced to it.
+Many other problems, like maximum bipartite matching, edge disjoint paths, circulation demand, as well as various scheduling and resource allocating problems can be reduced to it.
 The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. 
-The basic idea is to repeatedly search for \emph{augmenting path} along which more flow can be moved from the source to the sink.
+The basic idea is to repeatedly search for \emph{augmenting paths} along which more flow can be moved from the source to the sink.
 An important instance is the Edmonds-Karp algorithm~\cite{EK72},
@@
 which we then refine to Edmonds-Karp algorithm and finally to an efficient implementation. 
-Similarly, the generic push-relabel algorithm is refined to both, the relabel-to-front algorithm~\cite{CLRS09} and the FIFO push-relabel algorithm~\cite{GoTa88},
+Similarly, the generic push-relabel algorithm is refined to both the relabel-to-front algorithm~\cite{CLRS09} and the FIFO push-relabel algorithm~\cite{GoTa88},
 which, in turn, are refined to efficient implementations. The efficiency of our verified implementations compare well to unverified reference implementations of 
@@
 
-In the rest of this paper, we give a short informal introduction to flow networks and the Ford-Fulkerson method (\S\ref{sec:background}).
-We then present our formalization of these material (\S\ref{sec:abs-formalization}).
-After introducing the Isabelle Refinement Framework~\cite{LaTu12,La12} (\S\ref{sec:refinement}), we report on the instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm, and on the complexity proof (\S\ref{sec:edka}).
+In the rest of this paper, we give a short informal introduction the Ford-Fulkerson method (\S\ref{sec:background})
+and present our formalization of it (\S\ref{sec:abs-formalization}).
+After introducing the Isabelle Refinement Framework~\cite{LaTu12,La12} (\S\ref{sec:refinement}), we report on the instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm, and on its complexity proof (\S\ref{sec:edka}).
 Next, we describe the correctness and complexity proof of the generic push-relabel algorithm (\S\ref{sec:gen-prpu}), and its instantiations 
 to the relabel-to-front and FIFO push-relabel algorithms (\S\ref{sec:prpu-inst}).
-We then report on the refinement of all three algorithms to efficient implementations (\S\ref{sec:executable}), and their comparison to unverified reference 
+Finally, we report on the refinement of all three algorithms to efficient implementations (\S\ref{sec:executable}), and their comparison to unverified reference 
 implementations (\S\ref{sec:benchmark}).
@@
 
-A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real-valued capacity $c(u,v)>0$.
-Moreover, there are two distinct vertices $s,t\in V$, which are called \emph{source} and \emph{sink}.
-
-A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of incoming flows equals the
+A (flow) network is a directed graph over a finite set of nodes $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real-valued capacity $c(u,v)>0$.
+Moreover, there are two distinct nodes $s,t\in V$, which are called \emph{source} and \emph{sink}.
+
+A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all nodes except $s$ and $t$, the sum of incoming flows equals the
 sum of outgoing flows.
 % 
-% \emph{Non-generation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is not smaller than the sum of flows over all outgoing edges.
+% \emph{Non-generation constraint}: For all nodes except $s$ and $t$, the sum of flows over all incoming edges is not smaller than the sum of flows over all outgoing edges.
 % We define the \emph{excess} flow $x_f(u)$ of a node $u$ to be the difference of the incoming and outgoing flow of $u$.
@@
 % The excess flow is the amount of flow that is stuck at a node, \ie which flows in but does not flow out.
-% A \emph{flow} is a \emph{preflow} that satisfies the \emph{Conservation constraint}: For all vertices except $s$ and $t$, the excess flow is zero.
+% A \emph{flow} is a \emph{preflow} that satisfies the \emph{Conservation constraint}: For all nodes except $s$ and $t$, the excess flow is zero.
 % Intuitively, only the source may generate flow, and only the sink may consume flow, and for all other nodes, the incoming flow must be equal to the outgoing flow.
@@
 
-To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) the source only has outgoing edges while the sink only has incoming edges; 2) if the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the network must be on a path from $s$ to $t$.
+To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) the source only has outgoing edges while the sink only has incoming edges; 2) if the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every node of the network must be on a path from $s$ to $t$.
 Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.
 
-An important result is the relation between flows and cuts in a network. A \emph{cut} is a partitioning of the vertices into two sets, such that one set contains 
+An important result is the relation between flows and cuts in a network. A \emph{cut} is a partitioning of the nodes into two sets, such that one set contains 
 the source and the other set contains the sink. The capacity of a cut is the sum of the capacities of all edges going from the source's side to the sink's side of the cut.
 It is easy to see that the value of any flow cannot exceed the capacity of any cut, as all flow from the source must ultimately reach the sink, and thus go through the edges of the cut.
-The Ford-Fulkerson theorem tightens this bound and states that the value of the maximum flow is equal to the capacity of the minimum cut.
+The min-cut-max-flow theorem tightens this bound and states that the value of the maximum flow is equal to the capacity of the minimum cut.
 
@@
 Each edge is labeled by the amount of flow that can be effectively passed along this edge, by either increasing or decreasing the flow on a network edge.
-Formally, the residual graph $G_f$ of a (pre)flow $f$ is the graph induced by the 
+Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
 edges with positive labels according to the following labeling function $c_f$:
@@
 
-The correctness of the Ford-Fulkerson algorithm follows from the Ford-Fulkerson theorem, which is also known as min-cut-max-flow theorem.
-It states that the following three propositions are equivalent:
+The correctness of the Ford-Fulkerson algorithm follows from the min-cut-max-flow theorem, which states that the following three propositions are equivalent:
 \begin{enumerate}
@@
 operation \isai{bind m f} nondeterministically picks a result from \isai{m} and executes \isai{f} on it. 
-The bind operation fails iff either \isai{m = fail}, or \isai{f} may fail for a result in \isai{m},
+The bind operation fails iff either \isai{m = fail}, or \isai{f} may fail for a result in \isai{m}.
 
@@
 also possible results of \isai{m'}. 
-Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed-point construction~\cite{Kr10}.
+Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
 Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
@@
 
-Correctness of a program \isai{m} with precondition \isai{P} and postcondition \isai{Q} is expressed as \isai{P ==> m <= spec r. Q r} (or, eta-contracted, just \isai{spec Q}), which
+Correctness of a program \isai{m} with precondition \isai{P} and postcondition \isai{Q} is expressed as \isai{P ==> m <= spec r. Q r} (or, eta contracted, just \isai{spec Q}), which
 means that, if \isai{P} holds, \isai{m} does not fail and all possible results of \isai{m} satisfy \isai{Q}. Note that we provide different recursion constructs
@@
 Various tools, including a verification condition generator, assist the user in conducting the refinement proofs by breaking 
-them down to statements that do not contain monad constructs any more. In many cases, these verification conditions reflect the core idea of the proof precisely.
+them down to statements that do not contain monad constructs any more. In many cases, these verification conditions precisely reflect the core idea of the proof.
 
@@
 \section{The Generic Push-Relabel Algorithm}\label{sec:gen-prpu}
-  We present our formalization of the generic push relabel algorithm of Goldberg and Tarjan~\cite{GoTa88}. 
-  Again, our formalization closely follows the textbook presentation of Cormen et al.~\cite{CLRS09}.
-
-  Push-relabel algorithms are based on a generalization of flows called \emph{preflows}: 
-  The conservation constraint is relaxed to allow more incoming flow than outgoing flow on a vertex.
-  For a vertex $u\in V$, the \emph{excess} $x_f(u)$ is the difference between incoming and outgoing flow. 
+  We formalize the generic push relabel algorithm of Goldberg and Tarjan~\cite{GoTa88},
+  closely following the textbook presentation of Cormen et al.~\cite{CLRS09}.
+
+  Push-relabel algorithms are based on \emph{preflows}, a generalization of flows
+  by relaxing the conservation constraint to allow more incoming flow than outgoing flow on a node.
+  For a node $u\in V$, the \emph{excess} $x_f(u)$ is the difference between incoming and outgoing flow. 
   For a preflow, the excess of any node but the source must not be negative.
-  A non-sink vertex with positive excess is called \emph{active}. Thus, a preflow without any active vertices is actually a flow.
-
-  The push-relabel algorithm maintains a preflow $f$ and a height-labeling $l: V \to \mathbb {N}$ of the nodes, 
+  A non-sink node with positive excess is called \emph{active}. Thus, a preflow without any active nodes is actually a flow.
+
+  The push-relabel algorithm maintains a preflow $f$ and a height labeling $l: V \to \mathbb {N}$ of the nodes, 
   with the invariant that for every edge $(u, v)$ of the residual graph we have that $l(u) \le l(v) + 1$. 
@@
   By generalizing the labeling invariant to paths, we get that for each path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$.
-  This implies that there cannot be a path from $s$ to $t$: The length of any (simple) path is less than $|V|$, however, a valid labeling 
+  This implies that there cannot be a path from $s$ to $t$: the length of any (simple) path is less than $|V|$, however, a valid labeling 
   implies a length greater than or equal to $|V|$. Upon termination of the algorithm, there will be no active nodes. 
@@
   The push relabel algorithm initializes the heights of all nodes but the source to zero. Moreover, the outgoing edges of the source are 
-  set to carry maximum flow, all other edges get zero flow. Then, the algorithm repeats the following operation as long as they apply:
+  set to carry maximum flow, all other edges get zero flow. Then, the algorithm repeats the following operations as long as they apply:
   \begin{description}
-    \item[$\textrm{push}(u,v)$] for an edge $(u,v)$ of the residual graph with $l(u) = l(v)+1$ and $u$ active: Move as much flow as possible from $u$ to $v$.
+    \item[$\textrm{push}(u,v)$] for an edge $(u,v)$ of the residual graph with $l(u) = l(v)+1$ and $u$ active: move as much flow as possible from $u$ to $v$.
     The maximum possible flow is constrained by the residual capacity of the edge, and the excess flow available at $u$.
-    \item[$\textrm{relabel}(u)$] for an active node $u$ from which no push operation is possible: Increase the height of $u$ to the minimum value such that a push
+    \item[$\textrm{relabel}(u)$] for an active node $u$ from which no push operation is possible: increase the height of $u$ to the minimum value such that a push
       becomes possible.
@@
 \end{lstlisting}  
-where \isai$(f_0,l_0)$ is the initial flow and height-labeling and \isai{pr_lts^*} denotes the reflexive transitive closure of \isai{pr_lts}. 
-Intuitively, the cost of any (partial) run \isai$p$ from the initial state to $(f,l)$ is bounded by $O(V^2E)$, and if the 
-run is complete, \ie no more operations apply (\isai$(f,l)\<notin>Domain pr_lts$), the resulting flow is maximal.
+where \isai$(f_0,l_0)$ is the initial flow and height labeling and \isai{pr_lts^*} denotes the reflexive transitive closure of \isai{pr_lts}. 
+Intuitively, the cost of any (partial) run \isai$p$ from the initial state to \isai$(f,l)$ is bounded by $O(V^2E)$, and if the 
+run is complete, \ie no more operations apply (\isai$(f,l)\<notin>Domain pr_lts$), then the resulting flow is maximal.
 
@@
 
-The complexity argument is more tricky. Instead of defining a single measure function,
+The complexity argument is more tricky. Instead of defining a measure function,
 Cormen counts the different types of operations separately, using quite involved arguments.
-For our formalization, we have adapted some Cormen's proofs, trading precision of the constant factors for simplicity of the formalization. 
+For our formalization, we have adapted Cormen's proofs, trading precision of the constant factors for simplicity of the formalization. 
 Thus, our constant of $38$ is somewhat higher than the $17$ that would result from Cormen's proof.
@@
 by inductive arguments over the runs of the algorithm.
-Moreover, the powerful Isabelle/HOL standard library, in particular the lemmas for summations
-over finite sets, were very helpful to keep the formalization effort manageable. 
+Moreover, the powerful Isabelle/HOL standard library, in particular the support for summations
+over finite sets, was very helpful to keep the formalization effort manageable. 
 Anyway, the whole complexity proof is roughly 1000 lines of Isabelle text, and took a few weeks to complete.
@@
 The generic algorithm does not specify the order in which the operations are applied. 
-Choosing the right order decreases the worst-case time complexity, and, more important, the practical efficiency of the algorithm.
-
-We consider two common instantiations of the generic algorithm: The relabel-to-front algorithm~\cite{CLRS09}, and tzhe FIFO push-relabel algorithm~\cite{GoTa88}.
+Choosing the right order decreases the worst case time complexity, and, more important, the practical runtime.
+
+We consider two common instantiations of the generic algorithm: The relabel-to-front algorithm~\cite{CLRS09}, and the FIFO push-relabel algorithm~\cite{GoTa88}.
 Both have a theoretical time complexity of $O(V^3)$, which is better than the $O(V^2E)$ bound of the generic algorithm.
-However, the more complicated structure of these algorithms prevent a convenient specification as labeled transition system, and thus 
-a complexity proof in the style of the generic algorithm.
+However, the more complicated structure of these algorithms prevents a convenient specification as labeled transition system, and thus 
+a complexity proof in the style done for the generic algorithm.
 Instead, we use the Monadic Refinement Framework to specify the algorithms and only prove correctness and 
 termination.\footnote{Which we get for free from the bound of the generic algorithm.} We leave it to future work to integrate complexity reasoning tools into 
-the Refinement Framework and prove the tighter complexity bounds for the specific implementations.
+the Refinement Framework and to prove the tighter complexity bounds for the specific implementations.
 
 A general idea of most efficient implementations is to \emph{discharge} a node by repeatedly applying push operations until the node gets inactive or needs to be relabeled.
+The implementations mainly differ in the order in which the nodes are discharged.
 
 \subsection{Relabel-to-Front Algorithm}
-Cormen et al.\ present the relabel to front algorithm.
+Cormen et al.~\cite{CLRS09} present the relabel to front algorithm.
 It discharges a node until it gets inactive, applying relabel operations as necessary. 
@@
 which uses an additional gap heuristics: If there is a value $0<k<|V|$ such that there is no node that has height $k$, 
-all nodes with $k<l(v)<|V|$ may be relabeled to $|V|+1$, preserving a valid labeling. 
+all nodes $v$ with $k<l(v)<|V|$ may be relabeled to $|V|+1$, preserving a valid labeling. 
 This heuristics prevents unnecessary pushes of flow towards the sink when the path to the sink is already blocked.
-The correctness of the heuristics follows easily from the definitions of a valid height-labeling.
+The correctness of the heuristics follows easily from the definitions of a valid height labeling.
 In total, the correctness proof of the FIFO push-relabel algorithm is about 400 lines of code and, building on the experience from relabel-to-front, it 
@@
   However, we left open several algorithmic aspects. For example, we did not specify how to find a shortest augmenting path.
-  Moreover, some of the operations require additional auxiliary data structures to be maintained. For example, 
-  we maintain the current excess of each node explicitly, instead of summing up the incoming and outgoing flows each time 
-  we need to check whether a node is active.
-  Finally, we have to specify the concrete data structures that we want to use for implementation. 
-  For example, the residual graph is modeled as a function from edges to capacities on the abstract level. In the implementation, we use a two-dimensional array.
-
+  Moreover, efficient implementation of some operations requires additional auxiliary data to be maintained. 
+  Finally, we have to specify the concrete data structures to be used for the implementation. 
   In this section, we outline the further refinement steps that were necessary to obtain an efficient implementation.
@@
   \subsection{Algorithms}
-  The most complex problem for which we specified no algorithm is finding a shortest augmenting path.
+  The most complex problem for which we specified no algorithm yet is finding a shortest augmenting path.
   We use breadth first search (BFS) for this. We adapted a BFS formalization from the examples collection of the Isabelle Refinement Framework to
   our needs and added an efficient imperative implementation. The resulting algorithm 
-  is independent of flow networks, and can be re-used to find shortest path in directed graphs also in other applications.
-
-  Minor problems for which we had to develop algorithms are augmentation of the flow along the shortest path, and the relabel and discharge operations.
-  While it is straightforward to find the height to relabel a node with by iterating over the outgoing edges, the discharge operation is a bit more tricky:
-  Abstractly, we nondeterministically select a new edge to push, until we find no more edges.
+  is independent of flow networks and can be reused for other applications.
+
+  Simpler problems for which we had to develop algorithms are augmentation of the flow along the shortest path, and the relabel and discharge operations.
+  The abstract specification of augmentation (\cf Section~\ref{sec:background}) first maps the augmenting path to an augmenting flow, which is then added 
+  to the current flow. For the implementation, we skip the intermediate augmenting flow, and modify the current flow directly.
+  The relabel operation has to determine the new height for the relabeled node. It's straightforward to implement this by iterating over the 
+  successor nodes in the residual graph. Refinement of the discharge operation is more tricky: Abstractly, we nondeterministically select a new edge to push, 
+  until we find no more edges.
   In the concrete implementation, however, we fold over the list of successor nodes.
   This structural difference between abstract and concrete algorithm cannot be handled by the heuristics of the Refinement Framework's verification condition generator.
-  Thus, we had to manually prove some refinement rules that are specifically tailored to our case.
+  Thus, we had to manually prove some specialized refinement rules.
 
@@
   An initial version of our Edmonds-Karp algorithm worked on flows, and switching to residual graphs yielded a speedup of roughly factor 2.
-
-  Another important optimization is to pre-compute an adjacency map of the network:
+  Technically, we relate the concrete operations on residual graphs to the abstract operations on flows by the refinement 
+  relation $\{(c_f,f) \mid \text{$f$ is a flow}\}$.
+
+  Moreover, for the push-relabel algorithms, we introduce a map from nodes to their current excess value: 
+  A single lookup in this map is sufficient to compute the flow for a push operation --- instead of summing up all 
+  the incoming and outgoing flows. We combine the introduction of the excess map with the transition from preflows to residual 
+  graphs, using the refinement relation $\{((x_f,c_f),f) \mid \text{$f$ is a flow}\}$.
+
+  Another important optimization is to precompute an adjacency map of the network:
   To find the successors of a node in the residual graph, it is enough to search among the adjacent nodes in the network. 
@@
 
-  Moreover, for the push-relabel algorithms, we introduce a map from nodes to their current excess value: 
-  A single lookup in this map is sufficient to compute the flow for a push operation --- instead of summing up all 
-  the incoming and outgoing flows.
-
-  For the gap-heuristics, we introduce a height frequency map, which maps each height value to the number of nodes labeled with this value.
+  For the gap heuristics, we introduce a height frequency map, which maps each height value to the number of nodes labeled with this value.
   A single lookup in this map suffices to detect when we relabel the last node of a specific height such that the gap heuristics applies.
@@
   An augmenting path is represented by a list of edges, \ie a list of pairs of nodes. 
-  The height frequency map and the maps from nodes to heights and excess values are implemented as arrays.
+  The height frequency map and the maps from nodes to height labels and excess values are implemented as arrays.
   The queue of the FIFO push-relabel algorithm and the list of the relabel-to-front algorithm are implemented by two stacks, allowing for amortized constant time operations.
@@
   This calls for a deforestation optimization to get rid of the intermediate list, and iterate only two times over the predecessor map directly.
-  Also the FIFO queue could be implemented by a ring buffer based on a fixed size array, yielding true constant-time operations with less overhead.
+  Also the FIFO queue could be implemented by a ring buffer based on a fixed size array, yielding true constant time operations with less overhead.
   Fortunately, our profiling data shows that these operations do not significantly contribute to the runtime of our implementations, such that 
@@
 %   The queue of the FIFO push-relabel algorithm is implemented by two stacks, yielding amortized constant time enqueue and dequeue operations.
-%   Similarly, the list of the relabel-to-front algorithm is implemented by two lists, \emph{left} and \emph{right}, that contain the nodes left and right of the current position,
+%   Similarly, the list of the relabel-to-front algorithm is implemented by two lists, \emph{left} and \emph{right} that contain the nodes left and right of the current position,
 %   the left list being stored in reverse order. Advancing the position is implemented in constant time by popping the first element of the right list, and pushing 
@@
 %   After our profiling indicated a hot spot on accessing the capacity matrices of the network and the flow, we switched to 
-%   an algorithm that operates on a representation of the residual graph directly. This resulted in a speed-up of roughly a factor of two.
+%   an algorithm that operates on a representation of the residual graph directly. This resulted in a speedup of roughly a factor of two.
 %   As the residual graph uniquely determines the flow (and vice versa), it is straightforward to phrase the
@@
 %   This calls for a deforestation optimization to get rid of the intermediate list, and iterate only two times over the predecessor map directly.
-%   Also the FIFO queue could be implemented by a ring buffer based on a fixed size array, yielding true constant-time operations with less overhead.
+%   Also the FIFO queue could be implemented by a ring buffer based on a fixed size array, yielding true constant time operations with less overhead.
 %   Fortunately, our profiling data shows that these operations do not significantly contribute to the runtime of our implementations, such that 
@@
   the first to leave the augmenting path idea in research on finding the
-  maximum-flow in networks. He came up with an $O(V^3)$ maximum-flow algorithm
+  maximum flow in networks. He came up with an $O(V^3)$ maximum flow algorithm
   by relaxing some of the invariants in Dinitz's algorithm~\cite{Di06}. Years
@@
   %Karzanov's algorithm for constructing a feasible flow at the end of each
-  %phase. He proved that balancing the flow at some vertex is equivalent to
+  %phase. He proved that balancing the flow at some node is equivalent to
   %pushing from it in future when it will appear in some layared-network at a
@@
 
-  The Push-relabel algorithms maintain a preflow as well as a height-labeling
+  The Push-relabel algorithms maintain a preflow as well as a height labeling
   during the execution. A preflow in a network with source $s$ and sink $t$ is a
@@
     related definition, like flow's value or flow's corresponding residual
-    graph can be generalized for preflows.}. That is, for all vertices except
+    graph can be generalized for preflows.}. That is, for all nodes except
   the source $s$ the sum of flows over all incoming edges is greater equal than
-  the sum of flows over all outgoing edges. The excess flow $e(v)$ of a vertex
+  the sum of flows over all outgoing edges. The excess flow $e(v)$ of a node
   $v$ in the network is the amount by which in flow exceeds the out flow, and a
-  $v$ is overflowing if $e(v) > 0$. Moreover, a height-labeling in a $s$-$t$
+  $v$ is overflowing if $e(v) > 0$. Moreover, a height labeling in a $s$-$t$
   network is a funtion $l: V \mapsto \mathbb{N}$ where $l(s) =
@@
   We summerize Cormen's intutition behind Push-relabel algorithms to clarify why
-  such a labeling function is required. Consider vertices of the network to be
+  such a labeling function is required. Consider nodes of the network to be
   liquid containers and residual network's edges to be pipes that are connecting
@@
   lying on a platform which elevates it to higher altitudes as the algorithm
-  makes progress. The height-labeling function keeps track of the heights of
+  makes progress. The height labeling function keeps track of the heights of
   each container, and the flow can only go from higher altitude containers to
@@
   may happen that a non-empty container(resembeling the idea of an overflowing
-  vertex) is only attached to pipes with left-over capacities that are
+  node) is only attached to pipes with leftover capacities that are
   connecting it to a containers at higher altitude. In such a case, the
@@
     increases $e(v)$ at the same time.}. However, it can be applied only if $u$
-  is an overflowing vertex and $l(u) = l(v) + 1$. This restriction is due to the
-  fact that no residual edges exists between two vertices with hight-labeling
+  is an overflowing node and $l(u) = l(v) + 1$. This restriction is due to the
+  fact that no residual edges exists between two nodes with height labeling
   difference of more that 1. Hence, we would gain nothing by allowing such a
@@
 the edge $(u, v)$ is removed from the residual graph after the push operation,
-i.e. $e(u) \ge c_f(u, v)$. Otherwise, the push is called non-saturating, and the
+\ie $e(u) \ge c_f(u, v)$. Otherwise, the push is called non-saturating, and the
 edge $(u, v)$ remains in the residual graph after the push operation.
@@
   \newcommand{\relabel}{{\mathbin\uparrow_L}}%
-Next, the relabel operation $\relabel$ which increases the height-labeling of an
-overflowing vertex $u$, when for all edges $(u, v)$ that are going out of $u$ in
+Next, the relabel operation $\relabel$ which increases the height labeling of an
+overflowing node $u$, when for all edges $(u, v)$ that are going out of $u$ in
 the residual graph we have $l(u) \le l(v)$. The overflowing precondition is
@@
 operation $\push$ flow can only be pushed downhill. So, with relabel operation
-we enable an overflowing vertex to push its excess flow into one of its
+we enable an overflowing node to push its excess flow into one of its
 neighbours. By definition, neither the source $s$ nor the sink $t$ can change
-their height-labeling, so, the relabel operation does not apply to them.
+their height labeling, so, the relabel operation does not apply to them.
 \[ (l \relabel u) (u') = 
@@
 The generic Push-relabel algorithm is then defined by iterative application of the aformentioned basic
-operations. The algorithm starts by setting up an initial height-lableling and
+operations. The algorithm starts by setting up an initial height labeling and
 preflow function. \isai{init-labeling} operation outputs a function $l$ such that
@@
 relabel operation is applicable, and according to correctness theorem of the
-algorithm the resulting preflow upon termination is indeed a maximum-flow of the network.
+algorithm the resulting preflow upon termination is indeed a maximum flow of the network.
 
@@
 We start with the first step, and try to prove computation of a maximum flow
-upon termination of the algorithm. Maintaining helight-labeling function as well
+upon termination of the algorithm. Maintaining height labeling function as well
 as a preflow during the execution of the algorithm is crucial for its
@@
 
-We proceed by proving that generic Push-relabel maintains a height-labeling
+We proceed by proving that generic Push-relabel maintains a height labeling
 function. Using the definition of relabel operation $\relabel$, we observe that
-the height-labeling $l(u)$ of any vertex $u \in V$ never decreases during the
+the height labeling $l(u)$ of any node $u \in V$ never decreases during the
 execution of the algorithm, and $(l \relabel u)$ increases $u$'s height by at
-least one. These two basic results prove that $l$ remains a height-labeling
+least one. These two basic results prove that $l$ remains a height labeling
 whenever it is modified by applying the relabel operation. Note taht, we also
@@
 also it can add $(v, u)$ to it. However, both of these modifications still
-maintain the height-labeling function because of the preconditions that push
+maintain the height labeling function because of the preconditions that push
 operator has.
@@
 maximum flow upon its termination. We use the fact about maintanance of
-height-labeling function to build an isabelle context \isa{Labeling}. We embed
+height labeling function to build an isabelle context \isa{Labeling}. We embed
 the rest of the correctness theorems inside this context, as it simplifies the
 proofs and gives us a cleaner formalization. That is, as we can access the
-height-labeling assumptions whenever they are required in some proof step.
+height labeling assumptions whenever they are required in some proof step.
 
 As an intermediate step, we should fist prove that for any preflow $f$ and
-height-labeling $l$, the residual graph corresponding to $f$ does not contain a
+height labeling $l$, the residual graph corresponding to $f$ does not contain a
 path from the source $s$ to the sink $t$. In order to see this, let's by
@@
 $l(t) = 0$ and $p$ is assumed to be a simple path in the residual graph.
-However, this contradicts the assumption that $l$ is a height-labeling function,
+However, this contradicts the assumption that $l$ is a height labeling function,
 and completes the proof.
@@
 to see that \isai{init-preflow} and \isa{init-labeling} produce a valid preflow
-as well as a valid height-labeling function. We have also shown the maintanance
+as well as a valid height labeling function. We have also shown the maintanance
 of these properties earlier. When the algorithm terminates, the excess value
-$e(v)$ of each vertex $v \in V$ is equal to $0$. So, the preflow that was
+$e(v)$ of each node $v \in V$ is equal to $0$. So, the preflow that was
 maintained during the execution becomes a valid flow. Given the fact that the
-height-labeling assumptions are also maintained and using the latest auxillary
+height labeling assumptions are also maintained and using the latest auxillary
 lemma, we can conclude that there is no path from source to sink in the residual
@@
   We have benchmarked our verified algorithms against unverified reference implementations.
-  We extract SML code for our verified Edmonds-Karp and FIFO push-relabel algorithms, and compile with MLton 20100608~\cite{MLton}.
+  We extract SML code for our verified Edmonds-Karp and FIFO push-relabel algorithms, and compile it with MLton~\cite{MLton} version 20100608.
+  We have not included the relabel-to-front algorithm, as it was three orders of magnitude slower in preliminary tests.
 
@@
 
-  The results are displayed in Figure~\ref{fig:benchmark}, using a double-logarithmic scale.
-  Note that we only measure the time required to run the algorithms themselves, excluding the time required for parsing the input or the network checker.
+  The results are displayed in Figure~\ref{fig:benchmark}, using a log-log scale.
+  Note that we only measure the time required to run the algorithms themselves, excluding the time required for parsing the input and running the network checker.
 
@@
   The highly optimized C implementation is, depending on the test case, one to two orders of magnitudes faster. 
-  On the ``Goldberg Bad Case'' networks, which are designed as worst-case problems for push-relabel algorithms, 
-  even Edmonds-Karp algorithm is competitive to the simple FIFO push-relabel algorithms, while the highly optimized one shows no weakness on these networks.
+  On the ``goldberg bad case'' networks, which are designed as worst case problems for push-relabel algorithms, 
+  even Edmonds-Karp algorithm is competitive to the simple FIFO push-relabel algorithms, while the highly optimized one shows no weaknesses on these networks.
+
+  The performance gap between our verified SML code and the C++ code that implements essentially the same algorithm is not surprising: We compare a 
+  memory safe functional language with a slightly aged compiler against an unsafe language with a very recent and actively developed compiler.
 
@@
 \section{Conclusion}\label{sec:concl}
-  We have presented a mechanical verification of the correctness and time complexity of maximum flow algorithms based on both, 
+  We have presented a mechanical verification of the correctness and time complexity of maximum flow algorithms based on both 
   the augmenting path and the push-relabel approaches, as well as the Ford-Fulkerson (min-cut-max-flow) theorem, which underlies all these algorithms.
@@
   For the push-relabel approaches, we start with a verification of the generic push-relabel algorithm of Goldberg and Tarjan, show its time complexity of $O(V^2E)$, and
-  instantiate it to both, the relabel-to-front and the FIFO push-relabel algorithm.
+  instantiate it to both the relabel-to-front and the FIFO push-relabel algorithm.
   We then conduct several refinement steps to derive efficiently executable implementations of all three algorithm, from which we extract SML code.
@@
   which can be reused as building block for other algorithms.
-  The data structures are re-usable, too: although we had to implement the array representation of (capacity) matrices for this project, it will be added to the growing library of verified imperative data structures
-  supported by the Sepref tool, such that it can be re-used for future formalizations.
+  The data structures are reusable, too: although we had to implement the array representation of (capacity) matrices for this project, it is now part of the 
+  growing library of verified imperative data structures supported by the Sepref tool, such that it can be reused for future formalizations.
 
@@
     For example, when implementing the capacity function with arrays, one needs to show that it is only applied to valid nodes.
-    However, proving that, e.g., augmenting paths only contain valid nodes is hard at this low level. 
+    However, proving that, \eg augmenting paths only contain valid nodes is hard at this low level. 
     Instead, one can protect the application of the capacity function by an assertion --- already on a high abstraction level where it can be easily discharged. 
@@
   \item ``Efficiency bugs'' are as easy to introduce as for unverified software. For example, out of convenience, we implemented the successor list computation by
-    \emph{filter}. Profiling then indicated a hot-spot on this function. As the order of successors does not matter, we invested a bit more work to make the computation tail 
-    recursive and gained a significant speed-up. Moreover, we realized only lately that we had accidentally implemented and verified
-    matrices with column major ordering, which have a poor cache locality for our algorithm. Changing the order resulted in another significant speed-up.
+    \emph{filter}. Profiling then indicated a hot spot on this function. As the order of successors does not matter, we invested a bit more work to make the computation tail 
+    recursive and gained a significant speedup. Moreover, we realized only lately that we had accidentally implemented and verified
+    matrices with column major ordering, which have a poor cache locality for our algorithm. Changing the order resulted in another significant speedup.
   \end{itemize}
@@
   We are only aware of one other formalization of the Ford-Fulkerson method conducted in Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication
-  on this formalization except~\cite{Lee05}, which provides a Mizar proof script without any additional comments except that it ``defines and proves correctness of Ford/Fulkerson's Maximum Network-Flow algorithm at the level of graph manipulations''.
+  on this formalization except~\cite{Lee05}, which provides a Mizar proof script without any additional comments except that it ``defines and proves correctness of Ford/Fulkerson's Maximum Network Flow algorithm at the level of graph manipulations''.
 Moreover, in Lee et al.~\cite{LeRu07}, which is about graph representation in Mizar, the formalization is shortly mentioned, and it is clarified that it does not provide any implementation or data structure formalization.
@@
   \subsection{Future Work}
-  While we implemented a simple version of the push-relabel algorithm, there are more advanced implementations, like the one of Cherkassky et el.~\cite{ChGo97} that 
+  While we implemented a simple version of the push-relabel algorithm, there are more advanced implementations, like the one of Cherkassky and Goldberg~\cite{ChGo97} that 
   we use in the benchmarks. The same authors also provide an implementation based on the highest label selection rule, which is reported to be even more efficient 

### 2017-03-02 formalization/evaluation/bench_exec.sh
Improved benchmark scripts

@@ -2,8 +2,8 @@
 
 . bench_cmdline.inc
 
-for t in $tests; do
-  echo "*********************"
+for t in $fofu_tests; do
+  echo "** FOFU *************"
   echo "$t"
   echo "*********************"
 
@@ -15,6 +15,14 @@ for t in $tests; do
     ./run $name | tee "$t.$platform.log" | grep "^@@@"
     cd ..
   done
+done
+
+for t in $prpu_tests; do
+  echo "** PRPU *************"
+  echo "$t"
+  echo "*********************"
+
+  name="../data/samples/g-$t.txt"
 
   for platform in $platforms_prpu; do
     echo "##### prpu-$platform"

### 2017-03-02 formalization/evaluation/bench_print.sh

@@ -2,27 +2,40 @@
 
 . bench_cmdline.inc
 
-tabs -9
-echo -ne "Name \t& "
-for p in $platforms_fofu; do echo -ne "f-$p\t& "; done
-for p in $platforms_prpu; do echo -ne "p-$p\t& "; done
+
+test -t 1 && tabs 18,+10,+10,+10,+10,+10,+10,+10,+10,+10,+10,+10,+10,+10,+10,+10,+10,+10,+10,+10,+10,+10
+
+echo -ne "Name \t, "
+for p in $platforms_fofu; do echo -ne "f-$p\t, "; done
+for p in $platforms_prpu; do echo -ne "p-$p\t, "; done
 
 echo ""
 
+tests=$(for t in $fofu_tests $prpu_tests; do echo $t; done | awk '!a[$0]++')
+
 for t in $tests; do
-  echo -ne "$t\t& "
+  echo -ne "$t\t, "
   for p in $platforms_fofu; do
     log="fofu-$p/$t.$p.log"
-    time=`grep "@@@time:" $log | sed -re 's/.*@@@time: //g;s/ms//g'`
-    echo -ne "$time\t& "
+    
+    if test -f $log; then
+      time=`grep "@@@time:" $log | sed -re 's/.*@@@time: //g;s/ms//g'`
+    else
+      time="---"
+    fi
+    echo -ne "$time\t, "
   done
   
   for p in $platforms_prpu; do
     log="prpu-$p/$t.$p.log"
-    time=`grep "@@@time:" $log | sed -re 's/.*@@@time: //g;s/ms//g'`
-    echo -ne "$time\t& "
+    if test -f $log; then
+      time=`grep "@@@time:" $log | sed -re 's/.*@@@time: //g;s/ms//g'`
+    else
+      time="---"
+    fi
+    echo -ne "$time\t, "
   done
 
-  echo "\\\\"
+  echo ""
 done
 

### 2017-03-02 paper/root.tex

@@
 		  \addplot+[mark=square] table {
-		    1000 .024   
-		    1500 .098  
-		    2000 .211  
-		    2500 .693  
-		    3000 2.014 
-		    3500 3.153 
-		    4000 3.679 
-		    4500 8.364 
-		    5000 10.328 
-		    5500 11.749    
+        1000	0.024
+        1500	0.099
+        2000	0.211
+        2500	1.66
+        3000	1.616
+        3500	2.966
+        4000	3.758
+        4500	8.167
+        5000	10.222
+        5500	12.024
 		  }; %Sparse, Edka, SML
@@
 		  \addplot+[dashed, every mark/.append style={solid, fill=gray},mark=*] table {
-		    1000 .024   
-		    1500 .054   
-		    2000 .103  
-		    2500 .399  
-		    3000 .719  
-		    3500 1.362 
-		    4000 2.036 
-		    4500 3.334 
-		    5000 4.785 
-		    5500 6.079     
+        1000	0.024
+        1500	0.05
+        2000	0.241
+        2500	0.795
+        3000	0.689
+        3500	1.322
+        4000	2.022
+        4500	3.345
+        5000	4.675
+        5500	6.411
 		  }; %Sparse, Edka, Java
@@
 		  \addplot+[mark=square*] table {
-		    1000 .005   
-		    1500 .013  
-		    2000 .024  
-		    2500 .037  
-		    3000 .066 
-		    3500 .077 
-		    4000 .096 
-		    4500 .146 
-		    5000 .194 
-		    5500 .192    
+        1000	0.005
+        1500	0.018
+        2000	0.038
+        2500	0.07
+        3000	0.063
+        3500	0.074
+        4000	0.098
+        4500	0.148
+        5000	0.196
+        5500	0.191
 		  }; %Sparse, Prpu, SML 
@@
 		  \addplot+[mark=square] table {
-		    1000	2.794
-		    1050	2.9
-		    1100	4.405
-		    1150	4.435
-		    1200	5.232
-		    1250	6.029
-		    1300	7.699
-		    1350	8.91
-		    1400	7.225
-		    1450	10.042 
+        1000	2.777
+        1050	2.875
+        1100	4.43
+        1150	4.466
+        1200	5.2
+        1250	6.025
+        1300	7.654
+        1350	9.312
+        1400	7.462
+        1450	10.279
 		  }; %Dense, Edka, SML
@@
 		  \addplot+[dashed, every mark/.append style={solid, fill=gray},mark=*] table {
-		    1000	3.982
-		    1050	4.463
-		    1100	5.278
-		    1150	6.27
-		    1200	6.939
-		    1250	7.953
-		    1300	9.392
-		    1350	10.317
-		    1400	14.71
-		    1450	13.11  
+        1000	3.892
+        1050	4.773
+        1100	5.262
+        1150	6.02
+        1200	6.974
+        1250	7.947
+        1300	9.511
+        1350	10.754
+        1400	11.815
+        1450	13.034 
 		  }; %Dense, Edka, Java
@@
 		  \addplot+[mark=square*] table {
-		    1000	0.074
-		    1050	0.059
+        1000	0.058
+        1050	0.06
         1100	0.068
         1150	0.075
-		    1200	0.102
+        1200	0.103
         1250	0.085
-		    1300	0.096
-		    1350	0.103
-		    1400	0.088
+        1300	0.097
+        1350	0.104
+        1400	0.089
         1450	0.114
@@
     %start data for plotting 
-
 		  \addplot+[mark=square] table {
-        3000	0.356
-        4500	0.805
-        6000	1.539
-        7500	2.43
-        9000	3.28
-        10500	4.639
-        12000	5.61
-        13500	7.198
-        15000	8.904
-        16500	11.302
+        3000	0.36
+        4500	0.813
+        6000	1.565
+        7500	2.435
+        9000	3.313
+        10500	4.69
+        12000	5.693
+        13500	9.676
+        15000	9.284
+        16500	11.31
 		  }; %Wash, Edka, SML
@@
       \addplot+[mark=square*] table {
-        3000	1.185
-        4500	3.524
-        6000	6.457
-        7500	11.509
-        9000	17.529
-        10500	26.185
-        12000	36.828
-        13500	48.614
-        15000	86.449
-        16500	89.661
+        3000	1.187
+        4500	3.654
+        6000	7.491
+        7500	11.455
+        9000	18.457
+        10500	27.253
+        12000	38.922
+        13500	51.136
+        15000	86.261
+        16500	84.913
       }; %WASH, Prpu, SML
@@
       \addplot+[mark=o] table {
-        3000	0.906
-        4500	2.708
-        6000	5.083
-        7500	8.792
-        9000	12.951
-        10500	19.075
-        12000	27.466
-        13500	35.964
-        15000	61.839
-        16500	67.039
+        3000	0.183
+        4500	0.521
+        6000	0.974
+        7500	1.73
+        9000	2.579
+        10500	3.819
+        12000	5.562
+        13500	7.184
+        15000	12.398
+        16500	12.58
       }; %WASH, Prpu, C++
@@
       \addplot+[mark=otimes*] table {
-        3000	0.006
+        3000	0.008
         4500	0.015
-        6000	0.029
-        7500	0.042
-        9000	0.063
-        10500	0.084
-        12000	0.115
-        13500	0.144
-        15000	0.183
-        16500	0.223
+        6000	0.027
+        7500	0.043
+        9000	0.064
+        10500	0.095
+        12000	0.124
+        13500	0.16
+        15000	0.2
+        16500	0.259
       }; %WASH, Prpu, C
@@
       \addplot+[mark=square*] table {
-        1000	0.141
-        1050	0.202
-        1100	0.157
-        1150	0.273
-        1200	0.24
-        1250	0.408
-        1300	0.465
-        1350	1.416
-        1400	0.657
-        1450	0.806
+        1000	0.353
+        1050	0.213
+        1100	0.247
+        1150	0.301
+        1200	0.672
+        1250	0.703
+        1300	0.273
+        1350	0.71
+        1400	0.277
+        1450	1.141
       }; %ACNET, Prpu, SML
@@
       \addplot+[mark=o] table {
-        1000	0.121
-        1050	0.17
-        1100	0.123
-        1150	0.216
-        1200	0.197
-        1250	0.237
-        1300	0.298
-        1350	0.621
-        1400	0.399
-        1450	0.486
+        1000	0.046
+        1050	0.031
+        1100	0.034
+        1150	0.039
+        1200	0.073
+        1250	0.065
+        1300	0.038
+        1350	0.071
+        1400	0.036
+        1450	0.104
       }; %ACNET, Prpu, C++
@@
       \addplot+[mark=otimes*] table {
-        1000	0.056
-        1050	0.09
-        1100	0.069
-        1150	0.106
-        1200	0.09
-        1250	0.152
-        1300	0.163
-        1350	0.176
-        1400	0.223
-        1450	0.209
+        1000	0.092
+        1050	0.122
+        1100	0.116
+        1150	0.143
+        1200	0.179
+        1250	0.108
+        1300	0.202
+        1350	0.184
+        1400	0.146
+        1450	0.179
       }; %ACNET, Prpu, C
@@
       \addplot+[mark=square*] table {
-        16000	3.602
-        17600	4.223
-        19200	4.839
-        20800	8.201
-        22400	8.273
-        24000	13.426
-        25600	10.089
-        27200	15.889
-        28800	16.431
-        30400	21.437
+        16000	3.334
+        17600	4.538
+        19200	5.844
+        20800	6.391
+        22400	8.308
+        24000	8.242
+        25600	18.534
+        27200	12.174
+        28800	16.653
+        30400	19.968
       }; %RPFGEN, Prpu, SML
@@
       \addplot+[mark=o] table {
-        16000	4.51
-        17600	4.804
-        19200	5.899
-        20800	8.052
-        22400	8.448
-        24000	11.555
-        25600	8.878
-        27200	13.999
-        28800	12.99
-        30400	13.791
+        16000	0.775
+        17600	0.966
+        19200	1.496
+        20800	1.29
+        22400	1.598
+        24000	1.525
+        25600	2.145
+        27200	2.035
+        28800	2.478
+        30400	2.745
       }; %RPFGEN, Prpu, C++
@@
         16000	0.009
-        17600	0.012
-        19200	0.015
-        20800	0.021
-        22400	0.021
-        24000	0.019
-        25600	0.023
-        27200	0.028
-        28800	0.026
-        30400	0.026
+        17600	0.011
+        19200	0.013
+        20800	0.012
+        22400	0.017
+        24000	0.017
+        25600	0.021
+        27200	0.024
+        28800	0.024
+        30400	0.022
       }; %RPFGEN, Prpu, C  

### 2017-03-02 paper/root.tex
fixed code and listing indentations
Fixed the diagrams

--- 2017-03-02 paper/root.tex
@@
 \lstset{autogobble=true}
+\lstset{xleftmargin=.15in, xrightmargin=.15in}
+
+%change both above and following margins to have consistent indentation
+\newenvironment{lstquote}%
+  {\list{}{\leftmargin=0.15in\rightmargin=0.15in}\item[]}%
+  {\endlist}
 
@@
    \expandafter\newcommand\csname snippet--#1\endcsname{%
-     \begin{quote}
+     \begin{lstquote} 
      \begin{isabelle}\footnotesize
@@
      \end{isabelle}
-     \end{quote}}}
+     \end{lstquote}}}
 \newcommand{\Snippet}[1]{\ifcsname snippet--#1\endcsname\csname snippet--#1\endcsname\else\PackageError{}{No snippet '#1' defined.}{}\fi}
@@
     % legend information
-      \addlegendimage{fill=gray, mark=square}
       \addlegendimage{fill=gray, mark=square*}
-      \addlegendimage{dashed, every mark/.append style={solid, fill=gray}, mark=*}
-      \addlegendimage{fill=gray, mark=o}
-      \addlegendimage{fill=gray, mark=otimes*}
+      \addlegendimage{fill=gray, mark=*}
+      \addlegendimage{mark=square}
+      \addlegendimage{every mark/.append style={solid, fill=gray}, mark=o}
+      \addlegendimage{dashed, every mark/.append style={solid, fill=gray}, mark=otimes*}
       \addlegendentry{Edmonds-Karp, SML}
@@
 		%start data for plotting
+		  \addplot+[mark=square*] table {
+        1000	0.024
+        1500	0.099
+        2000	0.211
+        2500	1.66
+        3000	1.616
+        3500	2.966
+        4000	3.758
+        4500	8.167
+        5000	10.222
+        5500	12.024
+		  }; %Sparse, Edka, SML
+
 		  \addplot+[mark=square] table {
-		    1000 .024   
-		    1500 .098  
-		    2000 .211  
-		    2500 .693  
-		    3000 2.014 
-		    3500 3.153 
-		    4000 3.679 
-		    4500 8.364 
-		    5000 10.328 
-		    5500 11.749    
-		  }; %Sparse, Edka, SML
-
-		  \addplot+[dashed, every mark/.append style={solid, fill=gray},mark=*] table {
-		    1000 .024   
-		    1500 .054   
-		    2000 .103  
-		    2500 .399  
-		    3000 .719  
-		    3500 1.362 
-		    4000 2.036 
-		    4500 3.334 
-		    5000 4.785 
-		    5500 6.079     
+        1000	0.024
+        1500	0.05
+        2000	0.241
+        2500	0.795
+        3000	0.689
+        3500	1.322
+        4000	2.022
+        4500	3.345
+        5000	4.675
+        5500	6.411
 		  }; %Sparse, Edka, Java
 
-		  \addplot+[mark=square*] table {
-		    1000 .005   
-		    1500 .013  
-		    2000 .024  
-		    2500 .037  
-		    3000 .066 
-		    3500 .077 
-		    4000 .096 
-		    4500 .146 
-		    5000 .194 
-		    5500 .192    
+		  \addplot+[mark=*] table {
+        1000	0.005
+        1500	0.018
+        2000	0.038
+        2500	0.07
+        3000	0.063
+        3500	0.074
+        4000	0.098
+        4500	0.148
+        5000	0.196
+        5500	0.191
 		  }; %Sparse, Prpu, SML 
@@
 		%start data for plotting 
+		  \addplot+[mark=square*] table {
+        1000	2.777
+        1050	2.875
+        1100	4.43
+        1150	4.466
+        1200	5.2
+        1250	6.025
+        1300	7.654
+        1350	9.312
+        1400	7.462
+        1450	10.279
+		  }; %Dense, Edka, SML
+
 		  \addplot+[mark=square] table {
-		    1000	2.794
-		    1050	2.9
-		    1100	4.405
-		    1150	4.435
-		    1200	5.232
-		    1250	6.029
-		    1300	7.699
-		    1350	8.91
-		    1400	7.225
-		    1450	10.042 
-		  }; %Dense, Edka, SML
-
-		  \addplot+[dashed, every mark/.append style={solid, fill=gray},mark=*] table {
-		    1000	3.982
-		    1050	4.463
-		    1100	5.278
-		    1150	6.27
-		    1200	6.939
-		    1250	7.953
-		    1300	9.392
-		    1350	10.317
-		    1400	14.71
-		    1450	13.11  
+        1000	3.892
+        1050	4.773
+        1100	5.262
+        1150	6.02
+        1200	6.974
+        1250	7.947
+        1300	9.511
+        1350	10.754
+        1400	11.815
+        1450	13.034 
 		  }; %Dense, Edka, Java
 
-		  \addplot+[mark=square*] table {
-		    1000	0.074
-		    1050	0.059
+		  \addplot+[mark=*] table {
+        1000	0.058
+        1050	0.06
         1100	0.068
         1150	0.075
-		    1200	0.102
+        1200	0.103
         1250	0.085
-		    1300	0.096
-		    1350	0.103
-		    1400	0.088
+        1300	0.097
+        1350	0.104
+        1400	0.089
         1450	0.114
@@
     %start data for plotting 
-
-		  \addplot+[mark=square] table {
-        3000	0.356
-        4500	0.805
-        6000	1.539
-        7500	2.43
-        9000	3.28
-        10500	4.639
-        12000	5.61
-        13500	7.198
-        15000	8.904
-        16500	11.302
+		  \addplot+[mark=square*] table {
+        3000	0.36
+        4500	0.813
+        6000	1.565
+        7500	2.435
+        9000	3.313
+        10500	4.69
+        12000	5.693
+        13500	9.676
+        15000	9.284
+        16500	11.31
 		  }; %Wash, Edka, SML
 
-      \addplot+[mark=square*] table {
-        3000	1.185
-        4500	3.524
-        6000	6.457
-        7500	11.509
-        9000	17.529
-        10500	26.185
-        12000	36.828
-        13500	48.614
-        15000	86.449
-        16500	89.661
+      \addplot+[mark=*] table {
+        3000	1.187
+        4500	3.654
+        6000	7.491
+        7500	11.455
+        9000	18.457
+        10500	27.253
+        12000	38.922
+        13500	51.136
+        15000	86.261
+        16500	84.913
       }; %WASH, Prpu, SML
 
-      \addplot+[mark=o] table {
-        3000	0.906
-        4500	2.708
-        6000	5.083
-        7500	8.792
-        9000	12.951
-        10500	19.075
-        12000	27.466
-        13500	35.964
-        15000	61.839
-        16500	67.039
+      \addplot+[every mark/.append style={solid, fill=gray}, mark=o] table {
+        3000	0.183
+        4500	0.521
+        6000	0.974
+        7500	1.73
+        9000	2.579
+        10500	3.819
+        12000	5.562
+        13500	7.184
+        15000	12.398
+        16500	12.58
       }; %WASH, Prpu, C++
 
-      \addplot+[mark=otimes*] table {
-        3000	0.006
+      \addplot+[dashed, every mark/.append style={solid, fill=gray}, mark=otimes*] table {
+        3000	0.008
         4500	0.015
-        6000	0.029
-        7500	0.042
-        9000	0.063
-        10500	0.084
-        12000	0.115
-        13500	0.144
-        15000	0.183
-        16500	0.223
+        6000	0.027
+        7500	0.043
+        9000	0.064
+        10500	0.095
+        12000	0.124
+        13500	0.16
+        15000	0.2
+        16500	0.259
       }; %WASH, Prpu, C
@@
     %start data for plotting 
-      \addplot+[mark=square*] table {
-        1000	0.141
-        1050	0.202
-        1100	0.157
-        1150	0.273
-        1200	0.24
-        1250	0.408
-        1300	0.465
-        1350	1.416
-        1400	0.657
-        1450	0.806
+      \addplot+[mark=*] table {
+        1000	0.353
+        1050	0.213
+        1100	0.247
+        1150	0.301
+        1200	0.672
+        1250	0.703
+        1300	0.273
+        1350	0.71
+        1400	0.277
+        1450	1.141
       }; %ACNET, Prpu, SML
 
-      \addplot+[mark=o] table {
-        1000	0.121
-        1050	0.17
-        1100	0.123
-        1150	0.216
-        1200	0.197
-        1250	0.237
-        1300	0.298
-        1350	0.621
-        1400	0.399
-        1450	0.486
+      \addplot+[every mark/.append style={solid, fill=gray}, mark=o] table {
+        1000	0.046
+        1050	0.031
+        1100	0.034
+        1150	0.039
+        1200	0.073
+        1250	0.065
+        1300	0.038
+        1350	0.071
+        1400	0.036
+        1450	0.104
       }; %ACNET, Prpu, C++
 
-      \addplot+[mark=otimes*] table {
-        1000	0.056
-        1050	0.09
-        1100	0.069
-        1150	0.106
-        1200	0.09
-        1250	0.152
-        1300	0.163
-        1350	0.176
-        1400	0.223
-        1450	0.209
+      \addplot+[dashed, every mark/.append style={solid, fill=gray}, mark=otimes*] table {
+        1000	0.092
+        1050	0.122
+        1100	0.116
+        1150	0.143
+        1200	0.179
+        1250	0.108
+        1300	0.202
+        1350	0.184
+        1400	0.146
+        1450	0.179
       }; %ACNET, Prpu, C
@@
     %start data for plotting
-      \addplot+[mark=square*] table {
-        16000	3.602
-        17600	4.223
-        19200	4.839
-        20800	8.201
-        22400	8.273
-        24000	13.426
-        25600	10.089
-        27200	15.889
-        28800	16.431
-        30400	21.437
+      \addplot+[mark=*] table {
+        16000	3.334
+        17600	4.538
+        19200	5.844
+        20800	6.391
+        22400	8.308
+        24000	8.242
+        25600	18.534
+        27200	12.174
+        28800	16.653
+        30400	19.968
       }; %RPFGEN, Prpu, SML
 
-      \addplot+[mark=o] table {
-        16000	4.51
-        17600	4.804
-        19200	5.899
-        20800	8.052
-        22400	8.448
-        24000	11.555
-        25600	8.878
-        27200	13.999
-        28800	12.99
-        30400	13.791
+      \addplot+[every mark/.append style={solid, fill=gray}, mark=o] table {
+        16000	0.775
+        17600	0.966
+        19200	1.496
+        20800	1.29
+        22400	1.598
+        24000	1.525
+        25600	2.145
+        27200	2.035
+        28800	2.478
+        30400	2.745
       }; %RPFGEN, Prpu, C++
 
-      \addplot+[mark=otimes*] table {
+      \addplot+[dashed, every mark/.append style={solid, fill=gray}, mark=otimes*] table {
         16000	0.009
-        17600	0.012
-        19200	0.015
-        20800	0.021
-        22400	0.021
-        24000	0.019
-        25600	0.023
-        27200	0.028
-        28800	0.026
-        30400	0.026
+        17600	0.011
+        19200	0.013
+        20800	0.012
+        22400	0.017
+        24000	0.017
+        25600	0.021
+        27200	0.024
+        28800	0.024
+        30400	0.022
       }; %RPFGEN, Prpu, C  

### 2017-03-02 paper/root.tex
polishing paper

--- 2017-03-02 paper/root.tex
@@
 Implementations of the push-relabel algorithm are among the most efficient maximum flow algorithms. While the generic algorithm has a time complexity of $O(V^2E)$,
-specific variants of the algorithms achieve even lower complexities up to $O(V^2\sqrt{E})$.
+specific variants of the algorithms achieve even lower complexities down to $O(V^2\sqrt{E})$.
 
@@
 the implementation details: After proving the min-cut-max-flow theorem, it is straightforward to verify a generic version of the Ford-Fulkerson algorithm,
-which we then refine to Edmonds-Karp algorithm and finally to an efficient implementation. 
+which we then refine to Edmonds-Karp algorithm, and finally to an efficient implementation. 
 Similarly, the generic push-relabel algorithm is refined to both the relabel-to-front algorithm~\cite{CLRS09} and the FIFO push-relabel algorithm~\cite{GoTa88},
-which, in turn, are refined to efficient implementations. The efficiency of our verified implementations compare well to unverified reference implementations of 
+which, in turn, are refined to efficient implementations. The efficiency of our verified implementations compares well to unverified reference implementations of 
 the algorithms.
@@
 
-In the rest of this paper, we give a short informal introduction the Ford-Fulkerson method (\S\ref{sec:background})
+In the rest of this paper, we give a short informal description the Ford-Fulkerson method (\S\ref{sec:background})
 and present our formalization of it (\S\ref{sec:abs-formalization}).
@@
 
-Networks are based on graphs, and add the source and sink nodes, as well as the network assumptions:
+Networks extend graphs by adding the source and sink nodes, as well as the network assumptions:
 \begin{lstlisting}
@@
   Push-relabel algorithms are based on \emph{preflows}, a generalization of flows
-  by relaxing the conservation constraint to allow more incoming flow than outgoing flow on a node.
-  For a node $u\in V$, the \emph{excess} $x_f(u)$ is the difference between incoming and outgoing flow. 
-  For a preflow, the excess of any node but the source must not be negative.
-  A non-sink node with positive excess is called \emph{active}. Thus, a preflow without any active nodes is actually a flow.
+  by relaxing the conservation constraint to allow more incoming flow than outgoing flow.
+  For a node $u\in V$, the difference between incoming and outgoing flow is called \emph{excess} and denoted by $x_f(u)$.
+  For a preflow, the excess of any node but the source is not negative.
+  A non-sink node with positive excess is called \emph{active}. A preflow without any active nodes is actually a flow.
 
@@
   Moreover, the height of the source is fixed to $\left\vert{V}\right\vert$, and the height of the sink is fixed to $0$.
-  By generalizing the labeling invariant to paths, we get that for each path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$.
+  By generalizing the labeling invariant to paths, we get that for every path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$.
   This implies that there cannot be a path from $s$ to $t$: the length of any (simple) path is less than $|V|$, however, a valid labeling 
@@
   assumes "((f_0, l_0), p, (f, l)) \<in> pr_lts^*" 
-  shows "(\<sum>x<-p. cost x) \<le> 38 * (card V)$^2$ * card E"
+  shows "(\<sum>x<-p. cost x) \<le> 26 * (card V)$^2$ * card E"
     and "(f,l)\<notin>Domain pr_lts --> isMaxFlow f"
 \end{lstlisting}  
-where \isai$(f_0,l_0)$ is the initial flow and height labeling and \isai{pr_lts^*} denotes the reflexive transitive closure of \isai{pr_lts}. 
+where \isai$(f_0,l_0)$ is the initial flow and height labeling, and \isai{pr_lts^*} denotes the reflexive transitive closure of \isai{pr_lts}. 
 Intuitively, the cost of any (partial) run \isai$p$ from the initial state to \isai$(f,l)$ is bounded by $O(V^2E)$, and if the 
@@
 Cormen counts the different types of operations separately, using quite involved arguments.
-For our formalization, we have adapted Cormen's proofs, trading precision of the constant factors for simplicity of the formalization. 
-Thus, our constant of $38$ is somewhat higher than the $17$ that would result from Cormen's proof.
+We have slightly modified Cormen's proofs, trading precision of the constant factors for simplicity of the formalization. 
+% Thus, our constant of $26$ is somewhat higher than the $17$ that would result from Cormen's proof.
 
@@
 For the push operations, we define a push to be \emph{saturating}, if the excess at the source node of the push is not less than the capacity of the edge.
-In this case, the edge is completely saturated by the push and disappears from the residual graph. However, the source node of the push remains active.
-On the other hand, a non-saturating push deactivates the source node of the push.
+In this case, the edge is saturated by the push and disappears from the residual graph. However, the source node of the push may remain active.
+On the other hand, a non-saturating push deactivates the source node of the push, but the edge remains in the residual graph.
 
@@
 there must be an operation that restores the edge. This can only be a push on the reverse edge $(v,e)$. 
-However, this push can only occur after the labeling of $v$ has increased. This can only happen $O(V)$ times, which yields a bound of $O(VE)$ on the number 
+However, this push can only occur after the height of $v$ has increased. As the height of any node is bounded by $O(V)$, we get a bound of $O(VE)$ on the number 
 of saturating push operations. 
-Finally, the number of non-saturating pushes is bounded by defining a potential function that is decreased by a non-saturating push.
+Finally, the number of non-saturating pushes is bounded by defining the potential function $\Phi = \sum_{\textrm{$u$ is active node}}l(u)$.
+It is decreased by a non-saturating push.
 A saturating push or relabel operation increases the potential by $O(V)$.
@@
 
-We consider two common instantiations of the generic algorithm: The relabel-to-front algorithm~\cite{CLRS09}, and the FIFO push-relabel algorithm~\cite{GoTa88}.
+We consider two instantiations of the generic algorithm: The relabel-to-front algorithm~\cite{CLRS09}, and the FIFO push-relabel algorithm~\cite{GoTa88}.
 Both have a theoretical time complexity of $O(V^3)$, which is better than the $O(V^2E)$ bound of the generic algorithm.
@@
 a complexity proof in the style done for the generic algorithm.
-Instead, we use the Monadic Refinement Framework to specify the algorithms and only prove correctness and 
-termination.\footnote{Which we get for free from the bound of the generic algorithm.} We leave it to future work to integrate complexity reasoning tools into 
+Instead, we use the Isabelle Refinement Framework to specify the algorithms and only prove correctness and 
+termination.\footnote{Which we easily derive from the bound for the generic algorithm.} We leave it to future work to integrate complexity reasoning tools into 
 the Refinement Framework and to prove the tighter complexity bounds for the specific implementations.
@@
 as the only required additional invariant is that the queue contains exactly the active nodes. 
-Moreover, it performs well in practice, and we have found several implementations of it~\cite{ChGo97,ICPC-notebook}.
+Moreover, it performs well in practice, and we have found several reference implementations of it~\cite{ChGo97,ICPC-notebook}.
 
@@
 The correctness of the heuristics follows easily from the definitions of a valid height labeling.
-In total, the correctness proof of the FIFO push-relabel algorithm is about 400 lines of code and, building on the experience from relabel-to-front, it 
+In total, the correctness proof of the FIFO push-relabel algorithm is about 400 lines of proof text and, building on the experience from relabel-to-front, it 
 took only a day to complete.
@@
   However, we left open several algorithmic aspects. For example, we did not specify how to find a shortest augmenting path.
-  Moreover, efficient implementation of some operations requires additional auxiliary data to be maintained. 
+  Moreover, efficient implementations of some operations require additional auxiliary data to be maintained. 
   Finally, we have to specify the concrete data structures to be used for the implementation. 
@@
   to the current flow. For the implementation, we skip the intermediate augmenting flow, and modify the current flow directly.
-  The relabel operation has to determine the new height for the relabeled node. It's straightforward to implement this by iterating over the 
+  The relabel operation has to determine the new height for the relabeled node. It is straightforward to implement this by iterating over the 
   successor nodes in the residual graph. Refinement of the discharge operation is more tricky: Abstractly, we nondeterministically select a new edge to push, 
@@
   An important optimization is to let the algorithms manipulate residual graphs instead of flows: 
-  Most operations can be phrased to only access one edge of the residual graph, while they would 
+  Most operations can be implemented to only access one edge of the residual graph, while they would 
   need to access multiple edges if phrased on flows. At the end of the algorithm, the flow can be 
@@
 
-  We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers.
+  We implement capacities by (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers.
 Thus, we could chose any executable number representation here.}.
-Note that an implementation as fixed precision numbers would also be possible,
+Note that an implementation by fixed precision numbers would also be possible,
   but requires additional checks on the network to ensure that no overflows can happen. 
@@
   are realized as arrays of size $N\times N$ with row-major indexing, such that the successors of a node are close together in memory.
-  The adjacency map of the network is implemented as an array of lists of nodes. 
+  The adjacency map of the network is implemented by an array of lists of nodes. 
   An augmenting path is represented by a list of edges, \ie a list of pairs of nodes. 
-  The height frequency map and the maps from nodes to height labels and excess values are implemented as arrays.
+  The height frequency map and the maps from nodes to height labels and excess values are implemented by arrays.
   The queue of the FIFO push-relabel algorithm and the list of the relabel-to-front algorithm are implemented by two stacks, allowing for amortized constant time operations.
 
-  Note that there is still some optimization potential left in the choice of our data structures: 
+  There is still some optimization potential left in the choice of our data structures: 
   For example, the BFS algorithm computes a predecessor map $P$. It then iterates over $P$ to extract the shortest path as a list of edges.
@@
 
-  Finally, we use the Isabelle code generator~\cite{Haft09,HaNi10} to extract programs in OCaml, SML, Scala, and Haskell.
+  Finally, we use the Isabelle code generator~\cite{Haft09,HaNi10} to extract our algorithm to an SML program.
 
@@
 
-  Combining the implementations of our algorithms with the network checker yields our final implementations for which we export code.
+  Combining the implementations of our algorithms with the network checker yields our final implementations for which we also export code.
   On any input, the implementation checks whether the input is well-formed and returns a maximal flow for well-formed inputs.
-  We display the theorem for the Edmonds-Karp algorithm here, the theorems for relabel-to-front and FIFO push-relabel are analogous.
+  We display the correctness theorem for the Edmonds-Karp algorithm here, the theorems for relabel-to-front and FIFO push-relabel are analogous.
 
@@
 
-  \noindent Note that this theorem is stated as a Hoare triple, using separation logic~\cite{Rey02,LaMe12} assertions. There are no preconditions on the input.
+  \noindent This theorem is stated as a Hoare triple, using separation logic~\cite{Rey02,LaMe12} assertions. There are no preconditions on the input.
   If the algorithm returns \isai{None}, then the edge list was malformed or described a graph that does not satisfy the network assumptions.
@@
   For the Edmonds-Karp algorithm, we use a reference implementation in Java from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}, which 
-  is compiled and ran with (TODO, Reza's machine?) OpenJDK Java 1.8.0.
+  is compiled and ran with OpenJDK Java 1.8.0.
   As data set, we use randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
@@
   a template for our verified FIFO push-relabel algorithm. Moreover, we use a highly optimized C implementation of a FIFO push-relabel 
-  algorithm with various additional heuristics by Cherkassky et al.~\cite{ChGo97}. These were compiled using (TODO: Reza's machine) gcc 5.4 at optimization level 4.
+  algorithm with various additional heuristics by Cherkassky et al.~\cite{ChGo97}. These were compiled using gcc 5.4 at optimization level 4.
   The data set consists of networks from the DIMACS implementation challenge~\cite{JGC93}: 
@@
 
-  The tests were run on a standard laptop machine with a (TODO: Reza's machine) 2.8GHz i7 quadcore processor and 16GiB of RAM.
+  The tests were run on a standard laptop machine with a 2.7GHz i7 quadcore processor and 16GiB of RAM.
 
@@
 
-  For the push-relabel algorithms, our verified algorithm is roughly a factor of 8 (TODO) slower than the unverified C++ implementation.
-  The highly optimized C implementation is, depending on the test case, one to two orders of magnitudes faster. 
+  For the push-relabel algorithms, our verified algorithm is, depending on the test case, between 5 and 10 times slower than the unverified C++ implementation.
+  The highly optimized C implementation is up to two orders of magnitude faster. 
   On the ``goldberg bad case'' networks, which are designed as worst case problems for push-relabel algorithms, 
@@
   We have presented a mechanical verification of the correctness and time complexity of maximum flow algorithms based on both 
-  the augmenting path and the push-relabel approaches, as well as the Ford-Fulkerson (min-cut-max-flow) theorem, which underlies all these algorithms.
+  the augmenting path and the push-relabel approaches, as well as the min-cut-max-flow theorem, which underlies all these algorithms.
 
@@
   instantiate it to both the relabel-to-front and the FIFO push-relabel algorithm.
-  We then conduct several refinement steps to derive efficiently executable implementations of all three algorithm, from which we extract SML code.
+  We then conduct several refinement steps to derive efficiently executable implementations of all three algorithm.
   The runtime of our verified implementations compares well to that of unverified reference implementations.
@@
   We conclude with some statistics:
-  The formalization consists of roughly 14000 lines of proof text, where the graph theory up to the Ford-Fulkerson algorithm requires 2500 lines.
+  The formalization consists of roughly 14000 lines of proof text, where the graph theory up to the min-cut-max-flow theorem requires 2500 lines.
   The abstract Edmonds-Karp algorithm and its complexity analysis contribute 800 lines, and its implementation (including BFS) another 1700 lines.

### 2017-03-02 paper/root.tex

@@
 		%start data for plotting
-		  \addplot+[mark=square] table {
+		  \addplot+[mark=square*] table {
         1000	0.024
@@
 
-		  \addplot+[dashed, every mark/.append style={solid, fill=gray},mark=*] table {
+		  \addplot+[mark=square] table {
         1000	0.024
@@
 
-		  \addplot+[mark=square*] table {
+		  \addplot+[mark=*] table {
         1000	0.005
@@
 		%start data for plotting 
-		  \addplot+[mark=square] table {
+		  \addplot+[mark=square*] table {
         1000	2.777
@@
 
-		  \addplot+[dashed, every mark/.append style={solid, fill=gray},mark=*] table {
+		  \addplot+[mark=square] table {
         1000	3.892
@@
 
-		  \addplot+[mark=square*] table {
+		  \addplot+[mark=*] table {
         1000	0.058
@@
     %start data for plotting 
-		  \addplot+[mark=square] table {
+		  \addplot+[mark=square*] table {
         3000	0.36
@@
 
-      \addplot+[mark=square*] table {
+      \addplot+[mark=*] table {
         3000	1.187
@@
 
-      \addplot+[mark=o] table {
+      \addplot+[every mark/.append style={solid, fill=gray}, mark=o] table {
         3000	0.183
@@
 
-      \addplot+[mark=otimes*] table {
+      \addplot+[dashed, every mark/.append style={solid, fill=gray}, mark=otimes*] table {
         3000	0.008
@@
     %start data for plotting 
-      \addplot+[mark=square*] table {
+      \addplot+[mark=*] table {
         1000	0.353
@@
 
-      \addplot+[mark=o] table {
+      \addplot+[every mark/.append style={solid, fill=gray}, mark=o] table {
         1000	0.046
@@
 
-      \addplot+[mark=otimes*] table {
+      \addplot+[dashed, every mark/.append style={solid, fill=gray}, mark=otimes*] table {
         1000	0.092
@@
     %start data for plotting
-      \addplot+[mark=square*] table {
+      \addplot+[mark=*] table {
         16000	3.334
@@
 
-      \addplot+[mark=o] table {
+      \addplot+[every mark/.append style={solid, fill=gray}, mark=o] table {
         16000	0.775
@@
 
-      \addplot+[mark=otimes*] table {
+      \addplot+[dashed, every mark/.append style={solid, fill=gray}, mark=otimes*] table {
         16000	0.009

### 2017-03-02 paper/root.tex
improvements to diagrams

--- 2017-03-02 paper/root.tex
@@
         xtick = {0,1000,...,5000},
-        ytick = {0.01,0.1,1,10},
+        ytick = {0.005, 0.05,1,5,10},
         xticklabels = {$1k$, $2k$, $3k$, $4k$, $5k$},
-        yticklabels = {$0.01$,$0.1$,$1$,$10$},
+        yticklabels = {$0.005$, $0.05$,$1$,$5$,$10$},
         smooth,
@@
       xtick = {0,1000,1100,...,1400},
-      ytick = {2,4,6,10,16},
+      ytick = {0.1,2,5,10,15},
       xticklabels = {$1k$, $1.1k$, $1.2k$, $1.3k$, $1.4k$}, 
-      yticklabels = {$2$,$4$,$6$,$10$,$16$}, 
+      yticklabels = {$0.1$,$2$,$5$,$10$,$15$}, 
       smooth,
@@
       ymode = log,
-      xtick = {3000, 6000, 10500, 13500, 16500},
+      xtick = {3000, 6000, 9000, 12000, 15000},
       ytick = {0.1,1,10,50,100},
-      xticklabels = {$3k$, $6k$, $10k$, $13k$, $16k$},
+      xticklabels = {$3k$, $6k$, $9k$, $12k$, $15k$},
       yticklabels = {$0.1$,$1$,$10$,$50$,$100$}, 
@@
       xtick = {0,1000,1100,...,1400},
-      ytick = {0.01, 0.1, 0.2, 0.5, 1},
+      ytick = {0.05, 0.1, 0.2, 0.5, 1},
       xticklabels = {$1k$, $1.1k$, $1.2k$, $1.3k$, $1.4k$}, 
-      yticklabels = {$0.01$, $0.1$,$0.2$,$0.5$,$1$}, 
+      yticklabels = {$0.05$, $0.1$,$0.2$,$0.5$,$1$}, 
       smooth,
@@
       ymode = log,
-      xtick = {16000,19000,23000,27000,30000},
-      ytick = {0.1,3,6,12,24},
-      xticklabels = {$16k$, $19k$, $23k$, $27k$, $30k$}, 
-      yticklabels = {$0.1$, $3$,$6$,$12$,$24$}, 
+      xtick = {16000,19000,22000,25000,30000},
+      ytick = {0.1,1,5,10,25},
+      xticklabels = {$16k$, $19k$, $22k$, $25k$, $28k$}, 
+      yticklabels = {$0.1$, $1$,$5$,$10$,$25$}, 
       smooth,

### 2017-03-02 paper/root.tex
small fix

@@
 
+%%%%%%%%% IMPORTANT %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %change both above and following margins to have consistent indentation
@@
       xtick = {16000,19000,22000,25000,30000},
-      ytick = {0.1,1,5,10,25},
+      ytick = {0.1,1,5,10,20},
       xticklabels = {$16k$, $19k$, $22k$, $25k$, $28k$}, 
-      yticklabels = {$0.1$, $1$,$5$,$10$,$25$}, 
+      yticklabels = {$0.1$, $1$,$5$,$10$,$20$}, 
       smooth,

### 2017-03-02 paper/root.tex

@@
       xtick = {0,1000,1100,...,1400},
-      ytick = {0.1,2,5,10,15},
+      ytick = {0.1,2,4,8,12},
       xticklabels = {$1k$, $1.1k$, $1.2k$, $1.3k$, $1.4k$}, 
-      yticklabels = {$0.1$,$2$,$5$,$10$,$15$}, 
+      yticklabels = {$0.1$,$2$,$4$,$8$,$12$}, 
       smooth,
@@
       xtick = {3000, 6000, 9000, 12000, 15000},
-      ytick = {0.1,1,10,50,100},
+      ytick = {0.1,1,10,40,80},
       xticklabels = {$3k$, $6k$, $9k$, $12k$, $15k$},
-      yticklabels = {$0.1$,$1$,$10$,$50$,$100$}, 
+      yticklabels = {$0.1$,$1$,$10$,$40$,$80$}, 
       smooth,
@@
       xtick = {16000,19000,22000,25000,30000},
-      ytick = {0.1,1,5,10,20},
+      ytick = {0.1,1,4,8,16},
       xticklabels = {$16k$, $19k$, $22k$, $25k$, $28k$}, 
-      yticklabels = {$0.1$, $1$,$5$,$10$,$20$}, 
+      yticklabels = {$0.1$, $1$,$4$,$8$,$16$}, 
       smooth,

### 2017-03-02 formalization/evaluation/fofu-Java/Fofu.java
Fixed overflow bugs that we introduced in flow output

@@ -39,7 +39,7 @@ public class Fofu {
 	        long time = doTest(G);
           System.out.format("@@@time: %d ms\n", time);
           System.out.format("[Input (V E): %d %d]\n", G.V(), G.E());
-          System.out.format("@@@max-flow: %d\n", (int)maxflow.value());
+          System.out.format("@@@max-flow: %.0f\n", maxflow.value());
           System.out.format("outer_c: %d\n\tinner_c: %d\n", maxflow.outer_c, maxflow.inner_c);
 	        
 // 	        if (V_size > 0 && V_size <= 10000) {

### 2017-03-02 paper/root.tex
remark about overflows, fixed overful hbox (listing rightmargin setting caused spurious overful hbox warnings)

@@
 \lstset{autogobble=true}
-\lstset{xleftmargin=.15in, xrightmargin=.15in}
+\lstset{xleftmargin=.15in, xrightmargin=.05in}
 
@@
 \newenvironment{lstquote}%
-  {\list{}{\leftmargin=0.15in\rightmargin=0.15in}\item[]}%
+  {\list{}{\leftmargin=0.15in\rightmargin=.05in}\item[]}%
   {\endlist}
@@
 
-% \overfullrule=8pt
+\overfullrule=8pt
 
@@
 A push operation can be implemented in time $O(1)$, (only the flow on a single edge has to be modified) and a relabel operation can be implemented in time $O(V)$.
-(to find the new height, we have to consider the heights of all successor nodes in the residual graph) We define 
+(the new height depends on the heights of the successors in the residual graph) We define 
 \isai{cost PUSH == 1 | cost RELABEL == card(V)} and prove the following theorem:
@@
 Note that an implementation by fixed precision numbers would also be possible,
-  but requires additional checks on the network to ensure that no overflows can happen. 
+  but requires additional checks on the network to ensure that no overflows can happen.\footnote{Actually, the unverified reference 
+  implementations of the push-relabel algorithms~\cite{ChGo97,ICPC-notebook} that we use in our benchmarks (\cf Section~\ref{sec:benchmarks}) lack such checks, and silently report erroneous maximum flow values in case of overflow.}
   % TODO: Can we trick the Java reference implementation on integer overflow?

### 2017-03-02 paper/root.tex
moved the diagram to page 16
fixed a referencei error in one of the footnotes

@@
   but requires additional checks on the network to ensure that no overflows can happen.\footnote{Actually, the unverified reference 
-  implementations of the push-relabel algorithms~\cite{ChGo97,ICPC-notebook} that we use in our benchmarks (\cf Section~\ref{sec:benchmarks}) lack such checks, and silently report erroneous maximum flow values in case of overflow.}
+  implementations of the push-relabel algorithms~\cite{ChGo97,ICPC-notebook} that we use in our benchmarks (\cf Section~\ref{sec:benchmark}) lack such checks, and silently report erroneous maximum flow values in case of overflow.}
   % TODO: Can we trick the Java reference implementation on integer overflow?
@@
   the ``fully dense acyclic networks'' of Waissi, and the RMFGEN networks of Goldfarb and Grigoriadis.
-
-  The tests were run on a standard laptop machine with a 2.7GHz i7 quadcore processor and 16GiB of RAM.
-
-  The results are displayed in Figure~\ref{fig:benchmark}, using a log-log scale.
-  Note that we only measure the time required to run the algorithms themselves, excluding the time required for parsing the input and running the network checker.
-
-  For the Edmonds-Karp algorithm, we observe that, for sparse graphs, the Java implementation is roughly faster by a factor of $1.6$, while for dense graphs, 
-  our implementation is faster by a factor of $1.2$. Note that the Java implementation operates on flows, while our implementation 
-  operates on residual graphs (\cf Section~\ref{sec:impl_add_data}). Moreover, the Java implementation does not store the augmenting 
-  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).
-  Our push-relabel algorithm is more than one order of magnitude faster than Edmonds-Karp algorithm.
-
-  For the push-relabel algorithms, our verified algorithm is, depending on the test case, between 5 and 10 times slower than the unverified C++ implementation.
-  The highly optimized C implementation is up to two orders of magnitude faster. 
-  On the ``goldberg bad case'' networks, which are designed as worst case problems for push-relabel algorithms, 
-  even Edmonds-Karp algorithm is competitive to the simple FIFO push-relabel algorithms, while the highly optimized one shows no weaknesses on these networks.
-
-  The performance gap between our verified SML code and the C++ code that implements essentially the same algorithm is not surprising: We compare a 
-  memory safe functional language with a slightly aged compiler against an unsafe language with a very recent and actively developed compiler.
 
@@
 \end{figure}
+
+  The tests were run on a standard laptop machine with a 2.7GHz i7 quadcore processor and 16GiB of RAM.
+
+  The results are displayed in Figure~\ref{fig:benchmark}, using a log-log scale.
+  Note that we only measure the time required to run the algorithms themselves, excluding the time required for parsing the input and running the network checker.
+
+  For the Edmonds-Karp algorithm, we observe that, for sparse graphs, the Java implementation is roughly faster by a factor of $1.6$, while for dense graphs, 
+  our implementation is faster by a factor of $1.2$. Note that the Java implementation operates on flows, while our implementation 
+  operates on residual graphs (\cf Section~\ref{sec:impl_add_data}). Moreover, the Java implementation does not store the augmenting 
+  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).
+  Our push-relabel algorithm is more than one order of magnitude faster than Edmonds-Karp algorithm.
+
+  For the push-relabel algorithms, our verified algorithm is, depending on the test case, between 5 and 10 times slower than the unverified C++ implementation.
+  The highly optimized C implementation is up to two orders of magnitude faster. 
+  On the ``goldberg bad case'' networks, which are designed as worst case problems for push-relabel algorithms, 
+  even Edmonds-Karp algorithm is competitive to the simple FIFO push-relabel algorithms, while the highly optimized one shows no weaknesses on these networks.
+
+  The performance gap between our verified SML code and the C++ code that implements essentially the same algorithm is not surprising: We compare a 
+  memory safe functional language with a slightly aged compiler against an unsafe language with a very recent and actively developed compiler.
 

### 2017-03-02 paper/root.tex
small typo in diagram

@@
       ytick = {0.1,1,4,8,16},
-      xticklabels = {$16k$, $19k$, $22k$, $25k$, $28k$}, 
+      xticklabels = {$16k$, $19k$, $22k$, $25k$, $30k$}, 
       yticklabels = {$0.1$, $1$,$4$,$8$,$16$}, 

### 2017-03-03 paper/root.tex
Reza's comments, more precise citation of Cherkassky-Goldberg reference implementation

@@
 
-Using stepwise refinement techniques, we instantiate the generic Ford-Fulkerson method to Edmonds-Karp algorithm, 
+Using stepwise refinement techniques, we instantiate the generic Ford-Fulkerson algorithm to Edmonds-Karp algorithm, 
 and the generic push-relabel algorithm of Goldberg and Tarjan to both the relabel-to-front and the FIFO push-relabel algorithm.
@@
 \section{Introduction}
-Computing the maximum flow of a network is an important problem in graph theory.
+Computing the maximum flow in a network is an important problem in graph theory.
 Many other problems, like maximum bipartite matching, edge disjoint paths, circulation demand, as well as various scheduling and resource allocating problems can be reduced to it.
-The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. 
+The generic Ford-Fulkerson algorithm~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. 
 The basic idea is to repeatedly search for \emph{augmenting paths} along which more flow can be moved from the source to the sink.
@@
 This paper is an extended version of our conference paper~\cite{LaSe16}, which only covered the Edmonds-Karp algorithm.
-While there exists another formalization of the Ford-Fulkerson method in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial time maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation.
+While there exists another formalization of the Ford-Fulkerson algorithm in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial time maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation.
 Moreover, this paper is a case study on elegantly formalizing algorithms.
 
-In the rest of this paper, we give a short informal description the Ford-Fulkerson method (\S\ref{sec:background})
+In the rest of this paper, we give a short informal description of the Ford-Fulkerson algorithm (\S\ref{sec:background})
 and present our formalization of it (\S\ref{sec:abs-formalization}).
-After introducing the Isabelle Refinement Framework~\cite{LaTu12,La12} (\S\ref{sec:refinement}), we report on the instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm, and on its complexity proof (\S\ref{sec:edka}).
+After introducing the Isabelle Refinement Framework~\cite{LaTu12,La12} (\S\ref{sec:refinement}), we report on the instantiation of the Ford-Fulkerson algorithm to the Edmonds-Karp algorithm, and on its complexity proof (\S\ref{sec:edka}).
 Next, we describe the correctness and complexity proof of the generic push-relabel algorithm (\S\ref{sec:gen-prpu}), and its instantiations 
@@
 
-The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/edmonds_karp/}. 
+The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/max_flow/}. 
 The part on Edmonds-Karp algorithm is also available in the Archive of Formal Proofs~\cite{LaSe16_afp}.
@@
 
-\section{The Ford-Fulkerson Method}\label{sec:background}
+\section{The Ford-Fulkerson Algorithm}\label{sec:background}
 % \begin{comment}
@@
 
-In this section, we give a short introduction to the Ford-Fulkerson method, closely following the presentation by Cormen et al.~\cite{CLRS09}.
+In this section, we give a short introduction to the Ford-Fulkerson algorithm, closely following the presentation by Cormen et al.~\cite{CLRS09}.
 
@@
 The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
-Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.
+Given a network $G$, the maximum flow problem is to find a flow with the maximum value among all flows of the network.
 
@@
 
-The Ford-Fulkerson method is a corollary of this theorem.
+The Ford-Fulkerson algorithm is a corollary of this theorem.
 It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached.
 In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges.
-For this purpose the Ford-Fulkerson method defines the residual graph, which has edges in the same and opposite direction as the network edges.
+For this purpose the Ford-Fulkerson algorithm defines the residual graph, which has edges in the same and opposite direction as the network edges.
 % , which has edges both in parallel and reverse to the edges of the network. 
@@
 
-In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph.
+In each iteration, the algorithm tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph.
 It then pushes as much flow as possible along this path to increase the value of the current flow. 
@@
 
-The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph.
+The Ford-Fulkerson algorithm is generic in that it does not specify how to find an augmenting path in the residual graph.
 There are several possible implementations with different execution times.
-The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}.
+The generic algorithm is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}.
 When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real-valued capacities.
@@
 
-\section{Formalizing the Ford-Fulkerson Method}\label{sec:abs-formalization}
+\section{Formalizing the Ford-Fulkerson Algorithm}\label{sec:abs-formalization}
 %  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
@@
 
-In this section, we provide a brief overview of our formalization of the Ford-Fulkerson method. 
+In this section, we provide a brief overview of our formalization of the Ford-Fulkerson algorithm. 
 In order to develop theory in the context of a fixed graph or network, we use Isabelle's concept of \emph{locales}~\cite{Ballarin:2006:MKM}, which allows
@@
 For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that resemble standard mathematical textbook proofs, and are accessible, to a certain extent, even for those not familiar with Isabelle/HOL.
-We use Isar to present our proof of the Ford-Fulkerson method such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.
+We use Isar to present our proof of the Ford-Fulkerson algorithm such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.
 
@@
 As a formal equivalent to pseudocode, we use the monadic programming language provided by the Isabelle Refinement Framework~\cite{LaTu12,La12}.
-For example, we define the Ford-Fulkerson method as follows:
+For example, we define the Ford-Fulkerson algorithm as follows:
 \Snippet{ford_fulkerson_algo}
@@
 Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Isabelle Refinement Framework, it is straightforward to prove (partial) correctness of
-the Ford-Fulkerson method, which is stated in Isabelle/HOL by the following theorem:
+the Ford-Fulkerson algorithm, which is stated in Isabelle/HOL by the following theorem:
 \Snippet{ford_fulkerson_correct}
@@
 After having stated and proved correct an algorithm on the abstract level, the next step is to provide an (efficient) implementation. 
-In our case, we first specialize the Ford-Fulkerson method to use shortest augmenting paths, then implement the search for shortest augmenting paths by BFS, and finally use efficient data structures to represent the abstract objects modified by the algorithm.
+In our case, we first specialize the Ford-Fulkerson algorithm to use shortest augmenting paths, then implement the search for shortest augmenting paths by BFS, and finally use efficient data structures to represent the abstract objects modified by the algorithm.
 
@@
 \section{The Edmonds-Karp Algorithm}\label{sec:edka}
-  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm is straightforward, as finding a shortest augmenting path is a refinement of finding any augmenting path.
+  Specializing the Ford-Fulkerson algorithm to the Edmonds-Karp algorithm is straightforward, as finding a shortest augmenting path is a refinement of finding any augmenting path.
 
@@
   By generalizing the labeling invariant to paths, we get that for every path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$.
-  This implies that there cannot be a path from $s$ to $t$: the length of any (simple) path is less than $|V|$, however, a valid labeling 
+  This implies that there cannot be an augmenting path, \ie a simple path from $s$ to $t$: the length of any simple path is less than $|V|$, however, a valid labeling 
   implies a length greater than or equal to $|V|$. Upon termination of the algorithm, there will be no active nodes. 
-  Thus, the preflow is actually a flow, and as we have no $s$--$t$ paths, the Ford-Fulkerson theorem implies that it is maximal.
+  Thus, the preflow is actually a flow, and as there are no augmenting paths, the min-cut-max-flow theorem implies that the flow is maximal.
 
@@
 over finite sets, was very helpful to keep the formalization effort manageable. 
-Anyway, the whole complexity proof is roughly 1000 lines of Isabelle text, and took a few weeks to complete.
+Nevertheless, the whole complexity proof is roughly 1000 lines of Isabelle text, and took a few weeks to complete.
 This includes the time that we spent on an initial termination proof that used simpler arguments, but did not derive an exact complexity bound.
@@
 While the relabel-to-front algorithm has a theoretical worst case complexity of $O(V^3)$, it seems to be rather inefficient in practice.\footnote{We have not found any implementation based on relabel-to-front, and our own experiments indicate a rather poor performance.}
-An alternative implementation is the FIFO push-relabel algorithm, which maintains a first-in-first-out queue of the active nodes. 
-In each iteration, it dequeues a node from the FIFO, and tries to discharge it by only using push operations.
+An alternative implementation is the FIFO push-relabel algorithm, which maintains a queue (FIFO) of the active nodes. 
+In each iteration, it dequeues a node and tries to discharge it by only using push operations.
 If the node is still active, it is relabeled and enqueued.
@@
 as the only required additional invariant is that the queue contains exactly the active nodes. 
-Moreover, it performs well in practice, and we have found several reference implementations of it~\cite{ChGo97,ICPC-notebook}.
+Moreover, it performs well in practice, and we have found several reference implementations of it~\cite{Gold_Network_Opt,ICPC-notebook}.
 
@@
   \subsection{Algorithms}
-  The most complex problem for which we specified no algorithm yet is finding a shortest augmenting path.
+  The most complex problem for which we specified no algorithm yet is how to find a shortest augmenting path in the Edmonds-Karp algorithm.
   We use breadth first search (BFS) for this. We adapted a BFS formalization from the examples collection of the Isabelle Refinement Framework to
@@
   but requires additional checks on the network to ensure that no overflows can happen.\footnote{Actually, the unverified reference 
-  implementations of the push-relabel algorithms~\cite{ChGo97,ICPC-notebook} that we use in our benchmarks (\cf Section~\ref{sec:benchmark}) lack such checks, and silently report erroneous maximum flow values in case of overflow.}
+  implementations of the push-relabel algorithms~\cite{Gold_Network_Opt,ICPC-notebook} that we use in our benchmarks (\cf Section~\ref{sec:benchmark}) lack such checks, and silently report erroneous maximum flow values in case of overflow.}
   % TODO: Can we trick the Java reference implementation on integer overflow?
@@
   We have not included the relabel-to-front algorithm, as it was three orders of magnitude slower in preliminary tests.
-
-  For the Edmonds-Karp algorithm, we use a reference implementation in Java from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}, which 
-  is compiled and ran with OpenJDK Java 1.8.0.
-  As data set, we use randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
-Note that the maximum density for networks that satisfy our assumptions is $0.5$, as we allow no parallel edges.
-
-  For the push-relabel algorithm, we use the C++ implementation from the Stanford ACM-ICPC notebook~\cite{ICPC-notebook}, which served as 
-  a template for our verified FIFO push-relabel algorithm. Moreover, we use a highly optimized C implementation of a FIFO push-relabel 
-  algorithm with various additional heuristics by Cherkassky et al.~\cite{ChGo97}. These were compiled using gcc 5.4 at optimization level 4.
-  The data set consists of networks from the DIMACS implementation challenge~\cite{JGC93}: 
-  The ``Goldberg bad case networks'' from the ``Washington'' generator, which are tailored to be hard for push-relabel algorithms, 
-  the ``fully dense acyclic networks'' of Waissi, and the RMFGEN networks of Goldfarb and Grigoriadis.
 
@@
 
+  For the Edmonds-Karp algorithm, we use a reference implementation in Java from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}, which 
+  is compiled and ran with OpenJDK Java 1.8.0.
+  As data set, we use randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
+Note that the maximum density for networks that satisfy our assumptions is $0.5$, as we allow no parallel edges.
+
+  For the push-relabel algorithm, we use the C++ implementation from the Stanford ACM-ICPC notebook~\cite{ICPC-notebook}, which served as 
+  a template for our verified FIFO push-relabel algorithm. Moreover, we use the Q\_PRF program 
+  version $1.8$ from Andrew Goldberg's Network Optimization Library~\cite{Gold_Network_Opt}, which is a highly optimized C implementation of a FIFO push-relabel 
+  algorithm with various additional heuristics, originally developed by Cherkassky and Goldberg~\cite{ChGo97}. 
+  These were compiled using gcc 5.4 at optimization level 4.
+  The data set consists of networks from the DIMACS implementation challenge~\cite{JGC93}: 
+  The ``Goldberg bad case networks'' from the ``Washington'' generator, which are tailored to be hard for push-relabel algorithms, 
+  the ``fully dense acyclic networks'' of Waissi, and the RMFGEN networks of Goldfarb and Grigoriadis.
+
   The tests were run on a standard laptop machine with a 2.7GHz i7 quadcore processor and 16GiB of RAM.
@@
   For the push-relabel algorithms, our verified algorithm is, depending on the test case, between 5 and 10 times slower than the unverified C++ implementation.
-  The highly optimized C implementation is up to two orders of magnitude faster. 
-  On the ``goldberg bad case'' networks, which are designed as worst case problems for push-relabel algorithms, 
+  The highly optimized C implementation is significantly faster in most cases, only for the ``fully dense acyclic networks'' it is slower than the 
+  simple C++ implementation.
+  On the ``Goldberg bad case'' networks, which are designed as worst case problems for push-relabel algorithms, 
   even Edmonds-Karp algorithm is competitive to the simple FIFO push-relabel algorithms, while the highly optimized one shows no weaknesses on these networks.
@@
 
-  For the augmenting path algorithms, we start with a verification of the generic Ford-Fulkerson method, 
+  For the augmenting path algorithms, we start with a verification of the generic Ford-Fulkerson algorithm, 
   specialize it to the Edmonds-Karp algorithm, and prove its time complexity of $O(VE^2)$.
@@
   \subsection{Related Work}\label{sec:related_work}
-  We are only aware of one other formalization of the Ford-Fulkerson method conducted in Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication
+  We are only aware of one other formalization of the Ford-Fulkerson algorithm conducted in Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication
   on this formalization except~\cite{Lee05}, which provides a Mizar proof script without any additional comments except that it ``defines and proves correctness of Ford/Fulkerson's Maximum Network Flow algorithm at the level of graph manipulations''.
 Moreover, in Lee et al.~\cite{LeRu07}, which is about graph representation in Mizar, the formalization is shortly mentioned, and it is clarified that it does not provide any implementation or data structure formalization.
-  As far as we understood the Mizar proof script, it formalizes an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson method.
+  As far as we understood the Mizar proof script, it formalizes an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson algorithm.
   Termination is only proved for integer valued capacities.
@@
 Second, the algorithm is split in two phases:
-  The first phase ignores nodes with labels greater $|V|$, and already determines the exact value of the flow. The second phase only pushes excess flow back to the source. 
+  The first phase ignores nodes with labels greater than $|V|$, and already determines the exact value of the flow. The second phase only pushes excess flow back to the source.
   Using different implementations for the two phases is reported to improve practical performance~\cite{ChGo97}. 

### 2017-03-03 paper/root.tex
More precise description of overflow bugs

@@
   but requires additional checks on the network to ensure that no overflows can happen.\footnote{Actually, the unverified reference 
-  implementations of the push-relabel algorithms~\cite{Gold_Network_Opt,ICPC-notebook} that we use in our benchmarks (\cf Section~\ref{sec:benchmark}) lack such checks, and silently report erroneous maximum flow values in case of overflow.}
+  implementations of the push-relabel algorithms~\cite{Gold_Network_Opt,ICPC-notebook} that we use in our benchmarks (\cf Section~\ref{sec:benchmark}) lack such checks, and silently report erroneous maximum flow values in case of overflow.
+One implementation~\cite{ICPC-notebook} has a parser that silently ignores overflows,
+  while the other~\cite{Gold_Network_Opt} uses a too small integer type for the excess flow.}
   % TODO: Can we trick the Java reference implementation on integer overflow?

### 2017-03-03 formalization/make_submission_tgz.sh
Webpage, and minor fixes

@@ -1,3 +1,0 @@
-#!/bin/sh
-
-tar -czf EdmondsKarp_Maxflow.tgz --exclude="Dinic_Maxflow.tgz" --exclude-from=TGZ_EXCLUDE --exclude-tag-all=NOTGZ --exclude-backups --exclude-vcs --transform 's|^|Dinic_Maxflow/|' *

### 2017-03-03 formalization/mkpkg.sh

@@ -1,0 +1,3 @@
+#!/bin/sh
+
+tar -czf maxflow.tgz --exclude="maxflow.tgz" --exclude-from=TGZ_EXCLUDE --exclude-tag-all=NOTGZ --exclude-backups --exclude-vcs --transform 's|^|maxflow/|' *

### 2017-03-03 mkhtml.sh

@@ -1,0 +1,71 @@
+#!/bin/bash
+
+BASEDIR=$PWD
+
+REBUILD=false
+UPLOAD=false
+
+while [[ $# -gt 0 ]]
+do
+key="$1"
+
+case $key in
+    -r|--rebuild)
+    REBUILD=true
+    ;;
+    -u|--upload)
+    UPLOAD=true
+    ;;
+    *)
+    echo "Unknown command line argument $1"        # unknown option
+    exit 1
+    ;;
+esac
+shift # past argument or value
+done
+
+
+
+if $REBUILD; then
+
+cd formalization
+isabelle build -b -d '$AFP' -D .
+./mkpkg.sh
+cd $BASEDIR
+
+fi
+
+
+rm -rf html
+mkdir -p html
+
+
+cp -a README.md html/
+cp -a formalization/maxflow.tgz html/
+
+
+cp formalization/Flow_Networks/output/outline.pdf  html/fnet_outline.pdf
+cp formalization/Flow_Networks/output/document.pdf html/fnet_document.pdf
+
+cp formalization/Edka_Maxflow/output/outline.pdf  html/edka_outline.pdf
+cp formalization/Edka_Maxflow/output/document.pdf html/edka_document.pdf
+
+cp formalization/Prpu_Maxflow/output/outline.pdf  html/prpu_outline.pdf
+cp formalization/Prpu_Maxflow/output/document.pdf html/prpu_document.pdf
+
+
+pandoc -V pagetitle="Formalization of Network Flow Algorithms" -s index.md > html/index.html
+
+
+if $UPLOAD; then
+  LOCAL_DEST=~/devel/www21-lammich/max_flow
+  rm -rf $LOCAL_DEST
+  cp -a html $LOCAL_DEST
+  cd $LOCAL_DEST
+  hg add .
+  hg commit -m "Automatic update of Maxflow" .
+  hg push
+  cd $BASEDIR
+fi
+
+

### 2017-09-06 paper/rebuttal.txt
paper updates

Rebuttal and comments on reviewer comments:

Thank you to the reviewers for there reviews and very useful comments.

Please find in the following a short log on how we incorporated the comments in our paper, and reasons 
why we did not follow some of the reviewer's proposals.

@Reviewer #1:

Unfortunately, the paper is already scratching the 21 page limit, so that we were hardly able to add more detail.

We added a short explanation of reflexive transitive closure of LTSs, and explained Isabelle's non-standard set comprehension syntax.
We also added a more detailed explanation of the Hoare-triple and separation logic syntax used for the correctness theorem.

We changed the wording for the presentation of Isar, to not present it as unusual. 
We also added the argument of the community size behind MLton vs. gcc.

Chargueraud et al. actually used their methods to verify graph algorithms, while Chlipala et al. 
only applied refinement in a quite specific setting to data-structure refinement for databases. 
I'm not aware of any other application of Fiat so far. As this paper is mainly about human-readable formalizations of (graph) algorithms,
and not so much about automated refinement techniques, we did not include Fiat into the related work discussion.

@Reviewer2:

We tried to clarify in the introduction that we only formalize a O(V^2E) bound. 
We decided not to integrate this in the abstract, in order to keep it short.
Also, in the related work section, we added some discussion on our method of estimating complexity, 
compared to other methods of Nipkow and Chargueraud.

We clarified that the transformation to satisfy our constraints is linear time, and not part of our formalization.

We explained the non-standard set-comprehension syntax.

We re-phrased the paragraph mentioning that we need to provide more powerful combinators in the Refinement Framework.

We gave a pointer to our ITP'12 paper \sec 2.3, where a more detailed explanation of total vs partial correctness in the Refinement Framework can be found.

Unfortunately, the page-limit does not allow a more detailed explanation of the induction proof for the Edmonds-Karp complexity.

We slightly changed Sec 5, and added a discussion about our complexity analysis compared to other methods to related work.

We tried to re-phrase the informal explanation of push-relabel algorithms to make it easier to read. We think that 
"simple path" is standard enough that we do not need to define it explicitly. At least from the context it should be 
quite clear (now) that it is a path with no repeated vertexes.

We added a footnote to show how much we loosened the constant factors in our analysis, compared to Cormen.

We added some comments on our matrix implementation of graphs, and its deficits on sparse graphs.

We added some explanation of the syntax used in the main correctness statement (Hoare triple, sep-logic, ...)

We added some missing curves to the benchmark charts, and explained why some are still missing.

We added Chen and Levys work to related work, we also discuss Chargueraud and Pottier for their complexity analysis techniques.
However, we did not want to discuss any formalization of something like a graph algorithm, to keep the related work section concise 
(and within the page limit).

@Reviewer 3:

We did not want to over-emphasize these rather low-level and easy to fix parser and overflow errors in the reference 
implementations, thus we kept it as footnote.

We explained our Hoare-triple syntax. I, personally, am not a fan of too overloaded syntactic sugar, and there is already a lot of that in the Refinement Framework.

### 2017-09-06 paper/root.tex
Changed titles
Fixed my name spaceing in the paper + changed repository's .md title
paper updates

@@
 
-\title{Formalization of Network Flow Algorithms}
+\title{Formalizing Network Flow Algorithms}
 \subtitle{A Refinement Approach in Isabelle/HOL}
@@
 
-\author{Peter Lammich \and S. Reza Sefidgar}
+\author{Peter Lammich \and S.~Reza Sefidgar}
 
@@
 \and
-  S. Reza Sefidgar \at
+  S.~Reza Sefidgar \at
   Institute of Information Security,
@@
 Implementations of the push-relabel algorithm are among the most efficient maximum flow algorithms. While the generic algorithm has a time complexity of $O(V^2E)$,
-specific variants of the algorithms achieve even lower complexities down to $O(V^2\sqrt{E})$.
+specific variants of the algorithm achieve even lower complexities down to $O(V^2\sqrt{E})$.
 
@@
 
-The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/max_flow/}. 
-The part on Edmonds-Karp algorithm is also available in the Archive of Formal Proofs~\cite{LaSe16_afp}.
+The complete formalization is available as a set of three entries in the Archive of Formal Proofs~\cite{LaSe16_afp,LaSe17-flownet_afp,LaSe17-prpu_afp}.
+% The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/max_flow/}. 
+% The part on Edmonds-Karp algorithm is also available in the Archive of Formal Proofs~\cite{LaSe16_afp}.
 
@@
 % Intuitively, only the source may generate flow, and only the sink may consume flow, and for all other nodes, the incoming flow must be equal to the outgoing flow.
-The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
+The value $|f|$ of a flow $f$ is defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
 Given a network $G$, the maximum flow problem is to find a flow with the maximum value among all flows of the network.
@@
 To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) the source only has outgoing edges while the sink only has incoming edges; 2) if the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every node of the network must be on a path from $s$ to $t$.
-Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.
+Any network can be transformed in linear time to fulfill these properties, preserving the maximum flow.
+However, we did not formalize such a transformation.
+
+% to a network with the aforementioned properties and the same maximum flow 
+% in linear time~\cite{CLRS09}.\footnote{We leave the formalization of this transformation for future work.}
 
@@
 In each iteration, the algorithm tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph.
-It then pushes as much flow as possible along this path to increase the value of the current flow. 
+It then pushes as much flow as possible along this path, increasing the current flow. 
 Formally, for an augmenting path $p$, one first defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
@@
 The generic algorithm is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}.
-When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real-valued capacities.
+When always choosing a \emph{shortest}\footnote{\ie a path with a minimum number of edges.} augmenting path, the number of iterations is bounded by $O(VE)$, even for the general case of real-valued capacities.
 Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges if the intended meaning is clear from the context.
-A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.
+A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with a time complexity of $O(VE^2)$.
 
@@
 \end{lstlisting}
-Moreover, we define basic concepts like (simple, shortest) paths, and provide lemmas to reason about them.
+We use Isabelle/HOL's non-standard set-comprehension syntax \isai!{(x_1,...,x$_n$). P}! for the set 
+of all tuples \isai!(x_1,...,x$_n$)! satisfying $P$.
+% \isai!(x_1,...,x$_n$)! for the set 
+% 
+% The set of all tuples \isai!(x_1,...,x$_n$)! with property $P$ is written as \isai!{(x_1,...,x$_n$). P}!.
+The locale also provides basic concepts like (simple, shortest) paths and lemmas to reason about them.
+
+% We also define basic concepts like (simple, shortest) paths, and provide lemmas to reason about them.
 
@@
 \subsection{Presentation of Proofs}
-% (DONE: not sure about repeating citation of Cormen: I would definitely repeat the citation!) 
+% TODO: R1 critizises that Isar is presented as something unusual, although it is well-known in the TP community
 Informal proofs focus on the relevant thoughts by leaving out technical details and obvious steps. 
@@
 Although modern proof assistants provide high-level tactics to summarize some of these steps, formal proofs tend to be significantly more 
-verbose than informal proofs. Moreover, formal proofs are conducted in the tactic language of the proof assistant, which is often some dialect of ML. 
-Thus, many formal proofs are essentially programs that instruct the proof assistant how to conduct the proof.
+verbose than informal proofs. Moreover, many formal proofs are conducted in \emph{applicative} style, where a proof is merely a program that instructs 
+the proof assistant how to transform the theorem until it is proved. Those proofs tend to be inaccessible without a deep knowledge of the used 
+proof assistant, often requiring a replay of the proof in the proof assistant.
+In our proofs, we mainly used the \emph{declarative} proof language Isar~\cite{Wenzel99}, which allows to write formal proofs that resemble standard mathematical textbook proofs, and are accessible, to a certain extent, even for those not familiar with Isabelle/HOL.
+
+% 
+% Moreover, formal proofs are conducted in the tactic language of the proof assistant, which is often some dialect of ML. 
+% Thus, many formal proofs are essentially programs that instruct the proof assistant how to conduct the proof.
 They tend to be inaccessible without a deep knowledge of the used proof assistant, in many cases requiring to replay the proof in the proof assistant in order to understand the idea behind it.
-
-For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that resemble standard mathematical textbook proofs, and are accessible, to a certain extent, even for those not familiar with Isabelle/HOL.
-We use Isar to present our proof of the Ford-Fulkerson algorithm such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.
+% 
+% For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that resemble standard mathematical textbook proofs, and are accessible, to a certain extent, even for those not familiar with Isabelle/HOL.
+
+% We use Isar to present our proof of the Ford-Fulkerson algorithm such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.
 
@@
 Note that we had to use the available combinators of the Isabelle Refinement Framework, which made the code slightly more verbose than we would have liked.
-We leave it to future work to define a set of combinators and appropriate syntax that allows for more concise presentation of pseudocode.
+In particular we had to explicitly thread a \isai$brk$-flag through the state of the while loop, emulating a \isai{break} statement.
+We leave it to future work to extend the Isabelle Refinement Framework by more advanced combinators and syntactic sugar to allow for a more concise presentation of abstract algorithms.
+
+% 
+% define a set of combinators and appropriate syntax that allows for more concise presentation of pseudocode.
 
@@
 with results from other possible runs. On the other hand, a nonterminating partial correct recursion yields \isai|res {}|, which refines any specification 
-and disappears when joined with other results.
+and disappears when joined with other results. For a more detailed explanation of partial and total correctness in the Isabelle Refinement Framework, we refer the reader to~\cite[\S2.3]{LaTu12}.
 
@@
 
-  Formalizing the above intuitive argument was more tricky than it seemed on first glance:
+  Formalizing the above intuitive argument was trickier than it seemed on first glance:
   While it is easy to prove that, in a \emph{fixed graph}, an edge and its opposite cannot both lie on shortest paths, generalizing the argument
@@
   Having defined the measure function and shown that it decreases on augmentation, it is straightforward to refine the partial correct while loop to a total correct one.
-Moreover, to make explicit the bound on the number of loop iterations, we instrument the loop to count its iterations, and assert the upper bound after the loop.
+We have also instrumented the loop to count its iterations, and asserted an upper bound of $2VE+V$ = $O(VE)$ after the loop.
 
@@
   Moreover, the height of the source is fixed to $\left\vert{V}\right\vert$, and the height of the sink is fixed to $0$.
-  By generalizing the labeling invariant to paths, we get that for every path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$.
-  This implies that there cannot be an augmenting path, \ie a simple path from $s$ to $t$: the length of any simple path is less than $|V|$, however, a valid labeling 
-  implies a length greater than or equal to $|V|$. Upon termination of the algorithm, there will be no active nodes. 
+
+  The above invariant does not admit augmenting paths: By generalizing the invariant 
+  to paths, we get that for every path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$. 
+  In particular, the length of a path from the source to the sink must be greater than or equal to $\left\vert{V}\right\vert$. 
+  However, as an augmenting path is simple by definition, its length must be less than $\left\vert{V}\right\vert$.
+
+%   By generalizing the labeling invariant to paths, we get that for every path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$.
+%   This implies that there cannot be an augmenting path, \ie a simple path from $s$ to $t$: the length of any simple path is less than $|V|$, however, a valid labeling 
+%   implies a length greater than or equal to $|V|$.
+
+  Upon termination of the algorithm, there will be no active nodes. 
   Thus, the preflow is actually a flow, and as there are no augmenting paths, the min-cut-max-flow theorem implies that the flow is maximal.
@@
     The maximum possible flow is constrained by the residual capacity of the edge, and the excess flow available at $u$.
-    \item[$\textrm{relabel}(u)$] for an active node $u$ from which no push operation is possible: increase the height of $u$ to the minimum value such that a push
+    \item[$\textrm{relabel}(u)$] for an active node $u$ from which no push operation is possible: increase the height of $u$ to the minimum value such that a push from $u$
       becomes possible.
@@
 \end{lstlisting}  
-where \isai$(f_0,l_0)$ is the initial flow and height labeling, and \isai{pr_lts^*} denotes the reflexive transitive closure of \isai{pr_lts}. 
-Intuitively, the cost of any (partial) run \isai$p$ from the initial state to \isai$(f,l)$ is bounded by $O(V^2E)$, and if the 
+where \isai$(f_0,l_0)$ is the initial flow and height labeling, and \isai{pr_lts^*} denotes the reflexive transitive closure of \isai{pr_lts}.\footnote{
+  I.e., $(u_0,[l_1,\ldots,l_n],u_n)\in R^*$ iff $\forall 0\le i < n.~(u_i,l_{i+1},u_{i+1})\in R$.
+} 
+Intuitively, the cost of any (partial) run \isai$p$ from the initial state to \isai$(f,l)$ is bounded by $26V^2E = O(V^2E)$, and if the 
 run is complete, \ie no more operations apply (\isai$(f,l)\<notin>Domain pr_lts$), then the resulting flow is maximal.
@@
 
-The complexity argument is more tricky. Instead of defining a measure function,
+The complexity argument is trickier. Instead of defining a measure function,
 Cormen counts the different types of operations separately, using quite involved arguments.
-We have slightly modified Cormen's proofs, trading precision of the constant factors for simplicity of the formalization. 
+We have slightly modified Cormen's proofs, trading precision of the constant factors for simplicity of the formalization.\footnote{Using Cormen's technique, we would have been able to prove a bound of $17V^2E$.}
 % Thus, our constant of $26$ is somewhat higher than the $17$ that would result from Cormen's proof.
@@
 The generic algorithm does not specify the order in which the operations are applied. 
-Choosing the right order decreases the worst case time complexity, and, more important, the practical runtime.
+Choosing the right order decreases the worst case time complexity, and, more importantly, the practical runtime.
 
@@
 a complexity proof in the style done for the generic algorithm.
-Instead, we use the Isabelle Refinement Framework to specify the algorithms and only prove correctness and 
-termination.\footnote{Which we easily derive from the bound for the generic algorithm.} We leave it to future work to integrate complexity reasoning tools into 
+Instead, we use the Isabelle Refinement Framework to specify the algorithms and only prove total correctness. 
+% The termination proof is thereby easily derived from the termination of the generic algorithm.
+% % and 
+% % termination.\footnote{Which we easily derive from the bound for the generic algorithm.} 
+We leave it to future work to integrate complexity reasoning tools into 
 the Refinement Framework and to prove the tighter complexity bounds for the specific implementations.
 
-A general idea of most efficient implementations is to \emph{discharge} a node by repeatedly applying push operations until the node gets inactive or needs to be relabeled.
+A general idea of the most efficient implementations is to \emph{discharge} a node by repeatedly applying push operations until the node gets inactive or needs to be relabeled.
 The implementations mainly differ in the order in which the nodes are discharged.
+
+Obviously, a sequence of discharge and relabel operations is a special instance of the generic algorithm, which gives us termination and invariant preservation.
+The main work of the correctness proofs for the refined algorithms goes into showing that they only terminate if no active nodes are left.
 
@@
 \section{Refinement to Executable Code}\label{sec:executable}
-  In the previous section, we have presented our abstract formalization of the Edmonds-Karp and push-relabel algorithms.
+  In the previous sections, we have presented our abstract formalization of the Edmonds-Karp and push-relabel algorithms.
   However, we left open several algorithmic aspects. For example, we did not specify how to find a shortest augmenting path.
@@
   The relabel operation has to determine the new height for the relabeled node. It is straightforward to implement this by iterating over the 
-  successor nodes in the residual graph. Refinement of the discharge operation is more tricky: Abstractly, we nondeterministically select a new edge to push, 
+  successor nodes in the residual graph. Refinement of the discharge operation is trickier: Abstractly, we nondeterministically select a new edge to push, 
   until we find no more edges.
@@
   Also the FIFO queue could be implemented by a ring buffer based on a fixed size array, yielding true constant time operations with less overhead.
-  Fortunately, our profiling data shows that these operations do not significantly contribute to the runtime of our implementations, such that 
-  we leave these optimizations for future work.
-
-  Finally, we use the Isabelle code generator~\cite{Haft09,HaNi10} to extract our algorithm to an SML program.
+  Fortunately, our profiling data shows that these operations do not significantly contribute to the runtime of our implementations.
+  A more important optimization would be to refine our naive implementation of graphs as adjacency matrix. 
+  While this representation is well-suited for dense graphs, it wastes memory and cache locality for 
+  sparse graphs (\cf Section~\ref{sec:benchmark}). We leave it to future research to verify graph representations that are more suited to sparse graphs.
+
+  Finally, we use the Isabelle code generator~\cite{Haft09,HaNi10} to extract our algorithm to an SML program, which we 
+  link to a simple parser and command line interface to get a stand-alone program for benchmarking.
 
@@
 
-  \noindent This theorem is stated as a Hoare triple, using separation logic~\cite{Rey02,LaMe12} assertions. There are no preconditions on the input.
+  \noindent This theorem is stated as a Hoare triple, using separation logic~\cite{Rey02,LaMe12} assertions.\footnote{
+    A Hoare triple is written as ${<}P{>}~c~{<}\lambda r.~Q~r{>}_t$, where $P$ is the precondition, $c$ the program, and $Q$ the 
+    postcondition that also depends on the program's return value $r$. Pre- and postcondition are written as separation logic 
+    assertions, where $\textrm{emp}$ describes the empty heap, $*$ is the separating conjunction, $\uparrow\Phi$ indicates a pure assertion, 
+    \ie one that describes no heap content, and $\exists_\textrm{A}$ is the existential quantifier lifted to assertions.
+  } 
+  There are no preconditions on the input.
   If the algorithm returns \isai{None}, then the edge list was malformed or described a graph that does not satisfy the network assumptions.
@@
   We have benchmarked our verified algorithms against unverified reference implementations.
-  We extract SML code for our verified Edmonds-Karp and FIFO push-relabel algorithms, and compile it with MLton~\cite{MLton} version 20100608.
-  We have not included the relabel-to-front algorithm, as it was three orders of magnitude slower in preliminary tests.
+  The SML code for our verified Edmonds-Karp and FIFO push-relabel algorithms has been compiled with MLton~\cite{MLton}. % version 20100608.
+  We have not included the relabel-to-front algorithm, as it was three orders of magnitude slower than FIFO push-relabel in preliminary tests.
+%
 
@@
 		  }; %Sparse, Prpu, SML
+
+      \addplot+[every mark/.append style={solid, fill=gray}, mark=o] table {
+        1000	0.002
+        1500	0.003
+        2000	0.002
+        2500	0.008
+        3000	0.008
+        3500	0.006
+        4000	0.008
+        4500	0.014
+        5000	0.016
+        5500	0.015
+		  }; %Sparse, Prpu, C++
+
+      \addplot+[dashed, every mark/.append style={solid, fill=gray}, mark=otimes*] table {
+        1000	0.0005 % Modified from 0, otherwise not plotted on log-scale!
+        1500	0.001
+        2000	0.001
+        2500	0.004
+        3000	0.003
+        3500	0.004
+        4000	0.005
+        4500	0.007
+        5000	0.01
+        5500	0.009
+		  }; %Sparse, Prpu, C
+
 		%end data for plotting
@@
 		  }; %Dense, Prpu, SML
+
+      \addplot+[every mark/.append style={solid, fill=gray}, mark=o] table {
+        1000	0.006
+        1050	0.006
+        1100	0.007
+        1150	0.007
+        1200	0.009
+        1250	0.009
+        1300	0.009
+        1350	0.009
+        1400	0.01
+        1450	0.011
+		  }; %Dense, Prpu, C++
+
+      \addplot+[dashed, every mark/.append style={solid, fill=gray}, mark=otimes*] table {
+        1000	0.005
+        1050	0.004
+        1100	0.006
+        1150	0.005
+        1200	0.007
+        1250	0.007
+        1300	0.007
+        1350	0.007
+        1400	0.007
+        1450	0.008
+		  }; %Dense, Prpu, C
+
 		%end data for plotting
@@
 
+		  \addplot+[mark=square] table {
+        3000	0.099
+        4500	0.186
+        6000	0.326
+        7500	0.48
+        9000	0.84
+        10500	0.918
+        12000	1.162
+        13500	1.463
+        15000	2.006
+        16500	2.16
+		  }; %Wash, Edka, Java
+
       \addplot+[mark=*] table {
@@
   For the Edmonds-Karp algorithm, we use a reference implementation in Java from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}, which 
-  is compiled and ran with OpenJDK Java 1.8.0.
-  As data set, we use randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
-Note that the maximum density for networks that satisfy our assumptions is $0.5$, as we allow no parallel edges.
+  is compiled and run with OpenJDK Java 1.8.0.
+  It is similar to our implementation except that it uses adjacency lists to represent the graph, operates directly 
+  on flows (\cf Section~\ref{sec:impl_add_data}), and uses no intermediate list to store the augmenting path (\cf Section~\ref{sec:impl_data_structures}).
 
@@
   These were compiled using gcc 5.4 at optimization level 4.
-  The data set consists of networks from the DIMACS implementation challenge~\cite{JGC93}: 
-  The ``Goldberg bad case networks'' from the ``Washington'' generator, which are tailored to be hard for push-relabel algorithms, 
+  The tests were run on a standard laptop machine with a 2.7GHz i7 quadcore processor and 16GiB of RAM.
+
+  Our data set consists of randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, 
+  and the dense networks having a density of $0.25$. Note that the maximum density for networks that satisfy our assumptions is $0.5$, as we allow no 
+  parallel edges.
+  Moreover, we use some networks from the DIMACS implementation challenge~\cite{JGC93}: 
+  The ``Goldberg bad case networks'' from the ``Washington'' generator, 
   the ``fully dense acyclic networks'' of Waissi, and the RMFGEN networks of Goldfarb and Grigoriadis.
 
-  The tests were run on a standard laptop machine with a 2.7GHz i7 quadcore processor and 16GiB of RAM.
+%   
+%   For the latter two, we excluded the results for the Edmonds-Karp implementations, as their performance was to poor compared to push-relabel algorithms.
+%   
+%   
+%   As data set, we use randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$.
+Note that the maximum density for networks that satisfy our assumptions is $0.5$, as we allow no parallel edges.
+%   
+%   The data set consists of networks from the DIMACS implementation challenge~\cite{JGC93}: 
+%   The ``Goldberg bad case networks'' from the ``Washington'' generator, which are tailored to be hard for push-relabel algorithms, 
+%   the ``fully dense acyclic networks'' of Waissi, and the RMFGEN networks of Goldfarb and Grigoriadis.
 
@@
   Note that we only measure the time required to run the algorithms themselves, excluding the time required for parsing the input and running the network checker.
+  Moreover, for the sake of readability, we excluded the Edmonds-Karp implementations from the diagrams for fully dense acyclic networks and RMFGEN networks, as they were orders of magnitude slower than the push-relabel implementations.
 
   For the Edmonds-Karp algorithm, we observe that, for sparse graphs, the Java implementation is roughly faster by a factor of $1.6$, while for dense graphs, 
-  our implementation is faster by a factor of $1.2$. Note that the Java implementation operates on flows, while our implementation 
-  operates on residual graphs (\cf Section~\ref{sec:impl_add_data}). Moreover, the Java implementation does not store the augmenting 
-  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).
-  Our push-relabel algorithm is more than one order of magnitude faster than Edmonds-Karp algorithm.
+  our implementation is faster by a factor of $1.2$. We conjecture that this is mainly due to the Java implementation using adjacency lists which 
+  are better suited for sparse graphs than our adjacency matrices.
+This is supported by the results for the even sparser (density ranges from $10^{-4}$ to $10^{-5}$) Goldberg bad case networks, where the Java implementation is $4.8$ times faster on average.
+
+%   Note that the Java implementation operates on flows, while our implementation 
+%   operates on residual graphs (\cf Section~\ref{sec:impl_add_data}). Moreover, the Java implementation does not store the augmenting 
+%   path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).
+%   Our push-relabel algorithm is more than one order of magnitude faster than Edmonds-Karp algorithm.
 
@@
   simple C++ implementation.
-  On the ``Goldberg bad case'' networks, which are designed as worst case problems for push-relabel algorithms, 
-  even Edmonds-Karp algorithm is competitive to the simple FIFO push-relabel algorithms, while the highly optimized one shows no weaknesses on these networks.
-
   The performance gap between our verified SML code and the C++ code that implements essentially the same algorithm is not surprising: We compare a 
-  memory safe functional language with a slightly aged compiler against an unsafe language with a very recent and actively developed compiler.
+  memory safe functional language with a slightly aged compiler used by only a few people
+  against an unsafe language with a very recent and actively developed compiler and a huge user base.
+
+  The push-relabel implementations are significantly faster than the Edmonds-Karp implementations, except on the 
+  ``Goldberg bad case'' networks, which are designed as worst case problems for push-relabel algorithms: 
+  Here, the simple C++ implementation of FIFO push-relabel is slower than the Edmonds-Karp algorithm, while the highly optimized one 
+  in C shows no weaknesses.
+
+%   Here, the Edmonds-Karp implementations are competitive to the simple C++ implementation of FIFO push-relabel, while the highly optimized one 
+%   in C shows no weaknesses on these networks.
+
+%   On the ``Goldberg bad case'' networks, which are designed as worst case problems for push-relabel algorithms, 
+%   even Edmonds-Karp algorithm is competitive to the simple FIFO push-relabel algorithms, while the highly optimized one shows no weaknesses on these networks.
+%   Moreover, as these networks are very sparse ($10^{-4}$ to $10^{-5}$), the Java implementation of Edmonds-Karp is significantly faster 
+%   (5 times on average) than ours.
 
@@
 
+  More recently, Chen and Levy~\cite{ChLe17} used Why3~\cite{FiPa13} to verify a functional implementation of Tarjan's strongly connected components algorithm,
+  with emphasis on proof readability.
+
   Chargu\'eraud~\cite{char11} also uses a bottom-up approach to verify imperative programs written in a subset of OCaml, amongst them a version of Dijkstra's algorithm:
@@
 
+  We prove the complexity bounds of our algorithms on an abstract level, by 
+  instrumenting the loop with a counter to count abstract operations (in our case, loop iterations). 
+  This counter does not appear in the refined versions of the code, nor is there any formal model of runtime of the generated code. 
+  In this respect, our technique is similar to the approach used by Nipkow~\cite{Ni15}, where a complexity function is derived 
+  systematically but manually from the function to be analyzed. A more precise approach is taken by Chargu{\'e}raud and Pottier~\cite{ChPo15}, who 
+  argue about the number of $\beta$-reduction steps taken by the program.
+
   \subsection{Future Work}
-  While we implemented a simple version of the push-relabel algorithm, there are more advanced implementations, like the one of Cherkassky and Goldberg~\cite{ChGo97} that 
-  we use in the benchmarks. The same authors also provide an implementation based on the highest label selection rule, which is reported to be even more efficient 
-  in practice. 
-  Compared to these advanced algorithms, our formalization misses two important heuristics: First, the global relabeling heuristics periodically performs a breadth first search on the reversed residual graph, to determine exact node labels based on their distance to the sink.
+  The most obvious improvement to our formalization is to use a graph representation which is better suited for sparse graphs, or even switch between the two implementations depending on the input graph.
+Compared to the most advanced implementations of push-relabel algorithms, \eg \cite{ChGo97}, our formalization lacks two important heuristics: First, the global relabeling heuristics periodically performs a breadth first search on the reversed residual graph, to determine exact node labels based on their distance to the sink.
 Second, the algorithm is split in two phases:
-  The first phase ignores nodes with labels greater than $|V|$, and already determines the exact value of the flow. The second phase only pushes excess flow back to the source.
-  Using different implementations for the two phases is reported to improve practical performance~\cite{ChGo97}. 
-  Thus, the main focus of future work will be to verify these heuristics in order to obtain more efficient verified algorithms.
+  The first phase ignores nodes with labels greater than $|V|$, and already determines the exact value of the flow. The second phase only pushes excess 
+  flow back to the source.
+%   Using different implementations for the two phases is reported to improve practical performance~\cite{ChGo97}. 
+  The main focus of future work will be to verify these heuristics.
+
+%   While we implemented a simple version of the push-relabel algorithm, there are more advanced implementations, like the one of Cherkassky and Goldberg~\cite{ChGo97} that 
+%   we use in the benchmarks. The same authors also provide an implementation based on the highest label selection rule, which is reported to be even more efficient 
+%   in practice. 
+%   Compared to these advanced algorithms, our formalization misses two important heuristics: First, the global relabeling heuristics periodically performs a breadth first search on the reversed residual graph, to determine exact node labels based on their distance to the sink.
+Second, the algorithm is split in two phases:
+%   The first phase ignores nodes with labels greater than $|V|$, and already determines the exact value of the flow. The second phase only pushes excess flow back to the source.
+%   Using different implementations for the two phases is reported to improve practical performance~\cite{ChGo97}. 
+%   Thus, the main focus of future work will be to verify these heuristics in order to obtain more efficient verified algorithms.
+%
 

### 2017-09-06 formalization/mk_afp_submission.sh
tuned

@@ -1,0 +1,64 @@
+#!/bin/bash
+
+DIRS="Flow_Networks Prpu_Maxflow"
+ARCHIVE_NAME="afp-submission.tgz"
+AFP="$HOME/devel/isabelle/afp-2016-1-orig"
+SESSIONS="Prpu_Maxflow"
+
+EXCLUDE="Prpu_Maxflow/Prpu_Benchmark_Export.thy"
+
+
+set -e
+OLDDIR="$PWD"
+TMPDIR=`mktemp -d`
+echo "Preparing submission in $TMPDIR"
+
+TGZ_FILE="$OLDDIR/$ARCHIVE_NAME"
+
+# Collect files
+cp -a $DIRS "$TMPDIR"
+
+cd "$TMPDIR"
+
+# Clean up
+find . -type d -name obsolete | xargs rm -rf
+find . -type d -name output | xargs rm -rf
+find . -name '*~' | xargs rm -rf
+
+find . -name 'NOTGZ' | xargs rm -rf
+find . -name '.gitignore'  | xargs rm -rf
+
+rm -rf $EXCLUDE
+
+# Ughly post-processing
+find . -name "ROOT" | xargs sed -i -e '/^ *(\*No-AFP/d'
+find . -mindepth 2 -maxdepth 2 -name "*.thy" -or -name "ROOT" | xargs sed -i -e 's|"\$AFP|"..|g'
+find . -mindepth 3 -maxdepth 3 -name "*.thy" -or -name "ROOT" | xargs sed -i -e 's|"\$AFP|"../..|g'
+
+# Make tarfile
+tar -czf "$TGZ_FILE" $DIRS
+
+cd "$OLDDIR"
+
+# Remove temporary dir
+rm -rf "$TMPDIR"
+
+# Setup test
+mkdir "$TMPDIR"
+cd "$TMPDIR"
+
+# Prepare AFP environment
+find "$AFP/thys" -mindepth 1 -maxdepth 1 -type d | xargs -n1 ln -s
+cp -a "$AFP/thys/ROOTS" .
+
+# Extract files
+tar -xzf "$TGZ_FILE"
+
+# Add new folders to ROOTS
+for d in $DIRS; do echo $d >> ROOTS; done
+
+# Test build
+isabelle build -d. $SESSIONS
+
+cd $OLDDIR
+# rm -rf $TMPDIR
